### 2012-10-01 manuscript/cloudbreak.tex
manuscript

\documentclass[12pt]{article}
\usepackage{graphics}
\usepackage{subfigure}

\begin{document}

\title{A MapReduce Algorithm for Detecting Genomic Structural Variation from Complete Paired-End Sequence Data Sets}

\author{
Christopher Whelan\ Kemal S\"onmez\ \\
Biomedical Engineering \\
Oregon Health and Science University \\
20000 NW Walker Road, Beaverton, OR 97006, USA
}

\date{\today}

\maketitle

\begin{abstract}
This is the paper's abstract \ldots
\end{abstract}

\section{Introduction}
Genomic structural variants are important..

Difficulty of detection.. Lack of concordance of current algorithms...

Typical algorithms use only unambiguous discordant read pair mappings...

Hadoop/MapReduce as framework for processing complete data sets

\section{Methods}\label{Methods}

MapReduce algorithm description

Local deletion scores

\section{Results}\label{results}

Simulated data

1000 Genomes Data

AML Data set

\section{Discussion}\label{discussion}

\section{Conclusions}\label{conclusions}

\bibliographystyle{abbrv}
\bibliography{cloudbreak}

\end{document}

### 2012-10-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/command/CommandNovoalignSingleEnds.java
switch to cloudera hadoop distro

@@ -10,6 +10,7 @@ import org.apache.hadoop.filecache.DistributedCache;
 import org.apache.hadoop.fs.FileSystem;
 import org.apache.hadoop.fs.Path;
 import org.apache.hadoop.io.DoubleWritable;
+import org.apache.hadoop.io.LongWritable;
 import org.apache.hadoop.io.SequenceFile;
 import org.apache.hadoop.io.Text;
 import org.apache.hadoop.mapred.*;
@@ -70,7 +71,7 @@ public class CommandNovoalignSingleEnds extends BaseCloudbreakCommand {
         conf.set("novoalign.threshold", threshold);
         conf.set("novoalign.quality.format", qualityFormat);
 
-        conf.setInputFormat(TextInputFormat.class);
+        conf.setInputFormat(SequenceFileInputFormat.class);
 
         conf.setMapperClass(NovoalignSingleEndMapper.class);
         conf.setMapOutputKeyClass(Text.class);

### 2012-10-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/command/CommandReadPairedEndFilesIntoHDFS.java

@@ -7,6 +7,7 @@ import org.apache.hadoop.fs.FSDataOutputStream;
 import org.apache.hadoop.fs.FileSystem;
 import org.apache.hadoop.fs.Path;
 import org.apache.hadoop.io.IntWritable;
+import org.apache.hadoop.io.LongWritable;
 import org.apache.hadoop.io.SequenceFile;
 import org.apache.hadoop.io.Text;
 import org.apache.hadoop.io.compress.SnappyCodec;
@@ -49,7 +50,7 @@ public class CommandReadPairedEndFilesIntoHDFS implements CloudbreakCommand {
 
         HDFSWriter writer = new HDFSWriter();
         if ("snappy".equals(compress)) {
-            writer.seqFileWriter = SequenceFile.createWriter(hdfs, config, p, IntWritable.class, Text.class, SequenceFile.CompressionType.BLOCK, new SnappyCodec());
+            writer.seqFileWriter = SequenceFile.createWriter(hdfs, config, p, LongWritable.class, Text.class, SequenceFile.CompressionType.BLOCK, new SnappyCodec());
         } else {
             FSDataOutputStream outputStream = hdfs.create(p);
             BufferedWriter bufferedWriter = null;

### 2012-10-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/command/CommandReadPairedEndFilesIntoHDFS.java
switch to snappy compression for reads

@@ -156,7 +156,7 @@ public class CommandReadPairedEndFilesIntoHDFS implements CloudbreakCommand {
             if (textFileWriter != null) {
                 textFileWriter.write(line);
             } else {
-                seqFileWriter.append(new IntWritable(recordNum), new Text(line));
+                seqFileWriter.append(new LongWritable(recordNum), new Text(line));
             }
         }
 

### 2012-10-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/command/CommandReadPairedEndFilesIntoHDFS.java
switch to snappy compression for reads

@@ -6,7 +6,6 @@ import org.apache.hadoop.conf.Configuration;
 import org.apache.hadoop.fs.FSDataOutputStream;
 import org.apache.hadoop.fs.FileSystem;
 import org.apache.hadoop.fs.Path;
-import org.apache.hadoop.io.IntWritable;
 import org.apache.hadoop.io.LongWritable;
 import org.apache.hadoop.io.SequenceFile;
 import org.apache.hadoop.io.Text;

### 2012-10-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/command/CommandReadPairedEndFilesIntoHDFS.java
switch to snappy compression for reads

@@ -39,7 +39,7 @@ public class CommandReadPairedEndFilesIntoHDFS implements CloudbreakCommand {
     @Parameter(names = {"--compress"})
     String compress = "none";
 
-    private int numRecords;
+    private long numRecords;
 
     public void copyReadFilesToHdfs() throws IOException {
         Configuration config = new Configuration();
@@ -151,7 +151,7 @@ public class CommandReadPairedEndFilesIntoHDFS implements CloudbreakCommand {
         public BufferedWriter textFileWriter;
         public SequenceFile.Writer seqFileWriter;
 
-        public void write(int recordNum, String line) throws IOException {
+        public void write(long recordNum, String line) throws IOException {
             if (textFileWriter != null) {
                 textFileWriter.write(line);
             } else {

### 2012-10-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentMapper.java
switch to snappy compression for reads

@@ -1,5 +1,6 @@
 package edu.ohsu.sonmezsysbio.cloudbreak.mapper;
 
+import org.apache.hadoop.io.IntWritable;
 import org.apache.hadoop.io.LongWritable;
 import org.apache.hadoop.io.Text;
 import org.apache.hadoop.mapred.*;
@@ -13,7 +14,7 @@ import java.util.zip.GZIPOutputStream;
  * Date: 7/17/12
  * Time: 5:27 PM
  */
-public abstract class SingleEndAlignmentMapper extends MapReduceBase implements Mapper<LongWritable, Text, Text, Text> {
+public abstract class SingleEndAlignmentMapper extends MapReduceBase implements Mapper<IntWritable, Text, Text, Text> {
     private String localDir;
     protected Writer s1FileWriter;
     protected File s1File;
@@ -36,7 +37,7 @@ public abstract class SingleEndAlignmentMapper extends MapReduceBase implements 
 
     }
 
-    public void map(LongWritable key, Text value, OutputCollector<Text, Text> output, Reporter reporter) throws IOException {
+    public void map(IntWritable key, Text value, OutputCollector<Text, Text> output, Reporter reporter) throws IOException {
         if (this.output == null) {
             this.output = output;
         }

### 2012-10-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentMapper.java
switch to snappy compression for reads

@@ -48,6 +48,11 @@ public abstract class SingleEndAlignmentMapper extends MapReduceBase implements 
         String line = value.toString();
         String[] fields = line.split("\t");
 
+        if (fields[1].length() != fields[3].length()) {
+            System.err.println("Warning; mismatching seq and qual lengths in record " + key.toString() + "!");
+            System.err.println(fields[1]);
+            System.err.println(fields[3]);
+        }
         s1FileWriter.write(fields[0] + "\n");
         s1FileWriter.write(fields[1] + "\n");
         s1FileWriter.write(fields[2] + "\n");

### 2012-10-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentMapper.java
switch to snappy compression for reads

@@ -50,8 +50,11 @@ public abstract class SingleEndAlignmentMapper extends MapReduceBase implements 
 
         if (fields[1].length() != fields[3].length()) {
             System.err.println("Warning; mismatching seq and qual lengths in record " + key.toString() + "!");
+            System.err.println("Seq:");
             System.err.println(fields[1]);
+            System.err.println("Qual:");
             System.err.println(fields[3]);
+            System.err.println("DONE WARNING");
         }
         s1FileWriter.write(fields[0] + "\n");
         s1FileWriter.write(fields[1] + "\n");

### 2012-10-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentMapper.java
switch to snappy compression for reads

@@ -46,6 +46,15 @@ public abstract class SingleEndAlignmentMapper extends MapReduceBase implements 
         }
 
         String line = value.toString();
+        String[] reads = line.split("\n");
+        splitRead(key, reads[0]);
+        splitRead(key, reads[1]);
+
+        reporter.progress();
+        //System.out.println("Done with map method, real work will happen in close");
+    }
+
+    private void splitRead(IntWritable key, String line) throws IOException {
         String[] fields = line.split("\t");
 
         if (fields[1].length() != fields[3].length()) {
@@ -60,9 +69,6 @@ public abstract class SingleEndAlignmentMapper extends MapReduceBase implements 
         s1FileWriter.write(fields[1] + "\n");
         s1FileWriter.write(fields[2] + "\n");
         s1FileWriter.write(fields[3] + "\n");
-
-        reporter.progress();
-        //System.out.println("Done with map method, real work will happen in close");
     }
 
     public OutputCollector<Text, Text> getOutput() {

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -1,0 +1,63 @@
+#!/usr/bin/env python
+
+import sys
+
+sample_name = sys.argv[1]
+seq_index_file = sys.argv[2]
+hdfs_dir = sys.argv[3]
+data_dir = sys.argv[4]
+threshold = sys.argv[5]
+
+seq_index = open(seq_index_file, 'r')
+
+dag_file = open("align.dag", 'w')
+read_group_file = open("readGroups.txt", 'w')
+
+loaded_groups = set()
+
+int job_num = 1
+for line in seqIndex:
+    fields = line.split()
+    file_name = fields[1]
+    read_group = fields[2]
+    insert_size = fields[20]
+    pair_file_name = fields[24]
+    if (not loaded_groups.contains(read_group) and pair_file_name != ""):
+        print "processing read group {}".format(read_group)
+        dag_file.write("JOB {0} loadAndAlign.desc\n".format(job_num))
+        dag_file.write("VARS {0} read_group={1}\n".format(job_num, read_group))
+        dag_file.write("VARS {0} data_dir={1}\n".format(job_num, data_dir))
+        dag_file.write("VARS {0} hdfs_dir={1}\n".format(job_num, hdfs_dir))
+        dag_file.write("VARS {0} file1={1}\n".format(job_num, file_name))
+        dag_file.write("VARS {0} file2={1}\n".format(job_num, pair_file_name))
+        dag_file.write("VARS {0} threshold={1}\n".format(jon_num, threshold))
+
+job_desc_file = open("loadAndAlign.desc", 'w')
+job_desc_file.write("Executable = loadAndAlign.sh\n")
+job_desc_file.write("  Arguments  = \"$(read_group) $(data_dir) $(hdfs_dir) $(file1) $(file2) $(threshold)\"\n")
+job_desc_file.write("  Universe   = vanilla\n")
+job_desc_file.write("  output     = loadAndAlign.out.$(cluster).$(Process)\n")
+job_desc_file.write("  error      = loadAndAlign.err.$(cluster).$(Process)\n")
+job_desc_file.write("  Log        = loadAndAlign.log\n")
+job_desc_file.write("  get_env    = true\n")
+job_desc_file.write("  Notification = Never\n")
+job_desc_file.write("  Queue\n")
[…]

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -15,7 +15,7 @@ read_group_file = open("readGroups.txt", 'w')
 
 loaded_groups = set()
 
-int job_num = 1
+job_num = 1
 for line in seqIndex:
     fields = line.split()
     file_name = fields[1]
@@ -31,6 +31,7 @@ for line in seqIndex:
         dag_file.write("VARS {0} file1={1}\n".format(job_num, file_name))
         dag_file.write("VARS {0} file2={1}\n".format(job_num, pair_file_name))
         dag_file.write("VARS {0} threshold={1}\n".format(jon_num, threshold))
+        loaded_groups.add(read_group)
 
 job_desc_file = open("loadAndAlign.desc", 'w')
 job_desc_file.write("Executable = loadAndAlign.sh\n")

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -16,7 +16,7 @@ read_group_file = open("readGroups.txt", 'w')
 loaded_groups = set()
 
 job_num = 1
-for line in seqIndex:
+for line in seq_index:
     fields = line.split()
     file_name = fields[1]
     read_group = fields[2]

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -22,7 +22,7 @@ for line in seq_index:
     read_group = fields[2]
     insert_size = fields[20]
     pair_file_name = fields[24]
-    if (not loaded_groups.contains(read_group) and pair_file_name != ""):
+    if (not (read_group in loaded_groups) and pair_file_name != ""):
         print "processing read group {}".format(read_group)
         dag_file.write("JOB {0} loadAndAlign.desc\n".format(job_num))
         dag_file.write("VARS {0} read_group={1}\n".format(job_num, read_group))

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -23,7 +23,7 @@ for line in seq_index:
     insert_size = fields[20]
     pair_file_name = fields[24]
     if (not (read_group in loaded_groups) and pair_file_name != ""):
-        print "processing read group {}".format(read_group)
+        print "processing read group {0}".format(read_group)
         dag_file.write("JOB {0} loadAndAlign.desc\n".format(job_num))
         dag_file.write("VARS {0} read_group={1}\n".format(job_num, read_group))
         dag_file.write("VARS {0} data_dir={1}\n".format(job_num, data_dir))

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -30,7 +30,7 @@ for line in seq_index:
         dag_file.write("VARS {0} hdfs_dir={1}\n".format(job_num, hdfs_dir))
         dag_file.write("VARS {0} file1={1}\n".format(job_num, file_name))
         dag_file.write("VARS {0} file2={1}\n".format(job_num, pair_file_name))
-        dag_file.write("VARS {0} threshold={1}\n".format(jon_num, threshold))
+        dag_file.write("VARS {0} threshold={1}\n".format(job_num, threshold))
         loaded_groups.add(read_group)
 
 job_desc_file = open("loadAndAlign.desc", 'w')

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -43,7 +43,7 @@ job_desc_file.write("  Log        = loadAndAlign.log\n")
 job_desc_file.write("  get_env    = true\n")
 job_desc_file.write("  Notification = Never\n")
 job_desc_file.write("  Queue\n")
-close(job_desc_file)
+job_desc_file.close()
 
 exe_file = open("loadAndAlign.sh", 'w')
 exe_file.write("#!/bin/bash\n")
@@ -61,4 +61,4 @@ exe_file.write("\n")
 exe_file.write("hadoop jar /l2/users/whelanch/gene_rearrange/svpipeline/build/svpipeline/target/cloudbreak-1.0-SNAPSHOT-exe.jar readPairedEndFilesIntoHDFS --HDFSDataDir $HDFS_DIR/data/$READ_GROUP --fastqFile1 $DATA_DIR/$FILE1 --fastqFile2 $DATA_DIR/$FILE2 --compress snappy --outFileName $READ_GROUP\n")
 exe_file.write("hadoop jar /l2/users/whelanch/gene_rearrange/svpipeline/build/svpipeline/target/cloudbreak-1.0-SNAPSHOT-exe.jar alignSingleEnds --HDFSDataDir $HDFS_DIR/data/$READ_GROUP --HDFSAlignmentsDir $HDFS_DIR/alignments/$READ_GROUP --reference indices/hg19.fa.nix --threshold $THRESHOLD --qualityFormat STDFQ --HDFSPathToNovoalign executables/novoalign --HDFSPathToNovoalignLicense executables/novoalign.lic\n")
 exe_file.write("hadoop dfs -rmr $HDFS_DIR/data/$READ_GROUP\n")
-close(exe_file)
+exe_file.close()

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -20,6 +20,7 @@ for line in seq_index:
     fields = line.split()
     file_name = fields[1]
     read_group = fields[2]
+    library = fields[19]
     insert_size = fields[20]
     pair_file_name = fields[24]
     if (not (read_group in loaded_groups) and pair_file_name != ""):
@@ -32,6 +33,10 @@ for line in seq_index:
         dag_file.write("VARS {0} file2={1}\n".format(job_num, pair_file_name))
         dag_file.write("VARS {0} threshold={1}\n".format(job_num, threshold))
         loaded_groups.add(read_group)
+        job_num = job_num + 1
+
+        read_group_file.write("\t".join([read_group, library, insert_size, str(int(insert_size) * .15), "false", hdfs_dir + "/alignments/" + read_group]) + "\n")
+read_group_file.close()
 
 job_desc_file = open("loadAndAlign.desc", 'w')
 job_desc_file.write("Executable = loadAndAlign.sh\n")

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -20,9 +20,9 @@ for line in seq_index:
     fields = line.split()
     file_name = fields[1]
     read_group = fields[2]
-    library = fields[19]
-    insert_size = fields[20]
-    pair_file_name = fields[24]
+    library = fields[20]
+    insert_size = fields[21]
+    pair_file_name = fields[23]
     if (not (read_group in loaded_groups) and pair_file_name != ""):
         print "processing read group {0}".format(read_group)
         dag_file.write("JOB {0} loadAndAlign.desc\n".format(job_num))

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -17,7 +17,7 @@ loaded_groups = set()
 
 job_num = 1
 for line in seq_index:
-    fields = line.split()
+    fields = line.split("\t")
     file_name = fields[1]
     read_group = fields[2]
     library = fields[20]

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -18,11 +18,11 @@ loaded_groups = set()
 job_num = 1
 for line in seq_index:
     fields = line.split("\t")
-    file_name = fields[1]
+    file_name = fields[0]
     read_group = fields[2]
-    library = fields[20]
-    insert_size = fields[21]
-    pair_file_name = fields[23]
+    library = fields[18]
+    insert_size = fields[19]
+    pair_file_name = fields[21]
     if (not (read_group in loaded_groups) and pair_file_name != ""):
         print "processing read group {0}".format(read_group)
         dag_file.write("JOB {0} loadAndAlign.desc\n".format(job_num))

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -20,9 +20,9 @@ for line in seq_index:
     fields = line.split("\t")
     file_name = fields[0]
     read_group = fields[2]
-    library = fields[18]
-    insert_size = fields[19]
-    pair_file_name = fields[21]
+    library = fields[14]
+    insert_size = fields[17]
+    pair_file_name = fields[19]
     if (not (read_group in loaded_groups) and pair_file_name != ""):
         print "processing read group {0}".format(read_group)
         dag_file.write("JOB {0} loadAndAlign.desc\n".format(job_num))

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -35,7 +35,7 @@ for line in seq_index:
         loaded_groups.add(read_group)
         job_num = job_num + 1
 
-        read_group_file.write("\t".join([read_group, library, insert_size, str(int(insert_size) * .15), "false", hdfs_dir + "/alignments/" + read_group]) + "\n")
+        read_group_file.write("\t".join([read_group, library, insert_size, str(int(round(int(insert_size) * .15))), "false", hdfs_dir + "/alignments/" + read_group]) + "\n")
 read_group_file.close()
 
 job_desc_file = open("loadAndAlign.desc", 'w')

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -26,12 +26,12 @@ for line in seq_index:
     if (not (read_group in loaded_groups) and pair_file_name != ""):
         print "processing read group {0}".format(read_group)
         dag_file.write("JOB {0} loadAndAlign.desc\n".format(job_num))
-        dag_file.write("VARS {0} read_group={1}\n".format(job_num, read_group))
-        dag_file.write("VARS {0} data_dir={1}\n".format(job_num, data_dir))
-        dag_file.write("VARS {0} hdfs_dir={1}\n".format(job_num, hdfs_dir))
-        dag_file.write("VARS {0} file1={1}\n".format(job_num, file_name))
-        dag_file.write("VARS {0} file2={1}\n".format(job_num, pair_file_name))
-        dag_file.write("VARS {0} threshold={1}\n".format(job_num, threshold))
+        dag_file.write("VARS {0} read_group=\"{1}\"\n".format(job_num, read_group))
+        dag_file.write("VARS {0} data_dir=\"{1}\"\n".format(job_num, data_dir))
+        dag_file.write("VARS {0} hdfs_dir=\"{1}\"\n".format(job_num, hdfs_dir))
+        dag_file.write("VARS {0} file1=\"{1}\"\n".format(job_num, file_name))
+        dag_file.write("VARS {0} file2=\"{1}\"\n".format(job_num, pair_file_name))
+        dag_file.write("VARS {0} threshold=\"{1}\"\n".format(job_num, threshold))
         loaded_groups.add(read_group)
         job_num = job_num + 1
 

### 2012-10-03 src/main/scripts/buildAlignmentsFrom1KGSequenceIndex.py
script to create condor dag for loading and aligning 1KG sequences

@@ -44,7 +44,7 @@ job_desc_file.write("  Arguments  = \"$(read_group) $(data_dir) $(hdfs_dir) $(fi
 job_desc_file.write("  Universe   = vanilla\n")
 job_desc_file.write("  output     = loadAndAlign.out.$(cluster).$(Process)\n")
 job_desc_file.write("  error      = loadAndAlign.err.$(cluster).$(Process)\n")
-job_desc_file.write("  Log        = loadAndAlign.log\n")
+job_desc_file.write("  Log        = /tmp/loadAndAlign.log\n")
 job_desc_file.write("  get_env    = true\n")
 job_desc_file.write("  Notification = Never\n")
 job_desc_file.write("  Queue\n")
@@ -56,12 +56,12 @@ exe_file.write("\n")
 exe_file.write("set -e\n")
 exe_file.write("set -u\n")
 exe_file.write("\n")
-exe_file.write("$READ_GROUP=$1\n")
-exe_file.write("$DATA_DIR=$2\n")
-exe_file.write("$HDFS_DIR=$3\n")
-exe_file.write("$FILE1=$4\n")
-exe_file.write("$FILE2=$5\n")
-exe_file.write("$THRESHOLD=$6\n")
+exe_file.write("READ_GROUP=$1\n")
+exe_file.write("DATA_DIR=$2\n")
+exe_file.write("HDFS_DIR=$3\n")
+exe_file.write("FILE1=$4\n")
+exe_file.write("FILE2=$5\n")
+exe_file.write("THRESHOLD=$6\n")
 exe_file.write("\n")
 exe_file.write("hadoop jar /l2/users/whelanch/gene_rearrange/svpipeline/build/svpipeline/target/cloudbreak-1.0-SNAPSHOT-exe.jar readPairedEndFilesIntoHDFS --HDFSDataDir $HDFS_DIR/data/$READ_GROUP --fastqFile1 $DATA_DIR/$FILE1 --fastqFile2 $DATA_DIR/$FILE2 --compress snappy --outFileName $READ_GROUP\n")
[…]

### 2012-10-03 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentMapper.java
re-fix problem with snappy

@@ -14,7 +14,7 @@ import java.util.zip.GZIPOutputStream;
  * Date: 7/17/12
  * Time: 5:27 PM
  */
-public abstract class SingleEndAlignmentMapper extends MapReduceBase implements Mapper<IntWritable, Text, Text, Text> {
+public abstract class SingleEndAlignmentMapper extends MapReduceBase implements Mapper<LongWritable, Text, Text, Text> {
     private String localDir;
     protected Writer s1FileWriter;
     protected File s1File;
@@ -37,7 +37,7 @@ public abstract class SingleEndAlignmentMapper extends MapReduceBase implements 
 
     }
 
-    public void map(IntWritable key, Text value, OutputCollector<Text, Text> output, Reporter reporter) throws IOException {
+    public void map(LongWritable key, Text value, OutputCollector<Text, Text> output, Reporter reporter) throws IOException {
         if (this.output == null) {
             this.output = output;
         }
@@ -54,7 +54,7 @@ public abstract class SingleEndAlignmentMapper extends MapReduceBase implements 
         //System.out.println("Done with map method, real work will happen in close");
     }
 
-    private void splitRead(IntWritable key, String line) throws IOException {
+    private void splitRead(LongWritable key, String line) throws IOException {
         String[] fields = line.split("\t");
 
         if (fields[1].length() != fields[3].length()) {

### 2012-10-09 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/command/CommandMrFastSingleEnds.java
update mr fast alignment to use snappy

@@ -3,18 +3,16 @@ package edu.ohsu.sonmezsysbio.cloudbreak.command;
 import com.beust.jcommander.Parameter;
 import com.beust.jcommander.Parameters;
 import edu.ohsu.sonmezsysbio.cloudbreak.Cloudbreak;
-import edu.ohsu.sonmezsysbio.cloudbreak.reducer.SingleEndAlignmentsToPairsReducer;
 import edu.ohsu.sonmezsysbio.cloudbreak.mapper.MrFastSingleEndMapper;
+import edu.ohsu.sonmezsysbio.cloudbreak.reducer.SingleEndAlignmentsToPairsReducer;
 import org.apache.hadoop.conf.Configuration;
 import org.apache.hadoop.filecache.DistributedCache;
 import org.apache.hadoop.fs.FileSystem;
 import org.apache.hadoop.fs.Path;
 import org.apache.hadoop.io.Text;
-import org.apache.hadoop.io.compress.GzipCodec;
 import org.apache.hadoop.mapred.*;
 
-import java.io.*;
-import java.net.URI;
+import java.io.IOException;
 import java.net.URISyntaxException;
 
 /**
@@ -72,9 +70,9 @@ public class CommandMrFastSingleEnds extends BaseCloudbreakCommand {
         conf.setOutputKeyClass(Text.class);
         conf.setCompressMapOutput(true);
 
-        conf.setOutputFormat(TextOutputFormat.class);
-        TextOutputFormat.setCompressOutput(conf, true);
-        TextOutputFormat.setOutputCompressorClass(conf, GzipCodec.class);
+        conf.setOutputFormat(SequenceFileOutputFormat.class);
+        conf.set("mapred.output.compress", "true");
+        conf.set("mapred.output.compression", "org.apache.hadoop.io.compress.SnappyCodec");
 
         conf.setReducerClass(SingleEndAlignmentsToPairsReducer.class);
 

### 2012-10-09 manuscript/cloudbreak.tex
methods

@@
 \documentclass[12pt]{article}
+
 \usepackage{graphics}
 \usepackage{subfigure}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
+\usepackage{url}
 
@@
 \section{Introduction}
-Genomic structural variants are important..
+
+Genomic structural variation
 
@@
 
+Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping, \cite{Schatz:2009p278} querying variant databases, \cite{Oconnor:2010p1835} SNP calling, \cite{Langmead:2009p1225} RNA-seq differential expression analysis, \cite{Langmead:2010p1268} ChIP-seq peak calling, \cite{Feng:2011p1228} and computing genome mappability.
+\cite{Lee:2012bk}
+
 \section{Methods}\label{Methods}
 
-MapReduce algorithm description
+\subsection{A Hadoop/MapReduce Algorithm for SV Detection}
 
-Local deletion scores
+The MapReduce programming model divides computation for a job accross a cluster into three phases \cite{Dean:2008p277}.
+In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit $\langle key, value \rangle$ pairs for each block examined.
+The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
+Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or otherwise aggregate those values.
+Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system for storing large data sets across a cluster.
+
+\begin{figure}[p]
+\algrenewcommand\algorithmicprocedure{\textbf{job}}
+  \begin{algorithmic}[1]
+    \Procedure{Alignment}{}
+    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
+    \ForAll{$ \textrm{Alignments }a \in \textrm{Align}(<s,q>)$}
+    \State $\textsc{Emit}(\textrm{ReadPairId }rid, \textrm{Alignment }a)$
+    \EndFor
+    \EndFunction
+    \Function{Reduce}{$\textrm{ReadPairId }rid, \textrm{Alignments }a_{1,2,\ldots}$}
+    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{ValidAlignmentPairs}(a_{1,2,\ldots}))$
+    \EndFunction
+    \EndProcedure
+
+    \Procedure{Compute Deletion Scores}{}
+    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
+    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2> \in ap$}
+    \State $ \textrm{SpanningPairInfo }spi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
+    \ForAll{$ \textrm{GenomicLocations }g \in \textrm{Span}(a_1,a_2)$}
+    \State $\textsc{Emit}(\textrm{GenomicLocation }g, \textrm{SpanningPairInfo }spi)$
+    \EndFor
+    \EndFor
+    \EndFunction
+    \Function{Reduce}{$\textrm{GenomicLocation }g, \textrm{SpanningPairInfos }spi_{1,2,\ldots}$}
+    \State $\textrm{DeletionScore } d \gets \textrm{DeletionScore}(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
+    \State $\textsc{Emit}(\textrm{GenomicLocation }g, \textrm{DeletionScore } d)$
+    \EndFunction
+    \EndProcedure
+  \end{algorithmic}
+  \caption{MapReduce workflow of the Cloudbreak alorithm for deletion detection.
+Processing is dividied into two MapReduce jobs.
+In the first job, mappers find all alignments for each read pair end in the data set.
+Reducers then group the alignments for each pair and filter invalid alignment pairs.
+In the second job, mappers emit information about all possible read pair alignments that span a genomic location.
+Reducers compute a deletion score for each genomic location.}
+\label{algo1}
+\end{figure}
+
+Cloudbreak divides its processing into two distinct MapReduce jobs.
+The first job finds possible alignment locations for pairs of reads in FASTQ format.
+The second job computes local SV scores for a set of small windows completely covering the target genome.
+Finally, the SV scores are exported from the Hadoop file system and post-processed to produce a final set of SV calls which span multiple genomic windows.
+The Cloudbreak MapReduce algorithm is summarized in \ref{algo1}.
+
+The alignment job uses sensitive alignment tools and settings to discover as many mapping locations for each read pair as possible.
+In the mapping phase, mappers align reads in single-end mode to the reference genome in parallel, outputing each possible mapping location as a value under a key identifying the read pair.
+In the reduce phase, Cloudbreak combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found.
+
+The Cloudbreak algorithm could use any aligner that is capable of reporting multiple alignments for reads.
+We report results generated with alignments from Novoalign.
+\cite{novoalign} Novoalign exhibits higher sensitivity for mapping locations than most aligners, even in the presence of indels and errors, at the expense of greater runtime.
+\cite{Krawitz:2010iq} In addition, Novoalign can report multiple possible mapping locations for a read and has accurate mapping quality scores \cite{Ruffalo:2011p1758}, which are useful in assigning confidence scores to SV calls.
+The current implementation is also capable of using MrFast \cite{Alkan:2009cr} as an aligner, although MrFast does not provide mapping quality scores, which makes Cloudbreak predictions less accurate.
+
+In the second job, mappers collect information, namely the insert size and the pair mapping quality, about all of the possible alignments of read pairs that span each genomic locus, defined as a small (25bp) window in the genome reference.
+Reducers then integrate all of the read pair information for reads that span a given location to produce an SV score indicating the likelihood of an SV affecting that locus.
+Since scores are computed independently at each locus, this procedure is essentially constructing and analyzing a pileup of reads pairs at each location, making it similar to previous bioinformatic MapReduce algorithms for SNP calling implemented in Crossbow \cite{Langmead:2009p1225} and GATK.
+\cite{McKenna:2010p1051}
+
+\subsection{Computing Local Deletion Scores}
+
+As described in the previous section, reducers in the second Cloudbreak job compute an SV score indicating the presence or absence of a deletion for each genomic location $l$.
+Locations are defined as contiguous windows of 25bp covering the entire genome.
+For each location, Cloudbreak examines the set of possible alignments of read pairs that spans that location, meaning that either of the reads or the insert between them overlaps the location.
+We begin with the prior probabilities of having or not having a deletion that spans the location, $P(D)$ and $P(\bar{D}) = 1 - P(D)$.
+Based on the results reported by the 1000 Genomes Project pilot data \cite{Mills:2011p1611} we estimate $P(D)$ as approximately $9x10^{-7}$.
+Each possible spanning alignment indicates an implied insert size $s$ based on the distance between the alignments of the two ends, as well as a mapping quality that indicates $P(m)$, or the probability that the pair alignment is correct.
+Based on $s$, for each possible pair alignment we compute $P(s|D)$ and $P(s|\bar{D})$, or the probability of observing an aligning pair with insert size $s$ in the presence or absence of a deletion in the current window.
+$P(s|{\bar{D}})$ can be estimated using a normal distribution with a mean at the expected fragment size of the read group to which the current pair belongs.
+We estimate $P(s|D)$ using a lognormal distribution fit to the published range of deletion sizes from the 1000 Genomes Project pilot data.
+Using our estimated prior probabilities we then compute the quantities
+
+\[ 
+P(D|s) = \frac{P(s|D)P(D)} {P(s|D)P(D) + P(s|\bar{D})P(\bar{D})}
+\]
+
+\[ 
+P(\bar{D}|s) = \frac{P(s|\bar{D})P(\bar{D})}{P(s|D)P(D) + P(s|\bar{D})P(\bar{D})} 
+\]
+
+We then incorporate the probability that the mapping is correct into our current belief in the presence or absence of a deletion at $l$, reasoning that if the mapping is correct it provides evidence that should change the estimate of $P(D)$ at $l$, and if it is incorrect the estimate of $P(D)$ at $l$ is unchanged.
+The new probabilities $P'(D)$ and $P'(\bar{D})$ for $l$ are after examining a single pair alignment are:
+
+\[ P'(D) = P(m)P(D|s) + (1 - P(m))P(D) \]
+\[ P'(\bar{D}) = P(m)P(\bar{D}|s) + (1 - P(m))P(\bar{D}) \]
+
+We then use this procedure to update $P(D)$ and $P(\bar{D})$ for each alignment spanning $l$, and after all alignments have been considered use the log-likelihood ratio of the two values as a score indicating the presence or absence of a deletion that covers $l$.
+
+To increase the specificity of the algorithm, Cloudbreak can exclude alignments from regions of the genome that are particularly difficult to map.
+For example, in our current results we exclude possible alignments in which one or both of the reads maps to an annotated segmental duplication, or to a region known to have typical sequencing depth in the top 1\% of the distribution of read depths, likely due to mis-assembly of repetitive sequence in the genome reference.
+\cite{Pickrell:2011du}
 
@@
 
-Simulated data
+\subsection{Tests with Simulated Data}
+
+There is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference.
+Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics.
+On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data.
+To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous deletions from the genome of J.
+Craig Venter \cite{Levy:2007fb} and applied them to the human GRCh36 chromosome 2 and 17 reference sequences.
+We then simulated paired Illumina reads from this modified sequence using \emph{dwgsim} from the DNAA software package at varying levels of coverage, and compared results from Cloudbreak with the original list of deletions.
+A similar approach was recently used by \cite{Sindi:2012kk} to evaluate the performance of GASVPro.
+
+We compared the performance of Cloudbreak for detecting deletions to the previously published methods Breakdancer, \cite{Chen:2009p3} Hydra, \cite{Quinlan:2010gf} GASV, \cite{Sindi:2009gu} and DELLY.
+\cite{Rausch:2012he}
 

### 2012-10-09 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/Cloudbreak.java
remove unused deflate command

@@ -73,8 +73,6 @@ public class Cloudbreak extends Configured implements Tool
         CommandVESingleEndDeletionScores veSingleEndDeletionScores = new CommandVESingleEndDeletionScores();
         jc.addCommand("veSingleEndDeletionScores", veSingleEndDeletionScores);
 
-        jc.addCommand("splitDeflatedOutput", new CommandSplitDeflatedOutput());
-
         CommandAverageWigOverSlidingWindow averageWigOverSlidingWindow = new CommandAverageWigOverSlidingWindow();
         jc.addCommand("averageWigOverSlidingWindow", averageWigOverSlidingWindow);
 

### 2012-10-09 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/command/CommandSplitDeflatedOutput.java

@@ -1,52 +1,0 @@
-package edu.ohsu.sonmezsysbio.cloudbreak.command;
-
-import com.beust.jcommander.Parameter;
-import edu.ohsu.sonmezsysbio.cloudbreak.Cloudbreak;
-import org.apache.hadoop.conf.Configuration;
-import org.apache.hadoop.fs.FileSystem;
-import org.apache.hadoop.fs.Path;
-import org.apache.hadoop.mapred.*;
-import org.apache.hadoop.mapred.lib.IdentityMapper;
-import org.apache.hadoop.mapred.lib.IdentityReducer;
-
-/**
- * Created by IntelliJ IDEA.
- * User: cwhelan
- * Date: 2/6/12
- * Time: 10:29 PM
- */
-public class CommandSplitDeflatedOutput implements CloudbreakCommand {
-
-    @Parameter(names = {"--inputHDFSDir"}, required = true)
-    String inputHDFSDir;
-
-    @Parameter(names = {"--outputHDFSDir"}, required = true)
-    String ouptutHDFSDir;
-        
-    @Parameter(names = {"--numSplits"}, required = true)
-    int numSplits;
-
-    public void run(Configuration configuration) throws Exception {
-        JobConf conf = new JobConf(configuration);
-
-        conf.setJobName("Split deflated file");
-        conf.setJarByClass(Cloudbreak.class);
-        FileInputFormat.addInputPath(conf, new Path(inputHDFSDir));
-        Path outputDir = new Path(ouptutHDFSDir);
-        FileSystem.get(conf).delete(outputDir);
-
-        FileOutputFormat.setOutputPath(conf, outputDir);
-
-        conf.setInputFormat(TextInputFormat.class);
-
-        conf.setMapperClass(IdentityMapper.class);
-
-        //conf.setReducerClass(DeletionScorePileupReducer.class);
-        conf.setReducerClass(IdentityReducer.class);
-        conf.setNumReduceTasks(numSplits);
-        conf.setCompressMapOutput(true);
-
-        JobClient.runJob(conf);
-
-    }
-}

### 2012-10-11 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/VotingPairedAlignmentScorer.java
remove old voting scorer

@@ -1,29 +1,0 @@
-package edu.ohsu.sonmezsysbio.cloudbreak;
-
-import org.apache.commons.math3.distribution.NormalDistribution;
-import org.apache.commons.math3.exception.MathIllegalArgumentException;
-
-/**
- * Created by IntelliJ IDEA.
- * User: cwhelan
- * Date: 3/12/12
- * Time: 1:27 PM
- */
-public class VotingPairedAlignmentScorer extends PairedAlignmentScorer {
-    public double computeDeletionScore(int insertSize, Double targetIsize, Double targetIsizeSD, Double pMappingCorrect) {
-        NormalDistribution insertSizeDist = new NormalDistribution(targetIsize, targetIsizeSD);
-
-        double deletionProb;
-        try {
-            deletionProb = insertSizeDist.cumulativeProbability(Math.max(0, insertSize - 1.5 * targetIsizeSD));
-            //System.err.println("Deletion prob: " + deletionProb);
-        } catch (MathIllegalArgumentException e) {
-            e.printStackTrace();
-            throw new RuntimeException(e);
-        }
-
-        double vote = deletionProb > 0.5 ? 1 : -1;
-
-        return vote * Math.exp(pMappingCorrect);
-    }
-}

### 2012-10-11 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/VotingPairedAlignmentScorerTest.java

@@ -1,37 +1,0 @@
-package edu.ohsu.sonmezsysbio.cloudbreak;
-
-import edu.ohsu.sonmezsysbio.cloudbreak.io.NovoalignAlignmentReader;
-import org.junit.Test;
-
-import static org.junit.Assert.assertEquals;
-import static org.junit.Assert.assertTrue;
-
-/**
- * Created by IntelliJ IDEA.
- * User: cwhelan
- * Date: 3/12/12
- * Time: 1:32 PM
- */
-public class VotingPairedAlignmentScorerTest {
-
-    @Test
-    public void testComputeDeletionScore() throws Exception {
-        VotingPairedAlignmentScorer scorer = new VotingPairedAlignmentScorer();
-        // bigger insert size: more likely deletion
-        assertTrue(scorer.computeDeletionScore(5114, 3000.0, 300.0, NovoalignAlignmentReader.probabilityMappingIsCorrect(NovoalignNativeRecord.decodePosterior(185), NovoalignNativeRecord.decodePosterior(117))) >
-                scorer.computeDeletionScore(3114, 3000.0, 300.0, NovoalignAlignmentReader.probabilityMappingIsCorrect(NovoalignNativeRecord.decodePosterior(185), NovoalignNativeRecord.decodePosterior(117))));
-        // higher quality: more likely deletion
-        assertTrue(scorer.computeDeletionScore(3114, 3000.0, 300.0, NovoalignAlignmentReader.probabilityMappingIsCorrect(NovoalignNativeRecord.decodePosterior(185), NovoalignNativeRecord.decodePosterior(117))) <
-                scorer.computeDeletionScore(3114, 3000.0, 300.0, NovoalignAlignmentReader.probabilityMappingIsCorrect(NovoalignNativeRecord.decodePosterior(185), NovoalignNativeRecord.decodePosterior(20))));
-
-        assertEquals(9.999999999999969E-5, scorer.computeDeletionScore(10000,200.0,35.0, NovoalignAlignmentReader.probabilityMappingIsCorrect(NovoalignNativeRecord.decodePosterior(0), NovoalignNativeRecord.decodePosterior(145))), 0.000001);
-
[…]

### 2012-10-11 manuscript/cloudbreak.tex
More background

@@
 
-Hadoop/MapReduce as framework for processing complete data sets
-
-Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping, \cite{Schatz:2009p278} querying variant databases, \cite{Oconnor:2010p1835} SNP calling, \cite{Langmead:2009p1225} RNA-seq differential expression analysis, \cite{Langmead:2010p1268} ChIP-seq peak calling, \cite{Feng:2011p1228} and computing genome mappability.
-\cite{Lee:2012bk}
+Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping, \cite{Schatz:2009p278} querying variant databases, \cite{Oconnor:2010p1835} single nucleotide polymorphism (SNP) calling, \cite{Langmead:2009p1225} RNA-seq differential expression analysis, \cite{Langmead:2010p1268} ChIP-seq peak calling, \cite{Feng:2011p1228} and computing genome mappability.
+\cite{Lee:2012bk} The MapReduce programming style lends itself particularly well towards resequencing tasks such as SNP calling, in which a result can be computed independently for various locations along the genome.
+As mentioned previously, most algorithms for SV calling work by clustering discordant read pairs across the genome, an operation that does not lend itself to easy parallelization.
+Cloudbreak instead independently computes local scores for small contiguous regions along the genome.
+The recently published method forestSV \cite{Michaelson:2012fj} showed that SV detection from computation of local scores is a promising approach, since multiple features can be extracted from reads aligning to or near each region, and integrated with features built from genome annotations in machine learning frameworks.
+This approach also allows the use of a MapReduce programming style and opens the door to cluster computing for SV detection.
 
@@
 Craig Venter \cite{Levy:2007fb} and applied them to the human GRCh36 chromosome 2 and 17 reference sequences.
-We then simulated paired Illumina reads from this modified sequence using \emph{dwgsim} from the DNAA software package at varying levels of coverage, and compared results from Cloudbreak with the original list of deletions.
+We then simulated 50bp paired Illumina reads with a fragment size of 200bp from this modified sequence using \emph{dwgsim} from the DNAA software package \cite{DNAA} at varying levels of coverage, and compared results from Cloudbreak with the original list of deletions.
 A similar approach was recently used by \cite{Sindi:2012kk} to evaluate the performance of GASVPro.

### 2012-10-12 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/GenotypingGMMScorer.java
implementation of GMM scorer

@@ -1,0 +1,10 @@
+package edu.ohsu.sonmezsysbio.cloudbreak.reducer;
+
+/**
+ * Created by IntelliJ IDEA.
+ * User: cwhelan
+ * Date: 10/10/12
+ * Time: 9:17 PM
+ */
+public class GenotypingGMMScorer {
+}

### 2012-10-12 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/GenotypingGMMScorerTest.java

@@ -1,0 +1,16 @@
+package edu.ohsu.sonmezsysbio.cloudbreak.reducer;
+
+import org.junit.Test;
+
+/**
+ * Created by IntelliJ IDEA.
+ * User: cwhelan
+ * Date: 10/12/12
+ * Time: 10:11 AM
+ */
+public class GenotypingGMMScorerTest {
+    @Test
+    public void testNnclean() throws Exception {
+
+    }
+}

### 2012-10-12 src/main/scripts/prototypeEM.R
R prototype of GMM scorer

@@ -1,0 +1,241 @@
+sim.nodel <- function(coverage, insert.mean=200, insert.sd=30, max.insert=25000, num.noise=5) {
+    num.true.reads <- rpois(1, coverage)
+    true.inserts <- rnorm(num.true.reads, insert.mean, insert.sd)
+    noise.inserts <- runif(num.noise, insert.mean, max.insert)
+    c(true.inserts, noise.inserts)
+}
+
+sim.homdel <- function(coverage, del.size, insert.mean=200, insert.sd=30, max.insert=25000, num.noise=5) {
+  num.true.reads <- rpois(1, coverage)
+  true.inserts <- rnorm(num.true.reads, del.size + insert.mean, insert.sd)
+  noise.inserts <- runif(num.noise, insert.mean, max.insert)
+  c(true.inserts, noise.inserts)
+}
+
+sim.hetdel <- function(coverage, del.size, insert.mean=200, insert.sd=30, max.insert=25000, num.noise=5) {
+  num.true.reads <- rpois(1, coverage)
+  nodel.inserts <- rnorm(num.true.reads / 2, insert.mean, insert.sd)
+  del.inserts <- rnorm(num.true.reads / 2, del.size + insert.mean, insert.sd)
+  noise.inserts <- runif(num.noise, insert.mean, max.insert)
+  c(nodel.inserts, del.inserts, noise.inserts)
+}
+
+likelihood <- function(y, w, mu, sigma) {
+  m1.likelihoods <- dnorm(y, mu[1], sigma, log=TRUE)
+  m2.likelihoods <- dnorm(y, mu[2], sigma, log=TRUE)
+  
+  weighted.likelihoods <- exp(w[1]) *  m1.likelihoods + exp(w[2]) * m2.likelihoods
+  mean(weighted.likelihoods)
+}
+
+logsumexp <- function(x) {
+  m <- max(x)
+  s <- m + log(sum(exp(x - m)))
+  s
+}
+
+gamma <- function(y, w, mu, sigma) {
+  m1.likelihood <- w[1] + dnorm(y, mu[1], sigma, log=TRUE)
+  print("m1")
+  print(m1.likelihood)
+  print(exp(m1.likelihood))
+  m2.likelihood <- w[2] + dnorm(y, mu[2], sigma, log=TRUE)
+  print("m2")
+  print(m2.likelihood)
+  print(exp(m2.likelihood))
[…]

### 2012-10-12 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/GenotypingGMMScorer.java
gmm scorer

@@ -1,10 +1,182 @@
 package edu.ohsu.sonmezsysbio.cloudbreak.reducer;
 
+import com.google.common.primitives.Doubles;
+import edu.ohsu.sonmezsysbio.cloudbreak.ReadGroupInfo;
+import edu.ohsu.sonmezsysbio.cloudbreak.io.ReadPairInfo;
+import org.apache.commons.math3.distribution.NormalDistribution;
+
+import java.util.*;
+
 /**
  * Created by IntelliJ IDEA.
  * User: cwhelan
  * Date: 10/10/12
  * Time: 9:17 PM
  */
-public class GenotypingGMMScorer {
+public class GenotypingGMMScorer implements ReadPairInfoScorer {
+
+    private double[] pointLikelihoods(double[] y, NormalDistribution d) {
+        double[] pointLikelihoods = new double[y.length];
+        for (int i = 0; i < y.length; i++) {
+            pointLikelihoods[i] = Math.log(d.density(y[i]));
+        }
+        return pointLikelihoods;
+    }
+
+    private double likelihood(double[] y, double[] w, double[] mu, double sigma) {
+        NormalDistribution d1 = new NormalDistribution(mu[0], sigma);
+        NormalDistribution d2 = new NormalDistribution(mu[1], sigma);
+        double[] d1PointLikelihoods = pointLikelihoods(y, d1);
+        double[] d2PointLikelihoods = pointLikelihoods(y, d2);
+        double sumWeightedLikelihoods = 0;
+        int i = 0;
+        for (i = 0; i < y.length; i++) {
+            sumWeightedLikelihoods += Math.exp(w[0]) * d1PointLikelihoods[i] + Math.exp(w[1]) * d2PointLikelihoods[i];
+        }
+        return sumWeightedLikelihoods / y.length;
+    }
+
+    double logsumexp(double[] x) {
+        double m = Doubles.max(x);
+        double sumexp = 0;
+        for (int i = 0; i < x.length; i++) {
+            sumexp += Math.exp(x[i] - m);
+        }
+        double s = m + Math.log(sumexp);
+        return s;
+    }
+
+    private double[][] gamma(double[] y, double[] w, double[] mu, double sigma) {
+        NormalDistribution d1 = new NormalDistribution(mu[0], sigma);
+        NormalDistribution d2 = new NormalDistribution(mu[1], sigma);
+        double[] m1likelihoods = new double[y.length];
[…]

### 2012-10-12 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/ReadPairInfoToDeletionScoreReducer.java

@@ -26,7 +26,7 @@ public class ReadPairInfoToDeletionScoreReducer extends MapReduceBase implements
 
     //{ log.setLevel(Level.DEBUG); }
 
-    ReadPairInfoScorer readPairInfoScorer = new IncrementalDelBeliefUpdateReadPairInfoScorer();
+    ReadPairInfoScorer readPairInfoScorer = new GenotypingGMMScorer();
 
     public ReadPairInfoScorer getReadPairInfoScorer() {
         return readPairInfoScorer;

### 2012-10-12 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/GenotypingGMMScorerTest.java

@@ -1,6 +1,8 @@
 package edu.ohsu.sonmezsysbio.cloudbreak.reducer;
 
 import org.junit.Test;
+import static org.junit.Assert.assertArrayEquals;
+import static org.junit.Assert.assertEquals;
 
 /**
  * Created by IntelliJ IDEA.
@@ -8,9 +10,53 @@ import org.junit.Test;
  * Date: 10/12/12
  * Time: 10:11 AM
  */
+
+/**
+ * These tests are mostly direct ports from my R prototype
+ */
 public class GenotypingGMMScorerTest {
+
+    @Test
+    public void testLogsumexp() throws Exception {
+        double[] x = new double[] {-7.018195, -309.17497};
+        GenotypingGMMScorer scorer = new GenotypingGMMScorer();
+        assertEquals(-7.018195, scorer.logsumexp(x), 0.00001);
+    }
+
     @Test
     public void testNnclean() throws Exception {
-
+        double[] y = new double[] {260.0736, 197.4272,   194.8618,  1217.8588,
+                1228.2190,  1151.7017,  4511.5326, 19719.9700, 19707.2091, 16788.1891,
+                22556.1203,  9909.8687, 13709.7259,  9219.4829, 13076.1122,  8140.4713};
+        GenotypingGMMScorer scorer = new GenotypingGMMScorer();
+        double[] nonnoise = new double[] {260.0736, 197.4272,   194.8618,  1217.8588,
+                1228.2190,  1151.7017};
+        assertArrayEquals(nonnoise, scorer.nnclean(y, 30, 2), 0.000001);
     }
+
+    @Test
+    public void testEstimateW() throws Exception {
+        double[] y = new double[] {260.0736, 197.4272,   194.8618,  1217.8588,
+                1228.2190,  1151.7017,  4511.5326, 19719.9700, 19707.2091, 16788.1891,
+                22556.1203,  9909.8687, 13709.7259,  9219.4829, 13076.1122,  8140.4713};
+
+        double sigma = 30;
+        double[] initialW = new double[] {Math.log(.5),Math.log(.5)};
+        double[] initialMu = new double[] {200, 1000};
+
+        GenotypingGMMScorer scorer = new GenotypingGMMScorer();
+        assertEquals(.5, Math.exp(scorer.estimateW(y, initialW, initialMu, sigma)), 0.0001);
+    }
+
+    public void testEstimateWNoValues() throws Exception {
[…]

### 2012-10-13 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/GenotypingGMMScorer.java
oops, set initial mus

@@ -172,7 +172,7 @@ public class GenotypingGMMScorer implements ReadPairInfoScorer {
             }
         }
         double[] initialW = new double[]{.5,.5};
-        double[] initialMu = new double[]{0,};
+        double[] initialMu = new double[]{0,1000};
         double[] insertSizeArray = new double[insertSizes.size()];
         for (int i = 0; i < insertSizes.size(); i++) {
             insertSizeArray[i] = insertSizes.get(i);

### 2012-10-13 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/GenotypingGMMScorer.java
limit coverage

@@ -171,6 +171,9 @@ public class GenotypingGMMScorer implements ReadPairInfoScorer {
                 maxSD = readGroupInfo.isizeSD;
             }
         }
+        if (insertSizes.size() >= 100) {
+            return -1;
+        }
         double[] initialW = new double[]{.5,.5};
         double[] initialMu = new double[]{0,1000};
         double[] insertSizeArray = new double[insertSizes.size()];

### 2012-10-26 manuscript/cloudbreak.tex
Add new ref

@@
 
-Genomic structural variation
+Genomic structural variation.. as much or more variance as SNPS.. INDELS in exons, coding regions (Mills 2011)...
 
-Difficulty of detection.. Lack of concordance of current algorithms...
+Difficulty of detection.. Lack of concordance of current algorithms (described in 1KG SV paper)...
 
-Typical algorithms use only unambiguous discordant read pair mappings...
+Typical algorithms use only unambiguous discordant read pair mappings... New algos combine RP and RD or SP signals...
 
@@
 \cite{Lee:2012bk} The MapReduce programming style lends itself particularly well towards resequencing tasks such as SNP calling, in which a result can be computed independently for various locations along the genome.
-As mentioned previously, most algorithms for SV calling work by clustering discordant read pairs across the genome, an operation that does not lend itself to easy parallelization.
+As mentioned previously, most algorithms for SV and INDEL calling work by clustering discordant read pairs across the genome, an operation that does not lend itself to easy parallelization.
 Cloudbreak instead independently computes local scores for small contiguous regions along the genome.
@@
 
-\section{Methods}\label{Methods}
+\section{Results}\label{Results}
 
-\subsection{A Hadoop/MapReduce Algorithm for SV Detection}
+\subsection{A Hadoop/MapReduce Algorithm for INDEL Detection}
 
@@
 The first job finds possible alignment locations for pairs of reads in FASTQ format.
-The second job computes local SV scores for a set of small windows completely covering the target genome.
-Finally, the SV scores are exported from the Hadoop file system and post-processed to produce a final set of SV calls which span multiple genomic windows.
+The second job computes local deletion scores for a set of small windows completely covering the target genome.
+Finally, the scores are exported from the Hadoop file system and post-processed to produce a final set of variant calls which span multiple genomic windows.
 The Cloudbreak MapReduce algorithm is summarized in \ref{algo1}.
@@
 \cite{novoalign} Novoalign exhibits higher sensitivity for mapping locations than most aligners, even in the presence of indels and errors, at the expense of greater runtime.
-\cite{Krawitz:2010iq} In addition, Novoalign can report multiple possible mapping locations for a read and has accurate mapping quality scores \cite{Ruffalo:2011p1758}, which are useful in assigning confidence scores to SV calls.
+\cite{Krawitz:2010iq} In addition, Novoalign can report multiple possible mapping locations for a read and has accurate mapping quality scores \cite{Ruffalo:2011p1758}, which are useful in assigning confidence scores to variant calls.
 The current implementation is also capable of using MrFast \cite{Alkan:2009cr} as an aligner, although MrFast does not provide mapping quality scores, which makes Cloudbreak predictions less accurate.
@@
 In the second job, mappers collect information, namely the insert size and the pair mapping quality, about all of the possible alignments of read pairs that span each genomic locus, defined as a small (25bp) window in the genome reference.
-Reducers then integrate all of the read pair information for reads that span a given location to produce an SV score indicating the likelihood of an SV affecting that locus.
+Reducers then integrate all of the read pair information for reads that span a given location to produce a score indicating the likelihood of a deletion affecting that locus.
 Since scores are computed independently at each locus, this procedure is essentially constructing and analyzing a pileup of reads pairs at each location, making it similar to previous bioinformatic MapReduce algorithms for SNP calling implemented in Crossbow \cite{Langmead:2009p1225} and GATK.
@@
 
-As described in the previous section, reducers in the second Cloudbreak job compute an SV score indicating the presence or absence of a deletion for each genomic location $l$.
+As described in the previous section, reducers in the second Cloudbreak job compute a score indicating the presence or absence of a deletion for each genomic location $l$.
 Locations are defined as contiguous windows of 25bp covering the entire genome.
@@
 
-\section{Results}\label{results}
-
 \subsection{Tests with Simulated Data}
@@
 Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics.
-On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data.
+On the other hand, it is important that the simulated data contain realistic SVs that follow patterns observed in real data.
 To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous deletions from the genome of J.
@@
 
+\section{Methods}\label{methods}
+
 \section{Discussion}\label{discussion}

### 2012-10-26 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/WeightedBeliefDeltaScorer.java
turn off logging

@@ -22,7 +22,7 @@ public class WeightedBeliefDeltaScorer implements ReadPairInfoScorer {
     private static org.apache.log4j.Logger log = Logger
             .getLogger(WeightedBeliefDeltaScorer.class);
 
-    { log.setLevel(Level.DEBUG); }
+    // { log.setLevel(Level.DEBUG); }
 
     public double reduceReadPairInfos(Iterator<ReadPairInfo> values, Map<Short, ReadGroupInfo> readGroupInfos) {
         LogNormalDistribution logNormalDistribution = new LogNormalDistribution(6, 0.6);

### 2012-10-26 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/PairedAlignmentScorer.java
turn off logging

@@ -13,7 +13,7 @@ public abstract class PairedAlignmentScorer {
 
     private static org.apache.log4j.Logger logger = Logger.getLogger(PairedAlignmentScorer.class);
 
-    { logger.setLevel(Level.DEBUG); }
+    //{ logger.setLevel(Level.DEBUG); }
 
     public boolean validateInsertSize(int insertSize, String readPairId, Integer maxInsertSize1) {
         if (insertSize == 0) return false;

### 2012-10-26 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapper.java

@@ -32,7 +32,7 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
 
     private static org.apache.log4j.Logger logger = Logger.getLogger(SingleEndAlignmentsToReadPairInfoMapper.class);
 
-    { logger.setLevel(Level.DEBUG); }
+    //{ logger.setLevel(Level.DEBUG); }
 
     private boolean matePairs;
     private Integer maxInsertSize = Cloudbreak.DEFAULT_MAX_INSERT_SIZE;

### 2012-10-26 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/WeightedBeliefDeltaScorer.java

@@ -22,7 +22,7 @@ public class WeightedBeliefDeltaScorer implements ReadPairInfoScorer {
     private static org.apache.log4j.Logger log = Logger
             .getLogger(WeightedBeliefDeltaScorer.class);
 
-    { log.setLevel(Level.DEBUG); }
+    //{ log.setLevel(Level.DEBUG); }
 
     public double reduceReadPairInfos(Iterator<ReadPairInfo> values, Map<Short, ReadGroupInfo> readGroupInfos) {
         LogNormalDistribution logNormalDistribution = new LogNormalDistribution(6, 0.6);

### 2012-10-27 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapper.java
trying to debug

@@ -32,7 +32,7 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
 
     private static org.apache.log4j.Logger logger = Logger.getLogger(SingleEndAlignmentsToReadPairInfoMapper.class);
 
-    //{ logger.setLevel(Level.DEBUG); }
+    { logger.setLevel(Level.INFO); }
 
     private boolean matePairs;
     private Integer maxInsertSize = Cloudbreak.DEFAULT_MAX_INSERT_SIZE;
@@ -124,6 +124,9 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
 
     public void map(Text key, Text value, OutputCollector<GenomicLocationWithQuality, ReadPairInfo> output, Reporter reporter) throws IOException {
         String line = value.toString();
+        if (line.contains("1bf17d")) {
+            logger.info("got target line: " + line);
+        }
         ReadPairAlignments readPairAlignments = parsePairAlignmentLine(line);
         alignmentReader.resetForReadPairAlignemnts(readPairAlignments);
 
@@ -245,7 +248,7 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
                 }
             }
 
-            logger.debug("Emitting insert size " + insertSize);
+            logger.info("Emitting insert size " + insertSize);
             GenomicLocationWithQuality genomicLocation = new GenomicLocationWithQuality(chromosome, pos, readPairInfo.pMappingCorrect);
             output.collect(genomicLocation, readPairInfo);
 

### 2012-10-27 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapperTest.java

@@ -21,6 +21,7 @@ import java.util.ArrayList;
 import java.util.List;
 
 import static org.junit.Assert.assertEquals;
+import static org.junit.Assert.assertTrue;
 
 /**
  * Created by IntelliJ IDEA.
@@ -130,7 +131,7 @@ public class SingleEndAlignmentsToReadPairInfoMapperTest {
         assertEquals(101, collector.keys.size());
     }
 
-        private static class MockOutputCollector implements OutputCollector<GenomicLocationWithQuality, ReadPairInfo> {
+    private static class MockOutputCollector implements OutputCollector<GenomicLocationWithQuality, ReadPairInfo> {
 
         List<GenomicLocationWithQuality> keys = new ArrayList<GenomicLocationWithQuality>();
         List<ReadPairInfo> values = new ArrayList<ReadPairInfo>();
@@ -144,6 +145,30 @@ public class SingleEndAlignmentsToReadPairInfoMapperTest {
             keys.clear();
             values.clear();
         }
+    }
+
+    @Test
+    public void testMap_real1() throws Exception {
+        String key = "@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bf17d/";
+        String val = "@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bf17d//1\tS\tTAAAAAGCCGCGGCGACTAAAAGCCGCTGAGAGGGGGCAAAAAGCAGCGG\t66554410000////1.0000/----,/,.,.,,++++-----+*-****\tU\t25\t139.09267\t>2\t132512583\tF\t.\t.\t.\t18A>T\tSVP_READ\t@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bf17d//2\tS\tCCCCTGCCCCGCCGCGGCTTTTTGCGGCTTTCCGCCCCGGCCGCCGCGGA\t33324110000////...000//---,,...,,,++++++++++*****,\tU\t23\t135.68983\t>2\t132512814\tR\t.\t.\t.\t1G>T";
+
+        mapper.setFaix(new FaidxFileHelper("foo") {
+            @Override
+            public Short getKeyForChromName(String name) throws IOException {
+                assertEquals("2", name);
+                return (short) 0;
+            }
+        });
+
+        MockOutputCollector mockOutputCollector = new MockOutputCollector();
+        mapper.map(new Text(key), new Text(val), mockOutputCollector, null);
+
[…]

### 2012-10-27 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapper.java
trying to debug

@@ -126,6 +126,12 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
         String line = value.toString();
         if (line.contains("1bf17d")) {
             logger.info("got target line: " + line);
+            logger.setLevel(Level.DEBUG);
+            GenomicLocationWithQuality l = new GenomicLocationWithQuality((short) 0, 1, 0.0);
+            ReadPairInfo rpi = new ReadPairInfo(100, 0, (short) 0);
+            for (int i = 0; i < 100; i++) {
+                output.collect(l, rpi);
+            }
         }
         ReadPairAlignments readPairAlignments = parsePairAlignmentLine(line);
         alignmentReader.resetForReadPairAlignemnts(readPairAlignments);

### 2012-10-27 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapperTest.java

@@ -169,6 +169,7 @@ public class SingleEndAlignmentsToReadPairInfoMapperTest {
         gl132512700.pMappingCorrect = -3.9301895071730983E-14;
 
         assertTrue(mockOutputCollector.keys.contains(gl132512700));
+        assertEquals(281, mockOutputCollector.values.get(0).insertSize);
     }
 
     @Test

### 2012-10-27 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapperTest.java
trying to debug

@@ -149,8 +149,8 @@ public class SingleEndAlignmentsToReadPairInfoMapperTest {
 
     @Test
     public void testMap_real1() throws Exception {
-        String key = "@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bf17d/";
-        String val = "@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bf17d//1\tS\tTAAAAAGCCGCGGCGACTAAAAGCCGCTGAGAGGGGGCAAAAAGCAGCGG\t66554410000////1.0000/----,/,.,.,,++++-----+*-****\tU\t25\t139.09267\t>2\t132512583\tF\t.\t.\t.\t18A>T\tSVP_READ\t@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bf17d//2\tS\tCCCCTGCCCCGCCGCGGCTTTTTGCGGCTTTCCGCCCCGGCCGCCGCGGA\t33324110000////...000//---,,...,,,++++++++++*****,\tU\t23\t135.68983\t>2\t132512814\tR\t.\t.\t.\t1G>T";
+        String key = "@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bfnjd/";
+        String val = "@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bfnjd//1\tS\tTAAAAAGCCGCGGCGACTAAAAGCCGCTGAGAGGGGGCAAAAAGCAGCGG\t66554410000////1.0000/----,/,.,.,,++++-----+*-****\tU\t25\t139.09267\t>2\t132512583\tF\t.\t.\t.\t18A>T\tSVP_READ\t@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bfnjd//2\tS\tCCCCTGCCCCGCCGCGGCTTTTTGCGGCTTTCCGCCCCGGCCGCCGCGGA\t33324110000////...000//---,,...,,,++++++++++*****,\tU\t23\t135.68983\t>2\t132512814\tR\t.\t.\t.\t1G>T";
 
         mapper.setFaix(new FaidxFileHelper("foo") {
             @Override

### 2012-10-27 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapper.java
trying to debug

@@ -124,6 +124,7 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
 
     public void map(Text key, Text value, OutputCollector<GenomicLocationWithQuality, ReadPairInfo> output, Reporter reporter) throws IOException {
         String line = value.toString();
+        logger.setLevel(Level.INFO);
         if (line.contains("1bf17d")) {
             logger.info("got target line: " + line);
             logger.setLevel(Level.DEBUG);

### 2012-10-27 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/NovoalignNativeRecord.java
trying to debug

@@ -119,4 +119,16 @@ public class NovoalignNativeRecord implements AlignmentRecord {
         return codedPosterior == 0 ? 0.0001 : 1 - Math.pow(10.0, codedPosterior / -10.0);
     }
 
+    @Override
+    public String toString() {
+        return "NovoalignNativeRecord{" +
+                "chromosomeName='" + chromosomeName + '\'' +
+                ", position=" + position +
+                ", mappingStatus='" + mappingStatus + '\'' +
+                ", posteriorProb=" + posteriorProb +
+                ", forward=" + forward +
+                ", readId='" + readId + '\'' +
+                ", sequence='" + sequence + '\'' +
+                '}';
+    }
 }

### 2012-10-27 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapper.java

@@ -125,8 +125,10 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
     public void map(Text key, Text value, OutputCollector<GenomicLocationWithQuality, ReadPairInfo> output, Reporter reporter) throws IOException {
         String line = value.toString();
         logger.setLevel(Level.INFO);
+        boolean targetDebug = false;
         if (line.contains("1bf17d")) {
             logger.info("got target line: " + line);
+            targetDebug = true;
             logger.setLevel(Level.DEBUG);
             GenomicLocationWithQuality l = new GenomicLocationWithQuality((short) 0, 1, 0.0);
             ReadPairInfo rpi = new ReadPairInfo(100, 0, (short) 0);
@@ -135,6 +137,17 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
             }
         }
         ReadPairAlignments readPairAlignments = parsePairAlignmentLine(line);
+        if (targetDebug) {
+            logger.info("read 1 alignments");
+            for (AlignmentRecord a : readPairAlignments.getRead1Alignments()) {
+                logger.info(a);
+            }
+            logger.info("read 2 alignments");
+            for (AlignmentRecord a : readPairAlignments.getRead2Alignments()) {
+                logger.info(a);
+            }
+
+        }
         alignmentReader.resetForReadPairAlignemnts(readPairAlignments);
 
         Set<AlignmentRecord> recordsInExcludedAreas = new HashSet<AlignmentRecord>();

### 2012-10-27 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapperTest.java

@@ -151,6 +151,7 @@ public class SingleEndAlignmentsToReadPairInfoMapperTest {
     public void testMap_real1() throws Exception {
         String key = "@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bfnjd/";
         String val = "@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bfnjd//1\tS\tTAAAAAGCCGCGGCGACTAAAAGCCGCTGAGAGGGGGCAAAAAGCAGCGG\t66554410000////1.0000/----,/,.,.,,++++-----+*-****\tU\t25\t139.09267\t>2\t132512583\tF\t.\t.\t.\t18A>T\tSVP_READ\t@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bfnjd//2\tS\tCCCCTGCCCCGCCGCGGCTTTTTGCGGCTTTCCGCCCCGGCCGCCGCGGA\t33324110000////...000//---,,...,,,++++++++++*****,\tU\t23\t135.68983\t>2\t132512814\tR\t.\t.\t.\t1G>T";
+        String va2 = "@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bf17d//1\tS\tTAAAAAGCCGCGGCGACTAAAAGCCGCTGAGAGGGGGCAAAAAGCAGCGG\t66554410000////1.0000/----,/,.,.,,++++-----+*-****\tU\t25\t139.09267\t>2\t132512583\tF\t.\t.\t.\t18A>T\tSVP_READ\t@2_132385096_132385222_0_1_0_0_1:0:0_1:0:0_1bf17d//2\tS\tCCCCTGCCCCGCCGCGGCTTTTTGCGGCTTTCCGCCCCGGCCGCCGCGGA\t33324110000////...000//---,,...,,,++++++++++*****,\t23\t135.68983\t>2\t132512814\tR\t.\t.\t.\t1G>T";
 
         mapper.setFaix(new FaidxFileHelper("foo") {
             @Override

### 2012-10-27 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapper.java
trying to debug

@@ -156,6 +156,9 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
                 for (AlignmentRecord record : readPairAlignments.getRead1Alignments()) {
                     if (exclusionRegions.doesLocationOverlap(record.getChromosomeName(), record.getPosition(), record.getPosition() + record.getSequenceLength())) {
                         logger.debug("excluding record " + record);
+                        if (targetDebug) {
+                            logger.info("excluding record " + record);
+                        }
                         recordsInExcludedAreas.add(record);
                     }
                 }
@@ -163,6 +166,9 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
                 for (AlignmentRecord record : readPairAlignments.getRead2Alignments()) {
                     if (exclusionRegions.doesLocationOverlap(record.getChromosomeName(), record.getPosition(), record.getPosition() + record.getSequenceLength())) {
                         logger.debug("excluding record " + record);
+                        if (targetDebug) {
+                            logger.info("excluding record " + record);
+                        }
                         recordsInExcludedAreas.add(record);
                     }
                 }
@@ -174,34 +180,44 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
         }
 
         try {
-            emitReadPairInfoForAllPairs(readPairAlignments, output, recordsInExcludedAreas);
+            emitReadPairInfoForAllPairs(readPairAlignments, output, recordsInExcludedAreas, targetDebug);
         } catch (Exception e) {
             e.printStackTrace();
             throw new RuntimeException(e);
         }
     }
 
-    private void emitReadPairInfoForAllPairs(ReadPairAlignments readPairAlignments, OutputCollector<GenomicLocationWithQuality, ReadPairInfo> output, Set<AlignmentRecord> recordsInExcludedAreas) throws Exception {
+    private void emitReadPairInfoForAllPairs(ReadPairAlignments readPairAlignments, OutputCollector<GenomicLocationWithQuality, ReadPairInfo> output, Set<AlignmentRecord> recordsInExcludedAreas, boolean targetDebug) throws Exception {
         for (AlignmentRecord record1 : readPairAlignments.getRead1Alignments()) {
             for (AlignmentRecord record2 : readPairAlignments.getRead2Alignments()) {
[…]

### 2012-10-27 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapper.java
trying to debug

@@ -285,12 +285,19 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
             }
 
             int pos = genomeOffset + i;
+            if (targetDebug) {
+                logger.info("pos: "+ pos);
+            }
 
             if (getChromosomeFilter() != null) {
                 if (! record1.getChromosomeName().equals(getChromosomeFilter()) ||
                     pos < getStartFilter() || pos > getEndFilter()) {
                     if (targetDebug) {
                         logger.info("failed chrom,start,end filter");
+                        logger.info("chr filter: " + getChromosomeFilter());
+                        logger.info(" record1.getChromosomeName().equals(getChromosomeFilter()): " + record1.getChromosomeName().equals(getChromosomeFilter()));
+                        logger.info("startFilter: " + getStartFilter());
+                        logger.info("endFilter: " + getEndFilter());
                     }
                     return;
                 }

### 2012-10-27 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapperTest.java

@@ -163,6 +163,9 @@ public class SingleEndAlignmentsToReadPairInfoMapperTest {
 
         MockOutputCollector mockOutputCollector = new MockOutputCollector();
         mapper.map(new Text(key), new Text(val), mockOutputCollector, null);
+        mapper.setChromosomeFilter("2");
+        mapper.setStartFilter(132512600l);
+        mapper.setEndFilter(132512800l);
 
         GenomicLocationWithQuality gl132512700 = new GenomicLocationWithQuality();
         gl132512700.chromosome = 0;

### 2012-10-27 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapper.java
finally fix bug with chr filter?

@@ -299,7 +299,7 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
                         logger.info("startFilter: " + getStartFilter());
                         logger.info("endFilter: " + getEndFilter());
                     }
-                    return;
+                    continue;
                 }
             }
 

### 2012-10-27 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/SingleEndAlignmentsToReadPairInfoMapper.java
remove extra logging

@@ -124,30 +124,7 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
 
     public void map(Text key, Text value, OutputCollector<GenomicLocationWithQuality, ReadPairInfo> output, Reporter reporter) throws IOException {
         String line = value.toString();
-        logger.setLevel(Level.INFO);
-        boolean targetDebug = false;
-        if (line.contains("1bf17d")) {
-            logger.info("got target line: " + line);
-            targetDebug = true;
-            logger.setLevel(Level.DEBUG);
-            GenomicLocationWithQuality l = new GenomicLocationWithQuality((short) 0, 1, 0.0);
-            ReadPairInfo rpi = new ReadPairInfo(100, 0, (short) 0);
-            for (int i = 0; i < 100; i++) {
-                output.collect(l, rpi);
-            }
-        }
         ReadPairAlignments readPairAlignments = parsePairAlignmentLine(line);
-        if (targetDebug) {
-            logger.info("read 1 alignments");
-            for (AlignmentRecord a : readPairAlignments.getRead1Alignments()) {
-                logger.info(a);
-            }
-            logger.info("read 2 alignments");
-            for (AlignmentRecord a : readPairAlignments.getRead2Alignments()) {
-                logger.info(a);
-            }
-
-        }
         alignmentReader.resetForReadPairAlignemnts(readPairAlignments);
 
         Set<AlignmentRecord> recordsInExcludedAreas = new HashSet<AlignmentRecord>();
@@ -156,9 +133,6 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
                 for (AlignmentRecord record : readPairAlignments.getRead1Alignments()) {
                     if (exclusionRegions.doesLocationOverlap(record.getChromosomeName(), record.getPosition(), record.getPosition() + record.getSequenceLength())) {
                         logger.debug("excluding record " + record);
-                        if (targetDebug) {
-                            logger.info("excluding record " + record);
-                        }
                         recordsInExcludedAreas.add(record);
                     }
                 }
@@ -166,9 +140,6 @@ public class SingleEndAlignmentsToReadPairInfoMapper extends CloudbreakMapReduce
                 for (AlignmentRecord record : readPairAlignments.getRead2Alignments()) {
                     if (exclusionRegions.doesLocationOverlap(record.getChromosomeName(), record.getPosition(), record.getPosition() + record.getSequenceLength())) {
[…]

### 2012-10-29 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/file/WigFileHelper.java
Error reporting

@@ -125,6 +125,9 @@ public class WigFileHelper {
 
             } else {
                 String[] fields = line.split("\t");
+                if (fields.length < 2) {
+                    throw new RuntimeException("Failed to parse line: " + line);
+                }
                 long pos = Long.valueOf(fields[0]);
                 if (pos > faidx.getLengthForChromName(currentChromosome)) continue;
                 double val = Double.valueOf(fields[1]);

### 2012-10-29 src/main/scripts/runTest.sh

@@ -115,11 +115,17 @@ echo gzip *.wig
 gzip *.wig
 
 ~/software/IGVTools/igvtools tile ${NAME}_w0.wig.gz ${NAME}_w0.wig.gz.tdf hg18
+
 ~/software/IGVTools/igvtools tile ${NAME}_mu1.wig.gz ${NAME}_mu1.wig.gz.tdf hg18
+
 ~/software/IGVTools/igvtools tile ${NAME}_l1.wig.gz ${NAME}_l1.wig.gz.tdf hg18
+
 ~/software/IGVTools/igvtools tile ${NAME}_l2.wig.gz ${NAME}_l2.wig.gz.tdf hg18
+
 ~/software/IGVTools/igvtools tile ${NAME}_l1f.wig.gz ${NAME}_l1f.wig.gz.tdf hg18
+
 ~/software/IGVTools/igvtools tile ${NAME}_lrHet.wig.gz ${NAME}_lrHet.wig.gz.tdf hg18
+
 ~/software/IGVTools/igvtools tile ${NAME}_lrHom.wig.gz ${NAME}_lrHom.wig.gz.tdf hg18
 
 

### 2012-10-29 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/command/CommandExportGMMResults.java
Close file writers

@@ -88,10 +88,19 @@ public class CommandExportGMMResults implements CloudbreakCommand {
         BufferedWriter lrHomOutputFileWriter = createWriter(lrHomfileName);
         if (lrHomOutputFileWriter == null) return;
 
-        writeGMMResultWigFiles(conf, w0outputFileWriter, mu1outputFileWriter, l1outputFileWriter,
-                l2outputFileWriter, l1foutputFileWriter, lrHetOutputFileWriter, lrHomOutputFileWriter,
-                inputHDFSDir, faidx);
-        w0outputFileWriter.close();
+        try {
+            writeGMMResultWigFiles(conf, w0outputFileWriter, mu1outputFileWriter, l1outputFileWriter,
+                    l2outputFileWriter, l1foutputFileWriter, lrHetOutputFileWriter, lrHomOutputFileWriter,
+                    inputHDFSDir, faidx);
+        } finally {
+            w0outputFileWriter.close();
+            mu1outputFileWriter.close();
+            l1outputFileWriter.close();
+            l2outputFileWriter.close();
+            lrHetOutputFileWriter.close();
+            lrHomOutputFileWriter.close();
+        }
+
 
     }
 

### 2012-10-29 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/command/CommandGMMFitSingleEndInsertSizes.java
make max mapq diff a parameter

@@ -70,6 +70,9 @@ public class CommandGMMFitSingleEndInsertSizes extends BaseCloudbreakCommand {
     @Parameter(names = {"--aligner"})
     String aligner = Cloudbreak.ALIGNER_NOVOALIGN;
 
+    @Parameter(names = {"--maxLogMapqDiff"})
+    Double maxLogMapqDiff = 5.0;
+
     public void run(Configuration conf) throws IOException, URISyntaxException {
         runHadoopJob(conf);
     }
@@ -115,6 +118,8 @@ public class CommandGMMFitSingleEndInsertSizes extends BaseCloudbreakCommand {
         conf.set("cloudbreak.aligner", aligner);
 
         conf.set("pileupDeletionScore.maxInsertSize", String.valueOf(maxInsertSize));
+
+        conf.set("max.log.mapq.diff", String.valueOf(maxLogMapqDiff));
 
         if (chrFilter != null) {
             conf.set("alignments.filterchr", chrFilter);

### 2012-10-29 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/GenotypingGMMScorer.java

@@ -21,7 +21,15 @@ public class GenotypingGMMScorer {
 
     private static org.apache.log4j.Logger log = Logger.getLogger(GenotypingGMMScorer.class);
     public static final int MAX_COVERAGE = 200;
-    public static final int MAX_LOG_MAPQ_DIFF = 5;
+    private double maxLogMapqDiff;
+
+    public void setMaxLogMapqDiff(double maxLogMapqDiff) {
+        this.maxLogMapqDiff = maxLogMapqDiff;
+    }
+
+    public double getMaxLogMapqDiff() {
+        return maxLogMapqDiff;
+    }
 
     { log.setLevel(Level.INFO); }
 
@@ -276,7 +284,7 @@ public class GenotypingGMMScorer {
         double bestMappingQuality = 0;
         while (values.hasNext()) {
             ReadPairInfo rpi = values.next();
-            if (! first & (bestMappingQuality - rpi.pMappingCorrect > MAX_LOG_MAPQ_DIFF)) {
+            if (! first & (bestMappingQuality - rpi.pMappingCorrect > getMaxLogMapqDiff())) {
                 break;
             }
             int insertSize = rpi.insertSize;

### 2012-10-29 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/ReadPairInfoToGMMResultsReducer.java

@@ -64,6 +64,7 @@ public class ReadPairInfoToGMMResultsReducer extends MapReduceBase implements Re
             e.printStackTrace();
         }
 
+        readPairInfoScorer.setMaxLogMapqDiff(Double.parseDouble(job.get("max.log.mapq.diff")));
     }
 }
 

### 2012-10-29 src/main/scripts/runTest.sh

@@ -31,6 +31,7 @@ MAX_INSERT=$7
 TRUTH=$8
 THRESHOLD_MIN=$9
 SHORT_NAME=${10}
+MAX_MAPQ_DIFF=${11}
 
 pushd $BUILD_DIR
 git pull
@@ -39,7 +40,7 @@ mvn assembly:assembly
 SHORT_GIT_TAG=`git rev-parse --short HEAD`
 popd
 
-NAME=${SHORT_NAME}_`basename $READ_GROUP_FILE | awk -F'.' '{print $1}'`_`basename $MAPABILITY | awk -F'.' '{print $1}'`_`basename $FILTER | awk -F'.' '{print $1}'`_${MAX_INSERT}_${RESOLUTION}_${MEDIAN_FILTER_WINDOW}_${ALIGNER}_${SHORT_GIT_TAG}
+NAME=${SHORT_NAME}_`basename $READ_GROUP_FILE | awk -F'.' '{print $1}'`_`basename $MAPABILITY | awk -F'.' '{print $1}'`_`basename $FILTER | awk -F'.' '{print $1}'`_${MAX_INSERT}_${RESOLUTION}_${MEDIAN_FILTER_WINDOW}_${ALIGNER}_${MAX_MAPQ_DIFF}_${SHORT_GIT_TAG}
 
 echo Experiment name: $NAME
 
@@ -70,6 +71,7 @@ MAX_INSERT=$MAX_INSERT
 TRUTH=$TRUTH
 THRESHOLD_MIN=$THRESHOLD_MIN
 SHORT_NAME=$SHORT_NAME
+MAX_MAPQ_DIFF=$MAX_MAPQ_DIFF
 
 EOF
 
@@ -84,6 +86,7 @@ hadoop jar $BUILD_DIR/target/cloudbreak-1.0-SNAPSHOT-exe.jar -Dmapred.reduce.tas
    --excludePairsMappingIn $FILTER 
    --resolution $RESOLUTION 
    --aligner $ALIGNER
+   --maxLogMapqDiff $MAX_MAPQ_DIFF
 
 EOF
 
@@ -95,7 +98,8 @@ hadoop jar $BUILD_DIR/target/cloudbreak-1.0-SNAPSHOT-exe.jar -Dmapred.reduce.tas
     --mapabilityWeighting $MAPABILITY \
     --excludePairsMappingIn $FILTER \
     --resolution $RESOLUTION \
[…]

### 2012-10-29 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/reducer/GenotypingGMMScorerTest.java

@@ -2,6 +2,7 @@ package edu.ohsu.sonmezsysbio.cloudbreak.reducer;
 
 import edu.ohsu.sonmezsysbio.cloudbreak.ReadGroupInfo;
 import edu.ohsu.sonmezsysbio.cloudbreak.io.ReadPairInfo;
+import org.junit.Before;
 import org.junit.Test;
 
 import java.util.ArrayList;
@@ -37,17 +38,22 @@ public class GenotypingGMMScorerTest {
             1229.633, 1184.879, 1230.889, 1169.125, 1200.363, 1262.897, 1155.817, 1164.294
     };
 
+    private GenotypingGMMScorer scorer;
+
+    @Before
+    public void setup() {
+        scorer = new GenotypingGMMScorer();
+    }
+
     @Test
     public void testLogsumexp() throws Exception {
         double[] x = new double[] {-7.018195, -309.17497};
-        GenotypingGMMScorer scorer = new GenotypingGMMScorer();
         assertEquals(-7.018195, scorer.logsumexp(x), 0.00001);
     }
 
     @Test
     public void testNnclean() throws Exception {
         double[] y = HET_DEL_5X_5N;
-        GenotypingGMMScorer scorer = new GenotypingGMMScorer();
         double[] nonnoise = new double[] {260.0736, 197.4272,   194.8618,  1217.8588,
                 1228.2190,  1151.7017};
         assertArrayEquals(nonnoise, scorer.nnclean(y, 30, 2), 0.000001);
@@ -56,7 +62,6 @@ public class GenotypingGMMScorerTest {
     @Test
     public void testNncleanNotEnoughValues() throws Exception {
         double[] y = new double[] {260.0736, 1217.8588};
-        GenotypingGMMScorer scorer = new GenotypingGMMScorer();
         double[] nonnoise = new double[] {};
         assertArrayEquals(nonnoise, scorer.nnclean(y, 30, 2), 0.000001);
     }
@@ -68,7 +73,6 @@ public class GenotypingGMMScorerTest {
         double sigma = 30;
         double[] initialW = new double[] {Math.log(.5),Math.log(.5)};
 
-        GenotypingGMMScorer scorer = new GenotypingGMMScorer();
         assertEquals(.5, scorer.estimate(y, initialW, 200, sigma).w0, 0.0001);
     }
 
@@ -78,7 +82,6 @@ public class GenotypingGMMScorerTest {
         double sigma = 30;
[…]

### 2012-10-29 src/main/scripts/runTest.sh
make mapability and filter optional parameters

@@ -75,6 +75,16 @@ MAX_MAPQ_DIFF=$MAX_MAPQ_DIFF
 
 EOF
 
+MAPABILITY_PARAM=""
+if [ $MAPABILITY -ne "None" ]
+MAPABILITY_PARAM="--mapabilityWeighting $MAPABILITY"
+fi
+
+FILTER_PARAM=""
+if [ $FILTER -ne "None" ]
+FILTER_PARAM="--excludePairsMappingIn $FILTER"
+fi
+
 cat <<EOF
 
 hadoop jar $BUILD_DIR/target/cloudbreak-1.0-SNAPSHOT-exe.jar -Dmapred.reduce.tasks=200 GMMFitSingleEndInsertSizes
@@ -82,8 +92,8 @@ hadoop jar $BUILD_DIR/target/cloudbreak-1.0-SNAPSHOT-exe.jar -Dmapred.reduce.tas
    --outputHDFSDir $HDFS_SAMPLE_DIR/$NAME 
    --faidx $HDFS_FAI 
    --maxInsertSize $MAX_INSERT 
-   --mapabilityWeighting $MAPABILITY 
-   --excludePairsMappingIn $FILTER 
+   $MAPABILITY_PARAM
+   $FILTER_PARAM
    --resolution $RESOLUTION 
    --aligner $ALIGNER
    --maxLogMapqDiff $MAX_MAPQ_DIFF
@@ -95,8 +105,8 @@ hadoop jar $BUILD_DIR/target/cloudbreak-1.0-SNAPSHOT-exe.jar -Dmapred.reduce.tas
     --outputHDFSDir $HDFS_SAMPLE_DIR/$NAME \
     --faidx $HDFS_FAI \
     --maxInsertSize $MAX_INSERT \
-    --mapabilityWeighting $MAPABILITY \
-    --excludePairsMappingIn $FILTER \
+    $MAPABILITY_PARAM \
+    $FILTER_PARAM \
     --resolution $RESOLUTION \
     --aligner $ALIGNER \
     --maxLogMapqDiff $MAX_MAPQ_DIFF

### 2012-10-29 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/mapper/Bowtie2SingleEndMapper.java
Bowtie2 Aligner

@@ -1,0 +1,123 @@
+package edu.ohsu.sonmezsysbio.cloudbreak.mapper;
+
+import edu.ohsu.sonmezsysbio.cloudbreak.AlignmentRecord;
+import edu.ohsu.sonmezsysbio.cloudbreak.io.NovoalignAlignmentReader;
+import edu.ohsu.sonmezsysbio.cloudbreak.io.SAMAlignmentReader;
+import org.apache.hadoop.io.Text;
+import org.apache.hadoop.mapred.JobConf;
+import org.apache.hadoop.mapred.OutputCollector;
+import org.apache.log4j.Logger;
+
+import java.io.*;
+import java.util.Arrays;
+
+/**
+ * Created by IntelliJ IDEA.
+ * User: cwhelan
+ * Date: 5/21/11
+ * Time: 5:36 PM
+ */
+public class Bowtie2SingleEndMapper extends SingleEndAlignmentMapper {
+
+    private static org.apache.log4j.Logger logger = Logger.getLogger(Bowtie2SingleEndMapper.class);
+
+    private OutputCollector<Text, Text> output;
+    private String localDir;
+    private String reference;
+    private String numReports;
+    private String bowtie2Executable;
+
+    @Override
+    public void configure(JobConf job) {
+        super.configure(job);
+
+        logger.debug("Current dir: " + new File(".").getAbsolutePath());
+
+        this.localDir = job.get("mapred.child.tmp");
+        reference = job.get("bowtie2.reference");
+        numReports = job.get("bowtie2.num.reports");
+        bowtie2Executable = job.get("bowtie2.executable");
+
+    }
+
+    @Override
+    public void close() throws IOException {
+        super.close();
+
+        s1FileWriter.close();
+
+        if (! s1File.exists()) {
+            System.err.println("file does not exist: " + s1File.getPath());
+        } else {
+            System.err.println("read file length: " + s1File.length());
+        }
+
+        File indexFile = new File(reference);
+        if (! indexFile.exists()) {
+            System.err.println("index file does not exist: " + indexFile.getPath());
+        } else {
+            System.err.println("index file length: " + indexFile.length());
+        }
+
+        String referenceBaseName = new File(reference).getName();
+        String[] commandLine = buildCommandLine(bowtie2Executable, referenceBaseName, s1File.getPath(), numReports);
+        logger.debug("Executing command: " + Arrays.toString(commandLine));
[…]

### 2012-10-29 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/Cloudbreak.java
add bowtie2 command to CLI

@@ -66,6 +66,9 @@ public class Cloudbreak extends Configured implements Tool
         CommandNovoalignSingleEnds singleEnds  = new CommandNovoalignSingleEnds();
         jc.addCommand("alignSingleEnds", singleEnds);
 
+        CommandBowtie2SingleEnds bowtie2SingleEnds  = new CommandBowtie2SingleEnds();
+        jc.addCommand("bowtie2SingleEnds", bowtie2SingleEnds);
+
         CommandMrFastSingleEnds mrFastSingleEnds  = new CommandMrFastSingleEnds();
         jc.addCommand("mrfastSingleEnds", mrFastSingleEnds);
 

### 2013-01-02 src/main/scripts/recomb.R
adding..

@@ -1,20 +1,25 @@
 library(GenomicRanges)
 library(rtracklayer)
 library(plyr)
+library(RColorBrewer)
 
-drawPerfLine <- function(perf, col, maxFP) {
+drawPerfLine <- function(perf, col, lineType, maxFP) {
   calls <- c(perf$Calls, min(perf$Calls) - min(perf$TP), 0)
   tp <- c(perf$TP, 0, 0)
   fp <- calls - tp
-  lines(fp[fp <= maxFP], tp[fp <= maxFP], col=col, lwd=3)
+  lines(fp[fp <= maxFP], tp[fp <= maxFP], col=col, lty=lineType, lwd=3)
 }
 
-plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE) {
+plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legendLoc="bottomright") {
   maxFP <- totalDels
-  plot(0, type="n", ylim=c(0, totalDels), xlim=c(0,maxFP), xlab=ifelse(sim, "False Positives", "Novel Predictions"), ylab="True Positives", main=main)
-  perfCols <- rainbow(length(perfs))
-  mapply(drawPerfLine, perfs, perfCols, MoreArgs=list(maxFP=maxFP))  
-  legend("bottomright", legend=perfNames, col=perfCols, lwd=3, cex=.9)
+  if (is.na(maxTP)) {
+    maxTP <- totalDels
+  }
+  plot(0, type="n", ylim=c(0, maxTP), xlim=c(0,maxFP), xlab=ifelse(sim, "False Positives", "Novel Predictions"), ylab="True Positives", main=main)
+  perfCols <- brewer.pal(length(perfs), "Set1")
+  lineTypes <- seq(1,length(perfs))
+  mapply(drawPerfLine, perfs, perfCols, lineTypes, MoreArgs=list(maxFP=maxFP))  
+  legend(legendLoc, legend=perfNames, col=perfCols, lty=lineTypes, lwd=3, cex=.9)
 }
 
 overlap <- function(a,b,x,y) {
@@ -24,13 +29,12 @@ overlap <- function(a,b,x,y) {
 sizeclasses <- function(starts, ends) {
   sizelens <- ends - starts + 1
   sizeclasses <- rep(NA, length(sizelens))
-  sizeclasses[sizelens <= 50] <- "25-50bp"
[…]

### 2013-01-02 src/main/java/edu/ohsu/sonmezsysbio/cloudbreak/file/WigFileHelper.java
get ordering of extra wig files right

@@ -332,14 +332,9 @@ public class WigFileHelper {
         }
         for (String extraWigFile : extraFileNames) {
             bedFileWriter.write("\t" + extraWigValueSums.get(extraWigFile) * resolution / ((endPosition + 1) - peakStart));
-        }
-        for (String extraWigFile : extraFileNames) {
             bedFileWriter.write("\t" + extraWigValueMins.get(extraWigFile));
-        }
-        for (String extraWigFile : extraFileNames) {
             bedFileWriter.write("\t" + extraWigValueMaxes.get(extraWigFile));
         }
-
 
         bedFileWriter.write("\n");
     }

### 2013-01-02 src/test/java/edu/ohsu/sonmezsysbio/cloudbreak/file/WigFileHelperTest.java

@@ -117,8 +117,8 @@ public class WigFileHelperTest {
 
         String expectedOutput =
                 "track name = \"test peaks over " + threshold + "\"\n" +
-                        "chr1\t3000\t3999\t1\t5.0\t1.0\t3.0\t1.0\t3.0\t1.0\t3.0\n" +
-                        "chr2\t1000\t2999\t2\t7.0\t2.0\t4.0\t2.0\t3.0\t2.0\t5.0\n";
+                        "chr1\t3000\t3999\t1\t5.0\t1.0\t1.0\t1.0\t3.0\t3.0\t3.0\n" +
+                        "chr2\t1000\t2999\t2\t7.0\t2.0\t2.0\t2.0\t4.0\t3.0\t5.0\n";
         assertEquals(expectedOutput, stringWriter.getBuffer().toString());
     }
 

### 2013-01-03 manuscript/recomb.tex
recomb-seq manuscript

\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{url}
\usepackage{multirow}

\usepackage[style=authoryear,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}
\addbibresource{recomb.bib}

\begin{document}

\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation from Complete Paired-End Sequence Data Sets}

\author{Christopher W. Whelan \and Kemal S\"onmez}

\maketitle

\begin{abstract}
The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data. Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based methods. However, the computational requirements for storing and processing data sets with multiple mappings can be formidable. Meanwhile, the growing size and number of sequencing data sets has led to intense interest in distributing computations to cloud or commodity servers. MapReduce is becoming a standard framework for distributing computation across such compute clusters. We describe a conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome. We demonstrate an implementation of a deletion-finding algorithm in this framework based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome. On simulated and real data sets, our approach achieves performance equal to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes. In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome. Our implementation is available at \url{http://github.com/cwhelan/cloudbreak}.
\end{abstract}

\newpage

\section{Introduction}

Genomic structural variations such as deletions, insertions, and inversions of DNA, are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}. However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.

Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets. See \textcite{Alkan:2011p547} for a review. Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}. Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment, and then clustering the discordant mappings to find SVs with support from multiple pairs. Read-depth (RD) approaches, in contrast, consider the distribution of concordantly mapped reads to infer the presence of SVs from the depth of coverage of locations along the genome \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}. Finally, split-read (SP) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.

RP methods have begun to consider larger and larger data sets. Although the first RP methods used only unambiguously, discordantly mapped read pairs in their analyses, it has been demonstrated that including multiple mappings of discordant read pairs in their analysis improves sensitivity in repetitive and hard-to-map regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}. More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and be able to detect smaller events \autocite{Marschall:2012ek}. Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.

Google's MapReduce \autocite{Dean:2008p277} and its open-source implementation Hadoop\footnote{\url{http://hadoop.apache.org/}}, are designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers. Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping, \autocite{Schatz:2009p278} querying variant databases, \autocite{Oconnor:2010p1835} SNP calling, \autocite{Langmead:2009p1225} RNA-seq differential expression analysis, \autocite{Langmead:2010p1268} ChIP-seq peak calling, \autocite{Feng:2011p1228} and computing genome mappability \autocite{Lee:2012bk}. MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.

In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings. We then demonstrate an implementation of an SV algorithm, Cloudbreak, for discovering genomic deletions, which computes local features based on the insert sizes of spanning mapping. Cloudbreak computes features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM). We then characterize our algorithm's performance on simulated and real data sets.

\section{Results}\label{results}

\subsection{A MapReduce Framework for SV Detection Algorithms}

The MapReduce programming model \autocite{Dean:2008p277} divides computation across a cluster into three phases. In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined. The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key. Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values. Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system for storing large data sets across a cluster. In addition to a MapReduce programming framework, Hadoop also provides a distributed file system (HDFS), which facilitates storing large data sets redundantly across a cluster. MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.

The need to separate logic into mappers and reducers makes it difficult to implement traditional read-pair approaches in MapReduce. This is especially true because of the graph data structures and global clustering of paired end mappings at the heart of many read-pair approaches. MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel. In genomic sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome. The RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features at varying locations along the genome and then integrating those features into SV predictions in a post-processing step. Here we show that this formulation can be easily translated to the MapReduce programming model.

Our formulation divides processing into two distinct MapReduce jobs. The first job finds possible alignment locations for pairs of reads in FASTQ format. The second job computes features from the set of reads that are relevant to each of a set of small intervals covering the genome. In the final stage, the features for each location are exported from the Hadoop file system and post-processed to produce a set of SV calls which span multiple genomic intervals. The MapReduce algorithm is summarized below with a detailed explanation following.

\algrenewcommand\algorithmicprocedure{\textbf{job}}
  \begin{algorithmic}[1]
    \Procedure{Alignment}{}
    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
    \ForAll{$ \textrm{Alignments }a \in \textrm{Align}(<s,q>)$}
    \State $\textsc{Emit}(\textrm{ReadPairId }rid, \textrm{Alignment }a)$
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{ReadPairId }rid, \textrm{Alignments }a_{1,2,\ldots}$}
    \State $\textrm{AlignmentPairList }ap \gets \textrm{ValidAlignmentPairs}(a_{1,2,\ldots}$
    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList} ap)$
    \EndFunction
    \EndProcedure

    \Procedure{Compute SV Features}{}
    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2> \in ap$}
    \State $ \textrm{ReadPairInfo }spi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SpanningPairInfo }spi)$
    \EndFor
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }spi_{1,2,\ldots}$}
    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
    \EndFunction
    \EndProcedure
    \State $\textrm{StructuralVariationCalls } sv \gets \textsc{PostProcess }(\Phi)$ 
  \end{algorithmic}

The alignment job uses sensitive alignment tools and settings to discover as many mapping locations for each read pair as possible. In the mapping phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair. In the reduce phase, Cloudbreak combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found. Our algorithm could use any aligner that is capable of reporting multiple alignments for reads. We report results generated with alignments from RazerS 3 \autocite{Weese:2012by}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop. Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.

In the second job, we compute a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping intervals. For the experiments reported in this paper we use an interval size of 25bp. Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome. Let $R_1 = \left\{r_{1,1},r_{1,2},\ldots,r_{1,M}\right\}$ and $R_2 = \left\{r_{2,1},r_{2,2},\ldots,r_{2,M}\right\}$ be the input set of paired reads. Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,K}\right\}$ be the set of alignments for the left and right reads from read pair $m$. For any given pair of alignments of the two reads in a read pair $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ SpanningPairInfo} spi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g. the fragment size implied by the alignments and the likelihood the alignments are correct. We define three functions, $\textsc{Loci } : \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L$, $\Phi : \left\{\textrm{SpanningPairInfo }spi_{m,i,j}\right\} \rightarrow \mathbb{R}^N$, and $\textsc{PostProcess}$, which maps features across the genome into a set of SV calls. The first maps an alignment to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments. We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair. The second function maps a set of SpanningPairInfos relevant to a given location to a real-valued vector of features useful for SV detection.

We describe one possible set of functions below; however, we believe that there are many possible mappings that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage, heterogeneity of the sample, etc.

\subsection{Cloudbreak: A Hadoop Deletion Detection Implementation}

We have implemented a deletion detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak. 
\begin{description}
\item[\sc{Loci}] Because we are detecting deletions, we map SpanningPairInfo from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads. For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs which map more than 25kb apart, or in the incorrect orientation, to no genomic locations.
\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus. Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$. If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant. Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components. Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.

To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location. We rank the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$. In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$. This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.

We fit the means and of the GMM using the Expectation-Maximization algorithm. Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and the library to have mean fragment size $\mu$ with standard deviation $\sigma$. We initialize the two components to have means $\mu$ and $\bar{Y}$ and standard deviation $\sigma$, and set $\alpha = .5$. In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$. We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions. In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\sum_m{\gamma_{m,2} + y_m}$. We repeat E and M steps until convergence or a maximum number of steps has been taken.

Our features for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.

\item[\sc{PostProcess}] We convert our features along the genome to deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold. To eliminate noise we apply a median filter with window size 5. We end regions when $\mu'$ changes by more than 60bp, discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
\end{description}

\section{Results}\label{results}

We compared the performance of Cloudbreak for detecting deletions to a selection of popular tools: the RP methods Breakdancer \autocite{Chen:2009p3} and GASVPro, \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY. \autocite{Rausch:2012he} In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).

We use the following criteria to define a true prediction given a gold standard set of deletions to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been discovered by another prediction from the same method. Furthermore, we find that Pindel is very accurate at discovering deletions less than or equal to 40bp in length, so we exclude from consideration calls that match a true deletion of less than 40bp where the call is less than or equal to 75bp in length. It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods. Cloudbreak sacrifices additional resolution (by dividing the genome into 25bp windows); we strive however to increase sensitivity and specificity in the hopes that such calls could still be useful even if their resolution is limited. This seems reasonable to us given the possibility of pipelines in which RP calls are refined by SR mappings or local assembly methods.

\subsection{Tests with Simulated Data}

As has been observed elsewhere, there is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference. Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics. On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data. To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous deletions from the genome of J. Craig Venter \autocite{Levy:2007fb}. Since there are relatively few heterozygous deletions annotated in that set, we randomly assigned each deletion to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence. We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package\footnote{\url{http://sourceforge.net/apps/mediawiki/dnaa/}}. We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence. Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous deletions.

\begin{figure}[t]
\centering
\includegraphics[width=0.6\textwidth]{CHR2SIM_ROC_NEW.pdf}
\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with deletions from the Venter genome added to one or both haplotypes. Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer - score; DELLY, GASVPro, and Pindel - number of supporting read pairs.}
\label{chr2roc}
\end{figure}

Figure \ref{chr2roc} shows an ROC plot of the performance of each algorithm on the simulated data set. All approaches show excellent specificity at high thresholds in this simulation. Cloudbreak and DELLY provide the greatest sensitivity at high specificity. Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using choosing a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 1.68 log likelihood ratio for Cloudbreak, score 44 for Breakdancer, 8 supporting read pairs for GASVPro, four supporting read pairs for DELLY, and 18 supporting read pairs for Pindel. Cloudbreak shows excellent sensitivity at each size class, and contributes a number of exclusive true predictions in both small and medium size classes. In addition, Cloudbreak has the greatest sensitivity to repeats that intersect RepeatMasker annotated regions of the reference sequence, correctly identifying 12\% more repeat-overlapping deletions than the nearest competitor.

\begin{table}[b]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & 219 & 82 & 81 & 31 & 26 \\ 
  \hline
Cloudbreak &  \textbf{87} (12) &  56 (0) &  \textbf{61} (\textbf{4}) &  \textbf{10} (1) &  \textbf{16} (0) \\ 
  Breakdancer &  73 (11) &  58 (1) &  54 (0) &  7 (0) &  15 (0) \\ 
  GASVPro &  46 (3) &  27 (0) &  47 (1) &  5 (\textbf{2}) &  13 (0) \\ 
  DELLY &  61 (7) &  \textbf{66} (\textbf{4}) &  56 (0) &  9 (0) &  \textbf{16} (0) \\ 
  Pindel &  52 (\textbf{14}) &  11 (0) &  37 (0) &   4 (0) &  14 (0) \\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter deletions found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2preds}
\end{table}

\subsection{Tests with Biological Data}

We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive.\footnote{\url{http://sra.dnanexus.com/experiments/ERX009609}} This sample was sequenced on the Illumina Genome Analyzer II platform using with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp. The sample was sequenced to a depth of approximately 37X coverage.

To create a gold standard set of deletions to test against, we pooled annotated deletions discovered by three previous studies on the same sample. These included data from the Human Genome Structural Variation Project\footnote{\url{http://hgsv.washington.edu/general/download/SNPs_DIPs/}} reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507. We merged any overlapping calls into the region spanned by their unions. It should be noted that the 1000 Genomes call set was partially produced using DELLY and Breakdancer, and therefore those calls are ones that those tools are sensitive to.

\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{NA18507_ROC.pdf}
\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, test against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
\label{NA18507roc}
\end{figure}

Figure \ref{NA18507roc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard deletion set, using the same rules for a match as described in the previous section. All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to the missing deletions in the gold standard set.

We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted for the difference in coverage from 30X to 37X. The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507preds}. Although Cloudbreak is not the top performer in any category, it is the second-best in every category and provides a large number of exclusive predictions.

% latex table generated in R 2.15.2 by xtable 1.7-0 package
% Tue Jan  1 22:36:02 2013
\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
 & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
Total Number & & & 9,980 & 242 & 218 & 110 & 375 \\
  \hline
Cloudbreak & 0.094 & 0.115 & 601 (201) & 142 (8) & 171 (\textbf{7}) &  87 (\textbf{2}) & 254 (\textbf{7}) \\ 
Breakdancer & 0.118 & 0.094 & 337 (31) & 137 (4) & 169 (0) & 93 (0) & 293 (2) \\
  GASVPro & \textbf{0.127} & 0.048 & 189 (39) &  42 (1) & 102 (1) &  41 (0) & 147 (0) \\ 
  DELLY & 0.070 & 0.091 & 260 (15) & \textbf{156} (\textbf{13}) & \textbf{176} (5) &  \textbf{94} (0) & \textbf{304} (4) \\ 
  Pindel & 0.024 & \textbf{0.130} & \textbf{1,039} (\textbf{791}) & 51 (0) & 125 (1) &  53 (0) & 149 (1) \\ 
   \hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data set, as well as the number of deletions in the gold standard set of each size class found by each tool. The same cutoffs were used as for the simulated data, adjusted for the difference in coverage.}
\label{NA18507preds}
\end{table}

\subsection{Performance in Repetitive Areas}

We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance. We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements (Table \ref{repmaskpreds}) in both the simulated and NA18507 data sets, especially compared to the RP-based methods. Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes highly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.

\begin{table}
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & 122 & 317 & 604 & 10,321 \\ 
  \hline
Cloudbreak & \textbf{36} (5) & \textbf{194} (\textbf{12}) & \textbf{251} (\textbf{34}) & 1,004 (191) \\ 
  Breakdancer &  33 (3) & 174 (9) & 208 (12) & 821 (25) \\ 
  GASVPro & 20 (2) & 118 (4) & 93 (5) & 428 (36) \\ 
  DELLY & 35 (\textbf{6}) & 173 (5) & 229 (23) & 761 (14) \\ 
  Pindel & 12 (5) & 106 (9) & 148 (\textbf{31}) & \textbf{1,269} (\textbf{762}) \\ 
   \hline
\end{tabular}
\end{center}
\caption{True positive deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether or not the deletion overlaps with a RepeatMasker-annotated repetitive element.}
\label{repmaskpreds}
\end{table}

\subsection{Genotyping Deletions}

Since Cloudbreak explicitly fits a mixture of insert size distributions at each location, and each component of the mixture represents an allele containing or not containing a deletion, it should be possible to use the parameters of the fit GMM to infer the genotype of each predicted variant, assuming that our pipeline is capturing all relevant read mappings near the locus of the variant. To test this, we attempted to classify true positive Cloudbreak deletion predictions from the sets described above for both the simulated and biological data sets. Because we simulated a diploid genome, we had genotypes for the simulated data readily available. For the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}. We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes. By setting a simple cutoff of .2 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 87.8\% and 95.7\% accuracy in predicting the genotype of the true positive deletions we detected in the simulated and real data sets, respectively. Table \ref{genotypeaccuracy} shows confusion matrices for the two samples using this simple classifier.

\begin{table}
\begin{center}
\begin{tabular}{r|r|rr|rr|}
\multicolumn{2}{c}{}  & \multicolumn{4}{c}{Actual Genotypes} \\
\multicolumn{2}{c}{}  & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\cline{3-6}
\multicolumn{2}{c|}{} &  Homozygous & Heterozygous & Homozygous & Heterozygous \\ 
\cline{2-6}
\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 129 & 4 & 77 & 12 \\
 & Heterozygous & 24 & 73 & 4 & 280 \\
\cline{2-6}
\end{tabular}
\end{center}
\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
\label{genotypeaccuracy}
\end{table}

\section{Methods}\label{discussion}

Cloudbreak is a native Java Hadoop application. We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4, with snappy compression enabled.

For each evaluated SV detection program we attempt to follow the recommended alignment pipeline. Simulated reads were aligned to hg18 chromosome 2. Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap RazerS 3 version 3.1, with parameters \verb|-rr 99 -i 94 -m 1000|, requesting 99\% sensitivity, a maximum edit distance between read and reference of 6\%, and up to 1000 possible alignments reported for each read. Alignments for the other programs were first aligned using BWA \autocite{Li:2009p836} version 0.6.2-r126, with parameter \verb|-e 5| to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set. GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \verb|-a -r -Ex 1100 -t 250|.

We ran Breakdancer version 1.1\_2011\_02\_21 by first executing \verb|bam2cfg.pl| and then running \verb|breakdancer_max| with the default parameter values. We ran DELLY version 0.0.9 with the \verb|-p| parameter and default values for other parameters. We ran GASVPro version 1.2 using the GASVPro.sh script and default parameters. Finally, we ran Pindel 0.2.4t with default parameters.

\section{Discussion}\label{conclusions}

\section{Acknowledgements}

We would like to thank Izhak Shafran at the Center for Spoken Language Understanding and Lucia Carbone of the Department of Behavioral Neuroscience for their advice, support, and shared computational resources.

\newpage

\printbibliography

\end{document}

% LocalWords:  Abyzov 2011bk Yoon 2010gf 2012kk Michaelson 2012fj 2012ey Feng
% LocalWords:  Oconnor Langmead RNA-seq Langmead ChIP-seq 2012bk langle textbf
% LocalWords:  algrenewcommand algorithmicprocedure textrm rp textrm textrm ap
% LocalWords:  textrm textsc ldots spi RazerS Weese Novoalign Bowtie 2012jh sc
% LocalWords:  rightarrow subseteq mathbb th 2007fb dwgsim includegraphics rr
% LocalWords:  textwidth haplotypes chr2roc chr2preds rrrrrr hline Yoruban
% LocalWords:  indels NA18507preds printbibliography

### 2013-01-03 manuscript/recomb.tex
branching to kemal

@@
 \usepackage{amsfonts}
+\usepackage{amsmath}
 \usepackage{graphicx}
@@
 \usepackage{multirow}
+\usepackage{authblk}
 
@@
 
-\author{Christopher W. Whelan \and Kemal S\"onmez}
+\author[1,2,4]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
+\author[1,2,3,4]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
+\affil[1]{Institute on Development and Disability}
+\affil[2]{Center for Spoken Language Understanding}
+\affil[3]{Department of Medical Informatics \& Clinical Epidemiology}
+
+\affil[4]{Oregon Health \& Science University, Portland, OR, USA}
 
@@
 
-\section{Results}\label{results}
-
-\subsection{A MapReduce Framework for SV Detection Algorithms}
+\section{A MapReduce Framework for SV Detection Algorithms}
 
@@
     \Function{Reduce}{$\textrm{ReadPairId }rid, \textrm{Alignments }a_{1,2,\ldots}$}
-    \State $\textrm{AlignmentPairList }ap \gets \textrm{ValidAlignmentPairs}(a_{1,2,\ldots}$
-    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList} ap)$
+    \State $\textrm{AlignmentPairList }ap \gets \textrm{ValidAlignmentPairs}(a_{1,2,\ldots})$
+    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
     \EndFunction
@@
     \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2> \in ap$}
+    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
     \State $ \textrm{ReadPairInfo }spi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
-    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SpanningPairInfo }spi)$
+    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }spi)$
     \EndFor
@@
 In the mapping phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair.
-In the reduce phase, Cloudbreak combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found.
-Our algorithm could use any aligner that is capable of reporting multiple alignments for reads.
-We report results generated with alignments from RazerS 3 \autocite{Weese:2012by}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
+In the reduce phase, the framework combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found.
+Our framework could use any aligner that is capable of reporting multiple alignments for reads.
+We report results below generated with alignments from RazerS 3 \autocite{Weese:2012by}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
 Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
@@
 Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,K}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
-For any given pair of alignments of the two reads in a read pair $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ SpanningPairInfo} spi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
+For any given pair of alignments of the two reads in a read pair $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } spi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
 the fragment size implied by the alignments and the likelihood the alignments are correct.
-We define three functions, $\textsc{Loci } : \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L$, $\Phi : \left\{\textrm{SpanningPairInfo }spi_{m,i,j}\right\} \rightarrow \mathbb{R}^N$, and $\textsc{PostProcess}$, which maps features across the genome into a set of SV calls.
-The first maps an alignment to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
+We leave three functions to be implemented depending on the application:
+\begin{flalign*}
+ \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
+ \Phi :& \left\{\textrm{SpanningPairInfo }spi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
+ \textsc{PostProcess} :& \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\} 
+\end{flalign*}
+
+The first function, \textsc{Loci}, maps an alignment to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
 We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
-The second function maps a set of SpanningPairInfos relevant to a given location to a real-valued vector of features useful for SV detection.
+The second function, $\Phi$, maps a set of SpanningPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
+Finally, the \textsc{PostProcess} function maps the complete set of features for each locus across the genome into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
 
@@
 \centering
-\includegraphics[width=0.6\textwidth]{CHR2SIM_ROC_NEW.pdf}
+\includegraphics[width=0.75\textwidth]{CHR2SIM_ROC_NEW.pdf}
 \caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with deletions from the Venter genome added to one or both haplotypes.
@@
 
-\begin{table}
+\begin{table}[b]
 \begin{center}
@@
 Simulated reads were aligned to hg18 chromosome 2.
-Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap RazerS 3 version 3.1, with parameters \verb|-rr 99 -i 94 -m 1000|, requesting 99\% sensitivity, a maximum edit distance between read and reference of 6\%, and up to 1000 possible alignments reported for each read.
+Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap RazerS 3 version 3.1, with parameters \\ \verb|-rr 99 -i 94 -m 1000|, requesting 99\% sensitivity, a maximum edit distance between read and reference of 6\%, and up to 1000 possible alignments reported for each read.
 Alignments for the other programs were first aligned using BWA \autocite{Li:2009p836} version 0.6.2-r126, with parameter \verb|-e 5| to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.

### 2013-01-03 manuscript/recomb.tex
update to manuscript

@@
 
-Genomic structural variations such as deletions, insertions, and inversions of DNA, are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}.
+Genomic structural variations such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}.
 However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.
@@
 RP methods have begun to consider larger and larger data sets.
-Although the first RP methods used only unambiguously, discordantly mapped read pairs in their analyses, it has been demonstrated that including multiple mappings of discordant read pairs in their analysis improves sensitivity in repetitive and hard-to-map regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
+Although the first RP methods used only unambiguously discordantly mapped read pairs in their analyses, it has been demonstrated that including multiple mappings of discordant read pairs in the analysis improves sensitivity in repetitive and hard-to-map regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
 More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and be able to detect smaller events \autocite{Marschall:2012ek}.
@@
 This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
+To estimate the probability that the mapping is correct, we approximate the probability $p^{k}_c$ of the alignment of each end being correct by:
 
-We fit the means and of the GMM using the Expectation-Maximization algorithm.
-Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and the library to have mean fragment size $\mu$ with standard deviation $\sigma$.
-We initialize the two components to have means $\mu$ and $\bar{Y}$ and standard deviation $\sigma$, and set $\alpha = .5$.
+\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
+
+And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
+
+We fit the means and variances of the GMM using the Expectation-Maximization algorithm.
+Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
+We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both to $\sigma$, and set $\alpha = .5$.
 In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
@@
 In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\sum_m{\gamma_{m,2} + y_m}$.
-We repeat E and M steps until convergence or a maximum number of steps has been taken.
+We treat the variance as fixed and do not update it, since under our assumptons the standard deviation of each component should always be $\sigma$.
+We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.
 
-Our features for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
+The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
 
@@
 To eliminate noise we apply a median filter with window size 5.
-We end regions when $\mu'$ changes by more than 60bp, discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
+We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
 \end{description}

### 2013-01-03 src/main/scripts/evalBreakdancer.py
change evaulation of breakdancer to use number of pairs

@@ -20,7 +20,7 @@ for line in breakdancer_file:
     if line.startswith("#"):
         continue
     fields = line.split("\t")
-    score = float(fields[8])
+    score = float(fields[9])
     score_values.append(score)
 breakdancer_file.close()
 
@@ -32,16 +32,22 @@ for v in unique_score_values:
     calls_gte_threshold = []
     breakdancer_file = open(breakdancer_filename, "r")
     non_del_calls = 0
+    # sometimes Breakdancer makes nonsensical calls; deletions with ends on different chromosomes, we have to
+    # track those separately here
+    bad_calls = 0
     for line in breakdancer_file:
 #        print line
         fields = line.split("\t")
         if line.startswith("#"):
 #            print "comment!"
             continue
-        if float(fields[8]) >= v:
+        if float(fields[9]) >= v:
 #            print "gte v"
-            if not (fields[6] == "DEL") or (fields[0] != fields[3]):
+            if not (fields[6] == "DEL"):
                 non_del_calls += 1
+                continue
+            if (fields[0] != fields[3]):
+                bad_calls += 1
                 continue
             sv_len = int(fields[7])
             # sys.stderr.write("len: " + str(sv_len) + "\n")
@@ -55,6 +61,7 @@ for v in unique_score_values:
         bed_lines.append(bed_line)
 
     (qualified_calls, matches, short_calls) = evalBedFile.eval_bed(truth_filename, bed_lines)
+    qualified_calls += bad_calls
     tpr = float(matches) / (qualified_calls)
     print "\t".join(map(str, [v, qualified_calls, matches, non_del_calls, short_calls, tpr]))
     

### 2013-01-03 manuscript/recomb.tex
kemal changes

--- 2013-01-03 manuscript/recomb.tex
@@
 However, the computational requirements for storing and processing data sets with multiple mappings can be formidable.
-Meanwhile, the growing size and number of sequencing data sets has led to intense interest in distributing computations to cloud or commodity servers.
+Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computations to cloud or commodity servers.
 MapReduce is becoming a standard framework for distributing computation across such compute clusters.
-We describe a conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome.
-We demonstrate an implementation of a deletion-finding algorithm in this framework based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
+We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome.
+In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
 On simulated and real data sets, our approach achieves performance equal to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
@@
 In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings.
-We then demonstrate an implementation of an SV algorithm, Cloudbreak, for discovering genomic deletions, which computes local features based on the insert sizes of spanning mapping.
+We then demonstrate the development of an SV algorithm, Cloudbreak, for discovering genomic deletions, which computes local features based on the insert sizes of spanning mapping.
 Cloudbreak computes features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
-We then characterize our algorithm's performance on simulated and real data sets.
+We then characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.
 
@@
 
-\subsection{Cloudbreak: A Hadoop Deletion Detection Implementation}
+\subsection{Cloudbreak: A Hadoop Deletion Detection Algorithm}
 

### 2013-01-04 manuscript/recomb.tex
add discussion, other edits

@@
 In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome.
-Our implementation is available at \url{http://github.com/cwhelan/cloudbreak}.
+In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples.
+Our implementation and source files are available at \url{http://sonmezsysbio.org/software/cloudbreak}.
 \end{abstract}
@@
 Read-depth (RD) approaches, in contrast, consider the distribution of concordantly mapped reads to infer the presence of SVs from the depth of coverage of locations along the genome \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
-Finally, split-read (SP) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
+Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
 
-RP methods have begun to consider larger and larger data sets.
-Although the first RP methods used only unambiguously discordantly mapped read pairs in their analyses, it has been demonstrated that including multiple mappings of discordant read pairs in the analysis improves sensitivity in repetitive and hard-to-map regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
+The first RP methods used only unambiguously discordantly mapped read pairs in their analyses.
+However, it has been recently demonstrated that including multiple mappings of discordant read pairs in the analysis improves sensitivity in repetitive and hard-to-map regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
 More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and be able to detect smaller events \autocite{Marschall:2012ek}.
@@
 Google's MapReduce \autocite{Dean:2008p277} and its open-source implementation Hadoop\footnote{\url{http://hadoop.apache.org/}}, are designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
-Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping, \autocite{Schatz:2009p278} querying variant databases, \autocite{Oconnor:2010p1835} SNP calling, \autocite{Langmead:2009p1225} RNA-seq differential expression analysis, \autocite{Langmead:2010p1268} ChIP-seq peak calling, \autocite{Feng:2011p1228} and computing genome mappability \autocite{Lee:2012bk}.
+Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, querying variant databases \autocite{Oconnor:2010p1835}, SNP calling \autocite{Langmead:2009p1225}, RNA-seq differential expression analysis \autocite{Langmead:2010p1268}, ChIP-seq peak calling \autocite{Feng:2011p1228}, and computing genome mappability \autocite{Lee:2012bk}.
 MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.
@@
 In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings.
-We then demonstrate the development of an SV algorithm, Cloudbreak, for discovering genomic deletions, which computes local features based on the insert sizes of spanning mapping.
+We then demonstrate the development of an RP-based SV algorithm, Cloudbreak, for discovering genomic deletions, which computes local features based on the insert sizes of spanning mappings.
 Cloudbreak computes features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
+Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm.
 We then characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.
@@
 
-The need to separate logic into mappers and reducers makes it difficult to implement traditional read-pair approaches in MapReduce.
-This is especially true because of the graph data structures and global clustering of paired end mappings at the heart of many read-pair approaches.
+The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based approaches in MapReduce.
+This is especially true because of the graph data structures and global clustering of paired end mappings at the heart of many RP approaches.
 MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
@@
 The RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features at varying locations along the genome and then integrating those features into SV predictions in a post-processing step.
-Here we show that this formulation can be easily translated to the MapReduce programming model.
+Here we show that this technique can be easily translated to the MapReduce programming model.
 
@@
 The second job computes features from the set of reads that are relevant to each of a set of small intervals covering the genome.
-In the final stage, the features for each location are exported from the Hadoop file system and post-processed to produce a set of SV calls which span multiple genomic intervals.
+In a final third stage executed outside of MapReduce, the features for each location are exported from the Hadoop file system and post-processed to produce a set of SV calls which span multiple genomic intervals.
 The MapReduce algorithm is summarized below with a detailed explanation following.
@@
     \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
-    \State $ \textrm{ReadPairInfo }spi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }spi)$
+    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
+    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
     \EndFor
@@
     \EndFunction
-    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }spi_{1,2,\ldots}$}
+    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
     \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
@@
 Our framework could use any aligner that is capable of reporting multiple alignments for reads.
-We report results below generated with alignments from RazerS 3 \autocite{Weese:2012by}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
+The results we report below were generated with alignments from RazerS 3 \autocite{Weese:2012by}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
 Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
@@
 Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,K}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
-For any given pair of alignments of the two reads in a read pair $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } spi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
+For any given pair of alignments of the two reads in a read pair $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
 the fragment size implied by the alignments and the likelihood the alignments are correct.
@@
  \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
- \Phi :& \left\{\textrm{SpanningPairInfo }spi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
+ \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
  \textsc{PostProcess} :& \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\} 
@@
 In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\sum_m{\gamma_{m,2} + y_m}$.
-We treat the variance as fixed and do not update it, since under our assumptons the standard deviation of each component should always be $\sigma$.
+We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$.
 We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.
@@
 \caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with deletions from the Venter genome added to one or both haplotypes.
-Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer - score; DELLY, GASVPro, and Pindel - number of supporting read pairs.}
+Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro, and Pindel - number of supporting read pairs.}
 \label{chr2roc}
@@
 Cloudbreak and DELLY provide the greatest sensitivity at high specificity.
-Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using choosing a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 1.68 log likelihood ratio for Cloudbreak, score 44 for Breakdancer, 8 supporting read pairs for GASVPro, four supporting read pairs for DELLY, and 18 supporting read pairs for Pindel.
+Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using choosing a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 1.68 log likelihood ratio for Cloudbreak, 3 supporting read pairs for Breakdancer, 8 supporting read pairs for GASVPro, 4 supporting read pairs for DELLY, and 18 supporting read pairs for Pindel.
 Cloudbreak shows excellent sensitivity at each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
-In addition, Cloudbreak has the greatest sensitivity to repeats that intersect RepeatMasker annotated regions of the reference sequence, correctly identifying 12\% more repeat-overlapping deletions than the nearest competitor.
 
-\begin{table}[b]
+\begin{table}[t]
 \begin{center}
@@
 All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to the missing deletions in the gold standard set.
+Cloudbreak is the best performer at more stringent thresholds, and remains near the top at looser thresholds.
 
@@
 The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507preds}.
-Although Cloudbreak is not the top performer in any category, it is the second-best in every category and provides a large number of exclusive predictions.
+Cloudbreak scores well in every category,
 
@@
 We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements (Table \ref{repmaskpreds}) in both the simulated and NA18507 data sets, especially compared to the RP-based methods.
+In the simulated data set, Cloudbreak correctly identifies 10\% more repeat-overlapping deletions than the nearest competitor.
+In the real data set, Pindel finds many short deletions in repetitive regions; however, Cloudbreak identifies 18\% more repeat-overlapping deletions than the other RP methods tested.
 Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes highly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
@@
 
-\section{Discussion}\label{conclusions}
+\section{Discussion}\label{Discussion}
 
-\section{Acknowledgements}
+Over the next few years, due to advances in sequencing technology, genomics data are expected to increase both in volume and in coverage by several orders of magnitude.
+As a result, the field of genomics will properly join the Big Data phenomenon.
+In addition, the usage of existing data will increase drastically as translational research in genomics grows and the data is reprocessed and integrated into new pipelines.
+In this context, detection of variations in genomic data is a central problem and is likely to drive many translational applications as well as biomedical research.
+In order to capitalize on the emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of the data sets will have to be developed.
+
+Among existing Big Data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
+In this paper, we introduce a novel conceptual framework for structural variation detection algorithms in MapReduce, based on computing local features along the genome.
+This framework provides a scalable basis for developing SV detection algorithms that are capable of addressing high volumes and coverage.
+In particular, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
+
+On simulated and real data sets, our approach performs well across a wide range of deletion sizes and achieves performance similar to or better than a variety of state-of-the-art structural variation detection algorithms that run on traditional workload clusters and servers.
+Detection of deletions is an important subset of the SV detection problem is important for translational research; \textcite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
+In addition, an increased ability to detect variants in regions of the genome that are repetitive and difficult to align reads to opens up new possibilities in exploring human genetic variation.
+
+In addition to delivering state-of-the-art performance in a RP-based SV detection problem, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data either in a cloud or commodity server setting.
+Using tools like Apache Whirr\footnote{\url{http://whirr.apache.org/}}, it is becoming trivially easy to instantiate on-demand Hadoop compute clusters with cloud providers such as Amazon's Elastic Compute Cloud.
+Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who are unable to afford or unwilling to maintain large, specialized servers to analyze their data.
+
+\newpage
+
+\section*{Acknowledgements}
 
 We would like to thank Izhak Shafran at the Center for Spoken Language Understanding and Lucia Carbone of the Department of Behavioral Neuroscience for their advice, support, and shared computational resources.
-
-\newpage
 

### 2013-02-14 manuscript/manuscript.tex
Update calls to use the all indels sim

--- 2013-01-04 manuscript/recomb.tex
@@
 
-\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation from Complete Paired-End Sequence Data Sets}
+\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}
 
@@
 The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data.
-Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based methods.
+Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods.
 However, the computational requirements for storing and processing data sets with multiple mappings can be formidable.
-Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computations to cloud or commodity servers.
-MapReduce is becoming a standard framework for distributing computation across such compute clusters.
+Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers.
+MapReduce is becoming a standard framework for distributing processing across such compute clusters.
 We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome.
 In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
-On simulated and real data sets, our approach achieves performance equal to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
+On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
 In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome.
 In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples.
-Our implementation and source files are available at \url{http://sonmezsysbio.org/software/cloudbreak}.
+Our implementation and source code are available at \url{http://sonmezsysbio.org/software/cloudbreak}.
+
+\medskip
+\noindent\textbf{Keywords:} genomic structural variation; distributed computing; copy number variation; high-throughput sequencing; genotyping.
 \end{abstract}
@@
 Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
-Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment, and then clustering the discordant mappings to find SVs with support from multiple pairs.
-Read-depth (RD) approaches, in contrast, consider the distribution of concordantly mapped reads to infer the presence of SVs from the depth of coverage of locations along the genome \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
+Traditionally, this involves separating mappings into vthose that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment.
+These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs.
+Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
 Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
@@
 The first RP methods used only unambiguously discordantly mapped read pairs in their analyses.
-However, it has been recently demonstrated that including multiple mappings of discordant read pairs in the analysis improves sensitivity in repetitive and hard-to-map regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
-More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and be able to detect smaller events \autocite{Marschall:2012ek}.
-Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.
+However, it has been recently demonstrated that including multiple mappings of discordant read pairs in the analysis can improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
+More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and thus be able to detect smaller events \autocite{Marschall:2012ek}.
+Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, in part because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.
 
@@
 In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings.
-We then demonstrate the development of an RP-based SV algorithm, Cloudbreak, for discovering genomic deletions, which computes local features based on the insert sizes of spanning mappings.
-Cloudbreak computes features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
+We then demonstrate the development in this framework of an algorithm, Cloudbreak, for discovering genomic deletions.
+Cloudbreak computes local features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
 Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm.
@@
 Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
-Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system for storing large data sets across a cluster.
-In addition to a MapReduce programming framework, Hadoop also provides a distributed file system (HDFS), which facilitates storing large data sets redundantly across a cluster.
+Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
 MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
 
-The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based approaches in MapReduce.
-This is especially true because of the graph data structures and global clustering of paired end mappings at the heart of many RP approaches.
+The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based SV detection approaches in MapReduce.
+This is especially true because of the global clustering of paired end mappings at the heart of many RP approaches.
 MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
-In genomic sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome.
-The RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features at varying locations along the genome and then integrating those features into SV predictions in a post-processing step.
+In sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome.
+The RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features at locations along the genome and then producing SV predictions from those features in a post-processing step.
 Here we show that this technique can be easily translated to the MapReduce programming model.
@@
 The first job finds possible alignment locations for pairs of reads in FASTQ format.
-The second job computes features from the set of reads that are relevant to each of a set of small intervals covering the genome.
+The second job computes features from the set of alignment pairs that are relevant to each of a set of small intervals covering the genome.
 In a final third stage executed outside of MapReduce, the features for each location are exported from the Hadoop file system and post-processed to produce a set of SV calls which span multiple genomic intervals.
-The MapReduce algorithm is summarized below with a detailed explanation following.
+This MapReduce framework is summarized below with a detailed explanation following.
 
@@
     \Procedure{Alignment}{}
-    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
-    \ForAll{$ \textrm{Alignments }a \in \textrm{Align}(<s,q>)$}
-    \State $\textsc{Emit}(\textrm{ReadPairId }rid, \textrm{Alignment }a)$
+    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
+    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
+    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
     \EndFor
     \EndFunction
-    \Function{Reduce}{$\textrm{ReadPairId }rid, \textrm{Alignments }a_{1,2,\ldots}$}
-    \State $\textrm{AlignmentPairList }ap \gets \textrm{ValidAlignmentPairs}(a_{1,2,\ldots})$
+    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
+    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
     \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
@@
     \EndProcedure
-    \State $\textrm{StructuralVariationCalls } sv \gets \textsc{PostProcess }(\Phi)$ 
+    \State $\textrm{StructuralVariationCalls } svs \gets \textsc{PostProcess}(\phi_{1,2,\ldots})$ 
   \end{algorithmic}
 
-The alignment job uses sensitive alignment tools and settings to discover as many mapping locations for each read pair as possible.
-In the mapping phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair.
+The alignment job uses sensitive mapping tools and settings to discover as many mapping locations for each read pair as possible.
+In the map phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair.
 In the reduce phase, the framework combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found.
@@
 Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
-Let $R_1 = \left\{r_{1,1},r_{1,2},\ldots,r_{1,M}\right\}$ and $R_2 = \left\{r_{2,1},r_{2,2},\ldots,r_{2,M}\right\}$ be the input set of paired reads.
-Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,K}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
-For any given pair of alignments of the two reads in a read pair $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
+Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
+Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
+For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
 the fragment size implied by the alignments and the likelihood the alignments are correct.
-We leave three functions to be implemented depending on the application:
+We then leave three functions to be implemented depending on the application:
 \begin{flalign*}
@@
 
-The first function, \textsc{Loci}, maps an alignment to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
+The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
 We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
-The second function, $\Phi$, maps a set of SpanningPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
+The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
 Finally, the \textsc{PostProcess} function maps the complete set of features for each locus across the genome into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
 
-We describe one possible set of functions below; however, we believe that there are many possible mappings that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage, heterogeneity of the sample, etc.
+We describe one possible set of functions below; however, we believe that there are many possible definitions that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage of the experiment, heterogeneity of the sample, etc.
 
@@
 We have implemented a deletion detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak.
+Our implementation defines the three application-specific functions described above as:
 \begin{description}
-\item[\sc{Loci}] Because we are detecting deletions, we map SpanningPairInfo from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
-For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs which map more than 25kb apart, or in the incorrect orientation, to no genomic locations.
+\item[\sc{Loci}] Because we are detecting deletions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
+For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
 \item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus.
@@
 To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
-We rank the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
+We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
 In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
 This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
-To estimate the probability that the mapping is correct, we approximate the probability $p^{k}_c$ of the alignment of each end being correct by:
+We approximate the probability $p^{k}_c$ of each end alignment being correct by:
 
@@
 
-We fit the means and variances of the GMM using the Expectation-Maximization algorithm.
+We fit the parameters of the GMM using the Expectation-Maximization algorithm.
 Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
-We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both to $\sigma$, and set $\alpha = .5$.
+We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
 In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
 We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions.
-In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\sum_m{\gamma_{m,2} + y_m}$.
+In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$.
 We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$.
@@
 
-We compared the performance of Cloudbreak for detecting deletions to a selection of popular tools: the RP methods Breakdancer \autocite{Chen:2009p3} and GASVPro, \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY.
-\autocite{Rausch:2012he} In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).
+We compared the performance of Cloudbreak for detecting deletions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
+In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).
 
-We use the following criteria to define a true prediction given a gold standard set of deletions to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been discovered by another prediction from the same method.
+We use the following criteria to define a true prediction given a gold standard set of deletions to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method.
 Furthermore, we find that Pindel is very accurate at discovering deletions less than or equal to 40bp in length, so we exclude from consideration calls that match a true deletion of less than 40bp where the call is less than or equal to 75bp in length.
+We also discard all Pindel calls of less than or equal to 20bp.
 It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods.
@@
 On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data.
-To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous deletions from the genome of J.
+To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous insertions and deletions from the genome of J.
 Craig Venter \autocite{Levy:2007fb}.
-Since there are relatively few heterozygous deletions annotated in that set, we randomly assigned each deletion to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
+Since there are relatively few heterozygous insertions and deletions annotated in the Venter genome, we randomly assigned each variant to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
 We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package\footnote{\url{http://sourceforge.net/apps/mediawiki/dnaa/}}.
 We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence.
-Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous deletions.
+Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous insertions and deletions.
 
@@
 \centering
-\includegraphics[width=0.75\textwidth]{CHR2SIM_ROC_NEW.pdf}
-\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with deletions from the Venter genome added to one or both haplotypes.
+\includegraphics[width=0.75\textwidth]{CHR2SIM_DELS_ROC.pdf}
+\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes.
 Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro, and Pindel - number of supporting read pairs.}
@@
 All approaches show excellent specificity at high thresholds in this simulation.
-Cloudbreak and DELLY provide the greatest sensitivity at high specificity.
-Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using choosing a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 1.68 log likelihood ratio for Cloudbreak, 3 supporting read pairs for Breakdancer, 8 supporting read pairs for GASVPro, 4 supporting read pairs for DELLY, and 18 supporting read pairs for Pindel.
-Cloudbreak shows excellent sensitivity at each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
+Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.
+Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 2.1 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and 33 supporting read pairs for Pindel.
+Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
 
@@
  & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 219 & 82 & 81 & 31 & 26 \\ 
+ Total Number & 224 & 84 & 82 & 31 & 26 \\ 
   \hline
-Cloudbreak &  \textbf{87} (12) &  56 (0) &  \textbf{61} (\textbf{4}) &  \textbf{10} (1) &  \textbf{16} (0) \\ 
-  Breakdancer &  73 (11) &  58 (1) &  54 (0) &  7 (0) &  15 (0) \\ 
-  GASVPro &  46 (3) &  27 (0) &  47 (1) &  5 (\textbf{2}) &  13 (0) \\ 
-  DELLY &  61 (7) &  \textbf{66} (\textbf{4}) &  56 (0) &  9 (0) &  \textbf{16} (0) \\ 
-  Pindel &  52 (\textbf{14}) &  11 (0) &  37 (0) &   4 (0) &  14 (0) \\ 
+Cloudbreak &  51 (10) &  51 (\textbf{3}) &  \textbf{55} (\textbf{5}) &  \textbf{12} (\textbf{4}) &  \textbf{15} (\textbf{1}) \\ 
+  Breakdancer &  \textbf{52} (14) &  49 (2) &  49 (0) &  7 (0) &  14 (0) \\ 
+  GASVPro &  31 (4) &  25 (0) &  23 (0) &  2 (0) &  6 (0) \\ 
+  DELLY &  22 (2) &  \textbf{55} (2) &  40 (0) &  8 (0) &  12 (0) \\ 
+  Pindel &  29 (\textbf{16}) &  11 (0) &  21 (0) & 1 (0) &  8 (0) \\ 
    \hline
@@
 
-We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive.\footnote{\url{http://sra.dnanexus.com/experiments/ERX009609}} This sample was sequenced on the Illumina Genome Analyzer II platform using with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp.
+We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive.\footnote{\url{http://sra.dnanexus.com/experiments/ERX009609}} This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp.
 The sample was sequenced to a depth of approximately 37X coverage.
@@
 We merged any overlapping calls into the region spanned by their unions.
-It should be noted that the 1000 Genomes call set was partially produced using DELLY and Breakdancer, and therefore those calls are ones that those tools are sensitive to.
+It should be noted that the 1000 Genomes call set was partially produced using DELLY and Breakdancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor.
 
@@
 \includegraphics[width=0.75\textwidth]{NA18507_ROC.pdf}
-\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, test against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
+\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
 \label{NA18507roc}
@@
 Figure \ref{NA18507roc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard deletion set, using the same rules for a match as described in the previous section.
-All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to the missing deletions in the gold standard set.
-Cloudbreak is the best performer at more stringent thresholds, and remains near the top at looser thresholds.
+All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing deletions in the gold standard set that are actually present in the sample.
+Cloudbreak is the best performer at more stringent thresholds, and remains near the top at higher sensitivity levels.
 
-We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted for the difference in coverage from 30X to 37X.
+We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
 The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507preds}.
-Cloudbreak scores well in every category,
+Cloudbreak scores well in every category: it identifies more small deletions than any tool except for Pindel, which has low specificity at this cutoff, and contributes more exclusive true positives in larger size classes than any other tool.
 
@@
   \hline
-Cloudbreak & 0.094 & 0.115 & 601 (201) & 142 (8) & 171 (\textbf{7}) &  87 (\textbf{2}) & 254 (\textbf{7}) \\ 
-Breakdancer & 0.118 & 0.094 & 337 (31) & 137 (4) & 169 (0) & 93 (0) & 293 (2) \\
-  GASVPro & \textbf{0.127} & 0.048 & 189 (39) &  42 (1) & 102 (1) &  41 (0) & 147 (0) \\ 
-  DELLY & 0.070 & 0.091 & 260 (15) & \textbf{156} (\textbf{13}) & \textbf{176} (5) &  \textbf{94} (0) & \textbf{304} (4) \\ 
-  Pindel & 0.024 & \textbf{0.130} & \textbf{1,039} (\textbf{791}) & 51 (0) & 125 (1) &  53 (0) & 149 (1) \\ 
+Cloudbreak & 0.100 & \textbf{0.090} & \textbf{436} (174) & 128 (\textbf{9}) & 158 (\textbf{8}) &  70 (\textbf{4}) & 186 (12) \\ 
+Breakdancer & 0.120 & 0.086 & 260 (43) & \textbf{132} (8) & \textbf{167} (2) & \textbf{92} (1) & \textbf{288} (\textbf{14}) \\
+  GASVPro & \textbf{0.127} & 0.048 & 102 (11) &  37 (2) & 77 (0) &  26 (0) & 93 (0) \\ 
+  DELLY & 0.081 & 0.070 & 144 (10) & 125 (8) & 158 (1) &  83 (1) & 256 (3) \\ 
+  Pindel & 0.013 & 0.061 & 414 (\textbf{297}) & 40 (0) & 90 (0) &  37 (0) & 88 (0) \\ 
    \hline
@@
 \end{center}
-\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data set, as well as the number of deletions in the gold standard set of each size class found by each tool.
-The same cutoffs were used as for the simulated data, adjusted for the difference in coverage.}
+\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found.
+The same cutoffs were used as for the simulated data as in Table \ref{chr2preds}, adjusted for the difference in coverage from 30X to 37X.}
 \label{NA18507preds}
@@
 We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements (Table \ref{repmaskpreds}) in both the simulated and NA18507 data sets, especially compared to the RP-based methods.
-In the simulated data set, Cloudbreak correctly identifies 10\% more repeat-overlapping deletions than the nearest competitor.
-In the real data set, Pindel finds many short deletions in repetitive regions; however, Cloudbreak identifies 18\% more repeat-overlapping deletions than the other RP methods tested.
-Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes highly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
+In the simulated data set, Cloudbreak correctly identifies 9\% more repeat-overlapping deletions than the nearest competitor.
+Cloudbreak identifies more deletions in repeats than any other methods, and finds almost four times more exclusive hits than the other RP-based methods.
+Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
 
@@
   \hline
-Cloudbreak & \textbf{36} (5) & \textbf{194} (\textbf{12}) & \textbf{251} (\textbf{34}) & 1,004 (191) \\ 
-  Breakdancer &  33 (3) & 174 (9) & 208 (12) & 821 (25) \\ 
-  GASVPro & 20 (2) & 118 (4) & 93 (5) & 428 (36) \\ 
-  DELLY & 35 (\textbf{6}) & 173 (5) & 229 (23) & 761 (14) \\ 
-  Pindel & 12 (5) & 106 (9) & 148 (\textbf{31}) & \textbf{1,269} (\textbf{762}) \\ 
+Cloudbreak & \textbf{29} (4) & \textbf{155} (\textbf{19}) & \textbf{204} (\textbf{38}) & \textbf{774} (169) \\ 
+  Breakdancer &  \textbf{29} (\textbf{6}) & 142 (10) & 186 (25) & 753 (43) \\ 
+  GASVPro & 15 (2) & 72 (2) & 71 (5) & 264 (8) \\ 
+  DELLY & 21 (2) & 116 (2) & 147 (12) & 619 (11) \\ 
+  Pindel & 8 (4) & 62 (12) & 98 (26) & 571 (\textbf{271}) \\ 
    \hline
@@
 \end{center}
-\caption{True positive deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether or not the deletion overlaps with a RepeatMasker-annotated repetitive element.}
+\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
 \label{repmaskpreds}
@@
 Since Cloudbreak explicitly fits a mixture of insert size distributions at each location, and each component of the mixture represents an allele containing or not containing a deletion, it should be possible to use the parameters of the fit GMM to infer the genotype of each predicted variant, assuming that our pipeline is capturing all relevant read mappings near the locus of the variant.
-To test this, we attempted to classify true positive Cloudbreak deletion predictions from the sets described above for both the simulated and biological data sets.
+To test this, we attempted to classify the true positive Cloudbreak deletion predictions described above for both the simulated and biological data sets.
 Because we simulated a diploid genome, we had genotypes for the simulated data readily available.
@@
 We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes.
-By setting a simple cutoff of .2 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 87.8\% and 95.7\% accuracy in predicting the genotype of the true positive deletions we detected in the simulated and real data sets, respectively.
-Table \ref{genotypeaccuracy} shows confusion matrices for the two samples using this simple classifier.
+By setting a simple cutoff of .2 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.0\% and 94.9\% accuracy in predicting the genotype of the true positive deletions we detected in the simulated and real data sets, respectively.
+Table \ref{genotypeaccuracy} shows confusion matrices for the two samples using this classifier.
 
@@
 \cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 129 & 4 & 77 & 12 \\
- & Heterozygous & 24 & 73 & 4 & 280 \\
+\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 91 & 3 & 70 & 11 \\
+ & Heterozygous & 19 & 71 & 4 & 209 \\
 \cline{2-6}
@@
 
-Cloudbreak is a native Java Hadoop application. We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4, with snappy compression enabled.
+Cloudbreak is a native Java Hadoop application.
+We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4.
+We use snappy compression for MapReduce data.
+Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
 
-For each evaluated SV detection program we attempt to follow the recommended alignment pipeline.
+For each evaluated SV detection program we attempt to follow the authors' recommended alignment pipeline.
 Simulated reads were aligned to hg18 chromosome 2.
@@
 We ran DELLY version 0.0.9 with the \verb|-p| parameter and default values for other parameters.
-We ran GASVPro version 1.2 using the GASVPro.sh script and default parameters.
+We ran GASVPro version 1.2 using the \verb|GASVPro.sh| script and default parameters.
 Finally, we ran Pindel 0.2.4t with default parameters.
@@
 
-Over the next few years, due to advances in sequencing technology, genomics data are expected to increase both in volume and in coverage by several orders of magnitude.
-As a result, the field of genomics will properly join the Big Data phenomenon.
-In addition, the usage of existing data will increase drastically as translational research in genomics grows and the data is reprocessed and integrated into new pipelines.
-In this context, detection of variations in genomic data is a central problem and is likely to drive many translational applications as well as biomedical research.
-In order to capitalize on the emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of the data sets will have to be developed.
+Over the next few years, due to advances in sequencing technology, genomics data are expected to increase both in volume and in coverage by several orders of magnitude, expanding into the realm referred to as Big Data.
+In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines.
+In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.
 
-Among existing Big Data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
+Among Big Data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
 In this paper, we introduce a novel conceptual framework for structural variation detection algorithms in MapReduce, based on computing local features along the genome.
-This framework provides a scalable basis for developing SV detection algorithms that are capable of addressing high volumes and coverage.
-In particular, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
+This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of a distributed deletion-finding algorithm based on fitting a GMM to the distribution of mapped insert sizes spanning each location in the genome.
 
 On simulated and real data sets, our approach performs well across a wide range of deletion sizes and achieves performance similar to or better than a variety of state-of-the-art structural variation detection algorithms that run on traditional workload clusters and servers.
-Detection of deletions is an important subset of the SV detection problem is important for translational research; \textcite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
+Detection of deletions is an important area of research; \textcite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
 In addition, an increased ability to detect variants in regions of the genome that are repetitive and difficult to align reads to opens up new possibilities in exploring human genetic variation.
 
-In addition to delivering state-of-the-art performance in a RP-based SV detection problem, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data either in a cloud or commodity server setting.
-Using tools like Apache Whirr\footnote{\url{http://whirr.apache.org/}}, it is becoming trivially easy to instantiate on-demand Hadoop compute clusters with cloud providers such as Amazon's Elastic Compute Cloud.
+In addition to delivering state-of-the-art performance in a RP-based SV detection problem, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
+Using tools like Apache Whirr\footnote{\url{http://whirr.apache.org/}}, it is becoming trivially easy to instantiate on-demand Hadoop compute clusters with cloud providers such as Amazon's EC2.
 Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who are unable to afford or unwilling to maintain large, specialized servers to analyze their data.

### 2013-02-14 src/main/scripts/manuscript.R

@@ -1,0 +1,272 @@
+library(GenomicRanges)
+library(rtracklayer)
+library(plyr)
+library(RColorBrewer)
+library(VariantAnnotation)
+library(xtable)
+
+drawPerfLine <- function(perf, col, lineType, maxFP) {
+  calls <- c(perf$Calls, min(perf$Calls) - min(perf$TP), 0)
+  tp <- c(perf$TP, 0, 0)
+  fp <- calls - tp
+  lines(fp[fp <= maxFP], tp[fp <= maxFP], col=col, lty=lineType, lwd=3)
+}
+
+plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legendLoc="bottomright") {
+  maxFP <- totalDels
+  if (is.na(maxTP)) {
+    maxTP <- totalDels
+  }
+  plot(0, type="n", ylim=c(0, maxTP), xlim=c(0,maxFP), xlab=ifelse(sim, "False Positives", "Novel Predictions"), ylab="True Positives", main=main)
+  if (length(perfs) <= 9) {
+    perfCols <- brewer.pal(length(perfs), "Set1")
+  } else {
+    perfCols <- rainbow(length(perfs))
+  }
+  lineTypes <- seq(1,length(perfs))
+  mapply(drawPerfLine, perfs, perfCols, lineTypes, MoreArgs=list(maxFP=maxFP))  
+  legend(legendLoc, legend=perfNames, col=perfCols, lty=lineTypes, lwd=3, cex=.9)
+}
+
+overlap <- function(a,b,x,y) {
+  pmin(b,y) - pmax(a,x)
+}
+
+sizeclasses <- function(starts, ends) {
+  sizelens <- ends - starts + 1
+  sizeclasses <- rep(NA, length(sizelens))
+  sizeclasses[sizelens <= 100] <- "40-100bp"
+  sizeclasses[sizelens > 100 & sizelens <= 250] <- "101-250bp"
+  sizeclasses[sizelens > 250 & sizelens <= 500] <- "251-500bp"
+  sizeclasses[sizelens > 500 & sizelens <= 1000] <- "501-1000bp"
+  sizeclasses[sizelens > 1000] <- "> 1000bp"  
+  factor(sizeclasses, levels=c("40-100bp","101-250bp","251-500bp","501-1000bp","> 1000bp"),ordered=TRUE)
+}
+
+processPredictions <- function(name, predfile, trueDels) {
[…]

### 2013-02-15 src/main/scripts/manuscript.R
update r analysis to order lines right, use correct thresholds

@@ -68,11 +68,11 @@ gasv <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_
 delly <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_del.perf.txt', header=TRUE)
 pindel <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_pindel_D.perf.txt', header=TRUE)
 
-perfsList <- list(breakdancer=breakdancer, gasv=gasv, delly=delly, pindel=pindel, cloudbreak=cloudbreak)
+perfsList <- list(cloudbreak=cloudbreak, breakdancer=breakdancer, pindel=pindel, gasv=gasv, delly=delly, )
 pdf('~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf', width=10)
 par(xpd=T, mar=par()$mar+c(0,0,0,7))
 plotROC(perfsList, 
-        c("Breakdancer","GASVPro", "DELLY", "Pindel", "Cloudbreak"), 
+        c("Cloudbreak", "Breakdancer","Pindel", "GASVPro", "DELLY"), 
         totalDels, "Deletions in Venter diploid chr2 simulation",legendLoc=xy.coords(425,200), maxTP=350)
 dev.off()
 
@@ -82,11 +82,11 @@ cloudbreak <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_alli
 breakdancer <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/breakdancer_insertions.perf.txt', header=TRUE)
 pindel <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/pindel_insertions.perf.txt', header=TRUE)
 
-perfsList <- list(breakdancer=breakdancer, pindel=pindel, cloudbreak=cloudbreak)
+perfsList <- list(cloudbreak=cloudbreak, breakdancer=breakdancer, pindel=pindel)
[…]

### 2013-02-15 manuscript/recomb.tex
final version of recomb seq submission

@@
 Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
-Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment, and then clustering the discordant mappings to find SVs with support from multiple pairs.
-Read-depth (RD) approaches, in contrast, consider the distribution of concordantly mapped reads to infer the presence of SVs from the depth of coverage of locations along the genome \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
+Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment.
+These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs.
+Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
 Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
@@
 Craig Venter \autocite{Levy:2007fb}.
-Since there are relatively few heterozygous deletions annotated in that set, we randomly assigned each deletion to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
+Since there are relatively few heterozygous deletions annotated in the Venter genome, we randomly assigned each deletion to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
 We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package\footnote{\url{http://sourceforge.net/apps/mediawiki/dnaa/}}.
@@
 All approaches show excellent specificity at high thresholds in this simulation.
-Cloudbreak and DELLY provide the greatest sensitivity at high specificity.
-Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using choosing a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 1.68 log likelihood ratio for Cloudbreak, 3 supporting read pairs for Breakdancer, 8 supporting read pairs for GASVPro, 4 supporting read pairs for DELLY, and 18 supporting read pairs for Pindel.
-Cloudbreak shows excellent sensitivity at each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
+Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.
+Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 1.68 log likelihood ratio for Cloudbreak, 3 supporting read pairs for Breakdancer, 8 supporting read pairs for GASVPro, 4 supporting read pairs for DELLY, and 18 supporting read pairs for Pindel.
+Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
 
@@
 Figure \ref{NA18507roc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard deletion set, using the same rules for a match as described in the previous section.
-All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to the missing deletions in the gold standard set.
-Cloudbreak is the best performer at more stringent thresholds, and remains near the top at looser thresholds.
+All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing deletions in the gold standard set that are actually present in the sample.
+Cloudbreak is the best performer at more stringent thresholds, and remains near the top at higher sensitivity levels.
 
@@
 The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507preds}.
-Cloudbreak scores well in every category,
+Cloudbreak scores well in every category: it identifies more small deletions than any tool except for Pindel, which has low specificity at this cutoff, and contributes more exclusive true positives in larger size classes than any other tool.
 
@@
 In the simulated data set, Cloudbreak correctly identifies 10\% more repeat-overlapping deletions than the nearest competitor.
-In the real data set, Pindel finds many short deletions in repetitive regions; however, Cloudbreak identifies 18\% more repeat-overlapping deletions than the other RP methods tested.
-Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes highly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
+In the real data set, Pindel finds many short deletions in repetitive regions; however, Cloudbreak identifies 18\% more deletions in repeats than the other RP methods.
+Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
 
@@
 By setting a simple cutoff of .2 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 87.8\% and 95.7\% accuracy in predicting the genotype of the true positive deletions we detected in the simulated and real data sets, respectively.
-Table \ref{genotypeaccuracy} shows confusion matrices for the two samples using this simple classifier.
+Table \ref{genotypeaccuracy} shows confusion matrices for the two samples using this classifier.
 

### 2013-02-15 src/main/scripts/recomb.R

@@ -2,6 +2,7 @@ library(GenomicRanges)
 library(rtracklayer)
 library(plyr)
 library(RColorBrewer)
+library(VariantAnnotation)
 
 drawPerfLine <- function(perf, col, lineType, maxFP) {
   calls <- c(perf$Calls, min(perf$Calls) - min(perf$TP), 0)
@@ -57,13 +58,13 @@ processPredictions <- function(name, predfile, trueDels) {
 #chr2 gt 50 100bp diploid
 totalDels <- 400
 cloudbreak <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_100bp_dip/chr2mwf5_readGroupsRazerS3_new3_i94_s99_m1000_None_None_25000_25_sam_6_32fc04d_-1_-1_3_lrHeterozygous.short40.perf.txt', header=TRUE, sep="\t")
-breakdancer <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_100bp_dip/final_predictions/breakdancer/bwa_new3_sort_35_2_3.bd_out.small40.perf.txt', header=TRUE)
+breakdancer <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_100bp_dip/final_predictions/breakdancer/bwa_new3_sort_35_2_3.bd_out.short40.pairthresh.perf.txt', header=TRUE)
 gasv <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_100bp_dip/final_predictions/GASVPro/BAMToGASV.gasvpro.in.ALL.MCMCThreshold.clusters.pruned.clusters.short40.perf.txt', header=TRUE)
 delly <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_100bp_dip/final_predictions/delly/delly_q0_c5_del_bwa_new3.small40.perf.txt', header=TRUE)
 pindel <- read.table('~/Documents/gene_rearrange/svpipeline/venter_chr2_100bp_dip/final_predictions/pindel/bwa_new3.pindel.noshort.short40.perf.txt', header=TRUE)
 
 perfsList <- list(breakdancer=breakdancer, gasv=gasv, pindel=pindel, delly=delly, cloudbreak=cloudbreak)
-pdf('~/Documents/svpipeline/manuscript/CHR2SIM_ROC_NEW.pdf')
[…]

### 2013-02-15 manuscript/manuscript.tex
everything up-to-date with new simulation

@@
  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
-Total Number & & & 9,980 & 242 & 218 & 110 & 375 \\
+Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
   \hline
-Cloudbreak & 0.100 & \textbf{0.090} & \textbf{436} (174) & 128 (\textbf{9}) & 158 (\textbf{8}) &  70 (\textbf{4}) & 186 (12) \\ 
-Breakdancer & 0.120 & 0.086 & 260 (43) & \textbf{132} (8) & \textbf{167} (2) & \textbf{92} (1) & \textbf{288} (\textbf{14}) \\
-  GASVPro & \textbf{0.127} & 0.048 & 102 (11) &  37 (2) & 77 (0) &  26 (0) & 93 (0) \\ 
-  DELLY & 0.081 & 0.070 & 144 (10) & 125 (8) & 158 (1) &  83 (1) & 256 (3) \\ 
-  Pindel & 0.013 & 0.061 & 414 (\textbf{297}) & 40 (0) & 90 (0) &  37 (0) & 88 (0) \\ 
+Cloudbreak & 0.098 & \textbf{0.115} & \textbf{423} (163) & 128 (\textbf{9}) & 158 (\textbf{8}) &  70 (\textbf{4}) & 186 (12) \\ 
+Breakdancer & 0.097 & 0.111 & 258 (42) & \textbf{132} (8) & \textbf{167} (2) & \textbf{92} (1) & \textbf{288} (\textbf{14}) \\
+  GASVPro & \textbf{0.132} & 0.040 & 102 (11) &  37 (2) & 77 (0) &  26 (0) & 93 (0) \\ 
+  DELLY & 0.081 & 0.091 & 143 (10) & 125 (8) & 158 (1) &  83 (1) & 256 (3) \\ 
+  Pindel & 0.013 & 0.079 & 406 (\textbf{290}) & 40 (0) & 90 (0) &  37 (0) & 88 (0) \\ 
    \hline
@@
 In the simulated data set, Cloudbreak correctly identifies 9\% more repeat-overlapping deletions than the nearest competitor.
-Cloudbreak identifies more deletions in repeats than any other methods, and finds almost four times more exclusive hits than the other RP-based methods.
+Cloudbreak identifies more deletions in repeats than any other method, and in the NA18507 data set finds 3.75 times more exclusive hits than the other RP-based methods.
 Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
@@
   \hline
-Cloudbreak & \textbf{29} (4) & \textbf{155} (\textbf{19}) & \textbf{204} (\textbf{38}) & \textbf{774} (169) \\ 
-  Breakdancer &  \textbf{29} (\textbf{6}) & 142 (10) & 186 (25) & 753 (43) \\ 
+Cloudbreak & \textbf{29} (4) & \textbf{155} (\textbf{19}) & \textbf{204} (\textbf{38}) & \textbf{761} (158) \\ 
+  Breakdancer &  \textbf{29} (\textbf{6}) & 142 (10) & 186 (25) & 751 (42) \\ 
   GASVPro & 15 (2) & 72 (2) & 71 (5) & 264 (8) \\ 
-  DELLY & 21 (2) & 116 (2) & 147 (12) & 619 (11) \\ 
-  Pindel & 8 (4) & 62 (12) & 98 (26) & 571 (\textbf{271}) \\ 
+  DELLY & 21 (2) & 116 (2) & 147 (12) & 618 (11) \\ 
+  Pindel & 8 (4) & 62 (12) & 98 (26) & 563 (\textbf{264}) \\ 
    \hline

### 2013-02-27 manuscript/NA18507preds.table.brew.tex
brew latex tables

% latex table generated in R 2.15.2 by xtable 1.7-0 package
% Tue Jan  1 22:36:02 2013
\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
 & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
Total Number & & & <%= format(totals['40-100bp'], big.mark=",") %> & <%= format(totals['101-250bp'], big.mark=",") %> & <%= format(totals['251-500bp'], big.mark=",") %> & <%= format(totals['501-1000bp'], big.mark=",") %> & <%= format(totals['> 1000bp'], big.mark=",") %> \\
  \hline
Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', precision) %> & <%= printPredTableValueFromList('Cloudbreak', recall) %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\ 
Breakdancer & <%= printPredTableValueFromList('Breakdancer', precision) %> & <%= printPredTableValueFromList('Breakdancer', recall) %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\  
  GASVPro & <%= printPredTableValueFromList('GASVPro', precision) %> & <%= printPredTableValueFromList('GASVPro',recall) %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\ 
  DELLY & <%= printPredTableValueFromList('DELLY', precision) %> & <%= printPredTableValueFromList('DELLY', recall) %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '> 1000bp') %> \\ 
  Pindel & <%= printPredTableValueFromList('Pindel', precision) %> & <%= printPredTableValueFromList('Pindel', recall) %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %> \\ 
   \hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found. The same cutoffs were used as for the simulated data as in Table \ref{chr2preds}, adjusted for the difference in coverage from 30X to 37X.}
\label{NA18507preds}
\end{table}

### 2013-02-27 manuscript/chr2preds.table.brew.tex
brew latex tables

\begin{table}[t]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
  \hline
  Cloudbreak  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\ 
  Breakdancer & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\ 
  GASVPro     & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\ 
  DELLY       & <%= printPredTableCell(preds, exPreds, 'DELLY', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '> 1000bp') %> \\ 
  Pindel      & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter deletions found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2preds}
\end{table}

### 2013-02-27 manuscript/manuscript.tex
brew latex tables

@@
  & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 224 & 84 & 82 & 31 & 26 \\ 
+ Total Number & 224 &  84 & 82 &  31 & 26\\ 
   \hline
@@
   DELLY       & 22 (2)  & \textbf{55} (2) &  40 (0) & 8 (0) & 12 (0) \\ 
-  Pindel &  29 (\textbf{16}) &  11 (0) &  21 (0) & 1 (0) &  8 (0) \\ 
+  Pindel      & 29 (\textbf{16})  & 11 (0) &  21 (0) & 1 (0) & 8 (0)\\ 
    \hline
@@
  &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 122 & 317 & 604 & 10,321 \\ 
+ Total Number & 120 & 327 & 553 & 7851 \\ 
   \hline

### 2013-02-27 manuscript/repmask.table.brew.tex
brew latex tables

\begin{table}[b]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & <%= chr2Totals['0'] %> & <%= chr2Totals['1'] %> & <%= NA18507Totals['0'] %> & <%= NA18507Totals['1'] %> \\ 
  \hline
  Cloudbreak  & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Cloudbreak', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Cloudbreak', '1') %> \\ 
  Breakdancer & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Breakdancer', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Breakdancer', '1') %> \\
  GASVPro     & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'GASVPro', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'GASVPro', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'GASVPro', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'GASVPro', '1') %> \\
  DELLY       & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY', '1') %> \\
  Pindel      & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '1') %> \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
\label{repmaskpreds}
\end{table}

### 2013-02-27 src/main/scripts/manuscript.R

@@ -4,6 +4,7 @@ library(plyr)
 library(RColorBrewer)
 library(VariantAnnotation)
 library(xtable)
+library(brew)
 
 drawPerfLine <- function(perf, col, lineType, maxFP) {
   calls <- c(perf$Calls, min(perf$Calls) - min(perf$TP), 0)
@@ -32,18 +33,17 @@ overlap <- function(a,b,x,y) {
   pmin(b,y) - pmax(a,x)
 }
 
-sizeclasses <- function(starts, ends) {
-  sizelens <- ends - starts + 1
-  sizeclasses <- rep(NA, length(sizelens))
-  sizeclasses[sizelens <= 100] <- "40-100bp"
-  sizeclasses[sizelens > 100 & sizelens <= 250] <- "101-250bp"
-  sizeclasses[sizelens > 250 & sizelens <= 500] <- "251-500bp"
-  sizeclasses[sizelens > 500 & sizelens <= 1000] <- "501-1000bp"
-  sizeclasses[sizelens > 1000] <- "> 1000bp"  
-  factor(sizeclasses, levels=c("40-100bp","101-250bp","251-500bp","501-1000bp","> 1000bp"),ordered=TRUE)
+sizeclasses <- function(sizelens) {
+  sc <- rep("< 40bp", length(sizelens))
+  sc[sizelens >= 40 & sizelens <= 100] <- "40-100bp"
+  sc[sizelens > 100 & sizelens <= 250] <- "101-250bp"
+  sc[sizelens > 250 & sizelens <= 500] <- "251-500bp"
+  sc[sizelens > 500 & sizelens <= 1000] <- "501-1000bp"
+  sc[sizelens > 1000] <- "> 1000bp"  
+  factor(sc, levels=c("< 40bp", "40-100bp","101-250bp","251-500bp","501-1000bp","> 1000bp"),ordered=TRUE)
 }
 
-processPredictions <- function(name, predfile, trueDels) {
+processDeletionPredictions <- function(name, predfile, trueDels) {
   predswithhits <- read.table(predfile)
   names(predswithhits) <- c('predchrom', 'predstart', 'predend', 'truechrom', 'truestart', 'trueend')
   
@@ -59,6 +59,47 @@ processPredictions <- function(name, predfile, trueDels) {
   maxhits$name <- name
   maxhits
 }
+
+processInsertionPredictions <- function(name, predfile, trueInsertions) {
+  predswithhits <- read.table(predfile)
[…]

### 2013-02-28 manuscript/chr2DeletionPreds.table.brew.tex
everything is now up-to-speed with auto R output

--- 2013-02-27 manuscript/chr2preds.table.brew.tex
@@
 \end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter deletions found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
+\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
 The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2preds}
+\label{chr2DeletionPreds}
 \end{table}

### 2013-02-28 manuscript/chr2InsertionPredsBySize.table.brew.tex
everything is now up-to-speed with auto R output

\begin{table}[t]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
  \hline
  Cloudbreak  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\ 
  Breakdancer & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\ 
  Pindel      & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those insertions that were discovered exclusively by each tool (in parentheses). The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.}
\label{chr2InsertionPreds}
\end{table}

### 2013-02-28 manuscript/deletionGenotypeAccuracy.table.brew.tex
everything is now up-to-speed with auto R output

--- 2013-02-14 manuscript/manuscript.tex
@@
-\documentclass[11pt]{article}
-
-\usepackage[margin=1in]{geometry}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
-\usepackage{url}
-\usepackage{multirow}
-\usepackage{authblk}
-
-\usepackage[style=authoryear,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
-\renewbibmacro{in:}{%
-  \ifentrytype{article}{}{%
-  \printtext{\bibstring{in}\intitlepunct}}}
-\addbibresource{recomb.bib}
-
-\begin{document}
-
-\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}
-
-\author[1,2,4]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
-\author[1,2,3,4]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
-\affil[1]{Institute on Development and Disability}
-\affil[2]{Center for Spoken Language Understanding}
-\affil[3]{Department of Medical Informatics \& Clinical Epidemiology}
-
-\affil[4]{Oregon Health \& Science University, Portland, OR, USA}
-
-\maketitle
-
-\begin{abstract}
-The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data.
-Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods.
-However, the computational requirements for storing and processing data sets with multiple mappings can be formidable.
-Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers.
-MapReduce is becoming a standard framework for distributing processing across such compute clusters.
-We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome.
-In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
-On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
-In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome.
-In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples.
-Our implementation and source code are available at \url{http://sonmezsysbio.org/software/cloudbreak}.
-
-\medskip
-\noindent\textbf{Keywords:} genomic structural variation; distributed computing; copy number variation; high-throughput sequencing; genotyping.
-\end{abstract}
-
-\newpage
-
-\section{Introduction}
-
-Genomic structural variations such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}.
-However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.
-
-Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets.
-See \textcite{Alkan:2011p547} for a review.
-Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
-Traditionally, this involves separating mappings into vthose that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment.
-These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs.
-Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
-Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
-
-The first RP methods used only unambiguously discordantly mapped read pairs in their analyses.
-However, it has been recently demonstrated that including multiple mappings of discordant read pairs in the analysis can improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
-More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and thus be able to detect smaller events \autocite{Marschall:2012ek}.
-Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, in part because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.
-
-Google's MapReduce \autocite{Dean:2008p277} and its open-source implementation Hadoop\footnote{\url{http://hadoop.apache.org/}}, are designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
-Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, querying variant databases \autocite{Oconnor:2010p1835}, SNP calling \autocite{Langmead:2009p1225}, RNA-seq differential expression analysis \autocite{Langmead:2010p1268}, ChIP-seq peak calling \autocite{Feng:2011p1228}, and computing genome mappability \autocite{Lee:2012bk}.
-MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.
-
-In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings.
-We then demonstrate the development in this framework of an algorithm, Cloudbreak, for discovering genomic deletions.
-Cloudbreak computes local features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
-Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm.
-We then characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.
-
-\section{A MapReduce Framework for SV Detection Algorithms}
-
-The MapReduce programming model \autocite{Dean:2008p277} divides computation across a cluster into three phases.
-In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
-The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
-Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
-Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
-MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
-
-The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based SV detection approaches in MapReduce.
-This is especially true because of the global clustering of paired end mappings at the heart of many RP approaches.
-MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
-In sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome.
-The RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features at locations along the genome and then producing SV predictions from those features in a post-processing step.
-Here we show that this technique can be easily translated to the MapReduce programming model.
-
-Our formulation divides processing into two distinct MapReduce jobs.
-The first job finds possible alignment locations for pairs of reads in FASTQ format.
-The second job computes features from the set of alignment pairs that are relevant to each of a set of small intervals covering the genome.
-In a final third stage executed outside of MapReduce, the features for each location are exported from the Hadoop file system and post-processed to produce a set of SV calls which span multiple genomic intervals.
-This MapReduce framework is summarized below with a detailed explanation following.
-
-\algrenewcommand\algorithmicprocedure{\textbf{job}}
-  \begin{algorithmic}[1]
-    \Procedure{Alignment}{}
-    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
-    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
-    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
-    \EndFor
-    \EndFunction
-    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
-    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
-    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
-    \EndFunction
-    \EndProcedure
-
-    \Procedure{Compute SV Features}{}
-    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
-    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
-    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
-    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
-    \EndFor
-    \EndFor
-    \EndFunction
-    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
-    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
-    \EndFunction
-    \EndProcedure
-    \State $\textrm{StructuralVariationCalls } svs \gets \textsc{PostProcess}(\phi_{1,2,\ldots})$ 
-  \end{algorithmic}
-
-The alignment job uses sensitive mapping tools and settings to discover as many mapping locations for each read pair as possible.
-In the map phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair.
-In the reduce phase, the framework combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found.
-Our framework could use any aligner that is capable of reporting multiple alignments for reads.
-The results we report below were generated with alignments from RazerS 3 \autocite{Weese:2012by}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
-Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
-
-In the second job, we compute a set of features for each location in the genome.
-To begin, we tile the genome with small fixed-width, non-overlapping intervals.
-For the experiments reported in this paper we use an interval size of 25bp.
-Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
-Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
-Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
-For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
-the fragment size implied by the alignments and the likelihood the alignments are correct.
-We then leave three functions to be implemented depending on the application:
-\begin{flalign*}
- \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
- \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
- \textsc{PostProcess} :& \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\} 
-\end{flalign*}
-
-The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
-We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
-The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
-Finally, the \textsc{PostProcess} function maps the complete set of features for each locus across the genome into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
-
-We describe one possible set of functions below; however, we believe that there are many possible definitions that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage of the experiment, heterogeneity of the sample, etc.
-
-\subsection{Cloudbreak: A Hadoop Deletion Detection Algorithm}
-
-We have implemented a deletion detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak.
-Our implementation defines the three application-specific functions described above as:
-\begin{description}
-\item[\sc{Loci}] Because we are detecting deletions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
-For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
-\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus.
-Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
-If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant.
-Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components.
-Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
-
-To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
-We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
-In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
-This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
-We approximate the probability $p^{k}_c$ of each end alignment being correct by:
-
-\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
-
-And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
-
-We fit the parameters of the GMM using the Expectation-Maximization algorithm.
-Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
-We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
-In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
-We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions.
-In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$.
-We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$.
-We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.
-
-The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
-
-\item[\sc{PostProcess}] We convert our features along the genome to deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
-To eliminate noise we apply a median filter with window size 5.
-We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
-\end{description}
-
-\section{Results}\label{results}
-
-We compared the performance of Cloudbreak for detecting deletions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
-In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).
-
-We use the following criteria to define a true prediction given a gold standard set of deletions to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method.
-Furthermore, we find that Pindel is very accurate at discovering deletions less than or equal to 40bp in length, so we exclude from consideration calls that match a true deletion of less than 40bp where the call is less than or equal to 75bp in length.
-We also discard all Pindel calls of less than or equal to 20bp.
-It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods.
-Cloudbreak sacrifices additional resolution (by dividing the genome into 25bp windows); we strive however to increase sensitivity and specificity in the hopes that such calls could still be useful even if their resolution is limited.
-This seems reasonable to us given the possibility of pipelines in which RP calls are refined by SR mappings or local assembly methods.
-
-\subsection{Tests with Simulated Data}
-
-As has been observed elsewhere, there is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference.
-Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics.
-On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data.
-To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous insertions and deletions from the genome of J.
-Craig Venter \autocite{Levy:2007fb}.
-Since there are relatively few heterozygous insertions and deletions annotated in the Venter genome, we randomly assigned each variant to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
-We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package\footnote{\url{http://sourceforge.net/apps/mediawiki/dnaa/}}.
-We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence.
-Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous insertions and deletions.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.75\textwidth]{CHR2SIM_DELS_ROC.pdf}
-\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes.
-Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro, and Pindel - number of supporting read pairs.}
-\label{chr2roc}
-\end{figure}
-
-Figure \ref{chr2roc} shows an ROC plot of the performance of each algorithm on the simulated data set.
-All approaches show excellent specificity at high thresholds in this simulation.
-Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.
-Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 2.1 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and 33 supporting read pairs for Pindel.
-Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
-
-\begin{table}[t]
-\begin{center}
-\begin{tabular}{rrrrrr}
-  \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 224 & 84 & 82 & 31 & 26 \\ 
-  \hline
-Cloudbreak &  51 (10) &  51 (\textbf{3}) &  \textbf{55} (\textbf{5}) &  \textbf{12} (\textbf{4}) &  \textbf{15} (\textbf{1}) \\ 
-  Breakdancer &  \textbf{52} (14) &  49 (2) &  49 (0) &  7 (0) &  14 (0) \\ 
-  GASVPro &  31 (4) &  25 (0) &  23 (0) &  2 (0) &  6 (0) \\ 
-  DELLY &  22 (2) &  \textbf{55} (2) &  40 (0) &  8 (0) &  12 (0) \\ 
-  Pindel &  29 (\textbf{16}) &  11 (0) &  21 (0) & 1 (0) &  8 (0) \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter deletions found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2preds}
-\end{table}
-
-\subsection{Tests with Biological Data}
-
-We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive.\footnote{\url{http://sra.dnanexus.com/experiments/ERX009609}} This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp.
-The sample was sequenced to a depth of approximately 37X coverage.
-
-To create a gold standard set of deletions to test against, we pooled annotated deletions discovered by three previous studies on the same sample.
-These included data from the Human Genome Structural Variation Project\footnote{\url{http://hgsv.washington.edu/general/download/SNPs_DIPs/}} reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
-We merged any overlapping calls into the region spanned by their unions.
-It should be noted that the 1000 Genomes call set was partially produced using DELLY and Breakdancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.75\textwidth]{NA18507_ROC.pdf}
-\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
-\label{NA18507roc}
-\end{figure}
-
-Figure \ref{NA18507roc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard deletion set, using the same rules for a match as described in the previous section.
-All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing deletions in the gold standard set that are actually present in the sample.
-Cloudbreak is the best performer at more stringent thresholds, and remains near the top at higher sensitivity levels.
-
-We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
-The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507preds}.
-Cloudbreak scores well in every category: it identifies more small deletions than any tool except for Pindel, which has low specificity at this cutoff, and contributes more exclusive true positives in larger size classes than any other tool.
-
-% latex table generated in R 2.15.2 by xtable 1.7-0 package
-% Tue Jan  1 22:36:02 2013
-\begin{table}[t]
-\begin{center}
-\begin{tabular}{rrr|rrrrr}
-  \hline
- & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
-Total Number & & & 9,980 & 242 & 218 & 110 & 375 \\
-  \hline
-Cloudbreak & 0.100 & \textbf{0.090} & \textbf{436} (174) & 128 (\textbf{9}) & 158 (\textbf{8}) &  70 (\textbf{4}) & 186 (12) \\ 
-Breakdancer & 0.120 & 0.086 & 260 (43) & \textbf{132} (8) & \textbf{167} (2) & \textbf{92} (1) & \textbf{288} (\textbf{14}) \\
-  GASVPro & \textbf{0.127} & 0.048 & 102 (11) &  37 (2) & 77 (0) &  26 (0) & 93 (0) \\ 
-  DELLY & 0.081 & 0.070 & 144 (10) & 125 (8) & 158 (1) &  83 (1) & 256 (3) \\ 
-  Pindel & 0.013 & 0.061 & 414 (\textbf{297}) & 40 (0) & 90 (0) &  37 (0) & 88 (0) \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found.
-The same cutoffs were used as for the simulated data as in Table \ref{chr2preds}, adjusted for the difference in coverage from 30X to 37X.}
-\label{NA18507preds}
-\end{table}
-
-\subsection{Performance in Repetitive Areas}
-
-We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance.
-We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements (Table \ref{repmaskpreds}) in both the simulated and NA18507 data sets, especially compared to the RP-based methods.
-In the simulated data set, Cloudbreak correctly identifies 9\% more repeat-overlapping deletions than the nearest competitor.
-Cloudbreak identifies more deletions in repeats than any other methods, and finds almost four times more exclusive hits than the other RP-based methods.
-Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
-
-\begin{table}[b]
-\begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\hline
- &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 122 & 317 & 604 & 10,321 \\ 
-  \hline
-Cloudbreak & \textbf{29} (4) & \textbf{155} (\textbf{19}) & \textbf{204} (\textbf{38}) & \textbf{774} (169) \\ 
-  Breakdancer &  \textbf{29} (\textbf{6}) & 142 (10) & 186 (25) & 753 (43) \\ 
-  GASVPro & 15 (2) & 72 (2) & 71 (5) & 264 (8) \\ 
-  DELLY & 21 (2) & 116 (2) & 147 (12) & 619 (11) \\ 
-  Pindel & 8 (4) & 62 (12) & 98 (26) & 571 (\textbf{271}) \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
-\label{repmaskpreds}
-\end{table}
-
-\subsection{Genotyping Deletions}
-
-Since Cloudbreak explicitly fits a mixture of insert size distributions at each location, and each component of the mixture represents an allele containing or not containing a deletion, it should be possible to use the parameters of the fit GMM to infer the genotype of each predicted variant, assuming that our pipeline is capturing all relevant read mappings near the locus of the variant.
-To test this, we attempted to classify the true positive Cloudbreak deletion predictions described above for both the simulated and biological data sets.
-Because we simulated a diploid genome, we had genotypes for the simulated data readily available.
-For the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
-We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes.
-By setting a simple cutoff of .2 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.0\% and 94.9\% accuracy in predicting the genotype of the true positive deletions we detected in the simulated and real data sets, respectively.
-Table \ref{genotypeaccuracy} shows confusion matrices for the two samples using this classifier.
-
 \begin{table}
@@
 \cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 91 & 3 & 70 & 11 \\
- & Heterozygous & 19 & 71 & 4 & 209 \\
+\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & <%= chr2HapCM["2", "TRUE"] %> & <%= chr2HapCM["1", "TRUE"] %> &  <%= NA18507HapCM["2", "TRUE"] %> & <%= NA18507HapCM["1", "TRUE"] %> \\
+ & Heterozygous & <%= chr2HapCM["2", "FALSE"] %> & <%= chr2HapCM["1", "FALSE"] %> &  <%= NA18507HapCM["2", "FALSE"] %> & <%= NA18507HapCM["1", "FALSE"] %> \\
 \cline{2-6}
@@
 \end{table}
-
-\section{Methods}\label{discussion}
-
-Cloudbreak is a native Java Hadoop application.
-We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4.
-We use snappy compression for MapReduce data.
-Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
-
-For each evaluated SV detection program we attempt to follow the authors' recommended alignment pipeline.
-Simulated reads were aligned to hg18 chromosome 2.
-Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap RazerS 3 version 3.1, with parameters \\ \verb|-rr 99 -i 94 -m 1000|, requesting 99\% sensitivity, a maximum edit distance between read and reference of 6\%, and up to 1000 possible alignments reported for each read.
-Alignments for the other programs were first aligned using BWA \autocite{Li:2009p836} version 0.6.2-r126, with parameter \verb|-e 5| to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
-GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \verb|-a -r -Ex 1100 -t 250|.
-
-We ran Breakdancer version 1.1\_2011\_02\_21 by first executing \verb|bam2cfg.pl| and then running \verb|breakdancer_max| with the default parameter values.
-We ran DELLY version 0.0.9 with the \verb|-p| parameter and default values for other parameters.
-We ran GASVPro version 1.2 using the \verb|GASVPro.sh| script and default parameters.
-Finally, we ran Pindel 0.2.4t with default parameters.
-
-\section{Discussion}\label{Discussion}
-
-Over the next few years, due to advances in sequencing technology, genomics data are expected to increase both in volume and in coverage by several orders of magnitude, expanding into the realm referred to as Big Data.
-In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines.
-In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.
-
-Among Big Data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
-In this paper, we introduce a novel conceptual framework for structural variation detection algorithms in MapReduce, based on computing local features along the genome.
-This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of a distributed deletion-finding algorithm based on fitting a GMM to the distribution of mapped insert sizes spanning each location in the genome.
-
-On simulated and real data sets, our approach performs well across a wide range of deletion sizes and achieves performance similar to or better than a variety of state-of-the-art structural variation detection algorithms that run on traditional workload clusters and servers.
-Detection of deletions is an important area of research; \textcite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
-In addition, an increased ability to detect variants in regions of the genome that are repetitive and difficult to align reads to opens up new possibilities in exploring human genetic variation.
-
-In addition to delivering state-of-the-art performance in a RP-based SV detection problem, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
-Using tools like Apache Whirr\footnote{\url{http://whirr.apache.org/}}, it is becoming trivially easy to instantiate on-demand Hadoop compute clusters with cloud providers such as Amazon's EC2.
-Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who are unable to afford or unwilling to maintain large, specialized servers to analyze their data.
-
-\newpage
-
-\section*{Acknowledgements}
-
-We would like to thank Izhak Shafran at the Center for Spoken Language Understanding and Lucia Carbone of the Department of Behavioral Neuroscience for their advice, support, and shared computational resources.
-
-\printbibliography
-
-\end{document}
-
-% LocalWords:  Abyzov 2011bk Yoon 2010gf 2012kk Michaelson 2012fj 2012ey Feng
-% LocalWords:  Oconnor Langmead RNA-seq Langmead ChIP-seq 2012bk langle textbf
-% LocalWords:  algrenewcommand algorithmicprocedure textrm rp textrm textrm ap
-% LocalWords:  textrm textsc ldots spi RazerS Weese Novoalign Bowtie 2012jh sc
-% LocalWords:  rightarrow subseteq mathbb th 2007fb dwgsim includegraphics rr
-% LocalWords:  textwidth haplotypes chr2roc chr2preds rrrrrr hline Yoruban
-% LocalWords:  indels NA18507preds printbibliography

### 2013-02-28 manuscript/manuscript.tex
everything is now up-to-speed with auto R output

@@
 Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.
-Table \ref{chr2preds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 2.1 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and 33 supporting read pairs for Pindel.
+Table \ref{chr2DeletionPreds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 2.1 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and 33 supporting read pairs for Pindel.
 Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
@@
   \hline
-  Cloudbreak  & 51 (10)  & 51 (\textbf{3}) &  \textbf{55} (\textbf{5}) & \textbf{12} (\textbf{4}) & \textbf{15} (\textbf{1}) \\ 
-  Breakdancer & \textbf{52} (14)  & 49 (2) &  49 (0) & 7 (0) & 14 (0) \\ 
-  GASVPro     & 31 (4)  & 25 (0) &  23 (0) & 2 (0) & 6 (0) \\ 
-  DELLY       & 22 (2)  & \textbf{55} (2) &  40 (0) & 8 (0) & 12 (0) \\ 
-  Pindel      & 29 (\textbf{16})  & 11 (0) &  21 (0) & 1 (0) & 8 (0)\\ 
+  Cloudbreak  &   51 (  10)  &   51 (\textbf{   3}) &  \textbf{  55} (\textbf{   5}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{   1}) \\ 
+  Breakdancer & \textbf{  52} (  14)  &   49 (   2) &    49 (0) &    7 (0) &   14 (0) \\ 
+  GASVPro     &   31 (   4)  &   25 (0) &    23 (0) &    2 (0) &    6 (0) \\ 
+  DELLY       &   22 (   2)  & \textbf{  55} (   2) &    40 (0) &    8 (0) &   12 (0) \\ 
+  Pindel      &   29 (\textbf{  16})  &   11 (0) &    21 (0) &    1 (0) &    8 (0)\\ 
    \hline
@@
 \centering
-\includegraphics[width=0.75\textwidth]{NA18507_ROC.pdf}
+\includegraphics[width=0.75\textwidth]{NA18507_DELS_ROC.pdf}
 \caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
@@
 We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
-The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507preds}.
+The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507DeletionPreds}.
 Cloudbreak scores well in every category: it identifies more small deletions than any tool except for Pindel, which has low specificity at this cutoff, and contributes more exclusive true positives in larger size classes than any other tool.
 
-% latex table generated in R 2.15.2 by xtable 1.7-0 package
-% Tue Jan  1 22:36:02 2013
 \begin{table}[t]
@@
   \hline
-Cloudbreak & 0.098 & \textbf{0.115} & \textbf{423} (163) & 128 (\textbf{9}) & 158 (\textbf{8}) &  70 (\textbf{4}) & 186 (12) \\ 
-Breakdancer & 0.097 & 0.111 & 258 (42) & \textbf{132} (8) & \textbf{167} (2) & \textbf{92} (1) & \textbf{288} (\textbf{14}) \\
-  GASVPro & \textbf{0.132} & 0.040 & 102 (11) &  37 (2) & 77 (0) &  26 (0) & 93 (0) \\ 
-  DELLY & 0.081 & 0.091 & 143 (10) & 125 (8) & 158 (1) &  83 (1) & 256 (3) \\ 
-  Pindel & 0.013 & 0.079 & 406 (\textbf{290}) & 40 (0) & 90 (0) &  37 (0) & 88 (0) \\ 
+Cloudbreak & 0.0978 & \textbf{0.115} & \textbf{ 423} ( 163)  &  128 (\textbf{   9}) &   158 (\textbf{   8}) &   70 (\textbf{   4}) &  186 (  12) \\ 
+Breakdancer & 0.119 & 0.111 &  258 (  42)  & \textbf{ 132} (   8) &  \textbf{ 167} (   2) & \textbf{  92} (   1) & \textbf{ 288} (\textbf{  14}) \\  
+  GASVPro & \textbf{0.132} & 0.0399 &  102 (  11)  &   37 (   2) &    77 (0) &   26 (0) &   93 (0) \\ 
+  DELLY & 0.0813 & 0.091 &  143 (  10)  &  125 (   8) &   158 (   1) &   83 (   1) &  256 (   3) \\ 
+  Pindel & 0.0134 & 0.0787 &  406 (\textbf{ 290})  &   40 (0) &    90 (0) &   37 (0) &   88 (0) \\ 
    \hline
@@
 \caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found.
-The same cutoffs were used as for the simulated data as in Table \ref{chr2preds}, adjusted for the difference in coverage from 30X to 37X.}
-\label{NA18507preds}
+The same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPreds}, adjusted for the difference in coverage from 30X to 37X.}
+\label{NA18507DeletionPreds}
 \end{table}
@@
 We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance.
-We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements (Table \ref{repmaskpreds}) in both the simulated and NA18507 data sets, especially compared to the RP-based methods.
+We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements (Table \ref{deletionRepmaskpreds}) in both the simulated and NA18507 data sets, especially compared to the RP-based methods.
 In the simulated data set, Cloudbreak correctly identifies 9\% more repeat-overlapping deletions than the nearest competitor.
@@
   \hline
-  Cloudbreak  & \textbf{29} (4) & \textbf{155} (\textbf{19}) & \textbf{204} (\textbf{38}) & \textbf{761} (158) \\ 
-  Breakdancer & \textbf{29} (\textbf{6}) & 142 (10) & 186 (25) & 751 (42) \\
-  GASVPro     & 15 (2) & 72 (2) & 71 (5) & 264 (8) \\
-  DELLY       & 21 (2) & 116 (2) & 147 (12) & 618 (11) \\
-  Pindel      & 8 (4) & 62 (12) & 98 (26) & 563 (\textbf{264}) \\ 
+  Cloudbreak  & \textbf{  29} (   4) & \textbf{ 155} (\textbf{  19}) & \textbf{ 204} (\textbf{  38}) & \textbf{ 761} ( 158) \\ 
+  Breakdancer & \textbf{  29} (\textbf{   6}) &  142 (  10) &  186 (  25) &  751 (  42) \\
+  GASVPro     &   15 (   2) &   72 (   2) &   71 (   5) &  264 (   8) \\
+  DELLY       &   21 (   2) &  116 (   2) &  147 (  12) &  618 (  11) \\
+  Pindel      &    8 (   4) &   62 (  12) &   98 (  26) &  563 (\textbf{ 264}) \\ 
    \hline
@@
 \caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
-\label{repmaskpreds}
+\label{deletionRepmaskpreds}
 \end{table}

### 2013-02-28 src/main/scripts/manuscript.R

@@ -3,7 +3,6 @@ library(rtracklayer)
 library(plyr)
 library(RColorBrewer)
 library(VariantAnnotation)
-library(xtable)
 library(brew)
 
 drawPerfLine <- function(perf, col, lineType, maxFP) {
@@ -71,9 +70,7 @@ processInsertionPredictions <- function(name, predfile, trueInsertions) {
   maxhits <- ddply(predswithhits,.(predchrom,predstart,predend),function(x){x[which.max(x$overlaps),]})
   
   maxhits$tok <- paste(maxhits$truechrom, ':', maxhits$truestart, '-', maxhits$trueend, sep="")
-  print(head(maxhits))
   maxhits <- merge(maxhits,mcols(trueInsertions), by="tok")
-  print(head(maxhits))  
   maxhits$name <- name
   maxhits
 }
@@ -81,18 +78,18 @@ processInsertionPredictions <- function(name, predfile, trueInsertions) {
 printPredTableValue <- function(t, x, y) {
   d <- t[x,y]
   if (t[x,y] == max(t[,y])) {
-    return(paste0("\\textbf{",formatC(d,digits=3,format="f"),"}"))
+    return(paste0("\\textbf{",formatC(d,digits=3,format="fg"),"}"))
   } else {
-    return(format(d,digits=3,format="f"))
+    return(formatC(d,digits=3,format="fg"))
   }
 }
 
 printPredTableValueFromList <- function(x, l) {
   d <- l[x]
   if (d == max(l)) {
-    return(paste0("\\textbf{",formatC(d,digits=3,format="f"),"}"))
+    return(paste0("\\textbf{",formatC(d,digits=3,format="fg"),"}"))
   } else {
-    return(formatC(d,digits=3,format="f"))
+    return(formatC(d,digits=3,format="fg"))
   }
 }
 
@@ -100,16 +97,82 @@ printPredTableCell <- function(preds, ex, x, y) {
   return(paste0(printPredTableValue(preds, x, y), " (", printPredTableValue(ex, x, y), ")"))
 }
 
+# input files
+
+# chr2
+
+# deletions
[…]

### 2013-02-28 manuscript/insertionRepmask.table.brew.tex
finish adding first cut at insertion accurracy to manuscript

\begin{table}[b]
\begin{center}
\begin{tabular}{rrr|rr}
 & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
 Total Number & <%= chr2Totals['0'] %> & <%= chr2Totals['1'] %> & <%= NA18507Totals['0'] %> & <%= NA18507Totals['1'] %> \\ 
  \hline
  Cloudbreak  & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Cloudbreak', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Cloudbreak', '1') %> \\ 
  Breakdancer & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Breakdancer', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Breakdancer', '1') %> \\
  Pindel      & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '1') %> \\ 
   \hline
\end{tabular}
\end{center}
\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
\label{insertionRepmaskpreds}
\end{table}

### 2013-02-28 manuscript/manuscript.tex
finish adding first cut at insertion accurracy to manuscript

@@
 In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings.
-We then demonstrate the development in this framework of an algorithm, Cloudbreak, for discovering genomic deletions.
+We then demonstrate the development in this framework of an algorithm, Cloudbreak, for discovering genomic deletions and short insertions.
 Cloudbreak computes local features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
@@
 
-\subsection{Cloudbreak: A Hadoop Deletion Detection Algorithm}
+\subsection{Cloudbreak: A Hadoop Indel Detection Algorithm}
 
-We have implemented a deletion detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak.
+We have implemented an indel detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak.
 Our implementation defines the three application-specific functions described above as:
 \begin{description}
-\item[\sc{Loci}] Because we are detecting deletions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
+\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
 For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
@@
 
-\item[\sc{PostProcess}] We convert our features along the genome to deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
+\item[\sc{PostProcess}] We convert our features along the genome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
 To eliminate noise we apply a median filter with window size 5.
@@
 
-We compared the performance of Cloudbreak for detecting deletions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
+We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
+All of these methods detect deletions; in addition Breakdancer and Pindel are capable of detecting insertions.
 In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).
 
-We use the following criteria to define a true prediction given a gold standard set of deletions to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method.
+We use the following criteria to define a true prediction given a gold standard set of deletion and insertion variants to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method.
+For evaluating insertions, each algorithm produces insertion predictions that define an interval in which the insertion is predicted to have occurred with start and end coordinates $s$ and $e$ as well as the predicted length of the insertion, $l$.
+The true insertions are defined in terms of their actual insertion coordinate $i$ and their actual length $l_a$.
+Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$.
 Furthermore, we find that Pindel is very accurate at discovering deletions less than or equal to 40bp in length, so we exclude from consideration calls that match a true deletion of less than 40bp where the call is less than or equal to 75bp in length.
@@
 Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro, and Pindel - number of supporting read pairs.}
-\label{chr2roc}
+\label{chr2DeletionsRoc}
 \end{figure}
 
-Figure \ref{chr2roc} shows an ROC plot of the performance of each algorithm on the simulated data set.
+Figure \ref{chr2DeletionsRoc} shows an ROC plot of the performance of each algorithm for detecting deletions on the simulated data set.
 All approaches show excellent specificity at high thresholds in this simulation.
 Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.
-Table \ref{chr2DeletionPreds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 0.1 based on the ROC curve, representing cutoffs of 2.1 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and 33 supporting read pairs for Pindel.
+Table \ref{chr2DeletionPreds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 10\% based on the ROC curve, representing cutoffs of 2.1 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and 33 supporting read pairs for Pindel.
 Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
@@
 
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.75\textwidth]{CHR2SIM_INS_ROC.pdf}
+\caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
+\label{chr2InsertionsRoc}
+\end{figure}
+
+An ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}.
+Cloudbreak shows excellent specificity and greater sensitivity than Breakdancer, the other read-pair based method that predicts insertions.
+The number of insertions identified by each algorithm at maximum sensitivity, again broken down by size, is shown in Table \ref{chr2InsertionPreds}.
+
+\begin{table}[t]
+\begin{center}
+\begin{tabular}{rrrrrr}
+  \hline
+ & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number & 199 &  83 & 79 &  21 & 21\\ 
+  \hline
+  Cloudbreak  & \textbf{  49} (  25)  &    5 (   3) &     2 (   1) & 0 (0) & 0 (0) \\ 
+  Breakdancer &   23 (  10)  & \textbf{  14} (\textbf{  11}) &     2 (   1) & 0 (0) & 0 (0) \\ 
+  Pindel      &   47 (\textbf{  30})  &    7 (   4) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{The number of simulated insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those insertions that were discovered exclusively by each tool (in parentheses).
+The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.}
+\label{chr2InsertionPreds}
+\end{table}
+
 \subsection{Tests with Biological Data}
@@
 
-To create a gold standard set of deletions to test against, we pooled annotated deletions discovered by three previous studies on the same sample.
-These included data from the Human Genome Structural Variation Project\footnote{\url{http://hgsv.washington.edu/general/download/SNPs_DIPs/}} reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
-We merged any overlapping calls into the region spanned by their unions.
+To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same sample.
+These included data from the Human Genome Structural Variation Project\footnote{\url{http://hgsv.washington.edu/general/download/SNPs_DIPs/}} reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
+We merged any overlapping calls of the same type into the region spanned by their unions.
 It should be noted that the 1000 Genomes call set was partially produced using DELLY and Breakdancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor.
@@
 \caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
-\label{NA18507roc}
+\label{NA18507DeletionsRoc}
 \end{figure}
 
-Figure \ref{NA18507roc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard deletion set, using the same rules for a match as described in the previous section.
+Figure \ref{NA18507DeletionsRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard deletion set, using the same rules for a match as described in the previous section.
 All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing deletions in the gold standard set that are actually present in the sample.
@@
 
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.75\textwidth]{NA18507_INS_ROC.pdf}
+\caption{ROC curve of insertion prediction performance on the 37X NA18507 sample.}
+\label{NA18507InsertionsRoc}
+\end{figure}
+
+A similar ROC-like curve for insertions on the NA18507 data set is shown in Figure \ref{NA18507InsertionsRoc}.
+Breakdancer and Cloudbreak show similar performance at higher thresholds near the lower left corner of the plot; however, Cloudbreak is able to detect more variants at lower thresholds.
+The number of insertions detected at maximum sensitivity for the three algorithms is shown in Table \ref{NA18507InsertionPreds}, along with their precision and recall at that threshold.
+
+\begin{table}[t]
+\begin{center}
+\begin{tabular}{rrr|rrrrr}
+  \hline
+ & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp \\ 
+Total Number & & & 536 & 114 & 45 & 1 \\
+  \hline
+Cloudbreak & 0.000626 & \textbf{0.412} & \textbf{ 265} (\textbf{  95})  &   20 (   8) &     2 (   1) & 0 (0) \\ 
+Breakdancer & \textbf{0.0267} & 0.181 &   97 (  16)  & \textbf{  27} (\textbf{  13}) &     2 (   1) & 0 (0) \\  
+  Pindel & 0.00793 & 0.239 &  144 (  29)  &   14 (  12) &  \textbf{   7} (\textbf{   7}) & \textbf{   1} (\textbf{   1})  \\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{The precision and recall with respect to the gold standard set of insertions for each tool on the NA18507 data, as well as the number of insertions in each size class found, at maximum sensitivity.}
+\label{NA18507InsertionPreds}
+\end{table}
+
 \subsection{Performance in Repetitive Areas}
@@
 Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
+Performance on insertions is displayed in Table \ref{insertionRepmaskpreds}.
 
@@
 
-\subsection{Genotyping Deletions}
+\begin{table}[b]
+\begin{center}
+\begin{tabular}{rrr|rr}
+ & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
+\hline
+ &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
+ Total Number & 133 & 270 & 341 & 355 \\ 
+  \hline
+  Cloudbreak  &   16 (   4) &   40 (  25) & \textbf{ 146} (\textbf{  44}) & \textbf{ 141} (\textbf{  60}) \\ 
+  Breakdancer &   17 (   9) &   22 (  13) &   82 (  17) &   44 (  13) \\
+  Pindel      & \textbf{  26} (\textbf{  20}) & \textbf{  50} (\textbf{  36}) &   84 (  29) &   82 (  20) \\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
+\label{insertionRepmaskpreds}
+\end{table}
 
-Since Cloudbreak explicitly fits a mixture of insert size distributions at each location, and each component of the mixture represents an allele containing or not containing a deletion, it should be possible to use the parameters of the fit GMM to infer the genotype of each predicted variant, assuming that our pipeline is capturing all relevant read mappings near the locus of the variant.
-To test this, we attempted to classify the true positive Cloudbreak deletion predictions described above for both the simulated and biological data sets.
+\subsection{Genotyping Variants}
+
+Since Cloudbreak explicitly fits a mixture of insert size distributions at each location, and each component of the mixture represents an allele containing or not containing an insertion or deletion, it should be possible to use the parameters of the fit GMM to infer the genotype of each predicted variant, assuming that our pipeline is capturing all relevant read mappings near the locus of the variant.
+To test this, we attempted to classify the true positive Cloudbreak predictions described above for both the simulated and biological data sets.
 Because we simulated a diploid genome, we had genotypes for the simulated data readily available.
-For the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
+For benchmarking genotyyping of deletions in the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
 We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes.
 By setting a simple cutoff of .2 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.0\% and 94.9\% accuracy in predicting the genotype of the true positive deletions we detected in the simulated and real data sets, respectively.
-Table \ref{genotypeaccuracy} shows confusion matrices for the two samples using this classifier.
+Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
+For insertions, none of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertion rates that meet our size threshold that also had genotyping information.
+Of the 56 insertions detected by Cloudbreak on the simulated data set, only 4 were heterozygous.
+Using another simple cutoff on $\alpha$, this time with a value of .3, Cloudbreak correctly classified 50 of the 52 homozygous insertions and 3 of the four heterozygous insertions discovered.
 
@@
 \caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
-\label{genotypeaccuracy}
+\label{deletionGenotypeaccuracy}
 \end{table}

### 2013-02-28 src/main/scripts/manuscript.R

@@ -172,7 +172,7 @@ gasvChr2DeletionPerf <- read.table(gasvProChr2DeletionPerfFile, header=TRUE)
 dellyChr2DeletionPerf <- read.table(dellyChr2DeletionPerfFile, header=TRUE)
 pindelChr2DeletionPerf <- read.table(pindelChr2DeletionPerfFile, header=TRUE)
 
-perfsList <- list(cloudbreak=cloudbreakChr2DeletionPerf, breakdancer=breakdancerChr2DeletionPerf, pindel=pindelChr2DeletionPerf, gasv=gasvChr2DeletionPerf, delly=dellyChr2DeletionPerf, pindel=pindelChr2DeletionPerf)
+perfsList <- list(cloudbreak=cloudbreakChr2DeletionPerf, breakdancer=breakdancerChr2DeletionPerf, pindel=pindelChr2DeletionPerf, gasv=gasvChr2DeletionPerf, delly=dellyChr2DeletionPerf)
 pdf('~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf', width=10)
 par(xpd=T, mar=par()$mar+c(0,0,0,7))
 plotROC(perfsList, 
@@ -360,7 +360,7 @@ NA18507dellyDelsPerf <- read.table(dellyNA18507DeletionPerfFile, header=TRUE)
 NA18507pindelDelsPerf <- read.table(pindelNA18507DeletionPerfFile, header=TRUE)
 
 perfsList <- list(cloudbreak=NA18507cloudbreakDelsPerf, breakdancer=NA18507breakdancerDelsPerf, pindel=NA18507pindelDelsPerf, gasv=NA18507gasvDelsPerf, delly=NA18507dellyDelsPerf)
-pdf('~/Documents/svpipeline/manuscript/NA18507_ROC.pdf', width=10)
+pdf('~/Documents/svpipeline/manuscript/NA18507_DELS_ROC.pdf', width=10)
 par(xpd=T, mar=par()$mar+c(0,0,0,7))
 plotROC(perfsList, c("Cloudbreak", "Breakdancer","Pindel","GASVPro", "DELLY", "Cloudbreak"), totalDels, "NA18507", legendLoc=xy.coords(10750,700), maxTP=1300, sim=FALSE)
 dev.off()
[…]

### 2013-03-04 manuscript/chr2DeletionPredsMaxSensitivity.table.brew.tex
manuscript analysis complete runthru

\begin{table}[t]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
  \hline
  Cloudbreak  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\ 
  Breakdancer & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\ 
  GASVPro     & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\ 
  DELLY       & <%= printPredTableCell(preds, exPreds, 'DELLY', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '> 1000bp') %> \\ 
  Pindel      & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\ 
  MoDIL      & <%= printPredTableCell(preds, exPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '> 1000bp') %>\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPreds}
\end{table}

### 2013-03-04 src/main/scripts/manuscript.R

@@ -75,6 +75,10 @@ processInsertionPredictions <- function(name, predfile, trueInsertions) {
   maxhits
 }
 
+numPredictionsAtLevel <- function(perfTable, threshold) {
+  return(perfTable[perfTable$Thresh == min(perfTable[perfTable$Thresh >= threshold,'Thresh']),'Calls'])
+}
+
 printPredTableValue <- function(t, x, y) {
   d <- t[x,y]
   if (t[x,y] == max(t[,y])) {
@@ -102,67 +106,81 @@ printPredTableCell <- function(preds, ex, x, y) {
 # chr2
 
 # deletions
-cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/cloudbreak_venterchr2allindels100bpdip_readgroup1_mt0.8_del_perf.txt'
+cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/cloudbreak_venterchr2allindels100bpdip_readgroup1_del_mt0.8_mfw5_gte40.perf.txt'
 breakdancerChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_35_2_3_del.perf.txt'
-gasvProChr2DeletionPerfFile <-  '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gasvpro.perf.txt'
-dellyChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_del.perf.txt'
-pindelChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_pindel_D.perf.txt'
+gasvProChr2DeletionPerfFile <-  '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gasvpro_gte40.perf.txt'
[…]

### 2013-03-13 manuscript/manuscript.tex
updated numbers for all indels, insertions, max sensitivity

@@
 We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
-All of these methods detect deletions; in addition Breakdancer and Pindel are capable of detecting insertions.
+We also attempted to evaluate MoDIL on the same data.
+All of these methods detect deletions; in addition Breakdancer, Pindel, and MoDIL are capable of detecting insertions.
 In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).
@@
 \centering
-\includegraphics[width=0.75\textwidth]{CHR2SIM_DELS_ROC.pdf}
+\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf}
 \caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes.
-Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro, and Pindel - number of supporting read pairs.}
+Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.}
 \label{chr2DeletionsRoc}
@@
 Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.
-Table \ref{chr2DeletionPreds} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 10\% based on the ROC curve, representing cutoffs of 2.1 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and 33 supporting read pairs for Pindel.
+
+Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the perfomance characteristics of algorithms at varying levels.
+First, we used the results on simulated deletions to characterize the predictions made by each algorithm when a low false discovery rate is required.
+Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 10\% based on the ROC curve, representing cutoffs of 2.18 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 Pindel.
 Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
+The output which we obtained from MoDIL did not have a threshold or operating point that correlated with the tradeoff between precision and recall; therefore it is not included in this table.
 
@@
   \hline
-  Cloudbreak  &   51 (  10)  &   51 (\textbf{   3}) &  \textbf{  55} (\textbf{   5}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{   1}) \\ 
-  Breakdancer & \textbf{  52} (  14)  &   49 (   2) &    49 (0) &    7 (0) &   14 (0) \\ 
-  GASVPro     &   31 (   4)  &   25 (0) &    23 (0) &    2 (0) &    6 (0) \\ 
-  DELLY       &   22 (   2)  & \textbf{  55} (   2) &    40 (0) &    8 (0) &   12 (0) \\ 
-  Pindel      &   29 (\textbf{  16})  &   11 (0) &    21 (0) &    1 (0) &    8 (0)\\ 
+  Cloudbreak  &   47 (   7)  &   50 (   2) &  \textbf{  55} (\textbf{   4}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{0}) \\ 
+  Breakdancer &   52 (  10)  &   49 (   2) &    49 (0) &    7 (0) &   14 (\textbf{0}) \\ 
+  GASVPro     &   31 (   4)  &   25 (0) &    23 (0) &    2 (0) &    6 (\textbf{0}) \\ 
+  DELLY       &   22 (   2)  & \textbf{  56} (\textbf{   3}) &    40 (0) &    8 (0) &   12 (\textbf{0}) \\ 
+  Pindel      & \textbf{  60} (\textbf{  35})  &   16 (0) &    41 (   2) &    1 (0) &   12 (\textbf{0})\\ 
    \hline
@@
 The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPreds}
+\label{chr2DeletionPredsFDR10}
+\end{table}
+
+We also characterized the predictions made by each algorithm at the operating point which gives them maximum sensitivity.
+For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.
+The results are summarized in Table \ref{chr2DeletionPredsMaxSensitivity}.
+MoDIL has the greatest sensitivity but the precision is extremely low.
+
+\begin{table}[t]
+\begin{center}
+\begin{tabular}{rrrrrr}
+  \hline
+ & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number & 224 &  84 & 82 &  31 & 26\\ 
+  \hline
+  Cloudbreak  & \textbf{ 155} (\textbf{  11})  &   57 (\textbf{   3}) &    65 (   2) &   15 (0) &   15 (0) \\ 
+  Breakdancer &   89 (0)  &   54 (0) &    53 (0) &    8 (0) &   15 (0) \\ 
+  GASVPro     &   83 (   2)  &   32 (0) &    55 (0) &    8 (0) &   15 (0) \\ 
+  DELLY       &  114 (   7)  & \textbf{  68} (   2) &  \textbf{  66} (   1) &    9 (0) &   17 (   1) \\ 
+  Pindel      &   96 (  10)  &   24 (0) &    48 (0) &    5 (0) &   15 (0)\\ 
+  MoDIL      &  124 (   6)  &   66 (\textbf{   3}) &  \textbf{  66} (\textbf{   6}) & \textbf{  17} (\textbf{   4}) & \textbf{  23} (\textbf{   8})\\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
+The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.
+TODO: Add precision and recall!!!!!}
+\label{chr2DeletionPredsMaxSensitivity}
 \end{table}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{CHR2SIM_INS_ROC.pdf}
+\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf}
 \caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
@@
   \hline
-  Cloudbreak  & \textbf{  49} (  25)  &    5 (   3) &     2 (   1) & 0 (0) & 0 (0) \\ 
-  Breakdancer &   23 (  10)  & \textbf{  14} (\textbf{  11}) &     2 (   1) & 0 (0) & 0 (0) \\ 
-  Pindel      &   47 (\textbf{  30})  &    7 (   4) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
+  Cloudbreak  & \textbf{  65} (\textbf{  28})  &    9 (   5) &     3 (   2) &    1 (   1) & 0 (0) \\ 
+  Breakdancer &   23 (   8)  & \textbf{  14} (\textbf{  10}) &     2 (   1) & 0 (0) & 0 (0) \\ 
+  Pindel      &   52 (  26)  &    5 (   4) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
    \hline
@@
 \caption{The number of simulated insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those insertions that were discovered exclusively by each tool (in parentheses).
-The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.}
+The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.
+TODO: Add modil!!!!
+TODO: Add precision and recall!!!}
 \label{chr2InsertionPreds}
@@
 
+We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.
+
 \begin{figure}[t]
 \centering
-\includegraphics[width=0.75\textwidth]{NA18507_DELS_ROC.pdf}
+\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/NA18507/NA18507_DELS_ROC.pdf}
 \caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
@@
   \hline
-Cloudbreak & 0.0978 & \textbf{0.115} & \textbf{ 423} ( 163)  &  128 (\textbf{   9}) &   158 (\textbf{   8}) &   70 (\textbf{   4}) &  186 (  12) \\ 
-Breakdancer & 0.119 & 0.111 &  258 (  42)  & \textbf{ 132} (   8) &  \textbf{ 167} (   2) & \textbf{  92} (   1) & \textbf{ 288} (\textbf{  14}) \\  
-  GASVPro & \textbf{0.132} & 0.0399 &  102 (  11)  &   37 (   2) &    77 (0) &   26 (0) &   93 (0) \\ 
-  DELLY & 0.0813 & 0.091 &  143 (  10)  &  125 (   8) &   158 (   1) &   83 (   1) &  256 (   3) \\ 
-  Pindel & 0.0134 & 0.0787 &  406 (\textbf{ 290})  &   40 (0) &    90 (0) &   37 (0) &   88 (0) \\ 
+Cloudbreak & 0.0978 & \textbf{0.115} & \textbf{ 423} (\textbf{ 179})  &  128 (\textbf{   9}) &   158 (\textbf{   8}) &   70 (\textbf{   3}) &  186 (\textbf{  12}) \\ 
+Breakdancer & 0.122 & 0.112 &  261 (  41)  & \textbf{ 132} (   8) &  \textbf{ 167} (   1) & \textbf{  92} (0) & \textbf{ 288} (  10) \\  
+  GASVPro & 0.134 & 0.0401 &  104 (  17)  &   37 (   2) &    77 (0) &   26 (0) &   93 (0) \\ 
+  DELLY & 0.0824 & 0.091 &  143 (   9)  &  125 (   7) &   158 (   1) &   83 (   1) &  256 (   3) \\ 
+  Pindel & \textbf{0.16} & 0.0685 &  149 (  12)  &   57 (0) &   140 (   1) &   58 (0) &  172 (   2) \\ 
    \hline
@@
 \caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found.
-The same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPreds}, adjusted for the difference in coverage from 30X to 37X.}
+The same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.}
 \label{NA18507DeletionPreds}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{NA18507_INS_ROC.pdf}
+\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/NA18507/NA18507_INS_ROC.pdf}
 \caption{ROC curve of insertion prediction performance on the 37X NA18507 sample.}
@@
   \hline
-Cloudbreak & 0.000626 & \textbf{0.412} & \textbf{ 265} (\textbf{  95})  &   20 (   8) &     2 (   1) & 0 (0) \\ 
-Breakdancer & \textbf{0.0267} & 0.181 &   97 (  16)  & \textbf{  27} (\textbf{  13}) &     2 (   1) & 0 (0) \\  
-  Pindel & 0.00793 & 0.239 &  144 (  29)  &   14 (  12) &  \textbf{   7} (\textbf{   7}) & \textbf{   1} (\textbf{   1})  \\ 
+Cloudbreak & 0.0365 & 0.198 &  127 (  28)  &   11 (   3) &  0 (0) & 0 (0) \\ 
+Breakdancer & 0.0104 & 0.181 &   97 (  36)  & \textbf{  27} (\textbf{  17}) &     2 (   2) & 0 (0) \\  
+  Pindel & \textbf{0.0387} & \textbf{0.239} & \textbf{ 144} (\textbf{  76})  &   14 (  12) &  \textbf{   7} (\textbf{   7}) & \textbf{   1} (\textbf{   1})  \\ 
    \hline
@@
   \hline
-  Cloudbreak  & \textbf{  29} (   4) & \textbf{ 155} (\textbf{  19}) & \textbf{ 204} (\textbf{  38}) & \textbf{ 761} ( 158) \\ 
-  Breakdancer & \textbf{  29} (\textbf{   6}) &  142 (  10) &  186 (  25) &  751 (  42) \\
-  GASVPro     &   15 (   2) &   72 (   2) &   71 (   5) &  264 (   8) \\
-  DELLY       &   21 (   2) &  116 (   2) &  147 (  12) &  618 (  11) \\
-  Pindel      &    8 (   4) &   62 (  12) &   98 (  26) &  563 (\textbf{ 264}) \\ 
+  Cloudbreak  &   28 (   4) & \textbf{ 151} (  13) & \textbf{ 204} (\textbf{  46}) & \textbf{ 761} (\textbf{ 165}) \\ 
+  Breakdancer & \textbf{  29} (   5) &  142 (   7) &  186 (  21) &  754 (  39) \\
+  GASVPro     &   15 (   2) &   72 (   2) &   71 (   6) &  266 (  13) \\
+  DELLY       &   21 (   2) &  117 (   3) &  147 (  11) &  618 (  10) \\
+  Pindel      &   18 (\textbf{   9}) &  112 (\textbf{  28}) &  103 (   4) &  473 (  11) \\ 
    \hline
@@
   \hline
-  Cloudbreak  &   16 (   4) &   40 (  25) & \textbf{ 146} (\textbf{  44}) & \textbf{ 141} (\textbf{  60}) \\ 
-  Breakdancer &   17 (   9) &   22 (  13) &   82 (  17) &   44 (  13) \\
-  Pindel      & \textbf{  26} (\textbf{  20}) & \textbf{  50} (\textbf{  36}) &   84 (  29) &   82 (  20) \\ 
+  Cloudbreak  &   19 (   5) & \textbf{  59} (  31) &   73 (  12) &   65 (  19) \\ 
+  Breakdancer &   17 (   8) &   22 (  11) &   82 (  37) &   44 (  18) \\
+  Pindel      & \textbf{  25} (\textbf{  19}) &   54 (\textbf{  33}) & \textbf{  84} (\textbf{  48}) & \textbf{  82} (\textbf{  48}) \\ 
    \hline
@@
 We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes.
-By setting a simple cutoff of .2 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.0\% and 94.9\% accuracy in predicting the genotype of the true positive deletions we detected in the simulated and real data sets, respectively.
+By setting a simple cutoff of .2 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 86.7\% and 94.9\% accuracy in predicting the genotype of the true positive deletions we detected in the simulated and real data sets, respectively.
 Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
@@
 \cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 91 & 3 &  70 & 11 \\
- & Heterozygous & 19 & 71 &  4 & 209 \\
+\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 88 & 3 &  70 & 11 \\
+ & Heterozygous & 18 & 70 &  4 & 209 \\
 \cline{2-6}

### 2013-03-13 manuscript/chr2DeletionPredsMaxSensitivity.table.brew.tex
update chr2 max sensitivity table to have precision, recall

@@
 \begin{center}
-\begin{tabular}{rrrrrr}
+\begin{tabular}{rrr|rrrrr}
   \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
+                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number &          &           & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
   \hline
-  Cloudbreak  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\
-  Breakdancer & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\
-  GASVPro     & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\
-  DELLY       & <%= printPredTableCell(preds, exPreds, 'DELLY', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '> 1000bp') %> \\
-  Pindel      & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\
-  MoDIL      & <%= printPredTableCell(preds, exPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '> 1000bp') %>\\
+  Cloudbreak    &  <%= printPredTableValueFromList('Cloudbreak', precision) %> & <%= printPredTableValueFromList('Cloudbreak', recall) %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\
+  Breakdancer   &  <%= printPredTableValueFromList('Breakdancer', precision) %> & <%= printPredTableValueFromList('Breakdancer', recall) %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\
+  GASVPro        & <%= printPredTableValueFromList('GASVPro', precision) %> & <%= printPredTableValueFromList('GASVPro', recall) %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\
+  DELLY           & <%= printPredTableValueFromList('DELLY', precision) %> & <%= printPredTableValueFromList('DELLY', recall) %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '> 1000bp') %> \\
+  Pindel           & <%= printPredTableValueFromList('Pindel', precision) %> & <%= printPredTableValueFromList('Pindel', recall) %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\
+  MoDIL           & <%= printPredTableValueFromList('MoDIL', precision) %>  & <%= printPredTableValueFromList('MoDIL', recall) %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '> 1000bp') %>\\
    \hline
@@
 \end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
+\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivy, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
 The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPreds}
+\label{chr2DeletionPredsMaxSensitivity}
 \end{table}

### 2013-03-13 src/main/scripts/manuscript.R

@@ -106,7 +106,7 @@ printPredTableCell <- function(preds, ex, x, y) {
 # chr2
 
 # deletions
-cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/cloudbreak_venterchr2allindels100bpdip_readgroup1_del_mt0.8_mfw5_gte40.perf.txt'
+cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/extract_in_hadoop_dels_oldsplot_perf.txt'
 breakdancerChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_35_2_3_del.perf.txt'
 gasvProChr2DeletionPerfFile <-  '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gasvpro_gte40.perf.txt'
 dellyChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_gte40.perf.txt'
@@ -120,7 +120,10 @@ gasvProChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/vente
 dellyChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_gte40_gt7.hits.txt'
 pindelChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_pindel_D_gte40_gt13.hits.txt'
 
+chr2MaxSensitivityDelCutoffs=list(cloudbreak=0.84, breakdancer=2, GASVPro=4, DELLY=2, Pindel=4)
+
[…]

### 2013-03-14 manuscript/chr2InsertionPredsBySize.table.brew.tex
fixed up modil a bit; added to chr2 insertions table

@@
 \begin{center}
-\begin{tabular}{rrrrrr}
+\begin{tabular}{rrr|rrrrr}
   \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
+                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number &          &           & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
   \hline
-  Cloudbreak  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\
-  Breakdancer & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\
-  Pindel      & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\
+  Cloudbreak   &<%= printPredTableValueFromList('Cloudbreak', precision) %> & <%= printPredTableValueFromList('Cloudbreak', recall) %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\
+  Breakdancer & <%= printPredTableValueFromList('Breakdancer', precision) %> & <%= printPredTableValueFromList('Breakdancer', recall) %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\
+  Pindel          & <%= printPredTableValueFromList('Pindel', precision) %> & <%= printPredTableValueFromList('Pindel', recall) %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\
+  MoDIL          & <%= printPredTableValueFromList('MoDIL', precision) %> & <%= printPredTableValueFromList('MoDIL', recall) %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '> 1000bp') %>\\
    \hline

### 2013-03-14 src/main/scripts/manuscript.R

@@ -123,14 +123,14 @@ pindelChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter
 chr2MaxSensitivityDelCutoffs=list(cloudbreak=0.84, breakdancer=2, GASVPro=4, DELLY=2, Pindel=4)
 
 cloudbreakChr2DeletionPredictionsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/cloudbreak_venterchr2allindels100bpdip_readgroup1_deletions_0p84.bed'
-modilChr2DeletionPredictionsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/modil_deletions.bed'
+modilChr2DeletionPredictionsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/modil_del_gt1.bed'
 
 cloudbreakChr2DeletionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/cloudbreak_venterchr2allindels100bpdip_readgroup1_del_mfw5_gte40_gt0p50.hits.txt'
 breakdancerChr2DeletionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_35_2_3_del.gt2_gte40.hits.txt'
 gasvProChr2DeletionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gasvpro_gte40_gt4.hits.txt'
 dellyChr2DeletionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_gte40_gt2.hits.txt'
 pindelChr2DeletionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_pindel_D_gte40_gt4.hits.txt'
[…]

### 2013-03-14 src/main/scripts/manuscript.R
add modil to repmask

@@ -140,6 +140,8 @@ pindelChr2InsertionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr
 cloudbreakChr2InsertionPredictionsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/cloudbreak_venterchr2allindels100bpdip_readgroup1_insertions_0p27929.bed'
 modilChr2InsertionPredictionsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/modil_ins_gte1.bed'
 
+chr2MaxSensitivityInsCutoffs=list(cloudbreak=0.27929, breakdancer=2, Pindel=4)
+
 cloudbreakChr2InsertionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/cloudbreak_venterchr2allindels100bpdip_readgroup1_ins_mfw5_gte40_gt0p27929.hits.txt'
 breakdancerChr2InsertionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_35_2_3.bd_out.ins_gte40.gt2.hits.txt'
 pindelChr2InsertionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/pindel_insertions.gte40.gt4.hits.txt'

### 2013-03-18 manuscript/manuscript.tex
updated to use extract-in-hadoop calls

@@
   \hline
-  Cloudbreak  &   47 (   7)  &   50 (   2) &  \textbf{  55} (\textbf{   4}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{0}) \\ 
-  Breakdancer &   52 (  10)  &   49 (   2) &    49 (0) &    7 (0) &   14 (\textbf{0}) \\ 
+  Cloudbreak  &   50 (   8)  &   50 (\textbf{   3}) &  \textbf{  55} (\textbf{   4}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{0}) \\ 
+  Breakdancer &   52 (   9)  &   49 (   2) &    49 (0) &    7 (0) &   14 (\textbf{0}) \\ 
   GASVPro     &   31 (   4)  &   25 (0) &    23 (0) &    2 (0) &    6 (\textbf{0}) \\ 
   DELLY       &   22 (   2)  & \textbf{  56} (\textbf{   3}) &    40 (0) &    8 (0) &   12 (\textbf{0}) \\ 
-  Pindel      & \textbf{  60} (\textbf{  35})  &   16 (0) &    41 (   2) &    1 (0) &   12 (\textbf{0})\\ 
+  Pindel      & \textbf{  60} (\textbf{  34})  &   16 (0) &    41 (   2) &    1 (0) &   12 (\textbf{0})\\ 
    \hline
@@
 
-\begin{table}[t]
+\begin{table}[b]
 \begin{center}
-\begin{tabular}{rrrrrr}
+\begin{tabular}{rrr|rrrrr}
   \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 224 &  84 & 82 &  31 & 26\\ 
+                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
   \hline
-  Cloudbreak  & \textbf{ 155} (\textbf{  11})  &   57 (\textbf{   3}) &    65 (   2) &   15 (0) &   15 (0) \\ 
-  Breakdancer &   89 (0)  &   54 (0) &    53 (0) &    8 (0) &   15 (0) \\ 
-  GASVPro     &   83 (   2)  &   32 (0) &    55 (0) &    8 (0) &   15 (0) \\ 
-  DELLY       &  114 (   7)  & \textbf{  68} (   2) &  \textbf{  66} (   1) &    9 (0) &   17 (   1) \\ 
-  Pindel      &   96 (  10)  &   24 (0) &    48 (0) &    5 (0) &   15 (0)\\ 
-  MoDIL      &  124 (   6)  &   66 (\textbf{   3}) &  \textbf{  66} (\textbf{   6}) & \textbf{  17} (\textbf{   4}) & \textbf{  23} (\textbf{   8})\\ 
+  Cloudbreak    &  \textbf{0.497} & 0.642 & \textbf{ 140} (   5)  &   58 (\textbf{   4}) &    61 (   1) &   13 (0) &   15 (0) \\ 
+  Breakdancer   &  0.356 & 0.49 &   89 (0)  &   54 (0) &    53 (0) &    8 (0) &   15 (0) \\ 
+  GASVPro        & 0.146 & 0.432 &   83 (   2)  &   32 (0) &    55 (0) &    8 (0) &   15 (0) \\ 
+  DELLY           & 0.457 & 0.613 &  114 (   8)  & \textbf{  68} (   2) &  \textbf{  66} (   1) &    9 (0) &   17 (   1) \\ 
+  Pindel           & 0.462 & 0.421 &   96 (\textbf{  14})  &   24 (0) &    48 (0) &    5 (0) &   15 (0)\\ 
+  MoDIL           & 0.132  & \textbf{0.66} &  123 (   6)  &   66 (   2) &  \textbf{  66} (\textbf{   8}) & \textbf{  17} (\textbf{   6}) & \textbf{  23} (\textbf{   8})\\ 
    \hline
@@
 \end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.
-TODO: Add precision and recall!!!!!}
+\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivy, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
+The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
 \label{chr2DeletionPredsMaxSensitivity}
@@
 \begin{center}
-\begin{tabular}{rrrrrr}
+\begin{tabular}{rrr|rrrrr}
   \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 199 &  83 & 79 &  21 & 21\\ 
+                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
   \hline
-  Cloudbreak  & \textbf{  65} (\textbf{  28})  &    9 (   5) &     3 (   2) &    1 (   1) & 0 (0) \\ 
-  Breakdancer &   23 (   8)  & \textbf{  14} (\textbf{  10}) &     2 (   1) & 0 (0) & 0 (0) \\ 
-  Pindel      &   52 (  26)  &    5 (   4) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
+  Cloudbreak   &\textbf{0.624} & 0.194  & \textbf{  65} (  21)  &    9 (   4) &     3 (   1) &    1 (0) & 0 (0) \\ 
+  Breakdancer & 0.262 & 0.0968  &   23 (   8)  & \textbf{  14} (\textbf{  10}) &     2 (   1) & 0 (0) & 0 (0) \\ 
+  Pindel          & 0.572 & \textbf{0.196} &   52 (\textbf{  25})  &    5 (   3) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
+  MoDIL          & 0.186 & 0.0521 &   14 (   1)  &    4 (   1) &     1 (0) &    2 (   1) & 0 (0)\\ 
    \hline
@@
 \caption{The number of simulated insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those insertions that were discovered exclusively by each tool (in parentheses).
-The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.
-TODO: Add modil!!!!
-TODO: Add precision and recall!!!}
+The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.}
 \label{chr2InsertionPreds}
@@
   \hline
-Cloudbreak & 0.0978 & \textbf{0.115} & \textbf{ 423} (\textbf{ 179})  &  128 (\textbf{   9}) &   158 (\textbf{   8}) &   70 (\textbf{   3}) &  186 (\textbf{  12}) \\ 
-Breakdancer & 0.122 & 0.112 &  261 (  41)  & \textbf{ 132} (   8) &  \textbf{ 167} (   1) & \textbf{  92} (0) & \textbf{ 288} (  10) \\  
+Cloudbreak & 0.101 & \textbf{0.131} & \textbf{ 446} (\textbf{ 186})  & \textbf{ 141} (\textbf{  14}) &  \textbf{ 173} (\textbf{  12}) & \textbf{  93} (\textbf{   4}) &  248 (\textbf{  14}) \\ 
+Breakdancer & 0.122 & 0.112 &  261 (  35)  &  132 (   6) &   167 (   1) &   92 (0) & \textbf{ 288} (   6) \\  
   GASVPro & 0.134 & 0.0401 &  104 (  17)  &   37 (   2) &    77 (0) &   26 (0) &   93 (0) \\ 
-  DELLY & 0.0824 & 0.091 &  143 (   9)  &  125 (   7) &   158 (   1) &   83 (   1) &  256 (   3) \\ 
-  Pindel & \textbf{0.16} & 0.0685 &  149 (  12)  &   57 (0) &   140 (   1) &   58 (0) &  172 (   2) \\ 
+  DELLY & 0.0824 & 0.091 &  143 (   8)  &  125 (   6) &   158 (   1) &   83 (   1) &  256 (   3) \\ 
+  Pindel & \textbf{0.16} & 0.0685 &  149 (  12)  &   57 (0) &   140 (0) &   58 (0) &  172 (   1) \\ 
    \hline
@@
   \hline
-Cloudbreak & 0.0365 & 0.198 &  127 (  28)  &   11 (   3) &  0 (0) & 0 (0) \\ 
-Breakdancer & 0.0104 & 0.181 &   97 (  36)  & \textbf{  27} (\textbf{  17}) &     2 (   2) & 0 (0) \\  
-  Pindel & \textbf{0.0387} & \textbf{0.239} & \textbf{ 144} (\textbf{  76})  &   14 (  12) &  \textbf{   7} (\textbf{   7}) & \textbf{   1} (\textbf{   1})  \\ 
+Cloudbreak & \textbf{0.0596} & 0.197 &  124 (  28)  &   13 (   3) &  0 (0) & 0 (0) \\ 
+Breakdancer & 0.0281 & 0.181 &   97 (  37)  & \textbf{  27} (\textbf{  15}) &     2 (   2) & 0 (0) \\  
+  Pindel & 0.0387 & \textbf{0.239} & \textbf{ 144} (\textbf{  79})  &   14 (  12) &  \textbf{   7} (\textbf{   7}) & \textbf{   1} (\textbf{   1})  \\ 
    \hline
@@
   \hline
-  Cloudbreak  &   28 (   4) & \textbf{ 151} (  13) & \textbf{ 204} (\textbf{  46}) & \textbf{ 761} (\textbf{ 165}) \\ 
-  Breakdancer & \textbf{  29} (   5) &  142 (   7) &  186 (  21) &  754 (  39) \\
+  Cloudbreak  &   26 (   3) & \textbf{ 156} (  16) & \textbf{ 238} (\textbf{  61}) & \textbf{ 863} (\textbf{ 169}) \\ 
+  Breakdancer & \textbf{  29} (   5) &  142 (   6) &  186 (  18) &  754 (  30) \\
   GASVPro     &   15 (   2) &   72 (   2) &   71 (   6) &  266 (  13) \\
-  DELLY       &   21 (   2) &  117 (   3) &  147 (  11) &  618 (  10) \\
-  Pindel      &   18 (\textbf{   9}) &  112 (\textbf{  28}) &  103 (   4) &  473 (  11) \\ 
+  DELLY       &   21 (   2) &  117 (   3) &  147 (  10) &  618 (   9) \\
+  Pindel      &   18 (\textbf{   9}) &  112 (\textbf{  27}) &  103 (   3) &  473 (  10) \\ 
    \hline
@@
   \hline
-  Cloudbreak  &   19 (   5) & \textbf{  59} (  31) &   73 (  12) &   65 (  19) \\ 
-  Breakdancer &   17 (   8) &   22 (  11) &   82 (  37) &   44 (  18) \\
-  Pindel      & \textbf{  25} (\textbf{  19}) &   54 (\textbf{  33}) & \textbf{  84} (\textbf{  48}) & \textbf{  82} (\textbf{  48}) \\ 
+  Cloudbreak  &   19 (   3) & \textbf{  59} (  23) &   75 (  13) &   62 (  18) \\ 
+  Breakdancer &   17 (   8) &   22 (  11) &   82 (  36) &   44 (  18) \\
+  Pindel      & \textbf{  25} (\textbf{  18}) &   54 (\textbf{  32}) & \textbf{  84} (\textbf{  48}) & \textbf{  82} (\textbf{  51}) \\ 
+  MoDIL      &    5 (0) &   16 (   3) & NA & NA \\ 
    \hline
@@
 \cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 88 & 3 &  70 & 11 \\
- & Heterozygous & 18 & 70 &  4 & 209 \\
+\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 92 & 3 &  74 & 12 \\
+ & Heterozygous & 15 & 71 &  4 & 286 \\
 \cline{2-6}

### 2013-03-18 src/main/scripts/manuscript.R

@@ -112,9 +112,9 @@ gasvProChr2DeletionPerfFile <-  '~/Documents/gene_rearrange/svpipeline/venter_ch
 dellyChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_gte40.perf.txt'
 pindelChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_pindel_D_gte40.perf.txt'
 
-cloudbreakChr2DeletionPredictionsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/cloudbreak_venterchr2allindels100bpdip_readgroup1_deletions_2p18.bed'
+cloudbreakChr2DeletionPredictionsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/dels_called_in_hadoop_2p12_3.bed'
 
-cloudbreakChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/cloudbreak_venterchr2allindels100bpdip_readgroup1_del_mfw5_gte40_gt2p18.hits.txt'
+cloudbreakChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/dels_called_in_hadoop_2p12_3.hits.txt'
 breakdancerChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_35_2_3_del.gt4_gte40.hits.txt'
 gasvProChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gasvpro_gte40_gt12.hits.txt'
[…]

### 2013-03-23 manuscript/manuscript.tex
updated for calls extracted in hadoop

@@
 On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
-In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome.
+In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp, and those in repetitive regions of the genome.
 In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples.
@@
 Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
-Traditionally, this involves separating mappings into vthose that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment.
+Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment.
 These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs.
@@
 
-Our formulation divides processing into two distinct MapReduce jobs.
+Our formulation divides processing into three distinct MapReduce jobs.
 The first job finds possible alignment locations for pairs of reads in FASTQ format.
 The second job computes features from the set of alignment pairs that are relevant to each of a set of small intervals covering the genome.
-In a final third stage executed outside of MapReduce, the features for each location are exported from the Hadoop file system and post-processed to produce a set of SV calls which span multiple genomic intervals.
-This MapReduce framework is summarized below with a detailed explanation following.
+In the third stage, the features for all locations on each chromosome are examined to find contiguous blocks of locations that support an SV call.
+This MapReduce framework is summarized in Algorithm 1.
 
+\begin{algorithm}
 \algrenewcommand\algorithmicprocedure{\textbf{job}}
@@
     \EndProcedure
-    \State $\textrm{StructuralVariationCalls } svs \gets \textsc{PostProcess}(\phi_{1,2,\ldots})$ 
+
+    \Procedure{Call SVs}{}
+    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
+    \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
+    \EndFunction
+    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
+    \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
+    \EndFunction
+    \EndProcedure
   \end{algorithmic}
+\label{cloudbreakAlgorithm}
+\caption{The algorithmic framwork for SV calling in MapReduce.}
+\end{algorithm}
 
@@
 the fragment size implied by the alignments and the likelihood the alignments are correct.
-We then leave three functions to be implemented depending on the application:
+We then leave two functions to be implemented depending on the application:
 \begin{flalign*}
@@
  \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
- \textsc{PostProcess} :& \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\} 
 \end{flalign*}
@@
 The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
-Finally, the \textsc{PostProcess} function maps the complete set of features for each locus across the genome into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
 
-We describe one possible set of functions below; however, we believe that there are many possible definitions that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage of the experiment, heterogeneity of the sample, etc.
+Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
+It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
+We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
+
+We describe our implementations of the $\textsc{Loci}$, $\Phi$, and $\textsc{PostProcess}$ functions below.
+However, we believe that there are many possible definitions that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage of the experiment, heterogeneity of the sample, etc.
 
@@
 Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$.
-Furthermore, we find that Pindel is very accurate at discovering deletions less than or equal to 40bp in length, so we exclude from consideration calls that match a true deletion of less than 40bp where the call is less than or equal to 75bp in length.
-We also discard all Pindel calls of less than or equal to 20bp.
+In this study we are interested in detecting events larger than 40bp, because with longer reads, smaller events can be more easily discovered by examining gaps in individual reads.
+Both Pindel and MoDIL make many calls with a predicted event size of under 40bp, so we remove those calls from the output sets of those programs.
+Finally, we exclude from consideration calls from all approaches that match a true deletion of less than 40bp where the predicted variant length is less than or equal to 75bp in length.
+
 It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods.
@@
 
-\begin{figure}[t]
+\begin{figure}[b]
 \centering
@@
 Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.
+The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the tradeoff between precision and recall; therefore it is not included in this figure.
 
@@
 First, we used the results on simulated deletions to characterize the predictions made by each algorithm when a low false discovery rate is required.
-Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when using a threshold that gives an FDR of 10\% based on the ROC curve, representing cutoffs of 2.18 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 12 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 Pindel.
-Cloudbreak shows excellent sensitivity at detecting deletions from each size class, and contributes a number of exclusive true predictions in both small and medium size classes.
-The output which we obtained from MoDIL did not have a threshold or operating point that correlated with the tradeoff between precision and recall; therefore it is not included in this table.
+Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve, representing cutoffs of 2.22 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 11 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 for Pindel.
+Cloudbreak shows good performance at detecting deletions in all size classes, and finds more medium and large deletions than the other approaches.
+Pindel exlusively identifies a large number of small (40bp-100bp) deletions, as is expected for a split-read method.
 
-\begin{table}[t]
+\begin{table}[b]
 \begin{center}
@@
   \hline
-  Cloudbreak  &   50 (   8)  &   50 (\textbf{   3}) &  \textbf{  55} (\textbf{   4}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{0}) \\ 
-  Breakdancer &   52 (   9)  &   49 (   2) &    49 (0) &    7 (0) &   14 (\textbf{0}) \\ 
-  GASVPro     &   31 (   4)  &   25 (0) &    23 (0) &    2 (0) &    6 (\textbf{0}) \\ 
-  DELLY       &   22 (   2)  & \textbf{  56} (\textbf{   3}) &    40 (0) &    8 (0) &   12 (\textbf{0}) \\ 
+  Cloudbreak  &   48 (   8)  &   49 (\textbf{   2}) &  \textbf{  55} (\textbf{   4}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{0}) \\ 
+  Breakdancer &   52 (  10)  &   49 (\textbf{   2}) &    49 (0) &    7 (0) &   14 (\textbf{0}) \\ 
+  GASVPro     &   35 (   5)  &   26 (0) &    26 (0) &    2 (0) &    6 (\textbf{0}) \\ 
+  DELLY       &   22 (   2)  & \textbf{  56} (\textbf{   2}) &    40 (0) &    8 (0) &   12 (\textbf{0}) \\ 
   Pindel      & \textbf{  60} (\textbf{  34})  &   16 (0) &    41 (   2) &    1 (0) &   12 (\textbf{0})\\ 
@@
 The results are summarized in Table \ref{chr2DeletionPredsMaxSensitivity}.
-MoDIL has the greatest sensitivity but the precision is extremely low.
+MoDIL has the greatest recall but low precision.
+Cloudbreak has high precision and recall at this threshold, and surprisingly discovers more small deletions than Pindel.
 
-\begin{table}[b]
+\begin{table}[t]
 \begin{center}
@@
   \hline
-  Cloudbreak    &  \textbf{0.497} & 0.642 & \textbf{ 140} (   5)  &   58 (\textbf{   4}) &    61 (   1) &   13 (0) &   15 (0) \\ 
+  Cloudbreak    &  \textbf{0.507} & 0.649 & \textbf{ 144} (   6)  &   57 (\textbf{   4}) &    61 (   1) &   13 (0) &   15 (0) \\ 
   Breakdancer   &  0.356 & 0.49 &   89 (0)  &   54 (0) &    53 (0) &    8 (0) &   15 (0) \\ 
@@
   DELLY           & 0.457 & 0.613 &  114 (   8)  & \textbf{  68} (   2) &  \textbf{  66} (   1) &    9 (0) &   17 (   1) \\ 
-  Pindel           & 0.462 & 0.421 &   96 (\textbf{  14})  &   24 (0) &    48 (0) &    5 (0) &   15 (0)\\ 
+  Pindel           & 0.462 & 0.421 &   96 (\textbf{  13})  &   24 (0) &    48 (0) &    5 (0) &   15 (0)\\ 
   MoDIL           & 0.132  & \textbf{0.66} &  123 (   6)  &   66 (   2) &  \textbf{  66} (\textbf{   8}) & \textbf{  17} (\textbf{   6}) & \textbf{  23} (\textbf{   8})\\ 
@@
 An ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}.
-Cloudbreak shows excellent specificity and greater sensitivity than Breakdancer, the other read-pair based method that predicts insertions.
 The number of insertions identified by each algorithm at maximum sensitivity, again broken down by size, is shown in Table \ref{chr2InsertionPreds}.
+Cloudbreak shows excellent precision, although recall is low for all four aproaches.
+Cloudbreak again identifies the most small variants.
+Pindel is the only tool which can identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.
 
@@
   \hline
-  Cloudbreak   &\textbf{0.624} & 0.194  & \textbf{  65} (  21)  &    9 (   4) &     3 (   1) &    1 (0) & 0 (0) \\ 
-  Breakdancer & 0.262 & 0.0968  &   23 (   8)  & \textbf{  14} (\textbf{  10}) &     2 (   1) & 0 (0) & 0 (0) \\ 
-  Pindel          & 0.572 & \textbf{0.196} &   52 (\textbf{  25})  &    5 (   3) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
+  Cloudbreak   &\textbf{0.769} & \textbf{0.199}  & \textbf{  67} (  24)  &    9 (   3) &     3 (   1) &    1 (0) & 0 (0) \\ 
+  Breakdancer & 0.262 & 0.0968  &   23 (   8)  & \textbf{  14} (\textbf{   9}) &     2 (   1) & 0 (0) & 0 (0) \\ 
+  Pindel          & 0.572 & 0.196 &   52 (\textbf{  26})  &    5 (   3) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
   MoDIL          & 0.186 & 0.0521 &   14 (   1)  &    4 (   1) &     1 (0) &    2 (   1) & 0 (0)\\ 
@@
 It should be noted that the 1000 Genomes call set was partially produced using DELLY and Breakdancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor.
-
 We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.
@@
 We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
-The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those cutoffs, is shown in Table \ref{NA18507DeletionPreds}.
-Cloudbreak scores well in every category: it identifies more small deletions than any tool except for Pindel, which has low specificity at this cutoff, and contributes more exclusive true positives in larger size classes than any other tool.
+The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those thresholds, is shown in Table \ref{NA18507DeletionPreds}.
+Cloudbreak has the greatest sensitivity of any tool at these thresholds, identifying the most variants in each size class.
 
@@
   \hline
-Cloudbreak & 0.101 & \textbf{0.131} & \textbf{ 446} (\textbf{ 186})  & \textbf{ 141} (\textbf{  14}) &  \textbf{ 173} (\textbf{  12}) & \textbf{  93} (\textbf{   4}) &  248 (\textbf{  14}) \\ 
-Breakdancer & 0.122 & 0.112 &  261 (  35)  &  132 (   6) &   167 (   1) &   92 (0) & \textbf{ 288} (   6) \\  
-  GASVPro & 0.134 & 0.0401 &  104 (  17)  &   37 (   2) &    77 (0) &   26 (0) &   93 (0) \\ 
-  DELLY & 0.0824 & 0.091 &  143 (   8)  &  125 (   6) &   158 (   1) &   83 (   1) &  256 (   3) \\ 
-  Pindel & \textbf{0.16} & 0.0685 &  149 (  12)  &   57 (0) &   140 (0) &   58 (0) &  172 (   1) \\ 
+Cloudbreak & 0.111 & \textbf{0.131} & \textbf{ 383} (\textbf{ 139})  & \textbf{ 146} (\textbf{  13}) &  \textbf{ 178} (\textbf{  10}) & \textbf{  95} (\textbf{   3}) & \textbf{ 300} (\textbf{  16}) \\
+Breakdancer & 0.122 & 0.112 &  261 (  41)  &  132 (   7) &   167 (0) &   92 (0) &  288 (   2) \\  
+  GASVPro & 0.135 & 0.0449 &  120 (  29)  &   38 (   3) &    80 (0) &   29 (0) &  110 (0) \\ 
+  DELLY & 0.0824 & 0.091 &  143 (   7)  &  125 (   5) &   158 (   1) &   83 (   1) &  256 (   3) \\ 
+  Pindel & \textbf{0.16} & 0.0685 &  149 (  12)  &   57 (0) &   140 (0) &   58 (0) &  172 (0) \\ 
    \hline
@@
 A similar ROC-like curve for insertions on the NA18507 data set is shown in Figure \ref{NA18507InsertionsRoc}.
-Breakdancer and Cloudbreak show similar performance at higher thresholds near the lower left corner of the plot; however, Cloudbreak is able to detect more variants at lower thresholds.
 The number of insertions detected at maximum sensitivity for the three algorithms is shown in Table \ref{NA18507InsertionPreds}, along with their precision and recall at that threshold.
+Cloudbreak has the greatest recall, identifying twice as many short insertions as Pindel, while Pindel has the best precision.
 
@@
   \hline
-Cloudbreak & \textbf{0.0596} & 0.197 &  124 (  28)  &   13 (   3) &  0 (0) & 0 (0) \\ 
-Breakdancer & 0.0281 & 0.181 &   97 (  37)  & \textbf{  27} (\textbf{  15}) &     2 (   2) & 0 (0) \\  
-  Pindel & 0.0387 & \textbf{0.239} & \textbf{ 144} (\textbf{  79})  &   14 (  12) &  \textbf{   7} (\textbf{   7}) & \textbf{   1} (\textbf{   1})  \\ 
+Cloudbreak & 0.0258 & \textbf{0.395} & \textbf{ 249} (\textbf{  85})  &   25 (   9) &     1 (0) & 0 (0) \\ 
+Breakdancer & 0.0281 & 0.181 &   97 (  11)  & \textbf{  27} (   9) &     2 (   1) & 0 (0) \\  
+  Pindel & \textbf{0.0387} & 0.239 &  144 (  40)  &   14 (\textbf{  12}) &  \textbf{   7} (\textbf{   7}) & \textbf{   1} (\textbf{   1})  \\ 
    \hline
@@
 We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance.
-We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements (Table \ref{deletionRepmaskpreds}) in both the simulated and NA18507 data sets, especially compared to the RP-based methods.
-In the simulated data set, Cloudbreak correctly identifies 9\% more repeat-overlapping deletions than the nearest competitor.
-Cloudbreak identifies more deletions in repeats than any other method, and in the NA18507 data set finds 3.75 times more exclusive hits than the other RP-based methods.
+We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements in both the simulated and NA18507 data sets.
+Table \ref{deletionRepmaskpreds} shows the number of deletions that overlap repeats at the 10\% FDR cutoff described above.
+In the NA18507 data set, Cloudbreak identifies 16\% more repeat-overlapping variants than the nearest competitor, Breakdancer.
 Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
-Performance on insertions is displayed in Table \ref{insertionRepmaskpreds}.
+Performance on insertions is displayed in Table \ref{insertionRepmaskpreds}; again, Cloudbreak identifies more events that lie in repetitive regions than the other tools.
 
@@
   \hline
-  Cloudbreak  &   26 (   3) & \textbf{ 156} (  16) & \textbf{ 238} (\textbf{  61}) & \textbf{ 863} (\textbf{ 169}) \\ 
-  Breakdancer & \textbf{  29} (   5) &  142 (   6) &  186 (  18) &  754 (  30) \\
-  GASVPro     &   15 (   2) &   72 (   2) &   71 (   6) &  266 (  13) \\
-  DELLY       &   21 (   2) &  117 (   3) &  147 (  10) &  618 (   9) \\
-  Pindel      &   18 (\textbf{   9}) &  112 (\textbf{  27}) &  103 (   3) &  473 (  10) \\ 
+  Cloudbreak  &   27 (   4) & \textbf{ 152} (  14) & \textbf{ 222} (\textbf{  50}) & \textbf{ 880} (\textbf{ 131}) \\ 
+  Breakdancer & \textbf{  29} (   5) &  142 (   7) &  186 (  20) &  754 (  30) \\
+  GASVPro     &   16 (   2) &   79 (   3) &   75 (  10) &  302 (  22) \\
+  DELLY       &   21 (   2) &  117 (   2) &  147 (   8) &  618 (   9) \\
+  Pindel      &   18 (\textbf{   9}) &  112 (\textbf{  27}) &  103 (   4) &  473 (   8) \\ 
    \hline
@@
 \end{center}
-\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
+\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool at the 10\% FDR cutoff identified in the simulated data set, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
 \label{deletionRepmaskpreds}
@@
   \hline
-  Cloudbreak  &   19 (   3) & \textbf{  59} (  23) &   75 (  13) &   62 (  18) \\ 
-  Breakdancer &   17 (   8) &   22 (  11) &   82 (  36) &   44 (  18) \\
-  Pindel      & \textbf{  25} (\textbf{  18}) &   54 (\textbf{  32}) & \textbf{  84} (\textbf{  48}) & \textbf{  82} (\textbf{  51}) \\ 
+  Cloudbreak  &   21 (   4) & \textbf{  59} (  24) & \textbf{ 139} (\textbf{  39}) & \textbf{ 136} (\textbf{  55}) \\ 
+  Breakdancer &   17 (   7) &   22 (  11) &   82 (  12) &   44 (   9) \\
+  Pindel      & \textbf{  25} (\textbf{  18}) &   54 (\textbf{  33}) &   84 (  36) &   82 (  24) \\ 
   MoDIL      &    5 (0) &   16 (   3) & NA & NA \\ 
@@
 Because we simulated a diploid genome, we had genotypes for the simulated data readily available.
-For benchmarking genotyyping of deletions in the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
+For benchmarking genotyping of deletions in the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
 We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes.
-By setting a simple cutoff of .2 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 86.7\% and 94.9\% accuracy in predicting the genotype of the true positive deletions we detected in the simulated and real data sets, respectively.
+By setting a simple cutoff of .35 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.9\% and 96.2\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively.
 Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
 For insertions, none of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertion rates that meet our size threshold that also had genotyping information.
-Of the 56 insertions detected by Cloudbreak on the simulated data set, only 4 were heterozygous.
-Using another simple cutoff on $\alpha$, this time with a value of .3, Cloudbreak correctly classified 50 of the 52 homozygous insertions and 3 of the four heterozygous insertions discovered.
+Of the 80 insertions detected by Cloudbreak on the simulated data set, 15 were heterozygous.
+Using the same cutoff of .35 on $\alpha$, Cloudbreak correctly classified 62 of the 65 homozygous insertions and 10 of the 15 heterozygous insertions discovered, for an overall accuracy rate of 90\%.
 
-\begin{table}
+\begin{table}[b]
 \begin{center}
@@
 \cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 92 & 3 &  74 & 12 \\
- & Heterozygous & 15 & 71 &  4 & 286 \\
+\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 99 & 4 &  78 & 15 \\
+ & Heterozygous & 6 & 69 &  3 & 333 \\
 \cline{2-6}
@@
 \label{deletionGenotypeaccuracy}
+\end{table}
+
+\subsection{Runtimes}
+
+We implemented and executed Cloudbreak on a 56-node Hadoop cluster, with 636 map slots and 477 reduce slots.
+Not including alignment time, we were able to process the Chromosome 2 simulated data in under 10 minutes, and the the NA18507 data set in under an hour.
+For the simulated data set we used 100 reducers for the compute SV features job; for the real data set we used 300.
+The bulk of Cloudbreak's execution is spent in the feature generation step.
+Extracting deletion and insertion calls take under two minutes each for both the real and simulated data sets; the times are equal because each reducer is responsible for processing a single chromosome, and so the runtime is bounded by the length of time it takes to process the largest chromosome.
+
+In table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work.
+Each tool varies in the amount of parallelization supported.
+We report runtimes for tools run in their default single-threaded mode, as well as for levels of parallelization achievable with basic scripting, noting that one of the key advantages of Hadoop/MapReduce is the ability to scale parallel execution to the size of the available compute cluster without any custom programming.
+Pindel and Breakdancer allow processing of a single chromosome in one process, so it is possible to execute all chromosomes in parallel on a cluster that has a job scheduler and shared filesystem.
+Breakdancer has an additional preprocessing step (\verb|bam2cfg.pl|) which runs in a single thread.
+DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode.
+GASVPro allows paralellization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting so we do not report those results here.
+The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers.
+
+In parallel execution, the total time to execute is bounded by the runtime of the longest-running process.
+In the case of chromosome-parallelizable tools including Breakdancer, Pindel, and DELLY, this is typically the process working on the largest chromosome.\footnote{We note that one Breakdancer process, handling an unplaced contig in the hg19 reference genome, never completed in our runs and had to be killed manually; we exclude that process from our results.} In the case of MoDIL's run on the simulated data, we found that the different processes varied widely in their execution times, likely caused by regions of high coverage or with many ambiguously mapped reads.
+Cloudbreak mitigates this problem during the time-consuming feature generation process by using Hadoop partitioners to randomly assign each genomic location to one of the set of reducers, ensuring that the work is evenly distributed across all processes.
+
+\begin{table}[t]
+\begin{center}
+\begin{tabular}{rrr|rr}
+ & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
+\hline
+ &  Single CPU & Parallel (Max Proc.) &  Single CPU & Parallel (Max Proc.) \\ 
+  \hline
+  Cloudbreak  &   NA    & 426 (312)       & NA         & 2,481 (636) \\ 
+  Breakdancer &  653    & NA                 &              &  5,586 (84)  \\
+  GAVPPro       &  3,339  & NA                &              & NA  \\
+  DELLY          &  1,964  & NA                & 14,475  & 37,390 (84) \\
+  Pindel          &  1,336  &  NA               &              & 28,587 (84) \\ 
+  MoDIL         &  NA      & 52,547 (250)  & NA        & NA \\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{Runtimes (elapsed) on both data sets of each tool tested, in single-processor and parallel mode.
+For parallel runs, the maximum number of simultaneously running processes is shown in parentheses.
+All times are in seconds.}
+\label{runtimes}
 \end{table}
@@
 Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
+To execute the other tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine with directed acyclic graphs to describe dependencies between jobs.\footnote{\url{http://research.cs.wisc.edu/htcondor/}}
 
-For each evaluated SV detection program we attempt to follow the authors' recommended alignment pipeline.
-Simulated reads were aligned to hg18 chromosome 2.
+Simulated reads were aligned to hg18 chromosome 2, and NA18507 reads were aligned to the hg19 assembly.
+For each evaluated SV detection program we attempt to follow the authors' recommended alignment pipeline, except for MoDIL, for which we used reads aligned with BWA (see below).
 Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap RazerS 3 version 3.1, with parameters \\ \verb|-rr 99 -i 94 -m 1000|, requesting 99\% sensitivity, a maximum edit distance between read and reference of 6\%, and up to 1000 possible alignments reported for each read.
@@
 
-We ran Breakdancer version 1.1\_2011\_02\_21 by first executing \verb|bam2cfg.pl| and then running \verb|breakdancer_max| with the default parameter values.
+We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \verb|bam2cfg.pl| and then running \verb|breakdancer_max| with the default parameter values.
+To run Breakdancer in parallel mode we first ran \verb|bam2cfg.pl| and then launched parallel instances of \verb|breakdancer_max| for each chromosome using the \verb|-o| parameter.
 We ran DELLY version 0.0.9 with the \verb|-p| parameter and default values for other parameters.
+For the parallel run of DELLY we first split the original BAM file with BamTools \cite{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
 We ran GASVPro version 1.2 using the \verb|GASVPro.sh| script and default parameters.
-Finally, we ran Pindel 0.2.4t with default parameters.
+Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \verb|-c| option.
 

### 2013-03-23 src/main/scripts/manuscript.R

@@ -106,17 +106,17 @@ printPredTableCell <- function(preds, ex, x, y) {
 # chr2
 
 # deletions
-cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/extract_in_hadoop_dels_oldsplot_perf.txt'
+cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/extract_in_hadoop_del_postcombinefix_perf.txt'
 breakdancerChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_35_2_3_del.perf.txt'
 gasvProChr2DeletionPerfFile <-  '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gasvpro_gte40.perf.txt'
 dellyChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_gte40.perf.txt'
 pindelChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_pindel_D_gte40.perf.txt'
 
-cloudbreakChr2DeletionPredictionsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/dels_called_in_hadoop_2p12_3.bed'
+cloudbreakChr2DeletionPredictionsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/dels_called_in_hadoop_2p22.bed'
 
-cloudbreakChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/dels_called_in_hadoop_2p12_3.hits.txt'
[…]

### 2013-03-23 src/main/scripts/prototypeEM.R
generate example distributions for figures

@@ -158,3 +158,26 @@ testHomegrown <- function() {
   print(exp(ws))
   print(sum(exp(ws) < .25))
 }
+
+
+pdf('~/Documents/gene_rearrange/svpipeline/example_normal.pdf')
+h <- hist(sim.nodel(35, num.noise=5, max.insert=500), breaks=40, xlab="Observed Insert Size", main=NULL, xlim=c(0,500), ylim=c(0,7), yaxt='n', ann=FALSE)
+lines(seq(000,500), dnorm(seq(000,500), mean=200, sd=30)*300, col="blue", lwd=3)
+dev.off()
+
+pdf('~/Documents/gene_rearrange/svpipeline/example_homdel.pdf')
+h <- hist(sim.homdel(35, 75, num.noise=5, max.insert=500), breaks=40, xlab="Observed Insert Size", main=NULL, xlim=c(0,500), ylim=c(0,7), yaxt='n', ann=FALSE)
+lines(seq(000,500), dnorm(seq(000,500), mean=250, sd=30)*300, col="red", lwd=3)
+dev.off()
+
+pdf('~/Documents/gene_rearrange/svpipeline/example_hetdel.pdf')
+h <- hist(sim.hetdel(35, 75, num.noise=5, max.insert=500), breaks=40, xlab="Observed Insert Size", main=NULL, xlim=c(0,500), ylim=c(0,7), yaxt='n', ann=FALSE)
+lines(seq(000,500), dnorm(seq(000,500), mean=200, sd=30)*150, col="blue", lwd=3)
+lines(seq(000,500), dnorm(seq(000,500), mean=275, sd=30)*150, col="red", lwd=3)
+dev.off()
+
+pdf('~/Documents/gene_rearrange/svpipeline/example_ins.pdf')
+h <- hist(sim.hetdel(35, -75, num.noise=5, max.insert=500), breaks=40, xlab="Observed Insert Size", main=NULL, xlim=c(0,500), ylim=c(0,7), yaxt='n', ann=FALSE)
+lines(seq(000,500), dnorm(seq(000,500), mean=200, sd=30)*150, col="blue", lwd=3)
+lines(seq(000,500), dnorm(seq(000,500), mean=125, sd=30)*150, col="green", lwd=3)
+dev.off()

### 2013-03-25 manuscript/manuscript.tex
switch to GEM mapper in manuscript

@@
   \printtext{\bibstring{in}\intitlepunct}}}
-\addbibresource{recomb.bib}
+\addbibresource{manuscript.bib}
 
@@
 Our framework could use any aligner that is capable of reporting multiple alignments for reads.
-The results we report below were generated with alignments from RazerS 3 \autocite{Weese:2012by}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
-Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
+The results we report below were generated with alignments from the GEM mapper \autocite{MarcoSola:2012hm}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
+Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
 
@@
 
-\begin{table}[b]
+\begin{table}[t]
 \begin{center}
@@
   \hline
-  Cloudbreak  &   48 (   8)  &   49 (\textbf{   2}) &  \textbf{  55} (\textbf{   4}) & \textbf{  12} (\textbf{   4}) & \textbf{  15} (\textbf{0}) \\ 
-  Breakdancer &   52 (  10)  &   49 (\textbf{   2}) &    49 (0) &    7 (0) &   14 (\textbf{0}) \\ 
-  GASVPro     &   35 (   5)  &   26 (0) &    26 (0) &    2 (0) &    6 (\textbf{0}) \\ 
-  DELLY       &   22 (   2)  & \textbf{  56} (\textbf{   2}) &    40 (0) &    8 (0) &   12 (\textbf{0}) \\ 
-  Pindel      & \textbf{  60} (\textbf{  34})  &   16 (0) &    41 (   2) &    1 (0) &   12 (\textbf{0})\\ 
+  Cloudbreak  &  47 (  7)  &  52 (\textbf{  3}) &  \textbf{ 57} (\textbf{  5}) & \textbf{ 12} (\textbf{  4}) & \textbf{ 15} (\textbf{0}) \\ 
+  Breakdancer &  52 ( 10)  &  49 (\textbf{  3}) &   49 (0) &   7 (0) &  14 (\textbf{0}) \\ 
+  GASVPro     &  35 (  4)  &  26 (0) &   26 (0) &   2 (0) &   6 (\textbf{0}) \\ 
+  DELLY       &  22 (  2)  & \textbf{ 56} (  2) &   40 (0) &   8 (0) &  12 (\textbf{0}) \\ 
+  Pindel      & \textbf{ 60} (\textbf{ 34})  &  16 (0) &   41 (  1) &   1 (0) &  12 (\textbf{0})\\ 
    \hline
@@
   \hline
-  Cloudbreak    &  \textbf{0.507} & 0.649 & \textbf{ 144} (   6)  &   57 (\textbf{   4}) &    61 (   1) &   13 (0) &   15 (0) \\ 
+  Cloudbreak    &  \textbf{0.532} & \textbf{0.66} & \textbf{146} (  6)  &  58 (\textbf{  4}) &   62 (  1) &  14 (  1) &  15 (0) \\ 
   Breakdancer   &  0.356 & 0.49 &  89 (0)  &  54 (0) &   53 (0) &   8 (0) &  15 (0) \\ 
@@
   Pindel           & 0.462 & 0.421 &  96 (\textbf{ 13})  &  24 (0) &   48 (0) &   5 (0) &  15 (0)\\ 
-  MoDIL           & 0.132  & \textbf{0.66} &  123 (   6)  &   66 (   2) &  \textbf{  66} (\textbf{   8}) & \textbf{  17} (\textbf{   6}) & \textbf{  23} (\textbf{   8})\\ 
+  MoDIL           & 0.132  & \textbf{0.66} & 123 (  5)  &  66 (  2) &  \textbf{ 66} (\textbf{  7}) & \textbf{ 17} (\textbf{  6}) & \textbf{ 23} (\textbf{  8})\\ 
    \hline
@@
   \hline
-  Cloudbreak   &\textbf{0.769} & \textbf{0.199}  & \textbf{  67} (  24)  &    9 (   3) &     3 (   1) &    1 (0) & 0 (0) \\ 
+  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{ 67} ( 24)  &   8 (  3) &    3 (  1) &   1 (0) & 0 (0) \\ 
   Breakdancer & 0.262 & 0.0968  &  23 (  8)  & \textbf{ 14} (\textbf{  9}) &    2 (  1) & 0 (0) & 0 (0) \\ 
-  Pindel          & 0.572 & 0.196 &   52 (\textbf{  26})  &    5 (   3) &  \textbf{  10} (\textbf{  10}) & \textbf{   3} (\textbf{   3}) & \textbf{   9} (\textbf{   9})\\ 
+  Pindel          & 0.572 & \textbf{0.196} &  52 (\textbf{ 26})  &   5 (  3) &  \textbf{ 10} (\textbf{ 10}) & \textbf{  3} (\textbf{  3}) & \textbf{  9} (\textbf{  9})\\ 
   MoDIL          & 0.186 & 0.0521 &  14 (  1)  &   4 (  1) &    1 (0) &   2 (  1) & 0 (0)\\ 
@@
   \hline
-Cloudbreak & 0.111 & \textbf{0.131} & \textbf{ 383} (\textbf{ 139})  & \textbf{ 146} (\textbf{  13}) &  \textbf{ 178} (\textbf{  10}) & \textbf{  95} (\textbf{   3}) & \textbf{ 300} (\textbf{  16}) \\
-Breakdancer & 0.122 & 0.112 &  261 (  41)  &  132 (   7) &   167 (0) &   92 (0) &  288 (   2) \\  
-  GASVPro & 0.135 & 0.0449 &  120 (  29)  &   38 (   3) &    80 (0) &   29 (0) &  110 (0) \\ 
-  DELLY & 0.0824 & 0.091 &  143 (   7)  &  125 (   5) &   158 (   1) &   83 (   1) &  256 (   3) \\ 
+Cloudbreak & 0.11 & \textbf{0.13} & \textbf{371} (\textbf{127})  & \textbf{146} (\textbf{ 15}) &  \textbf{177} (\textbf{  9}) & \textbf{ 95} (\textbf{  3}) & \textbf{301} (\textbf{ 19}) \\ 
+Breakdancer & 0.122 & 0.112 & 261 ( 41)  & 132 (  6) &  167 (0) &  92 (0) & 288 (  2) \\  
+  GASVPro & 0.135 & 0.0449 & 120 ( 28)  &  38 (  3) &   80 (0) &  29 (0) & 110 (0) \\ 
+  DELLY & 0.0824 & 0.091 & 143 (  8)  & 125 (  4) &  158 (  1) &  83 (  1) & 256 (  3) \\ 
   Pindel & \textbf{0.16} & 0.0685 & 149 ( 12)  &  57 (0) &  140 (0) &  58 (0) & 172 (0) \\ 
@@
   \hline
-Cloudbreak & 0.0258 & \textbf{0.395} & \textbf{ 249} (\textbf{  85})  &   25 (   9) &     1 (0) & 0 (0) \\ 
+Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{ 83})  &  25 (  9) &    1 (0) & 0 (0) \\ 
 Breakdancer & 0.0281 & 0.181 &  97 ( 11)  & \textbf{ 27} (  9) &    2 (  1) & 0 (0) \\  
-  Pindel & \textbf{0.0387} & 0.239 &  144 (  40)  &   14 (\textbf{  12}) &  \textbf{   7} (\textbf{   7}) & \textbf{   1} (\textbf{   1})  \\ 
+  Pindel & \textbf{0.0387} & 0.239 & 144 ( 41)  &  14 (\textbf{ 12}) &  \textbf{  7} (\textbf{  7}) & \textbf{  1} (\textbf{  1})  \\ 
    \hline
@@
   \hline
-  Cloudbreak  &   27 (   4) & \textbf{ 152} (  14) & \textbf{ 222} (\textbf{  50}) & \textbf{ 880} (\textbf{ 131}) \\ 
-  Breakdancer & \textbf{  29} (   5) &  142 (   7) &  186 (  20) &  754 (  30) \\
-  GASVPro     &   16 (   2) &   79 (   3) &   75 (  10) &  302 (  22) \\
-  DELLY       &   21 (   2) &  117 (   2) &  147 (   8) &  618 (   9) \\
-  Pindel      &   18 (\textbf{   9}) &  112 (\textbf{  27}) &  103 (   4) &  473 (   8) \\ 
+  Cloudbreak  &  25 (  2) & \textbf{158} ( 17) & \textbf{220} (\textbf{ 49}) & \textbf{870} (\textbf{124}) \\ 
+  Breakdancer & \textbf{ 29} (  6) & 142 (  7) & 186 ( 19) & 754 ( 30) \\
+  GASVPro     &  16 (  2) &  79 (  2) &  75 (  9) & 302 ( 22) \\
+  DELLY       &  21 (  2) & 117 (  2) & 147 (  9) & 618 (  8) \\
+  Pindel      &  18 (\textbf{  9}) & 112 (\textbf{ 26}) & 103 (  4) & 473 (  8) \\ 
    \hline
@@
 \end{center}
-\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool at the 10\% FDR cutoff identified in the simulated data set, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
+\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
 \label{deletionRepmaskpreds}
@@
   \hline
-  Cloudbreak  &   21 (   4) & \textbf{  59} (  24) & \textbf{ 139} (\textbf{  39}) & \textbf{ 136} (\textbf{  55}) \\ 
+  Cloudbreak  &  21 (  4) & \textbf{ 58} ( 24) & \textbf{137} (\textbf{ 37}) & \textbf{135} (\textbf{ 55}) \\ 
   Breakdancer &  17 (  7) &  22 ( 11) &  82 ( 12) &  44 (  9) \\
-  Pindel      & \textbf{  25} (\textbf{  18}) &   54 (\textbf{  33}) &   84 (  36) &   82 (  24) \\ 
+  Pindel      & \textbf{ 25} (\textbf{ 18}) &  54 (\textbf{ 33}) &  84 ( 36) &  82 ( 25) \\ 
   MoDIL      &   5 (0) &  16 (  3) & NA & NA \\ 
@@
 
-\begin{table}[b]
+\begin{table}
 \begin{center}
@@
 \cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 99 & 4 &  78 & 15 \\
- & Heterozygous & 6 & 69 &  3 & 333 \\
+\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 100 & 4 &  77 & 12 \\
+ & Heterozygous & 7 & 72 &  3 & 338 \\
 \cline{2-6}
@@
 Pindel and Breakdancer allow processing of a single chromosome in one process, so it is possible to execute all chromosomes in parallel on a cluster that has a job scheduler and shared filesystem.
-Breakdancer has an additional preprocessing step (\verb|bam2cfg.pl|) which runs in a single thread.
+Breakdancer has an additional preprocessing step (\texttt{bam2cfg.pl}) which runs in a single thread.
 DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode.
@@
   Cloudbreak  &   NA    & 426 (312)       & NA         & 2,481 (636) \\ 
-  Breakdancer &  653    & NA                 &              &  5,586 (84)  \\
-  GAVPPro       &  3,339  & NA                &              & NA  \\
+  Breakdancer &  653    & NA                 & 134,170  &  5,586 (84)  \\
+  GAVPPro       &  3,339  & NA                & 52,385   & NA  \\
   DELLY          &  1,964  & NA                & 14,475   & 37,390 (84) \\
@@
 For each evaluated SV detection program we attempt to follow the authors' recommended alignment pipeline, except for MoDIL, for which we used reads aligned with BWA (see below).
-Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap RazerS 3 version 3.1, with parameters \\ \verb|-rr 99 -i 94 -m 1000|, requesting 99\% sensitivity, a maximum edit distance between read and reference of 6\%, and up to 1000 possible alignments reported for each read.
-Alignments for the other programs were first aligned using BWA \autocite{Li:2009p836} version 0.6.2-r126, with parameter \verb|-e 5| to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
-GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \verb|-a -r -Ex 1100 -t 250|.
+Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta) in single-ended mode, with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0},  requesting all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read.
+Alignments for the other programs were first aligned using BWA \autocite{Li:2009p836} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
+GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \texttt{-a -r -Ex 1100 -t 250}.
 
-We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \verb|bam2cfg.pl| and then running \verb|breakdancer_max| with the default parameter values.
-To run Breakdancer in parallel mode we first ran \verb|bam2cfg.pl| and then launched parallel instances of \verb|breakdancer_max| for each chromosome using the \verb|-o| parameter.
-We ran DELLY version 0.0.9 with the \verb|-p| parameter and default values for other parameters.
+We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer_max} with the default parameter values.
+To run Breakdancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer_max} for each chromosome using the \texttt{-o} parameter.
+We ran DELLY version 0.0.9 with the \texttt{-p} parameter and default values for other parameters.
 For the parallel run of DELLY we first split the original BAM file with BamTools \cite{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
-We ran GASVPro version 1.2 using the \verb|GASVPro.sh| script and default parameters.
-Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \verb|-c| option.
+We ran GASVPro version 1.2 using the \texttt{GASVPro.sh} script and default parameters.
+Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \texttt{-c} option.
 

### 2013-03-25 src/main/scripts/manuscript.R

@@ -82,18 +82,18 @@ numPredictionsAtLevel <- function(perfTable, threshold) {
 printPredTableValue <- function(t, x, y) {
   d <- t[x,y]
   if (t[x,y] == max(t[,y])) {
-    return(paste0("\\textbf{",formatC(d,digits=3,format="fg"),"}"))
+    return(paste0("\\textbf{",formatC(d,digits=3,format="fg",width=0),"}"))
   } else {
-    return(formatC(d,digits=3,format="fg"))
+    return(formatC(d,digits=3,format="fg",width=0))
   }
 }
 
 printPredTableValueFromList <- function(x, l) {
   d <- l[x]
   if (d == max(l)) {
-    return(paste0("\\textbf{",formatC(d,digits=3,format="fg"),"}"))
+    return(paste0("\\textbf{",formatC(d,digits=3,format="fg",width=0),"}"))
   } else {
-    return(formatC(d,digits=3,format="fg"))
+    return(formatC(d,digits=3,format="fg",width=0))
   }
 }
 
@@ -106,26 +106,26 @@ printPredTableCell <- function(preds, ex, x, y) {
 # chr2
 
 # deletions
-cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/extract_in_hadoop_del_postcombinefix_perf.txt'
+cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_e6_s2_del.perf.txt'
 breakdancerChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_35_2_3_del.perf.txt'
 gasvProChr2DeletionPerfFile <-  '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gasvpro_gte40.perf.txt'
[…]

### 2013-04-03 manuscript/manuscript.tex
update figures and tables

@@
 \label{cloudbreakAlgorithm}
-\caption{The algorithmic framwork for SV calling in MapReduce.}
+\caption{The algorithmic framework for SV calling in MapReduce.}
 \end{algorithm}
@@
 For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
-\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus.
+\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Figure \ref{insert_size_mixes}).
 Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
@@
 Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{/users/cwhelan/Documents/gene_rearrange/insert_size_mixtures.pdf}
+\caption{Illustration of insert size mixtures at individual genomic locations.
+A) there is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings.
+B) a homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes.
+C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected.
+D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
+\label{insert_size_mixes}
+\end{figure}
 
@@
 \end{description}
+
+A diagram of the Cloudbreak algorithm working on a simple example is shown in Figure \ref{algorithm_example}.
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{cloudbreak_mapred_diagram.pdf}
+\caption{An example of the Cloudbreak MapReduce algorithm.
+A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in single-ended mode to generate aligned reads.
+Reducers gather all alignments for both reads in each pair.
+B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) keys indicating the genomic location spanned by that pair.
+Only one genomic location is diagrammed here for simplicity.
+Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes.
+C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.}
+\label{algorithm_example}
+\end{figure}
 
@@
 
-\begin{figure}[b]
+\begin{figure}
 \centering
@@
 
-Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the perfomance characteristics of algorithms at varying levels.
+Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying levels.
 First, we used the results on simulated deletions to characterize the predictions made by each algorithm when a low false discovery rate is required.
@@
 Cloudbreak shows good performance at detecting deletions in all size classes, and finds more medium and large deletions than the other approaches.
-Pindel exlusively identifies a large number of small (40bp-100bp) deletions, as is expected for a split-read method.
+Pindel exclusively identifies a large number of small (40bp-100bp) deletions, as is expected for a split-read method.
 
-\begin{table}[t]
+\begin{table}
 \begin{center}
@@
 
-\begin{table}[t]
+\begin{table}
 \begin{center}
@@
 \end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivy, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
+\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
 The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
@@
 The number of insertions identified by each algorithm at maximum sensitivity, again broken down by size, is shown in Table \ref{chr2InsertionPreds}.
-Cloudbreak shows excellent precision, although recall is low for all four aproaches.
+Cloudbreak shows excellent precision, although recall is low for all four approaches.
 Cloudbreak again identifies the most small variants.
@@
 
-\begin{figure}[t]
+\begin{figure}
 \centering
@@
 Table \ref{deletionRepmaskpreds} shows the number of deletions that overlap repeats at the 10\% FDR cutoff described above.
-In the NA18507 data set, Cloudbreak identifies 16\% more repeat-overlapping variants than the nearest competitor, Breakdancer.
+In the NA18507 data set, Cloudbreak identifies 15\% more repeat-overlapping variants than the nearest competitor, Breakdancer.
 Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
@@
 We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes.
-By setting a simple cutoff of .35 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.9\% and 96.2\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively.
+By setting a simple cutoff of .35 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.1\% and 96.5\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively.
 Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
 For insertions, none of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertion rates that meet our size threshold that also had genotyping information.
-Of the 80 insertions detected by Cloudbreak on the simulated data set, 15 were heterozygous.
-Using the same cutoff of .35 on $\alpha$, Cloudbreak correctly classified 62 of the 65 homozygous insertions and 10 of the 15 heterozygous insertions discovered, for an overall accuracy rate of 90\%.
+Of the 79 insertions detected by Cloudbreak on the simulated data set, 13 were heterozygous.
+Using the same cutoff of .35 on $\alpha$, Cloudbreak correctly classified 62 of the 66 homozygous insertions and 10 of the 13 heterozygous insertions discovered, for an overall accuracy rate of 91\%.
 
@@
 We implemented and executed Cloudbreak on a 56-node Hadoop cluster, with 636 map slots and 477 reduce slots.
-Not including alignment time, we were able to process the Chromosome 2 simulated data in under 10 minutes, and the the NA18507 data set in under an hour.
+Not including alignment time, we were able to process the Chromosome 2 simulated data in under 6 minutes, and the the NA18507 data set in under 40 minutes.
 For the simulated data set we used 100 reducers for the compute SV features job; for the real data set we used 300.
@@
 We report runtimes for tools run in their default single-threaded mode, as well as for levels of parallelization achievable with basic scripting, noting that one of the key advantages of Hadoop/MapReduce is the ability to scale parallel execution to the size of the available compute cluster without any custom programming.
+Pindel allows multi-threaded operation on multicore servers.
 Pindel and Breakdancer allow processing of a single chromosome in one process, so it is possible to execute all chromosomes in parallel on a cluster that has a job scheduler and shared filesystem.
@@
 DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode.
-GASVPro allows paralellization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting so we do not report those results here.
-The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers.
+GASVPro allows parallelization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting, and we did not find that the MCMC module consumed most of the runtime in our experiments, so we do not attempt to parallelize this componen.
+The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers; we adapted these for use in our cluster.
 
@@
 Cloudbreak mitigates this problem during the time-consuming feature generation process by using Hadoop partitioners to randomly assign each genomic location to one of the set of reducers, ensuring that the work is evenly distributed across all processes.
+This distribution of processing across the entire cluster also server to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set in under 50 minutes during a run where one of the compute nodes was rebooted midway through the feature generation job.
 
-\begin{table}[t]
+\begin{table}
 \begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
+\begin{tabular}{r|r|rrr|rrr}
+\multicolumn{2}{c}{}  & \multicolumn{3}{c}{Simulated Data} & \multicolumn{3}{c}{NA18507} \\
 \hline
- &  Single CPU & Parallel (Max Proc.) &  Single CPU & Parallel (Max Proc.) \\ 
+ & SV Types &  Single CPU & Parallel & Proc. &  Single CPU & Parallel & Proc.  \\ 
   \hline
-  Cloudbreak  &   NA    & 426 (312)       & NA         & 2,481 (636) \\ 
-  Breakdancer &  653    & NA                 & 134,170  &  5,586 (84)  \\
-  GAVPPro       &  3,339  & NA                & 52,385   & NA  \\
-  DELLY          &  1,964  & NA                & 14,475   & 37,390 (84) \\
-  Pindel          &  1,336  &  NA               &               & 28,587 (84) \\ 
-  MoDIL         &  NA      & 52,547 (250)  & NA         & NA \\ 
+  Cloudbreak & D,I &   NA    & 308 & 312    & NA         & 2,310 & 636 \\ 
+  Breakdancer & D,I,V,T &  653   & NA       & NA          & 134,170 &  5,586 & 84 \\
+  GASVPro & D,V   &  3,339  & NA       & NA         & 52,385  & NA & NA \\
+  DELLY & D         &  1,964 & NA          & NA      & 14,475  & 37,390 & 84 \\
+  Pindel & D,I,V,P         & 37,006 &  4,885     & 8          &  284,932  & 28,587 & 84 \\ 
+  MoDIL & D,I        &  NA      & 52,547 & 250 & NA         & NA  & NA\\ 
    \hline
@@
 \caption{Runtimes (elapsed) on both data sets of each tool tested, in single-processor and parallel mode.
-For parallel runs, the maximum number of simultaneously running processes is shown in parentheses.
-All times are in seconds.}
+For parallel runs, Proc.
+is the maximum number of simultaneously running processes or threads.
+All times are in seconds.
+The types of variants detected by each program are listed with the abbreviations: D - deletion; I - insertion; V - Inversion; P - duplication; T - translocation.
+Interchromosomal translocations are only detected by Breakdancer in single CPU mode.
+}
 \label{runtimes}
@@
 
-We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer_max} with the default parameter values.
-To run Breakdancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer_max} for each chromosome using the \texttt{-o} parameter.
+We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
+To run Breakdancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer\_max} for each chromosome using the \texttt{-o} parameter.
 We ran DELLY version 0.0.9 with the \texttt{-p} parameter and default values for other parameters.
-For the parallel run of DELLY we first split the original BAM file with BamTools \cite{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
+For the parallel run of DELLY we first split the original BAM file with BamTools \autocite{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
 We ran GASVPro version 1.2 using the \texttt{GASVPro.sh} script and default parameters.

### 2013-04-03 src/main/scripts/manuscript.R

@@ -12,20 +12,31 @@ drawPerfLine <- function(perf, col, lineType, maxFP) {
   lines(fp[fp <= maxFP], tp[fp <= maxFP], col=col, lty=lineType, lwd=3)
 }
 
+rocColors <- function(numLines) {
+  if (numLines <= 9) { 
+    perfCols <- brewer.pal(numLines, "Set1")
+  } else {
+    perfCols <- rainbow(numLines)
+  }
+  return(perfCols)
+}
+
+rocLineTypes <- function(numLines) {
+  seq(1,numLines)
+}
+
 plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legendLoc="bottomright") {
   maxFP <- totalDels
   if (is.na(maxTP)) {
     maxTP <- totalDels
   }
   plot(0, type="n", ylim=c(0, maxTP), xlim=c(0,maxFP), xlab=ifelse(sim, "False Positives", "Novel Predictions"), ylab="True Positives", main=main)
-  if (length(perfs) <= 9) {
-    perfCols <- brewer.pal(length(perfs), "Set1")
-  } else {
-    perfCols <- rainbow(length(perfs))
-  }
-  lineTypes <- seq(1,length(perfs))
+  perfCols <- rocColors(length(perfs))
+  lineTypes <- rocLineTypes(length(perfs))
   mapply(drawPerfLine, perfs, perfCols, lineTypes, MoreArgs=list(maxFP=maxFP))  
-  legend(legendLoc, legend=perfNames, col=perfCols, lty=lineTypes, lwd=3, cex=.9)
+  if (! is.null(legendLoc)) {
+    legend(legendLoc, legend=perfNames, col=perfCols, lty=lineTypes, lwd=3, cex=.9)
+  }
 }
 
 overlap <- function(a,b,x,y) {
@@ -188,23 +199,25 @@ repMask <- import.bed('~/genomes/1kg/hg18/repeatmasker_b36_merged.bed.gz', asRan
 
 chr2DeletionsROCOutputFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf'
 chr2InsertionsROCOutputFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf'
[…]

### 2013-04-05 manuscript/chr2DeletionPredsFDR10.table.brew.tex
update chr2 max sensitivity table to have precision, recall
Separate delly calls into two sets

--- 2013-02-28 manuscript/chr2DeletionPreds.table.brew.tex
@@
   GASVPro     & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\
-  DELLY       & <%= printPredTableCell(preds, exPreds, 'DELLY', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '> 1000bp') %> \\
+  DELLY-RP       & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '> 1000bp') %> \\
+  DELLY-SR       & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '> 1000bp') %> \\
   Pindel      & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\
@@
 The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPreds}
+\label{chr2DeletionPredsFDR10}
 \end{table}

### 2013-04-05 manuscript/chr2DeletionPredsMaxSensitivity.table.brew.tex
Separate delly calls into two sets

--- 2013-03-04 manuscript/chr2DeletionPredsMaxSensitivity.table.brew.tex
@@
 \begin{center}
-\begin{tabular}{rrrrrr}
+\begin{tabular}{rrr|rrrrr}
   \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
+                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number &          &           & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
   \hline
-  Cloudbreak  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\
-  Breakdancer & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\
-  GASVPro     & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\
-  DELLY       & <%= printPredTableCell(preds, exPreds, 'DELLY', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '> 1000bp') %> \\
-  Pindel      & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\
-  MoDIL      & <%= printPredTableCell(preds, exPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '> 1000bp') %>\\
+  Cloudbreak    &  <%= printPredTableValueFromList('Cloudbreak', precision) %> & <%= printPredTableValueFromList('Cloudbreak', recall) %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\
+  Breakdancer   &  <%= printPredTableValueFromList('Breakdancer', precision) %> & <%= printPredTableValueFromList('Breakdancer', recall) %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\
+  GASVPro        & <%= printPredTableValueFromList('GASVPro', precision) %> & <%= printPredTableValueFromList('GASVPro', recall) %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\
+  DELLY-RP           & <%= printPredTableValueFromList('DELLY', precision) %> & <%= printPredTableValueFromList('DELLY', recall) %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '> 1000bp') %> \\
+  DELLY-SR           & <%= printPredTableValueFromList('DELLYSR', precision) %> & <%= printPredTableValueFromList('DELLYSR', recall) %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '> 1000bp') %> \\
+  Pindel           & <%= printPredTableValueFromList('Pindel', precision) %> & <%= printPredTableValueFromList('Pindel', recall) %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\
+  MoDIL           & <%= printPredTableValueFromList('MoDIL', precision) %>  & <%= printPredTableValueFromList('MoDIL', recall) %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '> 1000bp') %>\\
    \hline

### 2013-04-05 manuscript/manuscript.tex
Separate delly calls into two sets

@@
   GASVPro & D,V   &  3,339  & NA       & NA         & 52,385  & NA & NA \\
-  DELLY & D         &  1,964 & NA          & NA      & 14,475  & 37,390 & 84 \\
+  DELLY & D         &  1,964 & NA          & NA      & 30,311  & 20,224 & 84 \\
   Pindel & D,I,V,P         & 37,006 &  4,885     & 8          &  284,932  & 28,587 & 84 \\ 

### 2013-04-05 src/main/scripts/manuscript.R

@@ -13,8 +13,8 @@ drawPerfLine <- function(perf, col, lineType, maxFP) {
 }
 
 rocColors <- function(numLines) {
-  if (numLines <= 9) { 
-    perfCols <- brewer.pal(numLines, "Set1")
+  if (numLines <= 8) { 
+    perfCols <- brewer.pal(numLines, "Dark2")
   } else {
     perfCols <- rainbow(numLines)
   }
@@ -25,7 +25,7 @@ rocLineTypes <- function(numLines) {
   seq(1,numLines)
 }
 
-plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legendLoc="bottomright") {
+plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legendLoc="bottomright",legendCex=.9) {
   maxFP <- totalDels
   if (is.na(maxTP)) {
     maxTP <- totalDels
@@ -35,7 +35,7 @@ plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legen
   lineTypes <- rocLineTypes(length(perfs))
   mapply(drawPerfLine, perfs, perfCols, lineTypes, MoreArgs=list(maxFP=maxFP))  
   if (! is.null(legendLoc)) {
-    legend(legendLoc, legend=perfNames, col=perfCols, lty=lineTypes, lwd=3, cex=.9)
+    legend(legendLoc, legend=perfNames, col=perfCols, lty=lineTypes, lwd=3, cex=legendCex)
   }
 }
 
@@ -121,7 +121,16 @@ cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_
 breakdancerChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_35_2_3_del.perf.txt'
 gasvProChr2DeletionPerfFile <-  '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gasvpro_gte40.perf.txt'
[…]

### 2013-04-29 src/main/scripts/manuscript.R
add plot of breakpoint resolution

@@ -4,6 +4,7 @@ library(plyr)
 library(RColorBrewer)
 library(VariantAnnotation)
 library(brew)
+library(ggplot2)
 
 drawPerfLine <- function(perf, col, lineType, maxFP) {
   calls <- c(perf$Calls, min(perf$Calls) - min(perf$TP), 0)
@@ -811,7 +812,10 @@ assign("chr2HapCM", chr2HapCMFDR10, envir=tableEnv)
 assign("NA18507HapCM", NA18507HapCM, envir=tableEnv)
 brew('~/Documents/svpipeline/manuscript/deletionGenotypeAccuracy.table.brew.tex', env=tableEnv)
 
-#allpredsFDR10$lendiff <- with(allpredsFDR10, abs((predend - predstart) - (trueend - truestart)))
-#ggplot(aes(y=lendiff,x=name),data=allpredsFDR10) + geom_boxplot()
+
+pdf("~/Documents/svpipeline/manuscript/breakpoint_resolution.pdf")
+allpredsFDR10$lendiff <- with(allpredsFDR10, abs((predend - predstart) - (trueend - truestart)))
+ggplot(aes(y=lendiff,x=name),data=allpredsFDR10) + geom_boxplot() + xlab("") + ylab("Difference in length")
+dev.off()
 
 

### 2013-04-30 manuscript/manuscript.tex
add definition of mismatches
combined table for chr2 deletion and insertion predictions at max sensitivity

@@
 \usepackage{authblk}
+\usepackage{rotating}
 
@@
 \centering
-\includegraphics[width=1\textwidth]{/users/cwhelan/Documents/gene_rearrange/insert_size_mixtures.pdf}
+\includegraphics[width=1\textwidth]{/users/cwhelan/insert_size_mixtures.pdf}
 \caption{Illustration of insert size mixtures at individual genomic locations.
@@
 This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
-We approximate the probability $p^{k}_c$ of each end alignment being correct by:
+Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf}
+\caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
+\label{chr2InsertionsRoc}
+\end{figure}
+
 Figure \ref{chr2DeletionsRoc} shows an ROC plot of the performance of each algorithm for detecting deletions on the simulated data set.
@@
 The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the tradeoff between precision and recall; therefore it is not included in this figure.
+A ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}.
 
 Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying levels.
-First, we used the results on simulated deletions to characterize the predictions made by each algorithm when a low false discovery rate is required.
+First, we used the ROC curves to attempt to characterize the predictions made by each algorithm when a low false discovery rate is required.
 Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve, representing cutoffs of 2.22 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 11 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 for Pindel.
@@
 Pindel exclusively identifies a large number of small (40bp-100bp) deletions, as is expected for a split-read method.
+Performance on insertions never reached a level with an FDR of 10\%, so insertion predictions are not included in this table.
 
@@
 For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.
-The results are summarized in Table \ref{chr2DeletionPredsMaxSensitivity}.
-MoDIL has the greatest recall but low precision.
-Cloudbreak has high precision and recall at this threshold, and surprisingly discovers more small deletions than Pindel.
+The results for both deletion and insertion predictions are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}.
+MoDIL and Cloudbreak exhibited the greatest recall for deletions.
+Cloudbreak has high precision and recall for deletions at this threshold, and surprisingly discovers more small deletions than Pindel.
+For insertions, Cloudbreak shows excellent precision, although recall is low for all four approaches, with Cloudbreak and Pindel showing the highest recall rates.
+Cloudbreak again identifies the most small variants.
+Pindel is the only tool which can consitently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.
 
@@
 \begin{center}
-\begin{tabular}{rrr|rrrrr}
+\begin{tabular}{r|rrr|rrrrr}
+  \cline{2-9}
+   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 \hline
-                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
+\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
   \hline
-  Cloudbreak    &  \textbf{0.532} & \textbf{0.66} & \textbf{146} (  6)  &  58 (\textbf{  4}) &   62 (  1) &  14 (  1) &  15 (0) \\ 
-  Breakdancer   &  0.356 & 0.49 &  89 (0)  &  54 (0) &   53 (0) &   8 (0) &  15 (0) \\ 
-  GASVPro        & 0.146 & 0.432 &  83 (  2)  &  32 (0) &   55 (0) &   8 (0) &  15 (0) \\ 
-  DELLY           & 0.457 & 0.613 & 114 (  8)  & \textbf{ 68} (  2) &  \textbf{ 66} (  1) &   9 (0) &  17 (  1) \\ 
-  Pindel           & 0.462 & 0.421 &  96 (\textbf{ 13})  &  24 (0) &   48 (0) &   5 (0) &  15 (0)\\ 
-  MoDIL           & 0.132  & \textbf{0.66} & 123 (  5)  &  66 (  2) &  \textbf{ 66} (\textbf{  7}) & \textbf{ 17} (\textbf{  6}) & \textbf{ 23} (\textbf{  8})\\ 
+\cline{2-9}
+&  Cloudbreak    &  0.532 & \textbf{0.66} & \textbf{146} (  6)  &  58 (\textbf{  4}) &   62 (  1) &  14 (  1) &  15 (0) \\ 
+&  Breakdancer   &  0.356 & 0.49 &  89 (0)  &  54 (0) &   53 (0) &   8 (0) &  15 (0) \\ 
+&  GASVPro        & 0.146 & 0.432 &  83 (  2)  &  32 (0) &   55 (0) &   8 (0) &  15 (0) \\ 
+&  DELLY-RP           & 0.457 & 0.613 & 114 (  8)  & \textbf{ 68} (  2) &  \textbf{ 66} (0) &   9 (0) &  17 (0) \\ 
+&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  &   3 (0) &   49 (0) &   6 (0) &  16 (0) \\ 
+&  Pindel           & 0.462 & 0.421 &  96 (\textbf{ 13})  &  24 (0) &   48 (0) &   5 (0) &  15 (0)\\ 
+&  MoDIL           & 0.132  & \textbf{0.66} & 123 (  5)  &  66 (  2) &  \textbf{ 66} (\textbf{  7}) & \textbf{ 17} (\textbf{  6}) & \textbf{ 23} (\textbf{  8})\\ 
+   \hline
+\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
+\cline{2-9}
+&  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{ 67} ( 24)  &   8 (  3) &    3 (  1) &   1 (0) & 0 (0) \\ 
+&  Breakdancer & 0.262 & 0.0968  &  23 (  8)  & \textbf{ 14} (\textbf{  9}) &    2 (  1) & 0 (0) & 0 (0) \\ 
+&  Pindel          & 0.572 & \textbf{0.196} &  52 (\textbf{ 26})  &   5 (  3) &  \textbf{ 10} (\textbf{ 10}) & \textbf{  3} (\textbf{  3}) & \textbf{  9} (\textbf{  9})\\ 
+&  MoDIL          & 0.186 & 0.0521 &  14 (  1)  &   4 (  1) &    1 (0) &   2 (  1) & 0 (0)\\ 
 \hline
@@
 \end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPredsMaxSensitivity}
-\end{table}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf}
-\caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
-\label{chr2InsertionsRoc}
-\end{figure}
-
-An ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}.
-The number of insertions identified by each algorithm at maximum sensitivity, again broken down by size, is shown in Table \ref{chr2InsertionPreds}.
-Cloudbreak shows excellent precision, although recall is low for all four approaches.
-Cloudbreak again identifies the most small variants.
-Pindel is the only tool which can identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{rrr|rrrrr}
-  \hline
-                        & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
-  \hline
-  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{ 67} ( 24)  &   8 (  3) &    3 (  1) &   1 (0) & 0 (0) \\ 
-  Breakdancer & 0.262 & 0.0968  &  23 (  8)  & \textbf{ 14} (\textbf{  9}) &    2 (  1) & 0 (0) & 0 (0) \\ 
-  Pindel          & 0.572 & \textbf{0.196} &  52 (\textbf{ 26})  &   5 (  3) &  \textbf{ 10} (\textbf{ 10}) & \textbf{  3} (\textbf{  3}) & \textbf{  9} (\textbf{  9})\\ 
-  MoDIL          & 0.186 & 0.0521 &  14 (  1)  &   4 (  1) &    1 (0) &   2 (  1) & 0 (0)\\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The number of simulated insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those insertions that were discovered exclusively by each tool (in parentheses).
-The total number of insertions in each size class in the true set of insertions is shown in the second row of the header.}
-\label{chr2InsertionPreds}
+\caption{The precision and recall of each tool for the simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels at maximum sensitivity, as well as the number of those variants discovered in each size class.
+Variants exclusively predicted by one tool are shown in parentheses.
+The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section.}
+\label{chr2DeletionAndInsertionPredsMaxSensitivity}
 \end{table}

### 2013-04-30 src/main/scripts/manuscript.R

@@ -768,6 +768,20 @@ assign("precision", chr2MaxSensitivityInsPrecision, envir=tableEnv)
 assign("recall", chr2MaxSensitivityInsRecall, envir=tableEnv)
 brew('~/Documents/svpipeline/manuscript/chr2InsertionPredsBySize.table.brew.tex', env=tableEnv)
 
+# chr 2 Deletions and insertions by size at max sensitivity (combined table)
+tableEnv = new.env()
+assign("totalDels", trueDelSizes, envir=tableEnv)
+assign("delPreds", predsBySizeMaxSensitivity, envir=tableEnv)
+assign("exDelPreds", exBySizeMaxSensitivity, envir=tableEnv)
+assign("delPrecision", chr2MaxSensitivityDelPrecision, envir=tableEnv)
+assign("delRecall", chr2MaxSensitivityDelRecall, envir=tableEnv)
+assign("totalIns", trueInsertionsSizes, envir=tableEnv)
+assign("insPreds", insertionPredsBySizeMaxSensitivity, envir=tableEnv)
+assign("exInsPreds", insertionExBySizeMaxSensitivity, envir=tableEnv)
+assign("insPrecision", chr2MaxSensitivityInsPrecision, envir=tableEnv)
+assign("insRecall", chr2MaxSensitivityInsRecall, envir=tableEnv)
+brew('~/Documents/svpipeline/manuscript/chr2DeletionAndInsertionPredsMaxSensitivity.table.brew.tex', env=tableEnv)
+
 # NA18507 Deletions by size
 tableEnv = new.env()
 assign("totals", trueDelsNA18507Sizes, envir=tableEnv)

### 2013-04-30 manuscript/manuscript.tex
make all figures live in the same place

@@
 \centering
-\includegraphics[width=1\textwidth]{/users/cwhelan/insert_size_mixtures.pdf}
+\includegraphics[width=1\textwidth]{../figures/insert_size_mixtures.pdf}
 \caption{Illustration of insert size mixtures at individual genomic locations.
@@
 \centering
-\includegraphics[width=1\textwidth]{cloudbreak_mapred_diagram.pdf}
+\includegraphics[width=1\textwidth]{../figures/cloudbreak_mapred_diagram.pdf}
 \caption{An example of the Cloudbreak MapReduce algorithm.
@@
 \centering
-\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf}
+\includegraphics[width=0.75\textwidth]{../figures/CHR2SIM_DELS_ROC.pdf}
 \caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes.
@@
 \centering
-\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf}
+\includegraphics[width=0.75\textwidth]{../figures/CHR2SIM_INS_ROC.pdf}
 \caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/NA18507/NA18507_DELS_ROC.pdf}
+\includegraphics[width=0.75\textwidth]{../figures/NA18507_DELS_ROC.pdf}
 \caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{/Users/cwhelan/Documents/gene_rearrange/svpipeline/NA18507/NA18507_INS_ROC.pdf}
+\includegraphics[width=0.75\textwidth]{../figures/NA18507_INS_ROC.pdf}
 \caption{ROC curve of insertion prediction performance on the 37X NA18507 sample.}

### 2013-04-30 src/main/scripts/manuscript.R

@@ -211,15 +211,16 @@ repMask <- import.bed('~/genomes/1kg/hg18/repeatmasker_b36_merged.bed.gz', asRan
 
 # output files
 
-chr2DeletionsROCOutputFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_DELS_ROC.pdf'
-chr2InsertionsROCOutputFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_INS_ROC.pdf'
-chr2ROCsOutputFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_ROC_COMBINED_ROCS.pdf'
-chr2ROCsPosterOutputFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2SIM_ROC_COMBINED_ROCS_POSTER.pdf'
-NA18507DeletionsROCOutputFile <- '~/Documents/gene_rearrange/svpipeline/NA18507/NA18507_DELS_ROC.pdf'
-NA18507InsertionsROCOutputFile <- '~/Documents/gene_rearrange/svpipeline/NA18507/NA18507_INS_ROC.pdf'
-NA18507ROCsOutputFile <- '~/Documents/gene_rearrange/svpipeline/NA18507/NA18507_COMBINED_ROCS.pdf'
-NA18507ROCsPosterOutputFile <- '~/Documents/gene_rearrange/svpipeline/NA18507/NA18507_COMBINED_ROCS_POSTER.pdf'
-chr2DelReportsComparisonROCOutputFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/CHR2_SIM_DEL_REPORTSCOMPARISON_ROC.pdf'
+chr2DeletionsROCOutputFile <- '~/Documents/svpipeline/manuscript/figures/CHR2SIM_DELS_ROC.pdf'
+chr2InsertionsROCOutputFile <- '~/Documents/svpipeline/manuscript/figures/CHR2SIM_INS_ROC.pdf'
+chr2ROCsOutputFile <- '~/Documents/svpipeline/manuscript/figures/CHR2SIM_ROC_COMBINED_ROCS.pdf'
[…]

### 2013-04-30 manuscript/manuscript.tex
add cloudbreak workflow diagram

@@
 
-Our formulation divides processing into three distinct MapReduce jobs.
+Our formulation divides processing into three distinct MapReduce jobs, and is summarized in Figure \ref{cloudbreak_workflow}.
 The first job finds possible alignment locations for pairs of reads in FASTQ format.
@@
 This MapReduce framework is summarized in Algorithm 1.
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\textwidth]{../figures/workflow_with_whirr.png}
+\caption{An overview of the Cloudbreak workflow.
+Reads are first uploaded to a Hadoop cluster from local storage.
+Cloudbreak then executes three MapReduce jobs to process the data: 1) Mapping with sensitive settings, reporting multiple alignments.
+2) Computation of features across the genome.
+3) Calling structural variations based on the features computed in the previous step.
+Finally, SV predictions can be downloaded from the Hadoop cluster and examined.
+Cloudbreak can also use the Apache Whirr library to automatically provision Hadoop clusters on and deploy data to cloud providers such as Amazon Elastic Compute Cloud.}
+\label{cloudbreak_workflow}
+\end{figure}
 
@@
 Cloudbreak again identifies the most small variants.
-Pindel is the only tool which can consitently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.
+Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.
 

### 2013-05-01 src/main/scripts/manuscript.R
add runtime graphs

@@ -211,16 +211,17 @@ repMask <- import.bed('~/genomes/1kg/hg18/repeatmasker_b36_merged.bed.gz', asRan
 
 # output files
 
-chr2DeletionsROCOutputFile <- '~/Documents/svpipeline/manuscript/figures/CHR2SIM_DELS_ROC.pdf'
-chr2InsertionsROCOutputFile <- '~/Documents/svpipeline/manuscript/figures/CHR2SIM_INS_ROC.pdf'
-chr2ROCsOutputFile <- '~/Documents/svpipeline/manuscript/figures/CHR2SIM_ROC_COMBINED_ROCS.pdf'
-chr2ROCsPosterOutputFile <- '~/Documents/svpipeline/manuscript/figures/CHR2SIM_ROC_COMBINED_ROCS_POSTER.pdf'
-NA18507DeletionsROCOutputFile <- '~/Documents/svpipeline/manuscript/figures/NA18507/NA18507_DELS_ROC.pdf'
-NA18507InsertionsROCOutputFile <- '~/Documents/svpipeline/manuscript/figures/NA18507/NA18507_INS_ROC.pdf'
-NA18507ROCsOutputFile <- '~/Documents/svpipeline/manuscript/figures/NA18507_COMBINED_ROCS.pdf'
-NA18507ROCsPosterOutputFile <- '~/Documents/svpipeline/manuscript/figures/NA18507_COMBINED_ROCS_POSTER.pdf'
-chr2DelReportsComparisonROCOutputFile <- '~/Documents/svpipeline/manuscript/figures/CHR2_SIM_DEL_REPORTSCOMPARISON_ROC.pdf'
-breakpointResolutionOutputFile <- '~/Documents/svpipeline/manuscript/figures/breakpoint_resolution.pdf'
+chr2DeletionsROCOutputFile <- '~/Documents/svpipeline/figures/CHR2SIM_DELS_ROC.pdf'
+chr2InsertionsROCOutputFile <- '~/Documents/svpipeline/figures/CHR2SIM_INS_ROC.pdf'
+chr2ROCsOutputFile <- '~/Documents/svpipeline/figures/CHR2SIM_ROC_COMBINED_ROCS.pdf'
+chr2ROCsPosterOutputFile <- '~/Documents/svpipeline/figures/CHR2SIM_ROC_COMBINED_ROCS_POSTER.pdf'
+NA18507DeletionsROCOutputFile <- '~/Documents/svpipeline/figures/NA18507_DELS_ROC.pdf'
[…]

### 2013-05-01 manuscript/NA18507DeletionAndInsertionPredsBySize.table.brew.tex
consolidated insertion and deletion predictions to the same table

\begin{table}
\begin{center}
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & <%= format(totalDels['40-100bp'], big.mark=",") %> & <%= format(totalDels['101-250bp'], big.mark=",") %> & <%= format(totalDels['251-500bp'], big.mark=",") %> & <%= format(totalDels['501-1000bp'], big.mark=",") %> & <%= format(totalDels['> 1000bp'], big.mark=",") %> \\
  \hline
\cline{2-9}
& Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', delPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '> 1000bp') %> \\ 
& Breakdancer & <%= printPredTableValueFromList('Breakdancer', delPrecision) %> & <%= printPredTableValueFromList('Breakdancer', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '> 1000bp') %> \\  
&  GASVPro & <%= printPredTableValueFromList('GASVPro', delPrecision) %> & <%= printPredTableValueFromList('GASVPro',delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '> 1000bp') %> \\ 
&  DELLY-RP & <%= printPredTableValueFromList('DELLY', delPrecision) %> & <%= printPredTableValueFromList('DELLY', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '> 1000bp') %> \\ 
&  DELLY-SR & <%= printPredTableValueFromList('DELLYSR', delPrecision) %> & <%= printPredTableValueFromList('DELLYSR', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '> 1000bp') %> \\ 
&  Pindel & <%= printPredTableValueFromList('Pindel', delPrecision) %> & <%= printPredTableValueFromList('Pindel', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '> 1000bp') %> \\ 
\hline
\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & <%= format(totalIns['40-100bp'], big.mark=",") %> & <%= format(totalIns['101-250bp'], big.mark=",") %> & <%= format(totalIns['251-500bp'], big.mark=",") %> & <%= format(totalIns['501-1000bp'], big.mark=",") %> \\
\cline{2-9}
& Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', insPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '501-1000bp') %> \\ 
& Breakdancer & <%= printPredTableValueFromList('Breakdancer', insPrecision) %> & <%= printPredTableValueFromList('Breakdancer', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '501-1000bp') %> \\  
&  Pindel & <%= printPredTableValueFromList('Pindel', insPrecision) %> & <%= printPredTableValueFromList('Pindel', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '501-1000bp') %>  \\ 
\hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class found. Exclusive predictions are in parentheses. For deletions, the same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X. For insertions, the maximum sensitivity cutoff was used.}
\label{NA18507DeletionAndInsertionPreds}
\end{table}

### 2013-05-01 manuscript/manuscript.tex
consolidated insertion and deletion predictions to the same table

@@
 Cloudbreak is the best performer at more stringent thresholds, and remains near the top at higher sensitivity levels.
+A similar ROC-like curve for insertions on the NA18507 data set is shown in Figure \ref{NA18507InsertionsRoc}.
+Cloudbreak has optimal performance at all thresholds for insertions, although all algorithms make many novel predictions.
 
-We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
-The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those thresholds, is shown in Table \ref{NA18507DeletionPreds}.
-Cloudbreak has the greatest sensitivity of any tool at these thresholds, identifying the most variants in each size class.
+We then tested the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
+For insertions, again, we were forced to use the thresholds that gave maximum sensitivity for each tool due to the high observed FDR rates in the simulated data.
+The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those thresholds, is shown in Table \ref{NA18507DeletionAndInsertionPreds}.
+For deletions, Cloudbreak has the greatest sensitivity of any tool at these thresholds, identifying the most variants in each size class.
+Pindel exhibits the highest precision with respect to the gold standard set.
+For insertions, Pindel again has the highest precision at maximum sensitivity, although according the ROC curve it is possible to choose a threshold for Cloudbreak with higher precision and recall.
+At maximum sensitivity, Cloudbreak identifies 100 more insertions from the gold standard set than Pindel, giving it by far the highest recall.
 
@@
 \begin{center}
-\begin{tabular}{rrr|rrrrr}
+\begin{tabular}{r|rrr|rrrrr}
+  \cline{2-9}
+&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 \hline
- & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
-Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
+\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
   \hline
-Cloudbreak & 0.11 & \textbf{0.13} & \textbf{371} (\textbf{127})  & \textbf{146} (\textbf{ 15}) &  \textbf{177} (\textbf{  9}) & \textbf{ 95} (\textbf{  3}) & \textbf{301} (\textbf{ 19}) \\ 
-Breakdancer & 0.122 & 0.112 & 261 ( 41)  & 132 (  6) &  167 (0) &  92 (0) & 288 (  2) \\  
-  GASVPro & 0.135 & 0.0449 & 120 ( 28)  &  38 (  3) &   80 (0) &  29 (0) & 110 (0) \\ 
-  DELLY & 0.0824 & 0.091 & 143 (  8)  & 125 (  4) &  158 (  1) &  83 (  1) & 256 (  3) \\ 
-  Pindel & \textbf{0.16} & 0.0685 & 149 ( 12)  &  57 (0) &  140 (0) &  58 (0) & 172 (0) \\ 
+\cline{2-9}
+& Cloudbreak & 0.11 & \textbf{0.13} & \textbf{371} (\textbf{127})  & \textbf{146} (\textbf{ 15}) &  \textbf{177} (\textbf{  9}) & \textbf{ 95} (\textbf{  3}) & \textbf{301} (\textbf{ 17}) \\ 
+& Breakdancer & 0.122 & 0.112 & 261 ( 41)  & 132 (  6) &  167 (0) &  92 (0) & 288 (  2) \\  
+&  GASVPro & 0.135 & 0.0449 & 120 ( 28)  &  38 (  3) &   80 (0) &  29 (0) & 110 (0) \\ 
+&  DELLY-RP & 0.0824 & 0.091 & 143 (  8)  & 125 (  4) &  158 (  1) &  83 (  1) & 256 (  1) \\ 
+&  DELLY-SR & 0.135 & 0.0439 & 0 (0)  &  25 (0) &  117 (0) &  56 (0) & 171 (0) \\ 
+&  Pindel & \textbf{0.16} & 0.0685 & 149 ( 12)  &  57 (0) &  140 (0) &  58 (0) & 172 (0) \\ 
+\hline
+\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 \\
+\cline{2-9}
+& Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{ 83})  &  25 (  9) &    1 (0) & 0 (0) \\ 
+& Breakdancer & 0.0281 & 0.181 &  97 ( 11)  & \textbf{ 27} (  9) &    2 (  1) & 0 (0) \\  
+&  Pindel & \textbf{0.0387} & 0.239 & 144 ( 41)  &  14 (\textbf{ 12}) &  \textbf{  7} (\textbf{  7}) & \textbf{  1} (\textbf{  1})  \\ 
 \hline
@@
 \end{center}
-\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found.
-The same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.}
-\label{NA18507DeletionPreds}
+\caption{The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class found.
+Exclusive predictions are in parentheses.
+For deletions, the same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.
+For insertions, the maximum sensitivity cutoff was used.}
+\label{NA18507DeletionAndInsertionPreds}
 \end{table}
@@
 \end{figure}
-
-A similar ROC-like curve for insertions on the NA18507 data set is shown in Figure \ref{NA18507InsertionsRoc}.
-The number of insertions detected at maximum sensitivity for the three algorithms is shown in Table \ref{NA18507InsertionPreds}, along with their precision and recall at that threshold.
-Cloudbreak has the greatest recall, identifying twice as many short insertions as Pindel, while Pindel has the best precision.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{rrr|rrrrr}
-  \hline
- & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp \\ 
-Total Number & & & 536 & 114 & 45 & 1 \\
-  \hline
-Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{ 83})  &  25 (  9) &    1 (0) & 0 (0) \\ 
-Breakdancer & 0.0281 & 0.181 &  97 ( 11)  & \textbf{ 27} (  9) &    2 (  1) & 0 (0) \\  
-  Pindel & \textbf{0.0387} & 0.239 & 144 ( 41)  &  14 (\textbf{ 12}) &  \textbf{  7} (\textbf{  7}) & \textbf{  1} (\textbf{  1})  \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The precision and recall with respect to the gold standard set of insertions for each tool on the NA18507 data, as well as the number of insertions in each size class found, at maximum sensitivity.}
-\label{NA18507InsertionPreds}
-\end{table}
 
@@
 Cloudbreak mitigates this problem during the time-consuming feature generation process by using Hadoop partitioners to randomly assign each genomic location to one of the set of reducers, ensuring that the work is evenly distributed across all processes.
-This distribution of processing across the entire cluster also server to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set in under 50 minutes during a run where one of the compute nodes was rebooted midway through the feature generation job.
+This distribution of processing across the entire cluster also serves to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set in under 50 minutes during a run where one of the compute nodes was rebooted midway through the feature generation job.
 

### 2013-05-29 manuscript/NA18507DeletionAndInsertionPredsBySize.table.brew.tex
updates to NM manuscript

@@
 \hline
-\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & <%= format(totalIns['40-100bp'], big.mark=",") %> & <%= format(totalIns['101-250bp'], big.mark=",") %> & <%= format(totalIns['251-500bp'], big.mark=",") %> & <%= format(totalIns['501-1000bp'], big.mark=",") %> \\
+\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & <%= format(totalIns['40-100bp'], big.mark=",") %> & <%= format(totalIns['101-250bp'], big.mark=",") %> & <%= format(totalIns['251-500bp'], big.mark=",") %> & <%= format(totalIns['501-1000bp'], big.mark=",") %> & 0 \\
 \cline{2-9}
-& Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', insPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '501-1000bp') %> \\
-& Breakdancer & <%= printPredTableValueFromList('Breakdancer', insPrecision) %> & <%= printPredTableValueFromList('Breakdancer', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '501-1000bp') %> \\
-&  Pindel & <%= printPredTableValueFromList('Pindel', insPrecision) %> & <%= printPredTableValueFromList('Pindel', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '501-1000bp') %>  \\
+& Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', insPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '501-1000bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '> 1000bp') %>  \\
+& Breakdancer & <%= printPredTableValueFromList('Breakdancer', insPrecision) %> & <%= printPredTableValueFromList('Breakdancer', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '501-1000bp')  %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '> 1000bp') %> \\
+&  Pindel & <%= printPredTableValueFromList('Pindel', insPrecision) %> & <%= printPredTableValueFromList('Pindel', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '501-1000bp')  %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '> 1000bp')  %> \\
 \hline

### 2013-05-29 manuscript/nm.tex
updates to NM manuscript

--- 2013-05-01 manuscript/manuscript.tex
@@
 \usepackage{rotating}
+\usepackage{xr}
+\externaldocument[S]{nm_supp}
 
-\usepackage[style=authoryear,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
+\usepackage[style=nature,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
 \renewbibmacro{in:}{%
@@
 
-\section{A MapReduce Framework for SV Detection Algorithms}
+\section{Results}
+
+\subsection{A MapReduce Framework for SV Detection Algorithms}
 
@@
 In the third stage, the features for all locations on each chromosome are examined to find contiguous blocks of locations that support an SV call.
-This MapReduce framework is summarized in Algorithm 1.
+This MapReduce framework is summarized in Supplementary Algorithm \ref{Scb_algo}.
 
@@
 \end{figure}
-
-\begin{algorithm}
-\algrenewcommand\algorithmicprocedure{\textbf{job}}
-  \begin{algorithmic}[1]
-    \Procedure{Alignment}{}
-    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
-    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
-    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
-    \EndFor
-    \EndFunction
-    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
-    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
-    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
-    \EndFunction
-    \EndProcedure
-
-    \Procedure{Compute SV Features}{}
-    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
-    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
-    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
-    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
-    \EndFor
-    \EndFor
-    \EndFunction
-    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
-    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
-    \EndFunction
-    \EndProcedure
-
-    \Procedure{Call SVs}{}
-    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
-    \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
-    \EndFunction
-    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
-    \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
-    \EndFunction
-    \EndProcedure
-  \end{algorithmic}
-\label{cloudbreakAlgorithm}
-\caption{The algorithmic framework for SV calling in MapReduce.}
-\end{algorithm}
 
@@
 For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
-\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Figure \ref{insert_size_mixes}).
+\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure \ref{Sinsert_size_mixes}).
 Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
@@
 Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth]{../figures/insert_size_mixtures.pdf}
-\caption{Illustration of insert size mixtures at individual genomic locations.
-A) there is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings.
-B) a homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes.
-C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected.
-D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
-\label{insert_size_mixes}
-\end{figure}
 
@@
 
-A diagram of the Cloudbreak algorithm working on a simple example is shown in Figure \ref{algorithm_example}.
+An illustration of the Cloudbreak algorithm working on a simple example is shown in Figure S\ref{Salgorithm_example}.
 
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth]{../figures/cloudbreak_mapred_diagram.pdf}
-\caption{An example of the Cloudbreak MapReduce algorithm.
-A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in single-ended mode to generate aligned reads.
-Reducers gather all alignments for both reads in each pair.
-B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) keys indicating the genomic location spanned by that pair.
-Only one genomic location is diagrammed here for simplicity.
-Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes.
-C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.}
-\label{algorithm_example}
-\end{figure}
-
-\section{Results}\label{results}
+\subsection{Tests with Simulated Data}
 
 We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
+DELLY produces two sets of calls, one based solely on RP signals, and hte other based on RP calls that could be supported by SR evidence.
+We refer to these sets of calls as DELLY-RP and DELLY-SR in the following results.
 We also attempted to evaluate MoDIL on the same data.
@@
 
-We use the following criteria to define a true prediction given a gold standard set of deletion and insertion variants to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method.
-For evaluating insertions, each algorithm produces insertion predictions that define an interval in which the insertion is predicted to have occurred with start and end coordinates $s$ and $e$ as well as the predicted length of the insertion, $l$.
-The true insertions are defined in terms of their actual insertion coordinate $i$ and their actual length $l_a$.
-Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$.
-In this study we are interested in detecting events larger than 40bp, because with longer reads, smaller events can be more easily discovered by examining gaps in individual reads.
-Both Pindel and MoDIL make many calls with a predicted event size of under 40bp, so we remove those calls from the output sets of those programs.
-Finally, we exclude from consideration calls from all approaches that match a true deletion of less than 40bp where the predicted variant length is less than or equal to 75bp in length.
-
 It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods.
@@
 This seems reasonable to us given the possibility of pipelines in which RP calls are refined by SR mappings or local assembly methods.
-
-\subsection{Tests with Simulated Data}
 
@@
 \centering
-\includegraphics[width=0.75\textwidth]{../figures/CHR2SIM_DELS_ROC.pdf}
-\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes.
-Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.}
-\label{chr2DeletionsRoc}
+\includegraphics[width=1\textwidth]{../figures/chr2_sim_rocs_runtime.pdf}
+\caption{Accuracy and runtime performance on a simulated data set.
+(a) Receiver Operating Characteristic (ROC) curves showing the specificity and sensitivity of each tool to deletions larger than 40bp on a simulated set of reads giving diploid coverage of 30X on human chromosome 2.
+Deletions from the Venter genome were randomly added to one or both haplotypes.
+Each point on a curve represents a different threshold on the confidence of predictions made by that tool.
+Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.
+(b) ROC curves for insertion predictions.
+(c) Runtimes for each tool, not includding alignment time, parallelized when possible (see Supplementary Material).}
+\label{chr2CombinedRoc}
 \end{figure}
 
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth]{../figures/CHR2SIM_INS_ROC.pdf}
-\caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
-\label{chr2InsertionsRoc}
-\end{figure}
-
-Figure \ref{chr2DeletionsRoc} shows an ROC plot of the performance of each algorithm for detecting deletions on the simulated data set.
+Figure \ref{chr2CombinedRoc} shows ROC curves of the performance of each algorithm for detecting deletions and insertions on the simulated data set, as well as the runtimes of the approaches tested.
 All approaches show excellent specificity at high thresholds in this simulation.
-Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.
-The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the tradeoff between precision and recall; therefore it is not included in this figure.
-A ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}.
+Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY.
+For insertions, Cloudbreak clearly provides the best combinations of sensitivity and specificity.
+Cloudbreak's runtime is half that Breakdancer, the next fastest tool, albeit using many more CPUs.
+See Supplementary Material and Supplementary Table \ref{Sruntimes} for a discussion of runtimes and parallelization.
+The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the tradeoff between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster, so results are not included in the runtime figure.
 
@@
 
-\begin{table}
+\begin{table}[t]
 \begin{center}
@@
   \hline
-  Cloudbreak  &  47 (  7)  &  52 (\textbf{  3}) &  \textbf{ 57} (\textbf{  5}) & \textbf{ 12} (\textbf{  4}) & \textbf{ 15} (\textbf{0}) \\ 
-  Breakdancer &  52 ( 10)  &  49 (\textbf{  3}) &   49 (0) &   7 (0) &  14 (\textbf{0}) \\ 
-  GASVPro     &  35 (  4)  &  26 (0) &   26 (0) &   2 (0) &   6 (\textbf{0}) \\ 
-  DELLY       &  22 (  2)  & \textbf{ 56} (  2) &   40 (0) &   8 (0) &  12 (\textbf{0}) \\ 
-  Pindel      & \textbf{ 60} (\textbf{ 34})  &  16 (0) &   41 (  1) &   1 (0) &  12 (\textbf{0})\\ 
+  Cloudbreak  & 47 (7)  & 52 (\textbf{3}) &  \textbf{57} (\textbf{5}) & \textbf{12} (\textbf{4}) & \textbf{15} (\textbf{0}) \\ 
+  Breakdancer & 52 (10)  & 49 (\textbf{3}) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
+  GASVPro     & 35 (4)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
+  DELLY-RP       & 22 (2)  & \textbf{56} (2) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
+  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
+  Pindel      & \textbf{60} (\textbf{34})  & 16 (0) &  41 (1) & 1 (0) & 12 (\textbf{0})\\ 
    \hline
@@
 \cline{2-9}
-&  Cloudbreak    &  0.532 & \textbf{0.66} & \textbf{146} (  6)  &  58 (\textbf{  4}) &   62 (  1) &  14 (  1) &  15 (0) \\ 
+&  Cloudbreak    &  0.532 & \textbf{0.66} & \textbf{146} (6)  & 58 (\textbf{4}) &  62 (1) & 14 (1) & 15 (0) \\ 
 &  Breakdancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
-&  GASVPro        & 0.146 & 0.432 &  83 (  2)  &  32 (0) &   55 (0) &   8 (0) &  15 (0) \\ 
-&  DELLY-RP           & 0.457 & 0.613 & 114 (  8)  & \textbf{ 68} (  2) &  \textbf{ 66} (0) &   9 (0) &  17 (0) \\ 
+&  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
+&  DELLY-RP           & 0.457 & 0.613 & 114 (8)  & \textbf{68} (2) &  \textbf{66} (0) & 9 (0) & 17 (0) \\ 
 &  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
-&  Pindel           & 0.462 & 0.421 &  96 (\textbf{ 13})  &  24 (0) &   48 (0) &   5 (0) &  15 (0)\\ 
-&  MoDIL           & 0.132  & \textbf{0.66} & 123 (  5)  &  66 (  2) &  \textbf{ 66} (\textbf{  7}) & \textbf{ 17} (\textbf{  6}) & \textbf{ 23} (\textbf{  8})\\ 
+&  Pindel           & 0.462 & 0.421 & 96 (\textbf{13})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
+&  MoDIL           & 0.132  & \textbf{0.66} & 123 (5)  & 66 (2) &  \textbf{66} (\textbf{7}) & \textbf{17} (\textbf{6}) & \textbf{23} (\textbf{8})\\ 
    \hline
@@
 \cline{2-9}
-&  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{ 67} ( 24)  &   8 (  3) &    3 (  1) &   1 (0) & 0 (0) \\ 
-&  Breakdancer & 0.262 & 0.0968  &  23 (  8)  & \textbf{ 14} (\textbf{  9}) &    2 (  1) & 0 (0) & 0 (0) \\ 
-&  Pindel          & 0.572 & \textbf{0.196} &  52 (\textbf{ 26})  &   5 (  3) &  \textbf{ 10} (\textbf{ 10}) & \textbf{  3} (\textbf{  3}) & \textbf{  9} (\textbf{  9})\\ 
-&  MoDIL          & 0.186 & 0.0521 &  14 (  1)  &   4 (  1) &    1 (0) &   2 (  1) & 0 (0)\\ 
+&  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{67} (24)  & 8 (3) &  3 (1) & 1 (0) & 0 (0) \\ 
+&  Breakdancer & 0.262 & 0.0968  & 23 (8)  & \textbf{14} (\textbf{9}) &  2 (1) & 0 (0) & 0 (0) \\ 
+&  Pindel          & 0.572 & \textbf{0.196} & 52 (\textbf{26})  & 5 (3) &  \textbf{10} (\textbf{10}) & \textbf{3} (\textbf{3}) & \textbf{9} (\textbf{9})\\ 
+&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (1) &  1 (0) & 2 (1) & 0 (0)\\ 
 \hline
@@
 \end{center}
-\caption{The precision and recall of each tool for the simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels at maximum sensitivity, as well as the number of those variants discovered in each size class.
-Variants exclusively predicted by one tool are shown in parentheses.
+\caption{The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses).
 The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section.}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{../figures/NA18507_DELS_ROC.pdf}
-\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
-\label{NA18507DeletionsRoc}
+\includegraphics[width=1\textwidth]{../figures/NA18507_rocs_runtime.pdf}
+\caption{Accuracy and performance on the 37X NA18507 sample.
+(a) ROC curve for deletion prediction performance, tested against the combined gold standard sets of deletions taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.
+(b) ROC curve for insertion prediciton performance.
+(c) Runtimes for each tool, not includding alignment time, parallelized when possible (see Supplementary Material).
+}
+\label{NA18507CombinedRoc}
 \end{figure}
 
-Figure \ref{NA18507DeletionsRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard deletion set, using the same rules for a match as described in the previous section.
+Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard set for both deletions and insertions, using the same rules for a match as described in the previous section.
 All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing deletions in the gold standard set that are actually present in the sample.
-Cloudbreak is the best performer at more stringent thresholds, and remains near the top at higher sensitivity levels.
-A similar ROC-like curve for insertions on the NA18507 data set is shown in Figure \ref{NA18507InsertionsRoc}.
-Cloudbreak has optimal performance at all thresholds for insertions, although all algorithms make many novel predictions.
+Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
+Cloudbreak has optimal performance at all thresholds for insertions.
 
@@
 \hline
-\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
+\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,462 & 240 & 232 & 147 & 540 \\
   \hline
 \cline{2-9}
-& Cloudbreak & 0.11 & \textbf{0.13} & \textbf{371} (\textbf{127})  & \textbf{146} (\textbf{ 15}) &  \textbf{177} (\textbf{  9}) & \textbf{ 95} (\textbf{  3}) & \textbf{301} (\textbf{ 17}) \\ 
-& Breakdancer & 0.122 & 0.112 & 261 ( 41)  & 132 (  6) &  167 (0) &  92 (0) & 288 (  2) \\  
-&  GASVPro & 0.135 & 0.0449 & 120 ( 28)  &  38 (  3) &   80 (0) &  29 (0) & 110 (0) \\ 
-&  DELLY-RP & 0.0824 & 0.091 & 143 (  8)  & 125 (  4) &  158 (  1) &  83 (  1) & 256 (  1) \\ 
-&  DELLY-SR & 0.135 & 0.0439 & 0 (0)  &  25 (0) &  117 (0) &  56 (0) & 171 (0) \\ 
-&  Pindel & \textbf{0.16} & 0.0685 & 149 ( 12)  &  57 (0) &  140 (0) &  58 (0) & 172 (0) \\ 
+& Cloudbreak & 0.124 & \textbf{0.142} & \textbf{371} (\textbf{127})  & \textbf{150} (\textbf{15}) &  \textbf{187} (\textbf{9}) & \textbf{123} (\textbf{8}) & \textbf{396} (\textbf{27}) \\ 
+& Breakdancer & 0.137 & 0.123 & 261 (41)  & 136 (6) &  178 (0) & 114 (0) & 371 (2) \\  
+&  GASVPro & 0.147 & 0.0474 & 120 (28)  & 40 (3) &  85 (0) & 36 (0) & 128 (0) \\ 
+&  DELLY-RP & 0.0931 & 0.1 & 143 (8)  & 128 (4) &  167 (1) & 103 (1) & 323 (1) \\ 
+&  DELLY-SR & 0.153 & 0.0485 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
+&  Pindel & \textbf{0.179} & 0.0748 & 149 (12)  & 61 (0) &  149 (0) & 69 (0) & 217 (0) \\ 
 \hline
-\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 \\
+\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 & 0 \\
 \cline{2-9}
-& Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{ 83})  &  25 (  9) &    1 (0) & 0 (0) \\ 
-& Breakdancer & 0.0281 & 0.181 &  97 ( 11)  & \textbf{ 27} (  9) &    2 (  1) & 0 (0) \\  
-&  Pindel & \textbf{0.0387} & 0.239 & 144 ( 41)  &  14 (\textbf{ 12}) &  \textbf{  7} (\textbf{  7}) & \textbf{  1} (\textbf{  1})  \\ 
+& Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{83})  & 25 (9) &  1 (0) & 0 (0)  & 0 (0)  \\ 
+& Breakdancer & 0.0281 & 0.181 & 97 (11)  & \textbf{27} (9) &  2 (1) & 0 (0) & 0 (0) \\  
+&  Pindel & \textbf{0.0387} & 0.239 & 144 (41)  & 14 (\textbf{12}) &  \textbf{7} (\textbf{7}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
 \hline
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth]{../figures/NA18507_INS_ROC.pdf}
-\caption{ROC curve of insertion prediction performance on the 37X NA18507 sample.}
-\label{NA18507InsertionsRoc}
-\end{figure}
+\subsection{Performance on a Low-Coverage Cancer Data Set}
+
+We also tested Cloudbreak on a sequencing data set obtained from a patient with acute myeloid leukemia (AML).
+We sequenced CD14+ cells on an Illumina Genome Analyzer II, producing 128,819,200 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
+Using a pipeline consisting of Novoalign, Breakdancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions.
+Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even with lower coverage data sets with a greater variance of insert sizes.
+The variants identfied include deletions occurring in the genes CTDSPL/RBPS3, an AML tumor suppressor \autocite{Zheng:2012kk} and CAV2, a modulator of cancer progression \autocite{Lee:2011bz}.
+We are currently investigating these deletions to determine their functional impact on this patient.
 
@@
 Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
-Performance on insertions is displayed in Table \ref{insertionRepmaskpreds}; again, Cloudbreak identifies more events that lie in repetitive regions than the other tools.
+Performance on insertions is displayed in Supplementary Table \ref{SinsertionRepmaskpreds}; again, Cloudbreak identifies more events that lie in repetitive regions than the other tools.
 
-\begin{table}
+\begin{table}[b]
 \begin{center}
@@
  &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 120 & 327 & 553 & 7851 \\ 
+ Total Number & 120 & 327 & 562 & 8059 \\ 
   \hline
-  Cloudbreak  &  25 (  2) & \textbf{158} ( 17) & \textbf{220} (\textbf{ 49}) & \textbf{870} (\textbf{124}) \\ 
-  Breakdancer & \textbf{ 29} (  6) & 142 (  7) & 186 ( 19) & 754 ( 30) \\
-  GASVPro     &  16 (  2) &  79 (  2) &  75 (  9) & 302 ( 22) \\
-  DELLY       &  21 (  2) & 117 (  2) & 147 (  9) & 618 (  8) \\
-  Pindel      &  18 (\textbf{  9}) & 112 (\textbf{ 26}) & 103 (  4) & 473 (  8) \\ 
+  Cloudbreak  & 25 (2) & \textbf{158} (17) & \textbf{226} (\textbf{49}) & \textbf{1001} (\textbf{137}) \\ 
+  Breakdancer & \textbf{29} (6) & 142 (7) & 192 (19) & 868 (30) \\
+  GASVPro     & 16 (2) & 79 (2) & 79 (9) & 330 (22) \\
+  DELLY-RP       & 21 (2) & 117 (2) & 152 (9) & 712 (6) \\
+  DELLY-SR       & 0 (0) & 42 (0) & 27 (0) & 391 (0) \\
+  Pindel      & 18 (\textbf{9}) & 112 (\textbf{26}) & 109 (4) & 536 (8) \\ 
    \hline
@@
 \label{deletionRepmaskpreds}
-\end{table}
-
-\begin{table}
-\begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\hline
- &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 133 & 270 & 341 & 355 \\ 
-  \hline
-  Cloudbreak  &  21 (  4) & \textbf{ 58} ( 24) & \textbf{137} (\textbf{ 37}) & \textbf{135} (\textbf{ 55}) \\ 
-  Breakdancer &  17 (  7) &  22 ( 11) &  82 ( 12) &  44 (  9) \\
-  Pindel      & \textbf{ 25} (\textbf{ 18}) &  54 (\textbf{ 33}) &  84 ( 36) &  82 ( 25) \\ 
-  MoDIL      &   5 (0) &  16 (  3) & NA & NA \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
-\label{insertionRepmaskpreds}
 \end{table}
@@
 By setting a simple cutoff of .35 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.1\% and 96.5\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively.
-Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
+Supplementary Table \ref{SdeletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
 For insertions, none of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertion rates that meet our size threshold that also had genotyping information.
@@
 Using the same cutoff of .35 on $\alpha$, Cloudbreak correctly classified 62 of the 66 homozygous insertions and 10 of the 13 heterozygous insertions discovered, for an overall accuracy rate of 91\%.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{r|r|rr|rr|}
-\multicolumn{2}{c}{}  & \multicolumn{4}{c}{Actual Genotypes} \\
-\multicolumn{2}{c}{}  & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\cline{3-6}
-\multicolumn{2}{c|}{} &  Homozygous & Heterozygous & Homozygous & Heterozygous \\ 
-\cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 100 & 4 &  77 & 12 \\
- & Heterozygous & 7 & 72 &  3 & 338 \\
-\cline{2-6}
-\end{tabular}
-\end{center}
-\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
-\label{deletionGenotypeaccuracy}
-\end{table}
-
-\subsection{Runtimes}
-
-We implemented and executed Cloudbreak on a 56-node Hadoop cluster, with 636 map slots and 477 reduce slots.
-Not including alignment time, we were able to process the Chromosome 2 simulated data in under 6 minutes, and the the NA18507 data set in under 40 minutes.
-For the simulated data set we used 100 reducers for the compute SV features job; for the real data set we used 300.
-The bulk of Cloudbreak's execution is spent in the feature generation step.
-Extracting deletion and insertion calls take under two minutes each for both the real and simulated data sets; the times are equal because each reducer is responsible for processing a single chromosome, and so the runtime is bounded by the length of time it takes to process the largest chromosome.
-
-In table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work.
-Each tool varies in the amount of parallelization supported.
-We report runtimes for tools run in their default single-threaded mode, as well as for levels of parallelization achievable with basic scripting, noting that one of the key advantages of Hadoop/MapReduce is the ability to scale parallel execution to the size of the available compute cluster without any custom programming.
-Pindel allows multi-threaded operation on multicore servers.
-Pindel and Breakdancer allow processing of a single chromosome in one process, so it is possible to execute all chromosomes in parallel on a cluster that has a job scheduler and shared filesystem.
-Breakdancer has an additional preprocessing step (\texttt{bam2cfg.pl}) which runs in a single thread.
-DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode.
-GASVPro allows parallelization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting, and we did not find that the MCMC module consumed most of the runtime in our experiments, so we do not attempt to parallelize this componen.
-The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers; we adapted these for use in our cluster.
-
-In parallel execution, the total time to execute is bounded by the runtime of the longest-running process.
-In the case of chromosome-parallelizable tools including Breakdancer, Pindel, and DELLY, this is typically the process working on the largest chromosome.\footnote{We note that one Breakdancer process, handling an unplaced contig in the hg19 reference genome, never completed in our runs and had to be killed manually; we exclude that process from our results.} In the case of MoDIL's run on the simulated data, we found that the different processes varied widely in their execution times, likely caused by regions of high coverage or with many ambiguously mapped reads.
-Cloudbreak mitigates this problem during the time-consuming feature generation process by using Hadoop partitioners to randomly assign each genomic location to one of the set of reducers, ensuring that the work is evenly distributed across all processes.
-This distribution of processing across the entire cluster also serves to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set in under 50 minutes during a run where one of the compute nodes was rebooted midway through the feature generation job.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{r|r|rrr|rrr}
-\multicolumn{2}{c}{}  & \multicolumn{3}{c}{Simulated Data} & \multicolumn{3}{c}{NA18507} \\
-\hline
- & SV Types &  Single CPU & Parallel & Proc. &  Single CPU & Parallel & Proc.  \\ 
-  \hline
-  Cloudbreak & D,I &   NA    & 308 & 312    & NA         & 2,310 & 636 \\ 
-  Breakdancer & D,I,V,T &  653   & NA       & NA          & 134,170 &  5,586 & 84 \\
-  GASVPro & D,V   &  3,339  & NA       & NA         & 52,385  & NA & NA \\
-  DELLY & D         &  1,964 & NA          & NA      & 30,311  & 20,224 & 84 \\
-  Pindel & D,I,V,P         & 37,006 &  4,885     & 8          &  284,932  & 28,587 & 84 \\ 
-  MoDIL & D,I        &  NA      & 52,547 & 250 & NA         & NA  & NA\\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Runtimes (elapsed) on both data sets of each tool tested, in single-processor and parallel mode.
-For parallel runs, Proc.
-is the maximum number of simultaneously running processes or threads.
-All times are in seconds.
-The types of variants detected by each program are listed with the abbreviations: D - deletion; I - insertion; V - Inversion; P - duplication; T - translocation.
-Interchromosomal translocations are only detected by Breakdancer in single CPU mode.
-}
-\label{runtimes}
-\end{table}
-
-\section{Methods}\label{discussion}
-
-Cloudbreak is a native Java Hadoop application.
-We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4.
-We use snappy compression for MapReduce data.
-Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
-To execute the other tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine with directed acyclic graphs to describe dependencies between jobs.\footnote{\url{http://research.cs.wisc.edu/htcondor/}}
-
-Simulated reads were aligned to hg18 chromosome 2, and NA18507 reads were aligned to the hg19 assembly.
-For each evaluated SV detection program we attempt to follow the authors' recommended alignment pipeline, except for MoDIL, for which we used reads aligned with BWA (see below).
-Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta) in single-ended mode, with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0},  requesting all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read.
-Alignments for the other programs were first aligned using BWA \autocite{Li:2009p836} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
-GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \texttt{-a -r -Ex 1100 -t 250}.
-
-We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
-To run Breakdancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer\_max} for each chromosome using the \texttt{-o} parameter.
-We ran DELLY version 0.0.9 with the \texttt{-p} parameter and default values for other parameters.
-For the parallel run of DELLY we first split the original BAM file with BamTools \autocite{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
-We ran GASVPro version 1.2 using the \texttt{GASVPro.sh} script and default parameters.
-Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \texttt{-c} option.
 
@@
 
+\section{Methods}\label{discussion}
+
+\subsection{Cloudbreak Implementation}
+Cloudbreak is a native Java Hadoop application.
+We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4.
+We use snappy compression for MapReduce data.
+Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
+To execute the other tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine with directed acyclic graphs to describe dependencies between jobs.\footnote{\url{http://research.cs.wisc.edu/htcondor/}}
+
+\subsection{Read Alignments}
+Simulated reads were aligned to hg18 chromosome 2, and NA18507 reads were aligned to the hg19 assembly.
+For each evaluated SV detection program we attempt to follow the authors' recommended alignment pipeline, except for MoDIL, for which we used reads aligned with BWA (see below).
+Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta) in single-ended mode, with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0},  requesting all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read.
+Alignments for the other programs were first aligned using BWA \autocite{Li:2009p836} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
+GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \texttt{-a -r -Ex 1100 -t 250}.
+
+\subsection{SV Tool Executution}
+We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
+To run Breakdancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer\_max} for each chromosome using the \texttt{-o} parameter.
+We ran DELLY version 0.0.9 with the \texttt{-p} parameter and default values for other parameters.
+For the parallel run of DELLY we first split the original BAM file with BamTools \autocite{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
+We ran GASVPro version 1.2 using the \texttt{GASVPro.sh} script and default parameters.
+Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \texttt{-c} option.
+
+\subsection{SV Prediction Evaluation}
+We use the following criteria to define a true prediction given a gold standard set of deletion and insertion variants to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method.
+For evaluating insertions, each algorithm produces insertion predictions that define an interval in which the insertion is predicted to have occurred with start and end coordinates $s$ and $e$ as well as the predicted length of the insertion, $l$.
+The true insertions are defined in terms of their actual insertion coordinate $i$ and their actual length $l_a$.
+Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$.
+In this study we are interested in detecting events larger than 40bp, because with longer reads, smaller events can be more easily discovered by examining gaps in individual reads.
+Both Pindel and MoDIL make many calls with a predicted event size of under 40bp, so we remove those calls from the output sets of those programs.
+Finally, we exclude from consideration calls from all approaches that match a true deletion of less than 40bp where the predicted variant length is less than or equal to 75bp in length.
+
+\newpage
+
 \section*{Acknowledgements}

### 2013-05-29 manuscript/nm_supp.tex
updates to NM manuscript

--- 2013-04-30 manuscript/manuscript.tex
@@
 \usepackage{rotating}
+\usepackage{xr}
 
-\usepackage[style=authoryear,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
-\renewbibmacro{in:}{%
-  \ifentrytype{article}{}{%
-  \printtext{\bibstring{in}\intitlepunct}}}
-\addbibresource{manuscript.bib}
+% \renewcommand{\thetable}{S\arabic{table}}   
+% \renewcommand{\thefigure}{S\arabic{figure}}
+% \renewcommand{\thealgorithm}{\arabic{algorithm}}
+
+\renewcommand{\figurename}{Supplementary Figure}
+\renewcommand{\tablename}{Supplementary Table}
+
+\newfloat{suppalgorithm}{p}{cap}
+\floatname{suppalgorithm}{Supplementary Algorithm}
 
@@
 
-\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}
+\begin{figure}[h]
+\centering
+\includegraphics[width=1\textwidth]{../figures/cloudbreak_mapred_diagram.pdf}
+\caption{An example of the Cloudbreak MapReduce algorithm.
+A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in single-ended mode to generate aligned reads.
+Reducers gather all alignments for both reads in each pair.
+B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) keys indicating the genomic location spanned by that pair.
+Only one genomic location is diagrammed here for simplicity.
+Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes.
+C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.}
+\label{algorithm_example}
+\end{figure}
 
-\author[1,2,4]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
-\author[1,2,3,4]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
-\affil[1]{Institute on Development and Disability}
-\affil[2]{Center for Spoken Language Understanding}
-\affil[3]{Department of Medical Informatics \& Clinical Epidemiology}
+\clearpage
 
-\affil[4]{Oregon Health \& Science University, Portland, OR, USA}
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{../figures/insert_size_mixtures.pdf}
+\caption{Illustration of insert size mixtures at individual genomic locations.
+A) there is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings.
+B) a homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes.
+C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected.
+D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
+\label{insert_size_mixes}
+\end{figure}
 
-\maketitle
+\clearpage
 
-\begin{abstract}
-The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data.
-Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods.
-However, the computational requirements for storing and processing data sets with multiple mappings can be formidable.
-Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers.
-MapReduce is becoming a standard framework for distributing processing across such compute clusters.
-We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome.
-In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
-On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
-In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp, and those in repetitive regions of the genome.
-In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples.
-Our implementation and source code are available at \url{http://sonmezsysbio.org/software/cloudbreak}.
-
-\medskip
-\noindent\textbf{Keywords:} genomic structural variation; distributed computing; copy number variation; high-throughput sequencing; genotyping.
-\end{abstract}
-
-\newpage
-
-\section{Introduction}
-
-Genomic structural variations such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}.
-However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.
-
-Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets.
-See \textcite{Alkan:2011p547} for a review.
-Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
-Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment.
-These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs.
-Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
-Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
-
-The first RP methods used only unambiguously discordantly mapped read pairs in their analyses.
-However, it has been recently demonstrated that including multiple mappings of discordant read pairs in the analysis can improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
-More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and thus be able to detect smaller events \autocite{Marschall:2012ek}.
-Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, in part because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.
-
-Google's MapReduce \autocite{Dean:2008p277} and its open-source implementation Hadoop\footnote{\url{http://hadoop.apache.org/}}, are designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
-Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, querying variant databases \autocite{Oconnor:2010p1835}, SNP calling \autocite{Langmead:2009p1225}, RNA-seq differential expression analysis \autocite{Langmead:2010p1268}, ChIP-seq peak calling \autocite{Feng:2011p1228}, and computing genome mappability \autocite{Lee:2012bk}.
-MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.
-
-In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings.
-We then demonstrate the development in this framework of an algorithm, Cloudbreak, for discovering genomic deletions and short insertions.
-Cloudbreak computes local features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
-Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm.
-We then characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.
-
-\section{A MapReduce Framework for SV Detection Algorithms}
-
-The MapReduce programming model \autocite{Dean:2008p277} divides computation across a cluster into three phases.
-In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
-The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
-Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
-Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
-MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
-
-The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based SV detection approaches in MapReduce.
-This is especially true because of the global clustering of paired end mappings at the heart of many RP approaches.
-MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
-In sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome.
-The RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features at locations along the genome and then producing SV predictions from those features in a post-processing step.
-Here we show that this technique can be easily translated to the MapReduce programming model.
-
-Our formulation divides processing into three distinct MapReduce jobs.
-The first job finds possible alignment locations for pairs of reads in FASTQ format.
-The second job computes features from the set of alignment pairs that are relevant to each of a set of small intervals covering the genome.
-In the third stage, the features for all locations on each chromosome are examined to find contiguous blocks of locations that support an SV call.
-This MapReduce framework is summarized in Algorithm 1.
-
-\begin{algorithm}
+\begin{suppalgorithm}[h]
 \algrenewcommand\algorithmicprocedure{\textbf{job}}
@@
   \end{algorithmic}
-\label{cloudbreakAlgorithm}
 \caption{The algorithmic framework for SV calling in MapReduce.}
-\end{algorithm}
+\label{cb_algo}
+\end{suppalgorithm}
 
-The alignment job uses sensitive mapping tools and settings to discover as many mapping locations for each read pair as possible.
-In the map phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair.
-In the reduce phase, the framework combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found.
-Our framework could use any aligner that is capable of reporting multiple alignments for reads.
-The results we report below were generated with alignments from the GEM mapper \autocite{MarcoSola:2012hm}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
-Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
+\newpage
 
-In the second job, we compute a set of features for each location in the genome.
-To begin, we tile the genome with small fixed-width, non-overlapping intervals.
-For the experiments reported in this paper we use an interval size of 25bp.
-Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
-Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
-Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
-For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
-the fragment size implied by the alignments and the likelihood the alignments are correct.
-We then leave two functions to be implemented depending on the application:
-\begin{flalign*}
- \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
- \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
-\end{flalign*}
-
-The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
-We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
-The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
-
-Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
-It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
-We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
-
-We describe our implementations of the $\textsc{Loci}$, $\Phi$, and $\textsc{PostProcess}$ functions below.
-However, we believe that there are many possible definitions that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage of the experiment, heterogeneity of the sample, etc.
-
-\subsection{Cloudbreak: A Hadoop Indel Detection Algorithm}
-
-We have implemented an indel detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak.
-Our implementation defines the three application-specific functions described above as:
-\begin{description}
-\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
-For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
-\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Figure \ref{insert_size_mixes}).
-Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
-If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant.
-Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components.
-Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth]{../figures/insert_size_mixtures.pdf}
-\caption{Illustration of insert size mixtures at individual genomic locations.
-A) there is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings.
-B) a homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes.
-C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected.
-D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
-\label{insert_size_mixes}
-\end{figure}
-
-To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
-We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
-In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
-This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
-Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
-
-\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
-
-And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
-
-We fit the parameters of the GMM using the Expectation-Maximization algorithm.
-Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
-We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
-In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
-We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions.
-In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$.
-We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$.
-We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.
-
-The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
-
-\item[\sc{PostProcess}] We convert our features along the genome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
-To eliminate noise we apply a median filter with window size 5.
-We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
-\end{description}
-
-A diagram of the Cloudbreak algorithm working on a simple example is shown in Figure \ref{algorithm_example}.
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth]{../figures/cloudbreak_mapred_diagram.pdf}
-\caption{An example of the Cloudbreak MapReduce algorithm.
-A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in single-ended mode to generate aligned reads.
-Reducers gather all alignments for both reads in each pair.
-B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) keys indicating the genomic location spanned by that pair.
-Only one genomic location is diagrammed here for simplicity.
-Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes.
-C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.}
-\label{algorithm_example}
-\end{figure}
-
-\section{Results}\label{results}
-
-We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
-We also attempted to evaluate MoDIL on the same data.
-All of these methods detect deletions; in addition Breakdancer, Pindel, and MoDIL are capable of detecting insertions.
-In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).
-
-We use the following criteria to define a true prediction given a gold standard set of deletion and insertion variants to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method.
-For evaluating insertions, each algorithm produces insertion predictions that define an interval in which the insertion is predicted to have occurred with start and end coordinates $s$ and $e$ as well as the predicted length of the insertion, $l$.
-The true insertions are defined in terms of their actual insertion coordinate $i$ and their actual length $l_a$.
-Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$.
-In this study we are interested in detecting events larger than 40bp, because with longer reads, smaller events can be more easily discovered by examining gaps in individual reads.
-Both Pindel and MoDIL make many calls with a predicted event size of under 40bp, so we remove those calls from the output sets of those programs.
-Finally, we exclude from consideration calls from all approaches that match a true deletion of less than 40bp where the predicted variant length is less than or equal to 75bp in length.
-
-It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods.
-Cloudbreak sacrifices additional resolution (by dividing the genome into 25bp windows); we strive however to increase sensitivity and specificity in the hopes that such calls could still be useful even if their resolution is limited.
-This seems reasonable to us given the possibility of pipelines in which RP calls are refined by SR mappings or local assembly methods.
-
-\subsection{Tests with Simulated Data}
-
-As has been observed elsewhere, there is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference.
-Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics.
-On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data.
-To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous insertions and deletions from the genome of J.
-Craig Venter \autocite{Levy:2007fb}.
-Since there are relatively few heterozygous insertions and deletions annotated in the Venter genome, we randomly assigned each variant to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
-We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package\footnote{\url{http://sourceforge.net/apps/mediawiki/dnaa/}}.
-We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence.
-Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous insertions and deletions.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth]{../figures/CHR2SIM_DELS_ROC.pdf}
-\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes.
-Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.}
-\label{chr2DeletionsRoc}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth]{../figures/CHR2SIM_INS_ROC.pdf}
-\caption{ROC curve of insertion prediction performance on the chromosome 2 simulated data set.}
-\label{chr2InsertionsRoc}
-\end{figure}
-
-Figure \ref{chr2DeletionsRoc} shows an ROC plot of the performance of each algorithm for detecting deletions on the simulated data set.
-All approaches show excellent specificity at high thresholds in this simulation.
-Cloudbreak provides the greatest specificity at higher levels of sensitivity, followed by DELLY.
-The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the tradeoff between precision and recall; therefore it is not included in this figure.
-A ROC curve for insertion prediction performance is shown in Figure \ref{chr2InsertionsRoc}.
-
-Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying levels.
-First, we used the ROC curves to attempt to characterize the predictions made by each algorithm when a low false discovery rate is required.
-Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve, representing cutoffs of 2.22 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 11 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 for Pindel.
-Cloudbreak shows good performance at detecting deletions in all size classes, and finds more medium and large deletions than the other approaches.
-Pindel exclusively identifies a large number of small (40bp-100bp) deletions, as is expected for a split-read method.
-Performance on insertions never reached a level with an FDR of 10\%, so insertion predictions are not included in this table.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{rrrrrr}
-  \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 224 &  84 & 82 &  31 & 26\\ 
-  \hline
-  Cloudbreak  &  47 (  7)  &  52 (\textbf{  3}) &  \textbf{ 57} (\textbf{  5}) & \textbf{ 12} (\textbf{  4}) & \textbf{ 15} (\textbf{0}) \\ 
-  Breakdancer &  52 ( 10)  &  49 (\textbf{  3}) &   49 (0) &   7 (0) &  14 (\textbf{0}) \\ 
-  GASVPro     &  35 (  4)  &  26 (0) &   26 (0) &   2 (0) &   6 (\textbf{0}) \\ 
-  DELLY       &  22 (  2)  & \textbf{ 56} (  2) &   40 (0) &   8 (0) &  12 (\textbf{0}) \\ 
-  Pindel      & \textbf{ 60} (\textbf{ 34})  &  16 (0) &   41 (  1) &   1 (0) &  12 (\textbf{0})\\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPredsFDR10}
-\end{table}
-
-We also characterized the predictions made by each algorithm at the operating point which gives them maximum sensitivity.
-For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.
-The results for both deletion and insertion predictions are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}.
-MoDIL and Cloudbreak exhibited the greatest recall for deletions.
-Cloudbreak has high precision and recall for deletions at this threshold, and surprisingly discovers more small deletions than Pindel.
-For insertions, Cloudbreak shows excellent precision, although recall is low for all four approaches, with Cloudbreak and Pindel showing the highest recall rates.
-Cloudbreak again identifies the most small variants.
-Pindel is the only tool which can consitently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{r|rrr|rrrrr}
-  \cline{2-9}
-   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
-\hline
-\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
-  \hline
-\cline{2-9}
-&  Cloudbreak    &  0.532 & \textbf{0.66} & \textbf{146} (  6)  &  58 (\textbf{  4}) &   62 (  1) &  14 (  1) &  15 (0) \\ 
-&  Breakdancer   &  0.356 & 0.49 &  89 (0)  &  54 (0) &   53 (0) &   8 (0) &  15 (0) \\ 
-&  GASVPro        & 0.146 & 0.432 &  83 (  2)  &  32 (0) &   55 (0) &   8 (0) &  15 (0) \\ 
-&  DELLY-RP           & 0.457 & 0.613 & 114 (  8)  & \textbf{ 68} (  2) &  \textbf{ 66} (0) &   9 (0) &  17 (0) \\ 
-&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  &   3 (0) &   49 (0) &   6 (0) &  16 (0) \\ 
-&  Pindel           & 0.462 & 0.421 &  96 (\textbf{ 13})  &  24 (0) &   48 (0) &   5 (0) &  15 (0)\\ 
-&  MoDIL           & 0.132  & \textbf{0.66} & 123 (  5)  &  66 (  2) &  \textbf{ 66} (\textbf{  7}) & \textbf{ 17} (\textbf{  6}) & \textbf{ 23} (\textbf{  8})\\ 
-   \hline
-\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
-\cline{2-9}
-&  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{ 67} ( 24)  &   8 (  3) &    3 (  1) &   1 (0) & 0 (0) \\ 
-&  Breakdancer & 0.262 & 0.0968  &  23 (  8)  & \textbf{ 14} (\textbf{  9}) &    2 (  1) & 0 (0) & 0 (0) \\ 
-&  Pindel          & 0.572 & \textbf{0.196} &  52 (\textbf{ 26})  &   5 (  3) &  \textbf{ 10} (\textbf{ 10}) & \textbf{  3} (\textbf{  3}) & \textbf{  9} (\textbf{  9})\\ 
-&  MoDIL          & 0.186 & 0.0521 &  14 (  1)  &   4 (  1) &    1 (0) &   2 (  1) & 0 (0)\\ 
-\hline
-\end{tabular}
-\end{center}
-\caption{The precision and recall of each tool for the simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels at maximum sensitivity, as well as the number of those variants discovered in each size class.
-Variants exclusively predicted by one tool are shown in parentheses.
-The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section.}
-\label{chr2DeletionAndInsertionPredsMaxSensitivity}
-\end{table}
-
-\subsection{Tests with Biological Data}
-
-We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive.\footnote{\url{http://sra.dnanexus.com/experiments/ERX009609}} This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp.
-The sample was sequenced to a depth of approximately 37X coverage.
-
-To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same sample.
-These included data from the Human Genome Structural Variation Project\footnote{\url{http://hgsv.washington.edu/general/download/SNPs_DIPs/}} reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
-We merged any overlapping calls of the same type into the region spanned by their unions.
-It should be noted that the 1000 Genomes call set was partially produced using DELLY and Breakdancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor.
-We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth]{../figures/NA18507_DELS_ROC.pdf}
-\caption{ROC curve of deletion prediction performance on the 37X NA18507 sample, tested against the combined gold standard deletion set taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.}
-\label{NA18507DeletionsRoc}
-\end{figure}
-
-Figure \ref{NA18507DeletionsRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard deletion set, using the same rules for a match as described in the previous section.
-All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing deletions in the gold standard set that are actually present in the sample.
-Cloudbreak is the best performer at more stringent thresholds, and remains near the top at higher sensitivity levels.
-
-We then tested the predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
-The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those thresholds, is shown in Table \ref{NA18507DeletionPreds}.
-Cloudbreak has the greatest sensitivity of any tool at these thresholds, identifying the most variants in each size class.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{rrr|rrrrr}
-  \hline
- & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
-Total Number & & & 7,466 & 235 & 218 & 110 & 375 \\
-  \hline
-Cloudbreak & 0.11 & \textbf{0.13} & \textbf{371} (\textbf{127})  & \textbf{146} (\textbf{ 15}) &  \textbf{177} (\textbf{  9}) & \textbf{ 95} (\textbf{  3}) & \textbf{301} (\textbf{ 19}) \\ 
-Breakdancer & 0.122 & 0.112 & 261 ( 41)  & 132 (  6) &  167 (0) &  92 (0) & 288 (  2) \\  
-  GASVPro & 0.135 & 0.0449 & 120 ( 28)  &  38 (  3) &   80 (0) &  29 (0) & 110 (0) \\ 
-  DELLY & 0.0824 & 0.091 & 143 (  8)  & 125 (  4) &  158 (  1) &  83 (  1) & 256 (  3) \\ 
-  Pindel & \textbf{0.16} & 0.0685 & 149 ( 12)  &  57 (0) &  140 (0) &  58 (0) & 172 (0) \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found.
-The same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.}
-\label{NA18507DeletionPreds}
-\end{table}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth]{../figures/NA18507_INS_ROC.pdf}
-\caption{ROC curve of insertion prediction performance on the 37X NA18507 sample.}
-\label{NA18507InsertionsRoc}
-\end{figure}
-
-A similar ROC-like curve for insertions on the NA18507 data set is shown in Figure \ref{NA18507InsertionsRoc}.
-The number of insertions detected at maximum sensitivity for the three algorithms is shown in Table \ref{NA18507InsertionPreds}, along with their precision and recall at that threshold.
-Cloudbreak has the greatest recall, identifying twice as many short insertions as Pindel, while Pindel has the best precision.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{rrr|rrrrr}
-  \hline
- & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp \\ 
-Total Number & & & 536 & 114 & 45 & 1 \\
-  \hline
-Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{ 83})  &  25 (  9) &    1 (0) & 0 (0) \\ 
-Breakdancer & 0.0281 & 0.181 &  97 ( 11)  & \textbf{ 27} (  9) &    2 (  1) & 0 (0) \\  
-  Pindel & \textbf{0.0387} & 0.239 & 144 ( 41)  &  14 (\textbf{ 12}) &  \textbf{  7} (\textbf{  7}) & \textbf{  1} (\textbf{  1})  \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The precision and recall with respect to the gold standard set of insertions for each tool on the NA18507 data, as well as the number of insertions in each size class found, at maximum sensitivity.}
-\label{NA18507InsertionPreds}
-\end{table}
-
-\subsection{Performance in Repetitive Areas}
-
-We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance.
-We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements in both the simulated and NA18507 data sets.
-Table \ref{deletionRepmaskpreds} shows the number of deletions that overlap repeats at the 10\% FDR cutoff described above.
-In the NA18507 data set, Cloudbreak identifies 15\% more repeat-overlapping variants than the nearest competitor, Breakdancer.
-Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
-Performance on insertions is displayed in Table \ref{insertionRepmaskpreds}; again, Cloudbreak identifies more events that lie in repetitive regions than the other tools.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\hline
- &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 120 & 327 & 553 & 7851 \\ 
-  \hline
-  Cloudbreak  &  25 (  2) & \textbf{158} ( 17) & \textbf{220} (\textbf{ 49}) & \textbf{870} (\textbf{124}) \\ 
-  Breakdancer & \textbf{ 29} (  6) & 142 (  7) & 186 ( 19) & 754 ( 30) \\
-  GASVPro     &  16 (  2) &  79 (  2) &  75 (  9) & 302 ( 22) \\
-  DELLY       &  21 (  2) & 117 (  2) & 147 (  9) & 618 (  8) \\
-  Pindel      &  18 (\textbf{  9}) & 112 (\textbf{ 26}) & 103 (  4) & 473 (  8) \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
-\label{deletionRepmaskpreds}
-\end{table}
-
-\begin{table}
-\begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\hline
- &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 133 & 270 & 341 & 355 \\ 
-  \hline
-  Cloudbreak  &  21 (  4) & \textbf{ 58} ( 24) & \textbf{137} (\textbf{ 37}) & \textbf{135} (\textbf{ 55}) \\ 
-  Breakdancer &  17 (  7) &  22 ( 11) &  82 ( 12) &  44 (  9) \\
-  Pindel      & \textbf{ 25} (\textbf{ 18}) &  54 (\textbf{ 33}) &  84 ( 36) &  82 ( 25) \\ 
-  MoDIL      &   5 (0) &  16 (  3) & NA & NA \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
-\label{insertionRepmaskpreds}
-\end{table}
-
-\subsection{Genotyping Variants}
-
-Since Cloudbreak explicitly fits a mixture of insert size distributions at each location, and each component of the mixture represents an allele containing or not containing an insertion or deletion, it should be possible to use the parameters of the fit GMM to infer the genotype of each predicted variant, assuming that our pipeline is capturing all relevant read mappings near the locus of the variant.
-To test this, we attempted to classify the true positive Cloudbreak predictions described above for both the simulated and biological data sets.
-Because we simulated a diploid genome, we had genotypes for the simulated data readily available.
-For benchmarking genotyping of deletions in the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
-We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes.
-By setting a simple cutoff of .35 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.1\% and 96.5\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively.
-Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
-For insertions, none of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertion rates that meet our size threshold that also had genotyping information.
-Of the 79 insertions detected by Cloudbreak on the simulated data set, 13 were heterozygous.
-Using the same cutoff of .35 on $\alpha$, Cloudbreak correctly classified 62 of the 66 homozygous insertions and 10 of the 13 heterozygous insertions discovered, for an overall accuracy rate of 91\%.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{r|r|rr|rr|}
-\multicolumn{2}{c}{}  & \multicolumn{4}{c}{Actual Genotypes} \\
-\multicolumn{2}{c}{}  & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\cline{3-6}
-\multicolumn{2}{c|}{} &  Homozygous & Heterozygous & Homozygous & Heterozygous \\ 
-\cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 100 & 4 &  77 & 12 \\
- & Heterozygous & 7 & 72 &  3 & 338 \\
-\cline{2-6}
-\end{tabular}
-\end{center}
-\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
-\label{deletionGenotypeaccuracy}
-\end{table}
-
-\subsection{Runtimes}
-
-We implemented and executed Cloudbreak on a 56-node Hadoop cluster, with 636 map slots and 477 reduce slots.
-Not including alignment time, we were able to process the Chromosome 2 simulated data in under 6 minutes, and the the NA18507 data set in under 40 minutes.
-For the simulated data set we used 100 reducers for the compute SV features job; for the real data set we used 300.
-The bulk of Cloudbreak's execution is spent in the feature generation step.
-Extracting deletion and insertion calls take under two minutes each for both the real and simulated data sets; the times are equal because each reducer is responsible for processing a single chromosome, and so the runtime is bounded by the length of time it takes to process the largest chromosome.
-
-In table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work.
-Each tool varies in the amount of parallelization supported.
-We report runtimes for tools run in their default single-threaded mode, as well as for levels of parallelization achievable with basic scripting, noting that one of the key advantages of Hadoop/MapReduce is the ability to scale parallel execution to the size of the available compute cluster without any custom programming.
-Pindel allows multi-threaded operation on multicore servers.
-Pindel and Breakdancer allow processing of a single chromosome in one process, so it is possible to execute all chromosomes in parallel on a cluster that has a job scheduler and shared filesystem.
-Breakdancer has an additional preprocessing step (\texttt{bam2cfg.pl}) which runs in a single thread.
-DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode.
-GASVPro allows parallelization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting, and we did not find that the MCMC module consumed most of the runtime in our experiments, so we do not attempt to parallelize this componen.
-The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers; we adapted these for use in our cluster.
-
-In parallel execution, the total time to execute is bounded by the runtime of the longest-running process.
-In the case of chromosome-parallelizable tools including Breakdancer, Pindel, and DELLY, this is typically the process working on the largest chromosome.\footnote{We note that one Breakdancer process, handling an unplaced contig in the hg19 reference genome, never completed in our runs and had to be killed manually; we exclude that process from our results.} In the case of MoDIL's run on the simulated data, we found that the different processes varied widely in their execution times, likely caused by regions of high coverage or with many ambiguously mapped reads.
-Cloudbreak mitigates this problem during the time-consuming feature generation process by using Hadoop partitioners to randomly assign each genomic location to one of the set of reducers, ensuring that the work is evenly distributed across all processes.
-This distribution of processing across the entire cluster also server to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set in under 50 minutes during a run where one of the compute nodes was rebooted midway through the feature generation job.
-
-\begin{table}
+\begin{table}[h]
 \begin{center}
@@
 
-\section{Methods}\label{discussion}
+\newpage
 
-Cloudbreak is a native Java Hadoop application.
-We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4.
-We use snappy compression for MapReduce data.
-Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
-To execute the other tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine with directed acyclic graphs to describe dependencies between jobs.\footnote{\url{http://research.cs.wisc.edu/htcondor/}}
-
-Simulated reads were aligned to hg18 chromosome 2, and NA18507 reads were aligned to the hg19 assembly.
-For each evaluated SV detection program we attempt to follow the authors' recommended alignment pipeline, except for MoDIL, for which we used reads aligned with BWA (see below).
-Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta) in single-ended mode, with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0},  requesting all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read.
-Alignments for the other programs were first aligned using BWA \autocite{Li:2009p836} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
-GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \texttt{-a -r -Ex 1100 -t 250}.
-
-We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
-To run Breakdancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer\_max} for each chromosome using the \texttt{-o} parameter.
-We ran DELLY version 0.0.9 with the \texttt{-p} parameter and default values for other parameters.
-For the parallel run of DELLY we first split the original BAM file with BamTools \autocite{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
-We ran GASVPro version 1.2 using the \texttt{GASVPro.sh} script and default parameters.
-Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \texttt{-c} option.
-
-\section{Discussion}\label{Discussion}
-
-Over the next few years, due to advances in sequencing technology, genomics data are expected to increase both in volume and in coverage by several orders of magnitude, expanding into the realm referred to as Big Data.
-In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines.
-In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.
-
-Among Big Data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
-In this paper, we introduce a novel conceptual framework for structural variation detection algorithms in MapReduce, based on computing local features along the genome.
-This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of a distributed deletion-finding algorithm based on fitting a GMM to the distribution of mapped insert sizes spanning each location in the genome.
-
-On simulated and real data sets, our approach performs well across a wide range of deletion sizes and achieves performance similar to or better than a variety of state-of-the-art structural variation detection algorithms that run on traditional workload clusters and servers.
-Detection of deletions is an important area of research; \textcite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
-In addition, an increased ability to detect variants in regions of the genome that are repetitive and difficult to align reads to opens up new possibilities in exploring human genetic variation.
-
-In addition to delivering state-of-the-art performance in a RP-based SV detection problem, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
-Using tools like Apache Whirr\footnote{\url{http://whirr.apache.org/}}, it is becoming trivially easy to instantiate on-demand Hadoop compute clusters with cloud providers such as Amazon's EC2.
-Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who are unable to afford or unwilling to maintain large, specialized servers to analyze their data.
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{rrr|rr}
+ & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
+\hline
+ &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
+ Total Number & 133 & 270 & 341 & 355 \\ 
+  \hline
+  Cloudbreak  & 21 (4) & \textbf{58} (24) & \textbf{137} (\textbf{37}) & \textbf{135} (\textbf{55}) \\ 
+  Breakdancer & 17 (7) & 22 (11) & 82 (12) & 44 (9) \\
+  Pindel      & \textbf{25} (\textbf{18}) & 54 (\textbf{33}) & 84 (36) & 82 (25) \\ 
+  MoDIL      & 5 (0) & 16 (3) & NA & NA \\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
+\label{insertionRepmaskpreds}
+\end{table}
 
@@
 
-\section*{Acknowledgements}
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{r|r|rr|rr|}
+\multicolumn{2}{c}{}  & \multicolumn{4}{c}{Actual Genotypes} \\
+\multicolumn{2}{c}{}  & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
+\cline{3-6}
+\multicolumn{2}{c|}{} &  Homozygous & Heterozygous & Homozygous & Heterozygous \\ 
+\cline{2-6}
+\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 100 & 4 &  77 & 12 \\
+ & Heterozygous & 7 & 72 &  3 & 338 \\
+\cline{2-6}
+\end{tabular}
+\end{center}
+\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
+\label{deletionGenotypeaccuracy}
+\end{table}
 
-We would like to thank Izhak Shafran at the Center for Spoken Language Understanding and Lucia Carbone of the Department of Behavioral Neuroscience for their advice, support, and shared computational resources.
+\newpage
 
-\printbibliography
+\section{Supplementary Discussion}
+
+\subsection{A general framework for SV Detection in MapReduce}
+
+\subsection{Runtime Analysis}
+
+We implemented and executed Cloudbreak on a 56-node Hadoop cluster, with 636 map slots and 477 reduce slots.
+Not including alignment time, we were able to process the Chromosome 2 simulated data in under 6 minutes, and the the NA18507 data set in under 40 minutes.
+For the simulated data set we used 100 reducers for the compute SV features job; for the real data set we used 300.
+The bulk of Cloudbreak's execution is spent in the feature generation step.
+Extracting deletion and insertion calls take under two minutes each for both the real and simulated data sets; the times are equal because each reducer is responsible for processing a single chromosome, and so the runtime is bounded by the length of time it takes to process the largest chromosome.
+
+In table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work.
+Each tool varies in the amount of parallelization supported.
+We report runtimes for tools run in their default single-threaded mode, as well as for levels of parallelization achievable with basic scripting, noting that one of the key advantages of Hadoop/MapReduce is the ability to scale parallel execution to the size of the available compute cluster without any custom programming.
+Pindel allows multi-threaded operation on multicore servers.
+Pindel and Breakdancer allow processing of a single chromosome in one process, so it is possible to execute all chromosomes in parallel on a cluster that has a job scheduler and shared filesystem.
+Breakdancer has an additional preprocessing step (\texttt{bam2cfg.pl}) which runs in a single thread.
+DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode.
+GASVPro allows parallelization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting, and we did not find that the MCMC module consumed most of the runtime in our experiments, so we do not attempt to parallelize this componen.
+The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers; we adapted these for use in our cluster.
+
+In parallel execution, the total time to execute is bounded by the runtime of the longest-running process.
+In the case of chromosome-parallelizable tools including Breakdancer, Pindel, and DELLY, this is typically the process working on the largest chromosome.\footnote{We note that one Breakdancer process, handling an unplaced contig in the hg19 reference genome, never completed in our runs and had to be killed manually; we exclude that process from our results.} In the case of MoDIL's run on the simulated data, we found that the different processes varied widely in their execution times, likely caused by regions of high coverage or with many ambiguously mapped reads.
+Cloudbreak mitigates this problem during the time-consuming feature generation process by using Hadoop partitioners to randomly assign each genomic location to one of the set of reducers, ensuring that the work is evenly distributed across all processes.
+This distribution of processing across the entire cluster also serves to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set in under 50 minutes during a run where one of the compute nodes was rebooted midway through the feature generation job.
 
 \end{document}
-
-% LocalWords:  Abyzov 2011bk Yoon 2010gf 2012kk Michaelson 2012fj 2012ey Feng
-% LocalWords:  Oconnor Langmead RNA-seq Langmead ChIP-seq 2012bk langle textbf
-% LocalWords:  algrenewcommand algorithmicprocedure textrm rp textrm textrm ap
-% LocalWords:  textrm textsc ldots spi RazerS Weese Novoalign Bowtie 2012jh sc
-% LocalWords:  rightarrow subseteq mathbb th 2007fb dwgsim includegraphics rr
-% LocalWords:  textwidth haplotypes chr2roc chr2preds rrrrrr hline Yoruban
-% LocalWords:  indels NA18507preds printbibliography

### 2013-05-29 src/main/scripts/manuscript.R

@@ -91,21 +91,28 @@ numPredictionsAtLevel <- function(perfTable, threshold) {
   return(perfTable[perfTable$Thresh == min(perfTable[perfTable$Thresh >= threshold,'Thresh']),'Calls'])
 }
 
+formatNumber <- function(d) {
+  formatC(d,digits=3,format="fg",width=-1,drop0trailing=TRUE)
+}
+
 printPredTableValue <- function(t, x, y) {
+  if (! y %in% colnames(t)) {
+    return(0)
+  }
   d <- t[x,y]
-  if (t[x,y] == max(t[,y])) {
-    return(paste0("\\textbf{",formatC(d,digits=3,format="fg",width=0),"}"))
+  if (d == max(t[,y])) {
+    return(paste0("\\textbf{",formatNumber(d),"}"))
   } else {
-    return(formatC(d,digits=3,format="fg",width=0))
+    return(formatNumber(d))
   }
 }
 
 printPredTableValueFromList <- function(x, l) {
-  d <- l[x]
+  d <- ifelse(is.na(l[x]), 0, l[x])
   if (d == max(l)) {
-    return(paste0("\\textbf{",formatC(d,digits=3,format="fg",width=0),"}"))
+    return(paste0("\\textbf{",formatNumber(d),"}"))
   } else {
-    return(formatC(d,digits=3,format="fg",width=0))
+    return(formatNumber(d))
   }
 }
 
@@ -173,23 +180,23 @@ modilChr2InsertionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeli
 # NA18507
 
 # Deletions
-cloudbreakNA18507DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/NA18507/gem_gmm_300reducers_del_perf.txt'
-breakdancerNA18507DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/NA18507/NA18507_35_2_3_singlecpu_gte40_perf.txt'
-gasvProNA18507DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/NA18507/gasvpro.gte40.perf.txt'
-dellyNA18507DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/NA18507/NA18507.delly_q0_c5_del_gte40.perf.txt'
[…]

### 2013-06-02 manuscript/nm.tex
manuscript updates, nm draft ready

@@
   \printtext{\bibstring{in}\intitlepunct}}}
-\addbibresource{manuscript.bib}
+\addbibresource{nm.bib}
+
+\newcommand{\degree}{\ensuremath{^\circ}}
 
@@
 
-\author[1,2,4]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
-\author[1,2,3,4]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
-\affil[1]{Institute on Development and Disability}
-\affil[2]{Center for Spoken Language Understanding}
-\affil[3]{Department of Medical Informatics \& Clinical Epidemiology}
-
-\affil[4]{Oregon Health \& Science University, Portland, OR, USA}
+\author[1,5]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
+\author[4,5]{Elizabeth Terhune}
+\author[3,4]{Jeffrey Tyner}
+\author[6]{Alberto L'Abbate}
+\author[6]{Clelia Tiziana Storlazzi}
+\author[4,5]{Lucia Carbone}
+\author[1,2,5]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
+\affil[1]{Institute on Development and Disability and Center for Spoken Language Understanding}
+\affil[2]{Department of Medical Informatics \& Clinical Epidemiology}
+\affil[3]{Program in Molecular and Cellular Biosciences}
+\affil[4]{Behavioral Neuroscience Department}
+\affil[5]{Oregon Health \& Science University, Portland, OR, USA}
+\affil[6]{Department of Genetics and Microbiology, University of Bari, Bari, Italy}
 
@@
 
-Google's MapReduce \autocite{Dean:2008p277} and its open-source implementation Hadoop\footnote{\url{http://hadoop.apache.org/}}, are designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
-Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, querying variant databases \autocite{Oconnor:2010p1835}, SNP calling \autocite{Langmead:2009p1225}, RNA-seq differential expression analysis \autocite{Langmead:2010p1268}, ChIP-seq peak calling \autocite{Feng:2011p1228}, and computing genome mappability \autocite{Lee:2012bk}.
-MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.
+Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
+Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
+Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, RNA-seq differential expression analysis \autocite{Langmead:2010p1268}, ChIP-seq peak calling \autocite{Feng:2011p1228}, and computing genome mappability \autocite{Lee:2012bk}.
+
+Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.
+MapReduce \autocite{Dean:2008p277} divides computation across a cluster into three phases.
+In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
+The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
+Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
+MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
+
+The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches.
+MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
+In sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome.
+SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features along the genome and then producing SV predictions from those features in a post-processing step.
 
 In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings.
-We then demonstrate the development in this framework of an algorithm, Cloudbreak, for discovering genomic deletions and short insertions.
-Cloudbreak computes local features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
+We then demonstrate the development in this framework of a software package, Cloudbreak, for discovering genomic deletions up to 25,000bp long, and short insertions.
+Cloudbreak computes local features that build on the ideas implemented in MoDIL \autocite{Lee:2009da}, which modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
 Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm.
-We then characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.
+We characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.
 
@@
 
-\subsection{A MapReduce Framework for SV Detection Algorithms}
-
-The MapReduce programming model \autocite{Dean:2008p277} divides computation across a cluster into three phases.
-In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
-The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
-Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
-Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
-MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
-
-The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based SV detection approaches in MapReduce.
-This is especially true because of the global clustering of paired end mappings at the heart of many RP approaches.
-MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
-In sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome.
-The RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features at locations along the genome and then producing SV predictions from those features in a post-processing step.
-Here we show that this technique can be easily translated to the MapReduce programming model.
-
-Our formulation divides processing into three distinct MapReduce jobs, and is summarized in Figure \ref{cloudbreak_workflow}.
-The first job finds possible alignment locations for pairs of reads in FASTQ format.
-The second job computes features from the set of alignment pairs that are relevant to each of a set of small intervals covering the genome.
-In the third stage, the features for all locations on each chromosome are examined to find contiguous blocks of locations that support an SV call.
-This MapReduce framework is summarized in Supplementary Algorithm \ref{Scb_algo}.
+\subsection{Cloudbreak: A Hadoop/MapReduce Software Package for SV Detection}
 
@@
 
-The alignment job uses sensitive mapping tools and settings to discover as many mapping locations for each read pair as possible.
+Our framework for SV detection in MapReduce divides processing into three distinct MapReduce jobs.
+This framework is summarized in Figure \ref{cloudbreak_workflow}.
+Details of the MapReduce SV detection algorithmic framework and how Cloudbreak is implemented within that framework are provided in Supplementary Algorithm \ref{Scb_algo} and the Supplementary Materials; here we proceed with a high-level description of each job.
+
+Cloudbreak's alignment job uses sensitive mapping tools and settings to discover as many mapping locations for each read pair as possible.
 In the map phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair.
 In the reduce phase, the framework combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found.
-Our framework could use any aligner that is capable of reporting multiple alignments for reads.
-The results we report below were generated with alignments from the GEM mapper \autocite{MarcoSola:2012hm}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
-Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
+Our framework could use any aligner that is capable of reporting many multiple alignments for reads.
+For the results below we used the GEM mapper \autocite{MarcoSola:2012hm}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
 
-In the second job, we compute a set of features for each location in the genome.
-To begin, we tile the genome with small fixed-width, non-overlapping intervals.
-For the experiments reported in this paper we use an interval size of 25bp.
-Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
-Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
-Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
-For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
-the fragment size implied by the alignments and the likelihood the alignments are correct.
-We then leave two functions to be implemented depending on the application:
-\begin{flalign*}
- \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
- \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
-\end{flalign*}
-
-The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
-We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
-The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
+In the next job, Cloudbreak computes a set of features for each location in the genome.
+To begin, we tile the genome with small fixed-width, non-overlapping windows.
+For the experiments reported in this paper we use a window size of 25bp.
+Within each window, we examine the distribution of insert sizes of read pairs that span that window, and compute features by fitting a Gaussian Mixture Model to that distribution (Supplementary Materials, Supplementary Figure \ref{Sinsert_size_mixes}).
+To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials)
 
 Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
-It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
+In this job, we search for contiguous blocks of genomic locations with similar features and merge them into individual insertion and deletion SV calls, employing noise reduction techniques to increase specificity (Supplementary Materials).
 We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
+An illustration of the Cloudbreak algorithm working on a simple example is shown in Supplementary Figure \ref{Salgorithm_example}.
 
-We describe our implementations of the $\textsc{Loci}$, $\Phi$, and $\textsc{PostProcess}$ functions below.
-However, we believe that there are many possible definitions that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage of the experiment, heterogeneity of the sample, etc.
-
-\subsection{Cloudbreak: A Hadoop Indel Detection Algorithm}
-
-We have implemented an indel detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak.
-Our implementation defines the three application-specific functions described above as:
-\begin{description}
-\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
-For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
-\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure \ref{Sinsert_size_mixes}).
-Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
-If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant.
-Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components.
-Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
-
-To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
-We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
-In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
-This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
-Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
-
-\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
-
-And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
-
-We fit the parameters of the GMM using the Expectation-Maximization algorithm.
-Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
-We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
-In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
-We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions.
-In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$.
-We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$.
-We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.
-
-The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
-
-\item[\sc{PostProcess}] We convert our features along the genome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
-To eliminate noise we apply a median filter with window size 5.
-We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
-\end{description}
-
-An illustration of the Cloudbreak algorithm working on a simple example is shown in Figure S\ref{Salgorithm_example}.
+Cloudbreak can be executed on any Hadoop cluster; Hadoop abstracts away the details of cluster configuration, making distributed applications portable.
+In addition, our Cloudbreak implementation can leverage the Apache Whirr library to automatically create clusters with cloud service providers such as the Amazon Elastic Compute Cloud (EC2).
+This enables on demand provisioning of Hadoop clusters which can then be terminated when processing is complete, eliminating the need to invest in a standing cluster and allowing a model in which users can scale their computational infrastructure as their need for it varies over time.
 
@@
 We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
-DELLY produces two sets of calls, one based solely on RP signals, and hte other based on RP calls that could be supported by SR evidence.
+DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence.
 We refer to these sets of calls as DELLY-RP and DELLY-SR in the following results.
@@
 All of these methods detect deletions; in addition Breakdancer, Pindel, and MoDIL are capable of detecting insertions.
-In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).
-
-It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods.
-Cloudbreak sacrifices additional resolution (by dividing the genome into 25bp windows); we strive however to increase sensitivity and specificity in the hopes that such calls could still be useful even if their resolution is limited.
-This seems reasonable to us given the possibility of pipelines in which RP calls are refined by SR mappings or local assembly methods.
+See Methods for details on how reads were aligned and each program was invoked.
 
@@
 Craig Venter \autocite{Levy:2007fb}.
-Since there are relatively few heterozygous insertions and deletions annotated in the Venter genome, we randomly assigned each variant to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
-We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package\footnote{\url{http://sourceforge.net/apps/mediawiki/dnaa/}}.
-We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence.
-Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous insertions and deletions.
+Using these variants, we simulated a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants, with 100bp reads and a mean fragment size of 300bp (See Methods).
 
@@
 (b) ROC curves for insertion predictions.
-(c) Runtimes for each tool, not includding alignment time, parallelized when possible (see Supplementary Material).}
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).}
 \label{chr2CombinedRoc}
@@
 
-Figure \ref{chr2CombinedRoc} shows ROC curves of the performance of each algorithm for detecting deletions and insertions on the simulated data set, as well as the runtimes of the approaches tested.
+Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristics (ROC) curves of the performance of each algorithm for detecting deletions and insertions on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment.
+See Methods for a description of how we identified correct predictions.
 All approaches show excellent specificity at high thresholds in this simulation.
@@
 For insertions, Cloudbreak clearly provides the best combinations of sensitivity and specificity.
-Cloudbreak's runtime is half that Breakdancer, the next fastest tool, albeit using many more CPUs.
-See Supplementary Material and Supplementary Table \ref{Sruntimes} for a discussion of runtimes and parallelization.
-The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the tradeoff between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster, so results are not included in the runtime figure.
+Cloudbreak's runtime is half that of Breakdancer, the next fastest tool, processing the simulated data in under six minutes.
+(Of course, Cloudbreak uses many more CPUs as distributed algorithm.
+See Supplementary Material and Supplementary Table \ref{Sruntimes} for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster, so results are not included in the runtime figure.
+Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow.
+Therefore, to test scalability we measured the runtime of that job on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Figure \ref{Sscalability}).
 
-Choosing the correct operating point or threshold can be difficult when operating on a new data set, but the use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying levels.
-First, we used the ROC curves to attempt to characterize the predictions made by each algorithm when a low false discovery rate is required.
-Table \ref{chr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve, representing cutoffs of 2.22 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 11 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 for Pindel.
-Cloudbreak shows good performance at detecting deletions in all size classes, and finds more medium and large deletions than the other approaches.
-Pindel exclusively identifies a large number of small (40bp-100bp) deletions, as is expected for a split-read method.
-Performance on insertions never reached a level with an FDR of 10\%, so insertion predictions are not included in this table.
-
-\begin{table}[t]
-\begin{center}
-\begin{tabular}{rrrrrr}
-  \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 224 &  84 & 82 &  31 & 26\\ 
-  \hline
-  Cloudbreak  & 47 (7)  & 52 (\textbf{3}) &  \textbf{57} (\textbf{5}) & \textbf{12} (\textbf{4}) & \textbf{15} (\textbf{0}) \\ 
-  Breakdancer & 52 (10)  & 49 (\textbf{3}) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
-  GASVPro     & 35 (4)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
-  DELLY-RP       & 22 (2)  & \textbf{56} (2) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
-  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
-  Pindel      & \textbf{60} (\textbf{34})  & 16 (0) &  41 (1) & 1 (0) & 12 (\textbf{0})\\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPredsFDR10}
-\end{table}
-
-We also characterized the predictions made by each algorithm at the operating point which gives them maximum sensitivity.
+Choosing the correct operating point or threshold to set on the output of an SV calling algorithm can be difficult when operating on a new data set.
+The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying levels.
+First, we characterized the predictions made by each algorithm at the operating point which gives them maximum sensitivity.
 For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.
@@
 Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.
+We also used the ROC curves to attempt to characterize the predictions made by each algorithm when a low false discovery rate is required.
+Supplementary Table \ref{Schr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
+Cloudbreak shows good performance at detecting deletions in all size classes, and finds more medium and large deletions than the other approaches.
+Performance on insertions never reached an FDR of 10\%, so insertion predictions are not included in this table.
+
+It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Figure \ref{Sbreakpoint_resolution}): SR methods have higher resolution than RP methods.
+Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we strive however to increase sensitivity and specificity to actual variants in the hopes that such calls could still be useful even if their resolution is limited.
+This seems reasonable to us, especially given the possibility of pipelines in which RP calls are validated \emph{in silico} by local assembly methods or split-read methods.
 
@@
 
-We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive.\footnote{\url{http://sra.dnanexus.com/experiments/ERX009609}} This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp.
-The sample was sequenced to a depth of approximately 37X coverage.
+We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive.
+This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage.
 
 To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same sample.
-These included data from the Human Genome Structural Variation Project\footnote{\url{http://hgsv.washington.edu/general/download/SNPs_DIPs/}} reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
+These included data from the Human Genome Structural Variation Project reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
 We merged any overlapping calls of the same type into the region spanned by their unions.
@@
 (a) ROC curve for deletion prediction performance, tested against the combined gold standard sets of deletions taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.
-(b) ROC curve for insertion prediciton performance.
-(c) Runtimes for each tool, not includding alignment time, parallelized when possible (see Supplementary Material).
+(b) ROC curve for insertion prediction performance.
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).
 }
@@
 
-Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard set for both deletions and insertions, using the same rules for a match as described in the previous section.
-All algorithms show far less specificity for the gold standard set than they did for the true deletions in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing deletions in the gold standard set that are actually present in the sample.
-Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
+Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard set for both deletions and insertions.
+All algorithms show far less specificity for the gold standard set than they did for the true variants in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample.
+For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
 Cloudbreak has optimal performance at all thresholds for insertions.
+Cloudbreak is again approximately twice as fast as the next fastest program, Breakdancer, processing the sample in under 40 minutes, even when Breakdancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).
 
-We then tested the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
+Given the high number of false positives produced by all tools at maximum sensitivity indicated by the ROC curves, we decided to characterize the predictions made by each tool at more stringent thresholds.
+We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
 For insertions, again, we were forced to use the thresholds that gave maximum sensitivity for each tool due to the high observed FDR rates in the simulated data.
-The precision and recall with respect to the gold standard deletion set, as well as the performance of each algorithm at predicting each size class at those thresholds, is shown in Table \ref{NA18507DeletionAndInsertionPreds}.
+The precision and recall at these thresholds with respect to the gold standard set, as well as the performance of each algorithm at predicting variants of each size class at those thresholds, is shown in Table \ref{NA18507DeletionAndInsertionPreds}.
 For deletions, Cloudbreak has the greatest sensitivity of any tool at these thresholds, identifying the most variants in each size class.
@@
 Exclusive predictions are in parentheses.
-For deletions, the same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.
+For deletions, the same cutoffs those which yielded a 10\% FDR on the simulated data, as reported in in Supplementary Table \ref{Schr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.
 For insertions, the maximum sensitivity cutoff was used.}
@@
 We also tested Cloudbreak on a sequencing data set obtained from a patient with acute myeloid leukemia (AML).
-We sequenced CD14+ cells on an Illumina Genome Analyzer II, producing 128,819,200 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
+This data set consisted of 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
 Using a pipeline consisting of Novoalign, Breakdancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions.
-Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even with lower coverage data sets with a greater variance of insert sizes.
-The variants identfied include deletions occurring in the genes CTDSPL/RBPS3, an AML tumor suppressor \autocite{Zheng:2012kk} and CAV2, a modulator of cancer progression \autocite{Lee:2011bz}.
+Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even when using lower coverage data sets with a greater variance of insert sizes.
+The variants identified include deletions in the gene CTDSPL/RBPS3, an AML tumor suppressor \autocite{Zheng:2012kk}, and NBEAL1, a gene upregulated in some cancers \autocite{Chen:2004jo}.
 We are currently investigating these deletions to determine their functional impact on this patient.
@@
 
-Since Cloudbreak explicitly fits a mixture of insert size distributions at each location, and each component of the mixture represents an allele containing or not containing an insertion or deletion, it should be possible to use the parameters of the fit GMM to infer the genotype of each predicted variant, assuming that our pipeline is capturing all relevant read mappings near the locus of the variant.
-To test this, we attempted to classify the true positive Cloudbreak predictions described above for both the simulated and biological data sets.
+Because Cloudbreak explicitly models zygosity in its feature generation algorithm, it is possible to predict the genotypes of identified variants.
+We tested this on both the simulated and NA18507 data sets.
 Because we simulated a diploid genome, we had genotypes for the simulated data readily available.
-For benchmarking genotyping of deletions in the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
-We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes.
-By setting a simple cutoff of .35 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.1\% and 96.5\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively.
+For the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
+Cloudbreak was able to achieve 88.1\% and 96.5\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
 Supplementary Table \ref{SdeletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
-For insertions, none of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertion rates that meet our size threshold that also had genotyping information.
+None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that meet our size threshold that also had genotyping information.
 Of the 79 insertions detected by Cloudbreak on the simulated data set, 13 were heterozygous.
-Using the same cutoff of .35 on $\alpha$, Cloudbreak correctly classified 62 of the 66 homozygous insertions and 10 of the 13 heterozygous insertions discovered, for an overall accuracy rate of 91\%.
+Cloudbreak correctly classified 62 of the 66 homozygous insertions and 10 of the 13 heterozygous insertions discovered, for an overall accuracy rate of 91\%.
 
@@
 In addition to delivering state-of-the-art performance in a RP-based SV detection problem, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
-Using tools like Apache Whirr\footnote{\url{http://whirr.apache.org/}}, it is becoming trivially easy to instantiate on-demand Hadoop compute clusters with cloud providers such as Amazon's EC2.
+Using tools like Apache Whirr (\texttt{http://whirr.apache.org/}, it is becoming trivially easy to instantiate on-demand Hadoop compute clusters with cloud providers such as Amazon's EC2.
 Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who are unable to afford or unwilling to maintain large, specialized servers to analyze their data.
@@
 Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
-To execute the other tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine with directed acyclic graphs to describe dependencies between jobs.\footnote{\url{http://research.cs.wisc.edu/htcondor/}}
+To execute the other tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine (\url{http://research.cs.wisc.edu/htcondor/}) with directed acyclic graphs to describe dependencies between jobs.
+
+\subsection{Cloudbreak Genotyping}
+We used the parameters of the fit GMM to infer the genotype of each predicted variant.
+Assuming that our pipeline is capturing all relevant read mappings near the locus of the variant, the genotype should be indicated by the estimated parameter $\alpha$, the mixing parameter that controls the weight of the two components in the GMM.
+We setting a simple cutoff of .35 on the average value of $\alpha$ for each prediction to call the predicted variant homozygous or heterozygous.
+We used the same cutoff for deletion and insertion predictions.
+
+\subsection{Read Simulation}
+Since there are relatively few heterozygous insertions and deletions annotated in the Venter genome, we used the set of homozygous indels contained in the HuRef data (\texttt{HuRef.homozygous\_indels.061109.gff}) and randomly assigned each variant to be either homozygous or heterozygous.
+Based on this genotype, we applied each variant to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
+We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package (\url{http://sourceforge.net/apps/mediawiki/dnaa/}).
+We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence.
+Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous insertions and deletions.
 
@@
 
-\subsection{SV Tool Executution}
+\subsection{SV Tool Execution}
 We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
@@
 
+\subsection{Preparation of AML Sample}
+
+Peripheral blood was collected from a patient with acute myelomonocytic leukemia (previously designated acute myeloid leukemia FAB M4) under a written and oral informed consent process reviewed and approved by the Institutional Review Board of Oregon Health \& Science University.
+Known cytogenetic abnormalities associated with this specimen included trisomy 8 and internal tandem duplications within the FLT3 gene.
+Mononuclear cells were separated on a Ficoll gradient, followed by red cell lysis.
+Mononuclear cells were immunostained using antibodies specific for CD3, CD14, CD34, and CD117 (all from BD Biosciences) and cell fractions were sorted using a BD FACSAria flow cytometer.
+Cell fractions isolated included T-cells (CD3-positive), malignant monocytes (CD14-positive), and malignant blasts (CD34, CD117-positive).
+We sequenced CD14+ cells on an Illumina Genome Analyzer II, producing 128,819,200 paired-end reads.
+
+\subsection{Validation of AML Deletions by PCR}
+
+Deletions identified in the AML dataset were validated by PCR.
+The selection of the SVs, previously identified by BreakDancer analysis, to be validated by PCR, was carried out relying on the score and the number of the reads supporting each single event.
+Appropriate primers were designed with an internet-based interface, Primer3 (\url{http://frodo.wi.mit.edu/}), considering the chromosome localization and orientation of the interval involved in the candidate rearrangement.
+The primers were checked for specificity using the BLAT tool of the UCSC Human Genome Browser (\url{http://genome.ucsc.edu/cgi-bin/hgBlat}).
+All the primer pairs were preliminarily tested on the patient genomic DNA and a normal genomic DNA as control.
+The PCR conditions were as follows: 2 min at 95\degree C followed by 35 cycles of 30 sec at 95\degree C, 20 sec at 60\degree C, and 2 min at 72\degree C.
+All the obtained PCR products were sequenced and analyzed by BLAT for sequence specificity.
+
 \newpage
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 
-We would like to thank Izhak Shafran at the Center for Spoken Language Understanding and Lucia Carbone of the Department of Behavioral Neuroscience for their advice, support, and shared computational resources.
+We would like to thank Izhak Shafran at the Center for Spoken Language Understanding for advice, support, and shared computational resources, and Bob Handsaker for helpful conversations.
 
@@
 \end{document}
-
-% LocalWords:  Abyzov 2011bk Yoon 2010gf 2012kk Michaelson 2012fj 2012ey Feng
-% LocalWords:  Oconnor Langmead RNA-seq Langmead ChIP-seq 2012bk langle textbf
-% LocalWords:  algrenewcommand algorithmicprocedure textrm rp textrm textrm ap
-% LocalWords:  textrm textsc ldots spi RazerS Weese Novoalign Bowtie 2012jh sc
-% LocalWords:  rightarrow subseteq mathbb th 2007fb dwgsim includegraphics rr
-% LocalWords:  textwidth haplotypes chr2roc chr2preds rrrrrr hline Yoruban
-% LocalWords:  indels NA18507preds printbibliography

### 2013-06-02 src/main/scripts/manuscript.R

@@ -26,12 +26,12 @@ rocLineTypes <- function(numLines) {
   seq(1,numLines)
 }
 
-plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legendLoc="bottomright",legendCex=.9) {
+plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legendLoc="bottomright",legendCex=.9,...) {
   maxFP <- totalDels
   if (is.na(maxTP)) {
     maxTP <- totalDels
   }
-  plot(0, type="n", ylim=c(0, maxTP), xlim=c(0,maxFP), xlab=ifelse(sim, "False Positives", "Novel Predictions"), ylab="True Positives", main=main)
+  plot(0, type="n", ylim=c(0, maxTP), xlim=c(0,maxFP), xlab=ifelse(sim, "False Positives", "Novel Predictions"), ylab="True Positives", main=main, ...)
   perfCols <- rocColors(length(perfs))
   lineTypes <- rocLineTypes(length(perfs))
   mapply(drawPerfLine, perfs, perfCols, lineTypes, MoreArgs=list(maxFP=maxFP))  
@@ -243,8 +243,17 @@ perfsListDelsChr2 <- list(cloudbreak=cloudbreakChr2DeletionPerf, breakdancer=bre
 pdf(chr2DeletionsROCOutputFile, width=10)
 par(xpd=T, mar=par()$mar+c(0,0,0,7))
 plotROC(perfsListDelsChr2, 
-        c("Cloudbreak", "Breakdancer","Pindel", "GASVPro", "DELLY_RP", "DELLY-SR"), 
+        c("Cloudbreak", "Breakdancer","Pindel", "GASVPro", "DELLY-RP", "DELLY-SR"), 
         totalDelsChr2, "Deletions in Venter diploid chr2 simulation",legendLoc=xy.coords(425,200), maxTP=350)
+dev.off()
+
+perfsListDelsChr2 <- list(cloudbreak=cloudbreakChr2DeletionPerf, breakdancer=breakdancerChr2DeletionPerf, pindel=pindelChr2DeletionPerf, gasv=gasvChr2DeletionPerf, delly=dellyChr2DeletionPerf, dellyBR=dellyBRChr2DeletionPerf)
+cairo_pdf("~/Documents/svpipeline/figures/chr2_del_roc_cairo.pdf", width=10, height=7)
[…]

### 2013-06-02 manuscript/NA18507DeletionPredsBySize.table.brew.tex
everything is now up-to-speed with auto R output
update chr2 max sensitivity table to have precision, recall

--- 2013-02-27 manuscript/NA18507preds.table.brew.tex
@@
-% latex table generated in R 2.15.2 by xtable 1.7-0 package
-% Tue Jan  1 22:36:02 2013
 \begin{table}[t]
@@
   GASVPro & <%= printPredTableValueFromList('GASVPro', precision) %> & <%= printPredTableValueFromList('GASVPro',recall) %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\
-  DELLY & <%= printPredTableValueFromList('DELLY', precision) %> & <%= printPredTableValueFromList('DELLY', recall) %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '> 1000bp') %> \\
+  DELLY-RP & <%= printPredTableValueFromList('DELLY', precision) %> & <%= printPredTableValueFromList('DELLY', recall) %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '> 1000bp') %> \\
+  DELLY-SR & <%= printPredTableValueFromList('DELLYSR', precision) %> & <%= printPredTableValueFromList('DELLYSR', recall) %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '> 1000bp') %> \\
   Pindel & <%= printPredTableValueFromList('Pindel', precision) %> & <%= printPredTableValueFromList('Pindel', recall) %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %> \\
@@
 \caption{The precision and recall with respect to the gold standard set of each tool on the NA18507 data, as well as the number of deletions in each size class found.
-The same cutoffs were used as for the simulated data as in Table \ref{chr2preds}, adjusted for the difference in coverage from 30X to 37X.}
-\label{NA18507preds}
+The same cutoffs were used as for the simulated data as in Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.}
+\label{NA18507DeletionPreds}
 \end{table}

### 2013-06-02 manuscript/NA18507InsertionPredsBySize.table.brew.tex

\begin{table}[t]
\begin{center}
\begin{tabular}{rrr|rrrrr}
  \hline
 & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp \\ 
Total Number & & & <%= format(totals['40-100bp'], big.mark=",") %> & <%= format(totals['101-250bp'], big.mark=",") %> & <%= format(totals['251-500bp'], big.mark=",") %> & <%= format(totals['501-1000bp'], big.mark=",") %> \\
  \hline
Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', precision) %> & <%= printPredTableValueFromList('Cloudbreak', recall) %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> \\ 
Breakdancer & <%= printPredTableValueFromList('Breakdancer', precision) %> & <%= printPredTableValueFromList('Breakdancer', recall) %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> \\  
  Pindel & <%= printPredTableValueFromList('Pindel', precision) %> & <%= printPredTableValueFromList('Pindel', recall) %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %>  \\ 
   \hline
\end{tabular}
\end{center}
\caption{The precision and recall with respect to the gold standard set of insertions for each tool on the NA18507 data, as well as the number of insertions in each size class found, at maximum sensitivity.}
\label{NA18507InsertionPreds}
\end{table}

### 2013-06-02 manuscript/bmc_article.tex

%% BioMed_Central_Tex_Template_v1.05
%%                                      %
%  bmc_article.tex            ver: 1.05 %
%                                       %

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%         <27 January 2006>           %%
%%                                     %%
%%                                     %%
%% Uses:                               %%
%% cite.sty, url.sty, bmc_article.cls  %%
%% ifthen.sty. multicol.sty		       %%
%%									   %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%	
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.pdf and the instructions for    %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\NeedsTeXFormat{LaTeX2e}[1995/12/01]
\documentclass[10pt]{bmc_article}

% Load packages
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%   
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %% 
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %% 
%%  submitted article.                         %% 
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

\setlength{\topmargin}{0.0cm}
\setlength{\textheight}{21.5cm}
\setlength{\oddsidemargin}{0cm} 
\setlength{\textwidth}{16.5cm}
\setlength{\columnsep}{0.6cm}

\newboolean{publ}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                              %%
%% You may change the following style settings  %%
%% Should you wish to format your article       %%
%% in a publication style for printing out and  %%
%% sharing with colleagues, but ensure that     %%
%% before submitting to BMC that the style is   %%
%% returned to the Review style setting.        %%
%%                                              %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Review style settings
%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

% Begin ...
\begin{document}
\begin{bmcformat}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Ensure \and is entered between all but   %%
%% the last two authors. This will be       %%
%% replaced by a comma in the final article %%
%%                                          %%
%% Ensure there are no trailing spaces at   %% 
%% the ends of the lines                    %%     	
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author{Christopher W. Whelan\correspondingauthor$^{1}$%
       \email{Christopher W. Whelan\correspondingauthor -whelanch@ohsu.edu}%
      \and
         Kemal S\"onmez\correspondingauthor$^{1,2}$%
         \email{Kemal S\"onmez\correspondingauthor - sonmezk@ohsu.edu}
      }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK
}%

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the Instructions for       %%
%% authors on http://www.biomedcentral.com  %%
%% and change the section headings          %%
%% accordingly.                             %%   
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        \paragraph*{Background:} 
The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data. Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods. However, the computational requirements for storing and processing data sets with multiple mappings can be formidable. Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers. MapReduce is becoming a standard framework for distributing processing across such compute clusters.      
        \paragraph*{Results:} We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome. In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome. On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes. In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome. In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples. 
        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
\end{abstract}

\ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the instructions for       %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and change the section headings          %%
%% accordingly.                             %% 
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
 Text for this section.\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}
  \subsection*{Results sub-heading}
    \subsubsection*{This is a sub-sub-heading}
      Sub-sub-sub-headings are made with the \textsl{\\subsubsection} command. \pb
      pb at end of lines ensures correct paragraph spacing.\pb
	  Text for this sub-sub-section \ldots
    \subsubsection*{Another sub-sub-sub-heading}
      Text for this sub-sub-section \ldots

  \subsection*{Another results sub-heading}
    Text for this sub-section \ldots

  \subsection*{Yet another results sub-heading}
    Text for this sub-section.  More results \ldots

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
  Text for this section \ldots

%%%%%%%%%%%%%%%%%%
\section*{Methods}
  \subsection*{Methods sub-heading for this section}
    Text for this sub-section \ldots

  \subsection*{Another methods sub-heading for this section}
    Text for this sub-section \ldots

  \subsection*{Yet another sub-heading for this section}
    Text for this sub-section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Authors contributions}
    Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  \ifthenelse{\boolean{publ}}{\small}{}
  Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%              
%%  Bmc_article.bst  will be used to                       %%
%%  create a .BBL file for submission, which includes      %%
%%  XML structured for BMC.                                %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %% 
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %% 
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{bmc_article} }     % Bibliography file (usually '*.bib' )

%%%%%%%%%%%

\ifthenelse{\boolean{publ}}{\end{multicols}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \subsection*{Figure 1 - Sample figure title}
      A short description of the figure content
      should go here.

  \subsection*{Figure 2 - Sample figure title}
      Figure legend text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Sample table title}
    Here is an example of a \emph{small} table in \LaTeX\ using  
    \verb|\tabular{...}|. This is where the description of the table 
    should go. \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        A1 & B2  & C3 \\ \hline
        A2 & ... & .. \\ \hline
        A3 & ..  & .  \\ \hline
      \end{tabular}
      }
  \subsection*{Table 2 - Sample table title}
    Large tables are attached as separate files but should
    still be described here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{bmcformat}
\end{document}

### 2013-06-02 manuscript/chr2DeletionAndInsertionPredsMaxSensitivity.table.brew.tex

\begin{table}[t]
\begin{center}
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & <%= totalDels['40-100bp'] %> &  <%= totalDels['101-250bp'] %> & <%= totalDels['251-500bp'] %> &  <%= totalDels['501-1000bp'] %> & <%= totalDels['> 1000bp'] %>\\ 
  \hline
\cline{2-9}
&  Cloudbreak    &  <%= printPredTableValueFromList('Cloudbreak', delPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '> 1000bp') %> \\ 
&  Breakdancer   &  <%= printPredTableValueFromList('Breakdancer', delPrecision) %> & <%= printPredTableValueFromList('Breakdancer', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '> 1000bp') %> \\ 
&  GASVPro        & <%= printPredTableValueFromList('GASVPro', delPrecision) %> & <%= printPredTableValueFromList('GASVPro', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '> 1000bp') %> \\ 
&  DELLY-RP           & <%= printPredTableValueFromList('DELLY', delPrecision) %> & <%= printPredTableValueFromList('DELLY', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '> 1000bp') %> \\ 
&  DELLY-SR           & <%= printPredTableValueFromList('DELLYSR', delPrecision) %> & <%= printPredTableValueFromList('DELLYSR', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '> 1000bp') %> \\ 
&  Pindel           & <%= printPredTableValueFromList('Pindel', delPrecision) %> & <%= printPredTableValueFromList('Pindel', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '> 1000bp') %>\\ 
&  MoDIL           & <%= printPredTableValueFromList('MoDIL', delPrecision) %>  & <%= printPredTableValueFromList('MoDIL', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '> 1000bp') %>\\ 
   \hline
\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & <%= totalIns['40-100bp'] %> &  <%= totalIns['101-250bp'] %> & <%= totalIns['251-500bp'] %> &  <%= totalIns['501-1000bp'] %> & <%= totalIns['> 1000bp'] %>\\ 
\cline{2-9}
&  Cloudbreak   &<%= printPredTableValueFromList('Cloudbreak', insPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', insRecall) %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '> 1000bp') %> \\ 
&  Breakdancer & <%= printPredTableValueFromList('Breakdancer', insPrecision) %> & <%= printPredTableValueFromList('Breakdancer', insRecall) %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '> 1000bp') %> \\ 
&  Pindel          & <%= printPredTableValueFromList('Pindel', insPrecision) %> & <%= printPredTableValueFromList('Pindel', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '> 1000bp') %>\\ 
&  MoDIL          & <%= printPredTableValueFromList('MoDIL', insPrecision) %> & <%= printPredTableValueFromList('MoDIL', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '> 1000bp') %>\\ 
\hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses). The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section.}
\label{chr2DeletionAndInsertionPredsMaxSensitivity}
\end{table}

### 2013-06-02 manuscript/deletionRepmask.table.brew.tex
everything is now up-to-speed with auto R output

--- 2013-02-27 manuscript/repmask.table.brew.tex
@@
   GASVPro     & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'GASVPro', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'GASVPro', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'GASVPro', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'GASVPro', '1') %> \\
-  DELLY       & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY', '1') %> \\
+  DELLY-RP       & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY-RP', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY-RP', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY-RP', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY-RP', '1') %> \\
+  DELLY-SR       & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY-SR', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY-SR', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY-SR', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY-SR', '1') %> \\
   Pindel      & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '1') %> \\

### 2013-06-02 manuscript/insertionRepmask.table.brew.tex

@@
   Pindel      & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '1') %> \\
+  MoDIL      & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'MoDIL', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'MoDIL', '1') %> & NA & NA \\ 
    \hline

### 2013-06-02 manuscript/manuscript.tex

@@
 
+\subsection{Performance on a Low-Coverage Cancer Data Set}
+
+We also tested Cloudbreak on a sequencing data set obtained from a patient with acute myeloid leukemia (AML).
+We sequenced CD14+ cells on an Illumina Genome Analyzer II, producing 128,819,200 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
+Using a pipeline consisting of Novoalign, Breakdancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions.
+Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even with lower coverage data sets with a greater variance of insert sizes.
+The variants identfied include deletions occurring in the genes CTDSPL/RBPS3, an AML tumor suppressor \autocite{Zheng:2012kk} and CAV2, a modulator of cancer progression \autocite{Lee:2011bz}.
+We are currently investigating these deletions to determine their functional impact on this patient.
+
 \subsection{Performance in Repetitive Areas}
@@
   \hline
-  Cloudbreak  &  21 (  4) & \textbf{ 58} ( 24) & \textbf{137} (\textbf{ 37}) & \textbf{135} (\textbf{ 55}) \\ 
-  Breakdancer &  17 (  7) &  22 ( 11) &  82 ( 12) &  44 (  9) \\
-  Pindel      & \textbf{ 25} (\textbf{ 18}) &  54 (\textbf{ 33}) &  84 ( 36) &  82 ( 25) \\ 
-  MoDIL      &   5 (0) &  16 (  3) & NA & NA \\ 
+  Cloudbreak  & 21 (4) & \textbf{58} (24) & \textbf{137} (\textbf{37}) & \textbf{135} (\textbf{55}) \\ 
+  Breakdancer & 17 (7) & 22 (11) & 82 (12) & 44 (9) \\
+  Pindel      & \textbf{25} (\textbf{18}) & 54 (\textbf{33}) & 84 (36) & 82 (25) \\ 
+  MoDIL      & 5 (0) & 16 (3) & NA & NA \\ 
    \hline
@@
 We found that we could use $\alpha$, the mixing parameter that controls the weight of the two components in the GMM, to accurately predict deletion genotypes.
-By setting a simple cutoff of .35 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 88.1\% and 96.5\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively.
+By setting a simple cutoff of .25 on the average value of $\alpha$ in each Cloudbreak prediction, we were able to achieve 86.8\% and 96.0\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets respectively.
 Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
@@
 \cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 100 & 4 &  77 & 12 \\
- & Heterozygous & 7 & 72 &  3 & 338 \\
+\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 100 & 4 &  95 & 19 \\
+ & Heterozygous & 7 & 72 &  4 & 451 \\
 \cline{2-6}

### 2013-06-02 manuscript/nm_supp.tex

@@
 
+\usepackage[style=nature,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
+\renewbibmacro{in:}{%
+  \ifentrytype{article}{}{%
+  \printtext{\bibstring{in}\intitlepunct}}}
+\addbibresource{manuscript.bib}
+
 \renewcommand{\figurename}{Supplementary Figure}
@@
 \begin{document}
+
+\title{Supplementary Material for Cloudbreak}
+\author[1,2,4]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
+\author[4]{Elizabeth Terhune}
+\author[5]{Alberto L'Abbate}
+\author[5]{Clelia Tiziana Storlazzi}
+\author[4]{Jeffry Tyner}
+\author[4]{Lucia Carbone}
+\author[1,2,3,4]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
+\affil[1]{Institute on Development and Disability}
+\affil[2]{Center for Spoken Language Understanding}
+\affil[3]{Department of Medical Informatics \& Clinical Epidemiology}
+
+\affil[4]{Oregon Health \& Science University, Portland, OR, USA}
+\affil[5]{University of Bari, Bari, Italy}
+
+\maketitle
 
@@
 \label{insert_size_mixes}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{../figures/runtimeByNodes.pdf}
+\caption{Scalability of the Cloudbreak algorithm.
+Runtime of the Cloudbreak feature generation job for the simulated Chromosome 2 data is shown on Hadoop clusters consisting of varying numbers of compute nodes.
+Clusters were created in the Amazon Elastic Compute Cloud.}
+\label{scalability}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{../figures/breakpoint_resolution.pdf}
+\caption{Breakpoint resolution for each tool for deletions on the Chromosome 2 simulated data.
+For each correctly predicted deletion, we calculated the difference in length between the true deletion and the prediction.}
+\label{breakpoint_resolution}
 \end{figure}
@@
 
-\newpage
+\clearpage
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{rrrrrr}
+  \hline
+ & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number & 224 &  84 & 82 &  31 & 26\\ 
+  \hline
+  Cloudbreak  & 47 (7)  & 52 (\textbf{3}) &  \textbf{57} (\textbf{5}) & \textbf{12} (\textbf{4}) & \textbf{15} (\textbf{0}) \\ 
+  Breakdancer & 52 (10)  & 49 (\textbf{3}) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
+  GASVPro     & 35 (4)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
+  DELLY-RP       & 22 (2)  & \textbf{56} (2) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
+  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
+  Pindel      & \textbf{60} (\textbf{34})  & 16 (0) &  41 (1) & 1 (0) & 12 (\textbf{0})\\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
+The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.
+Thresholds for this FDR were derived from the ROC curves for each algorithm: representing cutoffs of 2.29 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 11 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 for Pindel.}
+\label{chr2DeletionPredsFDR10}
+\end{table}
+
+\clearpage
 
@@
 
+We have developed a conceptual algorithmic framework for SV detection in MapReduce, which is outlined in Supplementary Algorithm \ref{cb_algo}.
+As described in the main text, this framework divides processing into three separate MapReduce jobs: an alignment job, a feature computation job, and an SV calling job.
+
+The alignment job uses sensitive mapping tools and settings to discover as many mapping locations for each read pair as possible.
+Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in single ended mode and emits the possible locations under the $rpid$ key.
+Reducers then collect the alignments for each paired end, making them available under one key for the next job.
+Our implementation contains wrappers to execute the aligners GEM, Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
+
+In the second job, we compute a set of features for each location in the genome.
+To begin, we tile the genome with small fixed-width, non-overlapping intervals.
+For the experiments reported in this paper we use an interval size of 25bp.
+Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
+Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
+Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
+For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
+the fragment size implied by the alignments and the likelihood the alignments are correct.
+We then leave two functions to be implemented depending on the application:
+\begin{flalign*}
+ \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
+ \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
+\end{flalign*}
+
+The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
+We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
+The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
+
+Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
+It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
+We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
+
+\subsection{Cloudbreak: An Implementation of the MapReduce SV Detection Framework}
+
+Our Cloudbreak software package can be seen as an implementation of the general framework defined above. In particular, Cloudbreak implements the two user-defined function described above as follows:
+
+\begin{description}
+\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
+For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
+\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure \ref{insert_size_mixes}).
+Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
+If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant.
+Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components.
+Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
+
+To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
+We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
+In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
+This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
+Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
+
+\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
+
+And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
+
+We fit the parameters of the GMM using the Expectation-Maximization algorithm.
+Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
+We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
+In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
+We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions.
+In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$.
+We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$.
+We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.
+
+The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
+
+\item[\sc{PostProcess}] We convert our features along the genome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
+To eliminate noise we apply a median filter with window size 5.
+We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
+\end{description}
+
 \subsection{Runtime Analysis}
@@
 DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode.
-GASVPro allows parallelization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting, and we did not find that the MCMC module consumed most of the runtime in our experiments, so we do not attempt to parallelize this componen.
+GASVPro allows parallelization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting, and we did not find that the MCMC module consumed most of the runtime in our experiments, so we do not attempt to parallelize this component.
 The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers; we adapted these for use in our cluster.
@@
 
+\printbibliography
+
 \end{document}

### 2013-06-02 manuscript/recomb_abstract.tex

% Seventh M.I.T. Conference on Computational Fluid and Solid Mechanics
% One-Page Abstract Template
% 20081222PA
\documentclass[11pt,a4paper]{article}

% Page settings
%\usepackage{times}
%\usepackage{amsmath}
%
\usepackage{url}
\setlength{\voffset}{-1in}
\setlength{\topmargin}{1in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\textheight}{9in}
%
\setlength{\hoffset}{-1in}
\setlength{\oddsidemargin}{1in}
\setlength{\evensidemargin}{1in}
\setlength{\textwidth}{6.5in}
%
\setlength{\parskip}{0.175in}
\setlength{\parindent}{0in}

\begin{document}
\thispagestyle{empty}
%
\begin{center}
%%%
%%% Enter title in between the brackets and in lower case
%%%
\textbf{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}
\end{center}

\begin{center}
%%%
%%% Enter your name
%%%
Christopher W. Whelan$^1$ and Kemal S\"onmez$^{1,2}$
\\
%%%
%%% Enter your affiliation
%%%
$^1$Institute on Development and Disability, Center for Spoken Language Understanding \\
$^2$Department of Medical Informatics \& Clinical Epidemiology \\
Oregon Health \& Science University, Portland, OR, USA \\
%%%
%%% Enter your e-mail address
%%%
\texttt{cwhelan@gmail.com}, \texttt{sonmezk@ohsu.edu}

\end{center}

%%%
%%% Enter the text of your abstract
%%%
The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data. Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods. However, the computational requirements for storing and processing data sets with multiple mappings can be formidable. Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers.

MapReduce, via its Hadoop implementation, is becoming a standard
architecture for distributing processing across such compute
clusters. In this work we describe a novel conceptual
framework for structural variation detection in
MapReduce/Hadoop based on computing local features along the
genome. Our framework uses Hadoop to take advantage of distributed
computing to find all possible read alignments using modern short-read
aligners run with sensitive settings. We then provide an architecture
to first compute features for each
genomic location from the relevant alignments, and
then to call structural variants from the set of all features across
the genome.

In this framework, we have developed and evaluated a distributed
deletion-finding algorithm based on fitting a Gaussian mixture model
(GMM) to the distribution of mapped insert sizes spanning each
location in the genome. A similar method was used in MoDIL[1];
however, our algorithm and the Hadoop framework drastically reduce
the runtime requirements and overall difficulty of using this approach.

On simulated and real data sets of paired-end reads, our algorithm achieves performance
similar to or better than a variety of popular structural variation
detection algorithms, including read-pair, split-read, and hybrid
approaches. Cloudbreak performs well on both small and medium size
deletions, and in our simulations has greater sensitivity at most fixed levels of
specificity. We also show increased performance
in repetitive areas of the genome, identifying more deletions that
overlap repeats than other approaches in both simulated and real data.

In addition, our algorithm can accurately genotype heterozygous and
homozygous deletions from diploid samples. Using the parameters
computed in fitting the GMM and a simple thresholding procedure, we were able to achieve 88.0\% and 94.9\% accuracy in predicting the genotype of the true positive deletions we detected in simulated and real data sets, respectively.

Finally, we have recently added the ability to detect insertions to
Cloudbreak. Our implementation and source code are available at \url{https://github.com/cwhelan/cloudbreak}.

[1] Lee, S. et al., 2009. MoDIL: detecting small indels from clone-end
sequencing with mixtures of distributions. {\it Nat. Methods}, 6(7), pp.473–474.
\end{document}

### 2013-06-09 src/main/scripts/manuscript.R
characterize differences between gem and bwa alignments, insert size tweak

@@ -126,6 +126,7 @@ printPredTableCell <- function(preds, ex, x, y) {
 
 # deletions
 cloudbreakChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_e6_s2_del.perf.txt'
+cloudbreakBWAChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/bwa_del.perf.txt'
 breakdancerChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_35_2_3_del.perf.txt'
 gasvProChr2DeletionPerfFile <-  '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gasvpro_gte40.perf.txt'
 dellyChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_gte40.perf.txt'
@@ -141,20 +142,24 @@ cloudbreakChr2Perf50000ReportPerfFile <- '~/Documents/gene_rearrange/svpipeline/
 cloudbreakChr2Perf100000ReportPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_e6_s2_del_d100000.perf.txt'
 
 cloudbreakChr2DeletionPredictionsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_del_calls_gt2p29.bed'
+cloudbreakBWAChr2DeletionPredictionsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/bwa_del_1p89.bed'
 
 cloudbreakChr2DeletionHitsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_del_calls_gt2p29.hits.txt'
[…]

### 2013-06-11 manuscript/nm.tex
update to use BWA mappings

@@
 \author[1,5]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
-\author[4,5]{Elizabeth Terhune}
 \author[3,4]{Jeffrey Tyner}
@@
 Reads are first uploaded to a Hadoop cluster from local storage.
-Cloudbreak then executes three MapReduce jobs to process the data: 1) Mapping with sensitive settings, reporting multiple alignments.
+Cloudbreak then executes three MapReduce jobs to process the data: 1) Mapping with sensitive settings.
 2) Computation of features across the genome.
@@
 
-Cloudbreak's alignment job uses sensitive mapping tools and settings to discover as many mapping locations for each read pair as possible.
-In the map phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair.
-In the reduce phase, the framework combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found.
-Our framework could use any aligner that is capable of reporting many multiple alignments for reads.
-For the results below we used the GEM mapper \autocite{MarcoSola:2012hm}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop.
+Cloudbreak's alignment job can run a variety of alignment tools that report reads in SAM format (Supplementary Materials).
+In the map phase, mappers align reads in either single-end or paired-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair.
+In the reduce phase, the framework combines the reported mapping locations for the two ends of each read pair.
 
@@
 Within each window, we examine the distribution of insert sizes of read pairs that span that window, and compute features by fitting a Gaussian Mixture Model to that distribution (Supplementary Materials, Supplementary Figure \ref{Sinsert_size_mixes}).
-To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials)
+To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials); these procedures also allow us to process multiple mappings for each read if they are reported by the aligner.
 
@@
 See Methods for details on how reads were aligned and each program was invoked.
+For all alignments we used BWA \autocite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Figure \ref{Salignment_comparison}).
 
@@
 MoDIL and Cloudbreak exhibited the greatest recall for deletions.
-Cloudbreak has high precision and recall for deletions at this threshold, and surprisingly discovers more small deletions than Pindel.
-For insertions, Cloudbreak shows excellent precision, although recall is low for all four approaches, with Cloudbreak and Pindel showing the highest recall rates.
+Cloudbreak has high precision and recall for deletions at this threshold, and discovers many more small deletions.
+For insertions, Cloudbreak has the highest recall, although recall is low for all four approaches.
 Cloudbreak again identifies the most small variants.
@@
 Supplementary Table \ref{Schr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
-Cloudbreak shows good performance at detecting deletions in all size classes, and finds more medium and large deletions than the other approaches.
-Performance on insertions never reached an FDR of 10\%, so insertion predictions are not included in this table.
+At this more stringent threshold, Cloudbreak identifies more deletions in every size category than any other tool.
+Performance on insertions never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.
+We also examined Cloudbreak's ability to detect events in repetitive regions of the genome, and found that it was similar to the other methods tested (Supplementary Tables \ref{SdeletionRepmaskpreds} and \ref{SinsertionRepmaskpreds}).
 
@@
 \cline{2-9}
-&  Cloudbreak    &  0.532 & \textbf{0.66} & \textbf{146} (6)  & 58 (\textbf{4}) &  62 (1) & 14 (1) & 15 (0) \\ 
+&  Cloudbreak    &  0.638 & \textbf{0.678} & \textbf{153} (9)  & 61 (0) &  62 (0) & 12 (0) & 15 (0) \\ 
 &  Breakdancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
 &  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
-&  DELLY-RP           & 0.457 & 0.613 & 114 (8)  & \textbf{68} (2) &  \textbf{66} (0) & 9 (0) & 17 (0) \\ 
+&  DELLY-RP           & 0.457 & 0.613 & 114 (3)  & \textbf{68} (0) &  \textbf{66} (0) & 9 (1) & 17 (0) \\ 
 &  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
-&  Pindel           & 0.462 & 0.421 & 96 (\textbf{13})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
-&  MoDIL           & 0.132  & \textbf{0.66} & 123 (5)  & 66 (2) &  \textbf{66} (\textbf{7}) & \textbf{17} (\textbf{6}) & \textbf{23} (\textbf{8})\\ 
+&  Pindel           & 0.462 & 0.421 & 96 (\textbf{11})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
+&  MoDIL           & 0.132  & 0.66 & 123 (6)  & 66 (\textbf{3}) &  \textbf{66} (\textbf{11}) & \textbf{17} (\textbf{7}) & \textbf{23} (\textbf{8})\\ 
    \hline
@@
 \cline{2-9}
-&  Cloudbreak   &\textbf{0.705} & \textbf{0.196}  & \textbf{67} (24)  & 8 (3) &  3 (1) & 1 (0) & 0 (0) \\ 
-&  Breakdancer & 0.262 & 0.0968  & 23 (8)  & \textbf{14} (\textbf{9}) &  2 (1) & 0 (0) & 0 (0) \\ 
-&  Pindel          & 0.572 & \textbf{0.196} & 52 (\textbf{26})  & 5 (3) &  \textbf{10} (\textbf{10}) & \textbf{3} (\textbf{3}) & \textbf{9} (\textbf{9})\\ 
-&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (1) &  1 (0) & 2 (1) & 0 (0)\\ 
+&  Cloudbreak   &0.451 & \textbf{0.305}  & \textbf{79} (\textbf{32})  & \textbf{32} (\textbf{18}) &  \textbf{11} (8) & 1 (0) & 0 (0) \\ 
+&  Breakdancer & 0.262 & 0.0968  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
+&  Pindel          & \textbf{0.572} & 0.196 & 52 (25)  & 5 (1) &  10 (\textbf{9}) & \textbf{3} (\textbf{2}) & \textbf{9} (\textbf{9})\\ 
+&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (0) &  1 (0) & 2 (\textbf{2}) & 0 (0)\\ 
 \hline
@@
 For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
-Cloudbreak has optimal performance at all thresholds for insertions.
-Cloudbreak is again approximately twice as fast as the next fastest program, Breakdancer, processing the sample in under 40 minutes, even when Breakdancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).
+Cloudbreak has slightly lower accuracy for insertions than other tools, although it can identify the most variants at higher levels of sensitivity.
+Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, Breakdancer, even when Breakdancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).
 
@@
 For insertions, Pindel again has the highest precision at maximum sensitivity, although according the ROC curve it is possible to choose a threshold for Cloudbreak with higher precision and recall.
-At maximum sensitivity, Cloudbreak identifies 100 more insertions from the gold standard set than Pindel, giving it by far the highest recall.
+At maximum sensitivity, Cloudbreak identifies 120 more insertions from the gold standard set than Pindel, giving it by far the highest recall.
 
@@
 \cline{2-9}
-& Cloudbreak & 0.124 & \textbf{0.142} & \textbf{371} (\textbf{127})  & \textbf{150} (\textbf{15}) &  \textbf{187} (\textbf{9}) & \textbf{123} (\textbf{8}) & \textbf{396} (\textbf{27}) \\ 
-& Breakdancer & 0.137 & 0.123 & 261 (41)  & 136 (6) &  178 (0) & 114 (0) & 371 (2) \\  
-&  GASVPro & 0.147 & 0.0474 & 120 (28)  & 40 (3) &  85 (0) & 36 (0) & 128 (0) \\ 
-&  DELLY-RP & 0.0931 & 0.1 & 143 (8)  & 128 (4) &  167 (1) & 103 (1) & 323 (1) \\ 
+& Cloudbreak & 0.0943 & \textbf{0.17} & \textbf{573} (\textbf{277})  & \textbf{176} (\textbf{30}) &  \textbf{197} (\textbf{18}) & \textbf{121} (\textbf{6}) & \textbf{399} (\textbf{24}) \\ 
+& Breakdancer & 0.137 & 0.123 & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
+&  GASVPro & 0.147 & 0.0474 & 120 (21)  & 40 (2) &  85 (0) & 36 (0) & 128 (0) \\ 
+&  DELLY-RP & 0.0931 & 0.1 & 143 (6)  & 128 (3) &  167 (1) & 103 (0) & 323 (1) \\ 
 &  DELLY-SR & 0.153 & 0.0485 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
-&  Pindel & \textbf{0.179} & 0.0748 & 149 (12)  & 61 (0) &  149 (0) & 69 (0) & 217 (0) \\ 
+&  Pindel & \textbf{0.179} & 0.0748 & 149 (8)  & 61 (0) &  149 (0) & 69 (1) & 217 (0) \\ 
 \hline
@@
 \cline{2-9}
-& Cloudbreak & 0.0268 & \textbf{0.391} & \textbf{246} (\textbf{83})  & 25 (9) &  1 (0) & 0 (0)  & 0 (0)  \\ 
-& Breakdancer & 0.0281 & 0.181 & 97 (11)  & \textbf{27} (9) &  2 (1) & 0 (0) & 0 (0) \\  
-&  Pindel & \textbf{0.0387} & 0.239 & 144 (41)  & 14 (\textbf{12}) &  \textbf{7} (\textbf{7}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
+& Cloudbreak & 0.0323 & \textbf{0.455} & \textbf{265} (\textbf{104})  & \textbf{49} (\textbf{24}) &  3 (1) & 0 (0)  & 0 (0)  \\ 
+& Breakdancer & 0.0281 & 0.181 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
+&  Pindel & \textbf{0.0387} & 0.239 & 144 (45)  & 14 (7) &  \textbf{7} (\textbf{6}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
 \hline
@@
 Exclusive predictions are in parentheses.
-For deletions, the same cutoffs those which yielded a 10\% FDR on the simulated data, as reported in in Supplementary Table \ref{Schr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.
+For deletions, the same cutoffs were used as for the simulated data as in Supplementary Table \ref{Schr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.
 For insertions, the maximum sensitivity cutoff was used.}
@@
 
-\subsection{Performance in Repetitive Areas}
-
-We analyzed the predictions made by Cloudbreak to determine whether any particular class of deletion event was contributing to its performance.
-We found that Cloudbreak shows particularly strong performance in detecting deletions that overlap RepeatMasker-annotated elements in both the simulated and NA18507 data sets.
-Table \ref{deletionRepmaskpreds} shows the number of deletions that overlap repeats at the 10\% FDR cutoff described above.
-In the NA18507 data set, Cloudbreak identifies 15\% more repeat-overlapping variants than the nearest competitor, Breakdancer.
-Since deletions and other SVs are enriched in repetitive regions of the genome in both of our data sets, this contributes greatly to Cloudbreak's performance, and demonstrates the value in considering as many possible mappings of ambiguously mapping reads as possible.
-Performance on insertions is displayed in Supplementary Table \ref{SinsertionRepmaskpreds}; again, Cloudbreak identifies more events that lie in repetitive regions than the other tools.
-
-\begin{table}[b]
-\begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\hline
- &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 120 & 327 & 562 & 8059 \\ 
-  \hline
-  Cloudbreak  & 25 (2) & \textbf{158} (17) & \textbf{226} (\textbf{49}) & \textbf{1001} (\textbf{137}) \\ 
-  Breakdancer & \textbf{29} (6) & 142 (7) & 192 (19) & 868 (30) \\
-  GASVPro     & 16 (2) & 79 (2) & 79 (9) & 330 (22) \\
-  DELLY-RP       & 21 (2) & 117 (2) & 152 (9) & 712 (6) \\
-  DELLY-SR       & 0 (0) & 42 (0) & 27 (0) & 391 (0) \\
-  Pindel      & 18 (\textbf{9}) & 112 (\textbf{26}) & 109 (4) & 536 (8) \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.}
-\label{deletionRepmaskpreds}
-\end{table}
-
 \subsection{Genotyping Variants}
@@
 For the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
-Cloudbreak was able to achieve 88.1\% and 96.5\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
+Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
 Supplementary Table \ref{SdeletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
 None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that meet our size threshold that also had genotyping information.
-Of the 79 insertions detected by Cloudbreak on the simulated data set, 13 were heterozygous.
-Cloudbreak correctly classified 62 of the 66 homozygous insertions and 10 of the 13 heterozygous insertions discovered, for an overall accuracy rate of 91\%.
+Of the 123 insertions detected by Cloudbreak on the simulated data set, 43 were heterozygous.
+Cloudbreak correctly classified 78 of the 80 homozygous insertions and 31 of the 43 heterozygous insertions discovered, for an overall accuracy rate of 88.6\%.
 
@@
 Simulated reads were aligned to hg18 chromosome 2, and NA18507 reads were aligned to the hg19 assembly.
-For each evaluated SV detection program we attempt to follow the authors' recommended alignment pipeline, except for MoDIL, for which we used reads aligned with BWA (see below).
-Cloudbreak alignments were executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta) in single-ended mode, with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0},  requesting all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read.
-Alignments for the other programs were first aligned using BWA \autocite{Li:2009p836} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
+Alignments for all programs, unless otherwise noted, were aligned using BWA \autocite{Li:2009p836} \texttt{aln} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
 GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \texttt{-a -r -Ex 1100 -t 250}.
@@
 Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \texttt{-c} option.
+We executed MoDIL with default parameters except for a \texttt{MAX\_DEL\_SIZE} of 25000, and processed it in parallel on our cluster with a step size of 121475.
 

### 2013-06-11 manuscript/nm_supp.tex
update to use BWA mappings

@@
 \author[1,2,4]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
-\author[4]{Elizabeth Terhune}
 \author[5]{Alberto L'Abbate}
@@
 \label{insert_size_mixes}
+\end{figure}
+
+\clearpage
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{../figures/CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
+\caption{Cloudbreak performance on the chromsome 2 simulation using different alignment strategies.
+ROC curves show the number of true positives and false positives for each operating point for deletions and insertions.
+The Cloudbreak alignment strategies are: 1) ``Cloudbreak'': Alignments generated with BWA in paired-end mode, reporting the best hit for each pair.
+2) ``Cloudbreak-BWA-PE-MM'': Alignments generated with BWA in paired-end mode, reporting up to 25 additional hits for each mapping in SAM format using the \texttt{-n} and \texttt{-N} parameters for \texttt{bwa sampe} and the script \texttt{xa2multi.pl}.
+3) ``Cloudbreak-GEM-SE-MM'': Alignments generated by running the GEM aligner in single-ended mode, reporting up to 1000 additional hits per alignment.
+GEM was executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta), with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0},  requesting all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read.
+Considering multiple mappings improves Cloudbreak's specificity for insertions but decreases sensitivity to deletions.}
+\label{alignment_comparison}
 \end{figure}
@@
   \hline
-  Cloudbreak  & 47 (7)  & 52 (\textbf{3}) &  \textbf{57} (\textbf{5}) & \textbf{12} (\textbf{4}) & \textbf{15} (\textbf{0}) \\ 
-  Breakdancer & 52 (10)  & 49 (\textbf{3}) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
-  GASVPro     & 35 (4)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
-  DELLY-RP       & 22 (2)  & \textbf{56} (2) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
+  Cloudbreak  & \textbf{68} (17)  & \textbf{67} (\textbf{10}) &  \textbf{56} (\textbf{5}) & \textbf{11} (\textbf{3}) & \textbf{15} (\textbf{0}) \\ 
+  Breakdancer & 52 (8)  & 49 (2) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
+  GASVPro     & 35 (2)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
+  DELLY-RP       & 22 (1)  & 56 (1) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
   DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
-  Pindel      & \textbf{60} (\textbf{34})  & 16 (0) &  41 (1) & 1 (0) & 12 (\textbf{0})\\ 
+  Pindel      & 60 (\textbf{32})  & 16 (0) &  41 (2) & 1 (0) & 12 (\textbf{0})\\ 
    \hline
@@
 \caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.
-Thresholds for this FDR were derived from the ROC curves for each algorithm: representing cutoffs of 2.29 log likelihood ratio for Cloudbreak, 4 supporting read pairs for Breakdancer, 11 supporting read pairs for GASVPro, 7 supporting read pairs for DELLY, and a simple score of 13 for Pindel.}
+The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
 \label{chr2DeletionPredsFDR10}
@@
   \hline
-  Cloudbreak & D,I &   NA    & 308 & 312    & NA         & 2,310 & 636 \\ 
+  Cloudbreak & D,I &   NA    & 290 & 312    & NA         & 824 & 636 \\ 
   Breakdancer & D,I,V,T &  653   & NA       & NA          & 134,170 &  5,586 & 84 \\
@@
  &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
+ Total Number & 120 & 327 & 562 & 8059 \\ 
+  \hline
+  Cloudbreak  & \textbf{37} (\textbf{10}) & \textbf{180} (\textbf{25}) & \textbf{300} (\textbf{85}) & \textbf{1166} (\textbf{270}) \\ 
+  Breakdancer & 29 (7) & 142 (3) & 192 (11) & 868 (21) \\
+  GASVPro     & 16 (1) & 79 (1) & 79 (7) & 330 (16) \\
+  DELLY-RP       & 21 (1) & 117 (1) & 152 (6) & 712 (5) \\
+  DELLY-SR       & 0 (0) & 42 (0) & 27 (0) & 391 (0) \\
+  Pindel      & 18 (9) & 112 (\textbf{25}) & 109 (2) & 536 (7) \\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.
+For both data sets we used the thresholds derived from finding the 10\% FDR level in the simulated data set.
+The proportion of deletions that overlap repetitive elemements discovered by Cloudbreak is similar to that of the other methods.}
+\label{deletionRepmaskpreds}
+\end{table}
+
+\newpage
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{rrr|rr}
+ & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
+\hline
+ &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
  Total Number & 133 & 270 & 341 & 355 \\ 
   \hline
-  Cloudbreak  & 21 (4) & \textbf{58} (24) & \textbf{137} (\textbf{37}) & \textbf{135} (\textbf{55}) \\ 
-  Breakdancer & 17 (7) & 22 (11) & 82 (12) & 44 (9) \\
-  Pindel      & \textbf{25} (\textbf{18}) & 54 (\textbf{33}) & 84 (36) & 82 (25) \\ 
+  Cloudbreak  & \textbf{32} (11) & \textbf{91} (\textbf{47}) & \textbf{169} (\textbf{61}) & \textbf{148} (\textbf{68}) \\ 
+  Breakdancer & 17 (5) & 22 (6) & 82 (7) & 44 (9) \\
+  Pindel      & 25 (\textbf{16}) & 54 (30) & 84 (35) & 82 (24) \\ 
   MoDIL      & 5 (0) & 16 (3) & NA & NA \\ 
@@
 \end{center}
-\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.}
+\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.
+The maximum sensitivity cutoffs were used for both data sets.
+The proportion of insertions in repetitive elemements discovered by Cloudbreak is similar to that of the other methods.}
 \label{insertionRepmaskpreds}
@@
 \cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 100 & 4 &  77 & 12 \\
- & Heterozygous & 7 & 72 &  3 & 338 \\
+\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 35 & 2 &  96 & 21 \\
+ & Heterozygous & 0 & 39 &  2 & 448 \\
 \cline{2-6}
@@
 Reducers then collect the alignments for each paired end, making them available under one key for the next job.
-Our implementation contains wrappers to execute the aligners GEM, Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
+Our implementation contains wrappers to execute the aligners GEM \autocite{MarcoSola:2012hm}, Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
 

### 2013-06-11 src/main/scripts/manuscript.R

@@ -26,17 +26,35 @@ rocLineTypes <- function(numLines) {
   seq(1,numLines)
 }
 
-plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legendLoc="bottomright",legendCex=.9,...) {
+getColLtyMapping <- function(lineNames) {
+  myColors <- rocColors(length(lineNames))
+  myLtys <- rocLineTypes(length(lineNames))
+  colLtyMapping <- list()
+  for (n in 1:length(lineNames)) {
+    colLtyMapping[[lineNames[n]]] <- list(myColors[n], myLtys[n])
+  }
+  return(colLtyMapping)
+}
+
+getCols <- function(keys, colLtyMapping) {
+  unlist(lapply(colLtyMapping[keys], function(x) {x[1]}))
+}
+
+getLtys <- function(keys, colLtyMapping) {
+  unlist(lapply(colLtyMapping[keys], function(x) {x[2]}))
+}
+
+plotROC <- function(perfs, perfNames, totalDels, main, sim=TRUE, maxTP=NA, legendLoc="bottomright",legendCex=.9, 
+                    colLtyMapping=getColLtyMapping(perfNames), ...) {
   maxFP <- totalDels
   if (is.na(maxTP)) {
     maxTP <- totalDels
   }
   plot(0, type="n", ylim=c(0, maxTP), xlim=c(0,maxFP), xlab=ifelse(sim, "False Positives", "Novel Predictions"), ylab="True Positives", main=main, ...)
-  perfCols <- rocColors(length(perfs))
-  lineTypes <- rocLineTypes(length(perfs))
-  mapply(drawPerfLine, perfs, perfCols, lineTypes, MoreArgs=list(maxFP=maxFP))  
+  mapply(drawPerfLine, perfs, getCols(perfNames, colLtyMapping), getLtys(perfNames, colLtyMapping), MoreArgs=list(maxFP=maxFP))  
   if (! is.null(legendLoc)) {
-    legend(legendLoc, legend=perfNames, col=perfCols, lty=lineTypes, lwd=3, cex=legendCex)
+    legend(legendLoc, legend=perfNames, col=getCols(perfNames, colLtyMapping), 
[…]

### 2013-06-28 manuscript/nm_supp.tex
update algo description on multiple mappings

@@
 
-The alignment job uses sensitive mapping tools and settings to discover as many mapping locations for each read pair as possible.
-Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in single ended mode and emits the possible locations under the $rpid$ key.
+The alignment job uses sensitive mapping tools and settings to discover mapping locations for each read pair.
+Aligners can be executed to report multiple possible mapings for each read, or only the best possible mapping.
+Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits the possible locations under the $rpid$ key.
 Reducers then collect the alignments for each paired end, making them available under one key for the next job.
-Our implementation contains wrappers to execute the aligners GEM \autocite{MarcoSola:2012hm}, Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
+Our implementation contains wrappers to execute the aligners BWA, GEM \autocite{MarcoSola:2012hm}, Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
 

### 2013-06-28 manuscript/nm.tex
incorporated some of kemal's abstract; light editing

@@
 
-\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}
+%\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}
+\title{Fast, Accurate, and Scalable Genomic Structural Variation Detection on the Cloud: MapReduce based {\em Cloudbreak}}
 
@@
  \begin{abstract}
-The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data.
-Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods.
-However, the computational requirements for storing and processing data sets with multiple mappings can be formidable.
-Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers.
-MapReduce is becoming a standard framework for distributing processing across such compute clusters.
-We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome.
-In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
-On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
-In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp, and those in repetitive regions of the genome.
+The detection of genomic structural variations (SV) remains one of the the most difficult challenges in analyzing high-throughput sequencing data due to the uncertainty inherent in short-read sequencing data and the  scale of the data sets.
+The growing size and number of sequencing data sets have rendered SV detection, and high throughput sequencing analysis in general, a bona fide big data problem and led to intense interest in distributing computation to cloud or commodity servers.
+Meanwhile, MapReduce is becoming a standard framework for distributing processing across compute clusters and has convincingly demonstrated its effectiveness in offering scalable solutions to big data problems.
+We describe a novel conceptual framework for SV detection algorithms in MapReduce based on computing local features along the genome.
+In this framework, we develop and evaluate a distributed deletion and insertion detection algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
+On simulated and real data sets, our approach achieves performances meeting or exceeding those of a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
+In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp.
 In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples.
-Our implementation and source code are available at \url{http://sonmezsysbio.org/software/cloudbreak}.
+Finally, our software offers dramatically faster runtimes than existing tools, given a sufficiently sized compute cluster, and also includes an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, making cluster computing accessible to users who may not have the resources to maintain a MapReduce capable distributed compute server.
+Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.
 
@@
 More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and thus be able to detect smaller events \autocite{Marschall:2012ek}.
-Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, in part because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.
+Increasing the number of read mappings considered, however, increases the computational burden of SV detection.
+As large-scale sequencing projects and clinical sequencing applications emerge, fast runtimes and efficient storage of data are becoming increasingly important, although they have not been a focus of SV detection algorithm development.
 
@@
 Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
-Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, RNA-seq differential expression analysis \autocite{Langmead:2010p1268}, ChIP-seq peak calling \autocite{Feng:2011p1228}, and computing genome mappability \autocite{Lee:2012bk}.
+Use of the Hadoop framework has been demonstrated for several sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.
 
 Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.
-MapReduce \autocite{Dean:2008p277} divides computation across a cluster into three phases.
+MapReduce divides computation across a cluster into three phases.
 In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
@@
 
-We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
+We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
 DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence.
@@
 We also attempted to evaluate MoDIL on the same data.
-All of these methods detect deletions; in addition Breakdancer, Pindel, and MoDIL are capable of detecting insertions.
+All of these methods detect deletions; in addition BreakDancer, Pindel, and MoDIL are capable of detecting insertions.
 See Methods for details on how reads were aligned and each program was invoked.
@@
 Each point on a curve represents a different threshold on the confidence of predictions made by that tool.
-Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.
+Thresholds vary by: Cloudbreak - likelihood ratio; BreakDancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.
 (b) ROC curves for insertion predictions.
@@
 For insertions, Cloudbreak clearly provides the best combinations of sensitivity and specificity.
-Cloudbreak's runtime is half that of Breakdancer, the next fastest tool, processing the simulated data in under six minutes.
+Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes.
 (Of course, Cloudbreak uses many more CPUs as distributed algorithm.
@@
 \begin{center}
+\resizebox{\textwidth}{!}{
 \begin{tabular}{r|rrr|rrrrr}
@@
 &  Cloudbreak    &  0.638 & \textbf{0.678} & \textbf{153} (9)  & 61 (0) &  62 (0) & 12 (0) & 15 (0) \\ 
-&  Breakdancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
+&  BreakDancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
 &  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
@@
 &  Cloudbreak   &0.451 & \textbf{0.305}  & \textbf{79} (\textbf{32})  & \textbf{32} (\textbf{18}) &  \textbf{11} (8) & 1 (0) & 0 (0) \\ 
-&  Breakdancer & 0.262 & 0.0968  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
+&  BreakDancer & 0.262 & 0.0968  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
 &  Pindel          & \textbf{0.572} & 0.196 & 52 (25)  & 5 (1) &  10 (\textbf{9}) & \textbf{3} (\textbf{2}) & \textbf{9} (\textbf{9})\\ 
@@
 \end{tabular}
+}
 \end{center}
@@
 We merged any overlapping calls of the same type into the region spanned by their unions.
-It should be noted that the 1000 Genomes call set was partially produced using DELLY and Breakdancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor.
+It should be noted that the 1000 Genomes call set was partially produced using DELLY and BreakDancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor.
 We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.
@@
 Cloudbreak has slightly lower accuracy for insertions than other tools, although it can identify the most variants at higher levels of sensitivity.
-Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, Breakdancer, even when Breakdancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).
+Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, BreakDancer, even when BreakDancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).
 
@@
 & Cloudbreak & 0.0943 & \textbf{0.17} & \textbf{573} (\textbf{277})  & \textbf{176} (\textbf{30}) &  \textbf{197} (\textbf{18}) & \textbf{121} (\textbf{6}) & \textbf{399} (\textbf{24}) \\ 
-& Breakdancer & 0.137 & 0.123 & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
+& BreakDancer & 0.137 & 0.123 & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
 &  GASVPro & 0.147 & 0.0474 & 120 (21)  & 40 (2) &  85 (0) & 36 (0) & 128 (0) \\ 
@@
 & Cloudbreak & 0.0323 & \textbf{0.455} & \textbf{265} (\textbf{104})  & \textbf{49} (\textbf{24}) &  3 (1) & 0 (0)  & 0 (0)  \\ 
-& Breakdancer & 0.0281 & 0.181 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
+& BreakDancer & 0.0281 & 0.181 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
 &  Pindel & \textbf{0.0387} & 0.239 & 144 (45)  & 14 (7) &  \textbf{7} (\textbf{6}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
@@
 This data set consisted of 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
-Using a pipeline consisting of Novoalign, Breakdancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions.
+Using a pipeline consisting of Novoalign, BreakDancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions.
 Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even when using lower coverage data sets with a greater variance of insert sizes.
-The variants identified include deletions in the gene CTDSPL/RBPS3, an AML tumor suppressor \autocite{Zheng:2012kk}, and NBEAL1, a gene upregulated in some cancers \autocite{Chen:2004jo}.
+The variants identified include deletions in the gene CTDSPL/RBPS3, an AML tumor suppressor \autocite{Zheng:2012kk}, and NBEAL1, a gene up-regulated in some cancers \autocite{Chen:2004jo}.
 We are currently investigating these deletions to determine their functional impact on this patient.
@@
 \subsection{SV Tool Execution}
-We ran Breakdancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
+We ran BreakDancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
 To run Breakdancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer\_max} for each chromosome using the \texttt{-o} parameter.
@@
 Deletions identified in the AML dataset were validated by PCR.
-The selection of the SVs, previously identified by BreakDancer analysis, to be validated by PCR, was carried out relying on the score and the number of the reads supporting each single event.
+SVs to validate were selected based on calls made by BreakDancer, relying on the score and the number of the reads supporting each single event.
 Appropriate primers were designed with an internet-based interface, Primer3 (\url{http://frodo.wi.mit.edu/}), considering the chromosome localization and orientation of the interval involved in the candidate rearrangement.
@@
 
+\section{Author Contributions}
+
+C.W.W.
+and K.S.
+designed the algorithmic approaches.
+C.W.W.
+implemented the software and conducted the evaluations.
+J.T.
+obtained the AML sample and sorted the cells.
+L.C.
+sequenced the AML sample.
+C.T.S and A.L.A.
+performed PCR validations on the AML sample.
+C.W.W.
+and K.S.
+wrote the manuscript.
+
 \section*{Acknowledgments}
@@
 We would like to thank Izhak Shafran at the Center for Spoken Language Understanding for advice, support, and shared computational resources, and Bob Handsaker for helpful conversations.
+We also thank reviewers for the Third Annual RECOMB Satellite Workshop on Massively Parallel Sequencing (RECOMB-seq) for their comments on a preliminary description of Cloudbreak.
 

### 2013-06-28 manuscript/nm_supp.tex
incorporated some of kemal's abstract; light editing

@@
 
-\title{Supplementary Material for Cloudbreak}
-\author[1,2,4]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
-\author[5]{Alberto L'Abbate}
-\author[5]{Clelia Tiziana Storlazzi}
-\author[4]{Jeffry Tyner}
-\author[4]{Lucia Carbone}
-\author[1,2,3,4]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
-\affil[1]{Institute on Development and Disability}
-\affil[2]{Center for Spoken Language Understanding}
-\affil[3]{Department of Medical Informatics \& Clinical Epidemiology}
+\title{Fast, Accurate, and Scalable Genomic Structural Variation Detection on the Cloud: MapReduce based {\em Cloudbreak} \\
+Supplementary Material}
 
-\affil[4]{Oregon Health \& Science University, Portland, OR, USA}
-\affil[5]{University of Bari, Bari, Italy}
+\author[1,5]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
+\author[3,4]{Jeffrey Tyner}
+\author[6]{Alberto L'Abbate}
+\author[6]{Clelia Tiziana Storlazzi}
+\author[4,5]{Lucia Carbone}
+\author[1,2,5]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
+\affil[1]{Institute on Development and Disability and Center for Spoken Language Understanding}
+\affil[2]{Department of Medical Informatics \& Clinical Epidemiology}
+\affil[3]{Program in Molecular and Cellular Biosciences}
+\affil[4]{Behavioral Neuroscience Department}
+\affil[5]{Oregon Health \& Science University, Portland, OR, USA}
+\affil[6]{Department of Genetics and Microbiology, University of Bari, Bari, Italy}
 
@@
 \caption{An example of the Cloudbreak MapReduce algorithm.
-A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in single-ended mode to generate aligned reads.
+A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in either paired-end or single-ended mode to generate read mappings.
 Reducers gather all alignments for both reads in each pair.
-B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) keys indicating the genomic location spanned by that pair.
+B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) under keys indicating the genomic location spanned by that pair.
 Only one genomic location is diagrammed here for simplicity.
@@
 \caption{Illustration of insert size mixtures at individual genomic locations.
-A) there is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings.
-B) a homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes.
+A) There is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings.
+B) A homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes.
 C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected.
@@
 \includegraphics[width=1\textwidth]{../figures/CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
-\caption{Cloudbreak performance on the chromsome 2 simulation using different alignment strategies.
+\caption{Cloudbreak performance on the chromosome 2 simulation using different alignment strategies.
 ROC curves show the number of true positives and false positives for each operating point for deletions and insertions.
@@
 For both data sets we used the thresholds derived from finding the 10\% FDR level in the simulated data set.
-The proportion of deletions that overlap repetitive elemements discovered by Cloudbreak is similar to that of the other methods.}
+The proportion of deletions that overlap repetitive elements discovered by Cloudbreak is similar to that of the other methods.}
 \label{deletionRepmaskpreds}
@@
 The maximum sensitivity cutoffs were used for both data sets.
-The proportion of insertions in repetitive elemements discovered by Cloudbreak is similar to that of the other methods.}
+The proportion of insertions in repetitive elements discovered by Cloudbreak is similar to that of the other methods.}
 \label{insertionRepmaskpreds}
@@
 Aligners can be executed to report multiple possible mapings for each read, or only the best possible mapping.
-Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits the possible locations under the $rpid$ key.
+Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits possible mapping locations under the $rpid$ key.
 Reducers then collect the alignments for each paired end, making them available under one key for the next job.
@@
 We implemented and executed Cloudbreak on a 56-node Hadoop cluster, with 636 map slots and 477 reduce slots.
-Not including alignment time, we were able to process the Chromosome 2 simulated data in under 6 minutes, and the the NA18507 data set in under 40 minutes.
+Not including alignment time, we were able to process the Chromosome 2 simulated data in under five minutes, and the the NA18507 data set in under 15 minutes.
 For the simulated data set we used 100 reducers for the compute SV features job; for the real data set we used 300.
@@
 
-In table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work.
+In Supplementary Table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work.
 Each tool varies in the amount of parallelization supported.
@@
 Cloudbreak mitigates this problem during the time-consuming feature generation process by using Hadoop partitioners to randomly assign each genomic location to one of the set of reducers, ensuring that the work is evenly distributed across all processes.
-This distribution of processing across the entire cluster also serves to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set in under 50 minutes during a run where one of the compute nodes was rebooted midway through the feature generation job.
+This distribution of processing across the entire cluster also serves to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set during a run where one of the compute nodes was rebooted midway through the feature generation job.
 

### 2013-07-03 manuscript/nm.tex
update abstract, editing

@@
  \begin{abstract}
-The detection of genomic structural variations (SV) remains one of the the most difficult challenges in analyzing high-throughput sequencing data due to the uncertainty inherent in short-read sequencing data and the  scale of the data sets.
-The growing size and number of sequencing data sets have rendered SV detection, and high throughput sequencing analysis in general, a bona fide big data problem and led to intense interest in distributing computation to cloud or commodity servers.
-Meanwhile, MapReduce is becoming a standard framework for distributing processing across compute clusters and has convincingly demonstrated its effectiveness in offering scalable solutions to big data problems.
+
+The detection of genomic structural variations (SV) remains one of the the most difficult challenges in analyzing high-throughput sequencing data due to the uncertainty inherent in mapping short sequencing reads and the scale of the data.
+The growing size and number of sequencing data sets have rendered SV detection, and high throughput sequencing analysis in general, a bona fide big data problem and led to intense interest in distributing computation to clusters of commodity servers or on-demand cloud infrastructure.
+Meanwhile, MapReduce is becoming a standard framework for distributed computing and has convincingly demonstrated its effectiveness in offering scalable solutions to big data problems.
 We describe a novel conceptual framework for SV detection algorithms in MapReduce based on computing local features along the genome.
 In this framework, we develop and evaluate a distributed deletion and insertion detection algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
-On simulated and real data sets, our approach achieves performances meeting or exceeding those of a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
-In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp.
-In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples.
-Finally, our software offers dramatically faster runtimes than existing tools, given a sufficiently sized compute cluster, and also includes an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, making cluster computing accessible to users who may not have the resources to maintain a MapReduce capable distributed compute server.
+On simulated and real data sets, our approach achieves performances meeting or exceeding those of a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes, while providing dramatically shorter runtimes and the ability to readily scale up in data volume on large clusters.
+In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp and can accurately genotype heterozygous and homozygous deletions from diploid samples.
+Finally, our software distribution includes an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, enabling on-demand cluster computing.
 Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.
@@
 This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage.
-
 To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same sample.
@@
 For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
-Cloudbreak has slightly lower accuracy for insertions than other tools, although it can identify the most variants at higher levels of sensitivity.
+Cloudbreak has comparable accuracy for insertions to other tools, and can identify the most variants at higher levels of sensitivity.
 Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, BreakDancer, even when BreakDancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).
 
-Given the high number of false positives produced by all tools at maximum sensitivity indicated by the ROC curves, we decided to characterize the predictions made by each tool at more stringent thresholds.
+Given the high number of false positives produced by all tools at maximum sensitivity, we decided to characterize the predictions made at more stringent thresholds.
 We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
-For insertions, again, we were forced to use the thresholds that gave maximum sensitivity for each tool due to the high observed FDR rates in the simulated data.
-The precision and recall at these thresholds with respect to the gold standard set, as well as the performance of each algorithm at predicting variants of each size class at those thresholds, is shown in Table \ref{NA18507DeletionAndInsertionPreds}.
-For deletions, Cloudbreak has the greatest sensitivity of any tool at these thresholds, identifying the most variants in each size class.
+For insertions, we were again forced to use the maximum sensitivity thresholds for each tool due to the high observed FDRs in the simulated data.
+The precision and recall at these thresholds with respect to the gold standard set, as well as the performance of each algorithm at predicting variants of each size class, is shown in Table \ref{NA18507DeletionAndInsertionPreds}.
+For deletions, Cloudbreak has the greatest sensitivity of any tool, identifying the most variants in each size class.
 Pindel exhibits the highest precision with respect to the gold standard set.
@@
 
-Because Cloudbreak explicitly models zygosity in its feature generation algorithm, it is possible to predict the genotypes of identified variants.
+Because Cloudbreak explicitly models zygosity in its feature generation algorithm, it can predict the genotypes of identified variants.
 We tested this on both the simulated and NA18507 data sets.
-Because we simulated a diploid genome, we had genotypes for the simulated data readily available.
-For the real dataset, we only considered the deletions from the 1000 Genomes Project data set, which had been genotyped as part of the Phase 1 release using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
-Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the true positive deletions we detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
+Because we simulated a diploid genome, genotypes for the simulated data were readily available.
+For the real dataset, we considered the deletions from the 1000 Genomes Project data set, which had been genotyped using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
+Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the deletions it detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
 Supplementary Table \ref{SdeletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
-None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that meet our size threshold that also had genotyping information.
+None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that met our size threshold and also had genotyping information.
 Of the 123 insertions detected by Cloudbreak on the simulated data set, 43 were heterozygous.
-Cloudbreak correctly classified 78 of the 80 homozygous insertions and 31 of the 43 heterozygous insertions discovered, for an overall accuracy rate of 88.6\%.
+Cloudbreak correctly classified 78 of the 80 homozygous insertions and 31 of the 43 heterozygous insertions, for an overall accuracy of 88.6\%.
 

### 2013-07-03 manuscript/nm.tex
kemal instroduction

@@
 However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.
+Part of the problem stems from the fact that SV detection requires high coverage data sets and large computational resources.
+Thus, as the volume of massively parallel sequencing data approaches scales that are associated with big data, SV detection becomes a time consuming component of genomics pipelines on traditional cluster architectures, and presents a significant challenge for genomics groups or clinical operations that may not be able to scale up their computational resources readily in order to stay on top of their workloads.
+Here we present a distributed solution that is scalable and readily available on the cloud
+
+In the big data realm, challenges of this scale have been addressed through distributed processing, i.e.
+taking the computation to the data via many compute nodes working on local chunks of data distributed over the cluster.
+In particular, Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
+Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
+Use of the Hadoop framework has been demonstrated for several sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.
+
+Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.
+MapReduce divides computation across a cluster into three phases.
+In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
+The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
+Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
+MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
 
@@
 
-Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
-Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
-Use of the Hadoop framework has been demonstrated for several sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.
-
-Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.
-MapReduce divides computation across a cluster into three phases.
-In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
-The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
-Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
-MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
-
-The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches.
+Our goal is to leverage the strengths of the MapReduce computational framework in order to provide fast, accurate and readily scalable SV-detection pipelines.
+The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches.
 MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
@@
 We characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.
+FInally, our Cloudbreak distribution provides an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, making distributed Sv-detection accessible to groups that may not have the resources to maintain a MapReduce capable distributed compute server.
 

### 2013-07-03 manuscript/nm.tex
updates to intro

--- 2013-07-03 manuscript/nm.tex
@@
 However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.
+Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets and integrating them all is computationally difficult.
+As the volume of massively parallel sequencing data approaches scales that are associated with ``big data'', SV detection is becoming a time consuming component of genomics pipelines on traditional cluster architectures, and presents a significant challenge for genomics groups or clinical operations that may not be able to scale up their computational resources readily.
+Here we present a distributed software solution that is scalable and readily available on the cloud.
 
-Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets.
-See \textcite{Alkan:2011p547} for a review.
-Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
-Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment.
-These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs.
-Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
-Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
-
-The first RP methods used only unambiguously discordantly mapped read pairs in their analyses.
-However, it has been recently demonstrated that including multiple mappings of discordant read pairs in the analysis can improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
-More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and thus be able to detect smaller events \autocite{Marschall:2012ek}.
-Increasing the number of read mappings considered, however, increases the computational burden of SV detection.
-As large-scale sequencing projects and clinical sequencing applications emerge, fast runtimes and efficient storage of data are becoming increasingly important, although they have not been a focus of SV detection algorithm development.
-
-Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
+In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute computation to many compute nodes, each working on local copies of portions of the data.
+In particular, Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
 Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
+MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
 Use of the Hadoop framework has been demonstrated for several sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.
@@
 Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
-MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
 
-The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches.
+Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets \textcite{Alkan:2011p547}.
+Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
+Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment.
+These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs.
+Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
+Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
+
+Many RP methods use only unambiguously discordantly mapped read pairs in their analyses.
+Some approaches also consider ambiguous mappings of discordant read pairs in an attempt to improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
+More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and thus be able to detect smaller events \autocite{Marschall:2012ek}.
+Increasing the number of read mappings considered, however, increases the computational burden of SV detection.
+As large-scale sequencing projects and clinical sequencing applications emerge, fast runtimes and efficient storage of data are becoming increasingly important, although they have not been a focus of SV detection algorithm development.
+
+Our goal is to leverage the strengths of the MapReduce computational framework in order to provide fast, accurate and readily scalable SV-detection pipelines.
+The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches.
 MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
-In sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome.
+In sequencing applications, for example, MapReduce based SNV-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome.
 SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features along the genome and then producing SV predictions from those features in a post-processing step.
@@
 We characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.
+Cloudbreak also provides an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, making distributed SV detection accessible to those who wish to scale their computational infrastructure as their needs change.
 

### 2013-07-03 manuscript/nm.tex
shrink word count, more edits

@@
 \affil[5]{Oregon Health \& Science University, Portland, OR, USA}
-\affil[6]{Department of Genetics and Microbiology, University of Bari, Bari, Italy}
+\affil[6]{Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy}
 
@@
 
-The detection of genomic structural variations (SV) remains one of the the most difficult challenges in analyzing high-throughput sequencing data due to the uncertainty inherent in mapping short sequencing reads and the scale of the data.
-The growing size and number of sequencing data sets have rendered SV detection, and high throughput sequencing analysis in general, a bona fide big data problem and led to intense interest in distributing computation to clusters of commodity servers or on-demand cloud infrastructure.
-Meanwhile, MapReduce is becoming a standard framework for distributed computing and has convincingly demonstrated its effectiveness in offering scalable solutions to big data problems.
-We describe a novel conceptual framework for SV detection algorithms in MapReduce based on computing local features along the genome.
-In this framework, we develop and evaluate a distributed deletion and insertion detection algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
-On simulated and real data sets, our approach achieves performances meeting or exceeding those of a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes, while providing dramatically shorter runtimes and the ability to readily scale up in data volume on large clusters.
-In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp and can accurately genotype heterozygous and homozygous deletions from diploid samples.
-Finally, our software distribution includes an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, enabling on-demand cluster computing.
+The detection of genomic structural variations (SV) remains a difficult challenge in analyzing high-throughput sequencing data, and the growing size and number of sequencing data sets have rendered SV detection a bona fide big data problem.
+MapReduce is a proven, scalable solution for distributed computing on huge data sets.
+We describe a conceptual framework for SV detection algorithms in MapReduce based on computing local genomic features, and use it to develop a deletion and insertion detection algorithm, Cloudbreak.
+On simulated and real data sets, Cloudbreak achieves accuracy improvements over popular SV detection algorithms across a range of variant sizes, especially those 40bp-100bp.
+Cloudbreak can accurately genotype variants from diploid samples.
+On large clusters, it provides dramatically shorter runtimes and the ability to scale to big data volumes.
+Cloudbreak includes tools to set up and configure MapReduce (Hadoop) clusters on cloud services, enabling on-demand cluster computing.
 Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.
@@
 
-Genomic structural variations such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}.
-However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.
-Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets and integrating them all is computationally difficult.
-As the volume of massively parallel sequencing data approaches scales that are associated with ``big data'', SV detection is becoming a time consuming component of genomics pipelines on traditional cluster architectures, and presents a significant challenge for genomics groups or clinical operations that may not be able to scale up their computational resources readily.
+Genomic structural variations (SVs) such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}.
+However, detection of SVs with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.
+Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, and integrating them all is computationally difficult.
+As the volume of massively parallel sequencing data approaches ``big data'' scales, SV detection is becoming a time consuming component of genomics pipelines, and presents a significant challenge for research groups and clinical operations that may not be able to scale their computational infrastructure.
 Here we present a distributed software solution that is scalable and readily available on the cloud.
 
-In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute computation to many compute nodes, each working on local copies of portions of the data.
-In particular, Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers.
-Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.
-MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.
-Use of the Hadoop framework has been demonstrated for several sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.
+In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data.
+In particular, Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large data sets across clusters of commodity servers.
+Hadoop is an open source project of the Apache Foundation which provides an implementation of MapReduce.
+MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention.
+The Hadoop framework has been shown to be effective in sequencing-related tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.
 
-Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection.
+Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing problems like SV detection.
 MapReduce divides computation across a cluster into three phases.
 In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
-The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
+The framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
 Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
@@
 Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets \textcite{Alkan:2011p547}.
-Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
-Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment.
-These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs.
-Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
-Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
+Read-pair (RP) based methods use the distance between and orientation of the mappings of paired reads to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
+Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation, and then clustering the discordant mappings to find SVs supported by multiple discordantly mapped read pairs.
+Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
+Finally, split-read (SR) methods look for breakpoints by mapping portions of individual reads to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
 
-Many RP methods use only unambiguously discordantly mapped read pairs in their analyses.
-Some approaches also consider ambiguous mappings of discordant read pairs in an attempt to improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
-More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and thus be able to detect smaller events \autocite{Marschall:2012ek}.
+Many RP methods consider only unambiguously discordantly mapped read pairs.
+Some approaches also include ambiguous mappings of discordant read pairs to improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
+Several recent RP approaches have also considered concordant read pairs, either to integrate RD signals for improved accuracy \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds that separate concordant from discordant mappings and thus detect smaller events \autocite{Marschall:2012ek}.
 Increasing the number of read mappings considered, however, increases the computational burden of SV detection.
-As large-scale sequencing projects and clinical sequencing applications emerge, fast runtimes and efficient storage of data are becoming increasingly important, although they have not been a focus of SV detection algorithm development.
 
@@
 MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
-In sequencing applications, for example, MapReduce based SNV-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome.
-SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features along the genome and then producing SV predictions from those features in a post-processing step.
+In sequencing applications, for example, MapReduce based SNV-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations on partitions of the genome.
+SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} compute scores or features along the genome and then produce SV predictions from those features in a post-processing step.
+We will show that this strategy can be translated into the MapReduce architecture.
 
-In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings.
-We then demonstrate the development in this framework of a software package, Cloudbreak, for discovering genomic deletions up to 25,000bp long, and short insertions.
-Cloudbreak computes local features that build on the ideas implemented in MoDIL \autocite{Lee:2009da}, which modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM).
-Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm.
-We characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.
-Cloudbreak also provides an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, making distributed SV detection accessible to those who wish to scale their computational infrastructure as their needs change.
+In this paper, we describe a framework for solving SV detection problems in Hadoop based on the computation of local genomic features from paired end mappings.
+In this framework we have developed a software package, Cloudbreak, that discovers genomic deletions up to 25,000bp long, and short insertions.
+Cloudbreak computes local features based on modeling the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM), an idea first implemented in MoDIL \autocite{Lee:2009da}.
+The use of Hadoop enables scalable implementations of this class of algorithm.
+We characterize our algorithm's performance on simulated and real data sets and compare it to several popular methods.
+Our implementation of Cloudbreak also provides logic to set up and configure Hadoop clusters on cloud service providers, making distributed SV detection accessible to those who wish to dynamically scale their computational infrastructure as their needs change.
 
@@
 
-Our framework for SV detection in MapReduce divides processing into three distinct MapReduce jobs.
-This framework is summarized in Figure \ref{cloudbreak_workflow}.
-Details of the MapReduce SV detection algorithmic framework and how Cloudbreak is implemented within that framework are provided in Supplementary Algorithm \ref{Scb_algo} and the Supplementary Materials; here we proceed with a high-level description of each job.
+Our framework for SV detection in MapReduce divides processing into three distinct MapReduce jobs (Figure \ref{cloudbreak_workflow}): a job that can align reads to the reference using a variety of mapping algorithms; a job that computes a set of features along the genome; and a job which calls structural variations based on those features.
+Our description of a general MapReduce SV detection algorithmic framework and how Cloudbreak is implemented within that framework are provided in Supplementary Algorithm \ref{Scb_algo} and the Supplementary Materials; here we proceed with a high-level description of the Cloudbreak implementation.
 
 Cloudbreak's alignment job can run a variety of alignment tools that report reads in SAM format (Supplementary Materials).
-In the map phase, mappers align reads in either single-end or paired-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair.
+In the map phase, mappers align reads in either single-end or paired-end mode to the reference genome in parallel, outputting mapping locations as values under a key identifying the read pair.
 In the reduce phase, the framework combines the reported mapping locations for the two ends of each read pair.
+This job can also be skipped in favor of importing a pre-existing set of mappings directly into the Hadoop cluster.
 
@@
 For the experiments reported in this paper we use a window size of 25bp.
-Within each window, we examine the distribution of insert sizes of read pairs that span that window, and compute features by fitting a Gaussian Mixture Model to that distribution (Supplementary Materials, Supplementary Figure \ref{Sinsert_size_mixes}).
+Within each window, we examine the distribution of insert sizes of mappings that span that window, and compute features by fitting a GMM to that distribution (Supplementary Materials, Supplementary Figure \ref{Sinsert_size_mixes}).
 To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials); these procedures also allow us to process multiple mappings for each read if they are reported by the aligner.
 
-Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
-In this job, we search for contiguous blocks of genomic locations with similar features and merge them into individual insertion and deletion SV calls, employing noise reduction techniques to increase specificity (Supplementary Materials).
-We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
+Finally, the third MapReduce job is responsible for making SV calls based on these features.
+In this job, we search for contiguous blocks of genomic locations with similar features and merge them into individual insertion and deletion SV calls after applying noise reduction (Supplementary Materials).
+Reducers process each chromosome in parallel after mappers find and organize its features.
 An illustration of the Cloudbreak algorithm working on a simple example is shown in Supplementary Figure \ref{Salgorithm_example}.
@@
 
-We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
-DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence.
-We refer to these sets of calls as DELLY-RP and DELLY-SR in the following results.
+We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
+DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence; we refer to these sets of calls as DELLY-RP and DELLY-SR.
 We also attempted to evaluate MoDIL on the same data.
-All of these methods detect deletions; in addition BreakDancer, Pindel, and MoDIL are capable of detecting insertions.
+All of these methods detect deletions.
+Insertions can be detected by BreakDancer, Pindel, and MoDIL.
 See Methods for details on how reads were aligned and each program was invoked.
@@
 
-As has been observed elsewhere, there is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference.
+There is no available test set of real Illumina sequencing data from a sample that has a complete annotation of SVs.
 Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics.
-On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data.
-To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous insertions and deletions from the genome of J.
-Craig Venter \autocite{Levy:2007fb}.
-Using these variants, we simulated a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants, with 100bp reads and a mean fragment size of 300bp (See Methods).
+On the other hand, any simulated data should contain realistic SVs that follow patterns observed in real data.
+Therefore, we took one of the most complete lists of SVs from an individual, the list of homozygous insertions and deletions from the genome of J.
+Craig Venter \autocite{Levy:2007fb}, and used it to simulate a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants, with 100bp reads and a mean fragment size of 300bp (See Methods).
 
@@
 
-Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristics (ROC) curves of the performance of each algorithm for detecting deletions and insertions on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment.
+Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment.
 See Methods for a description of how we identified correct predictions.
@@
 Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY.
-For insertions, Cloudbreak clearly provides the best combinations of sensitivity and specificity.
+For insertions, Cloudbreak's performance is similar to or slightly better than Pindel.
 Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes.
 (Of course, Cloudbreak uses many more CPUs as distributed algorithm.
-See Supplementary Material and Supplementary Table \ref{Sruntimes} for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster, so results are not included in the runtime figure.
+See Supplementary Material and Supplementary Table \ref{Sruntimes} for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster.
 Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow.
-Therefore, to test scalability we measured the runtime of that job on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Figure \ref{Sscalability}).
+To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Figure \ref{Sscalability}).
 
-Choosing the correct operating point or threshold to set on the output of an SV calling algorithm can be difficult when operating on a new data set.
-The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying levels.
-First, we characterized the predictions made by each algorithm at the operating point which gives them maximum sensitivity.
+Choosing the correct threshold to set on the output of an SV calling algorithm can be difficult.
+The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying thresholds.
+First, we characterized the predictions made by each algorithm at the threshold that gives them maximum sensitivity.
 For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.
-The results for both deletion and insertion predictions are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}.
+The results are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}.
 MoDIL and Cloudbreak exhibited the greatest recall for deletions.
-Cloudbreak has high precision and recall for deletions at this threshold, and discovers many more small deletions.
+Cloudbreak has also has high precision at this threshold, and discovers many small variants.
 For insertions, Cloudbreak has the highest recall, although recall is low for all four approaches.
-Cloudbreak again identifies the most small variants.
-Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping.
-We also used the ROC curves to attempt to characterize the predictions made by each algorithm when a low false discovery rate is required.
-Supplementary Table \ref{Schr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
-At this more stringent threshold, Cloudbreak identifies more deletions in every size category than any other tool.
-Performance on insertions never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.
+Cloudbreak again identifies many small variants.
+Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by RP mapping.
+We also used the ROC curves to characterize algorithm performance when a low false discovery rate is required.
+Supplementary Table \ref{Schr2DeletionPredsFDR10} shows the total number of deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
+At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool.
+Insertions performance never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.
 We also examined Cloudbreak's ability to detect events in repetitive regions of the genome, and found that it was similar to the other methods tested (Supplementary Tables \ref{SdeletionRepmaskpreds} and \ref{SinsertionRepmaskpreds}).
@@
 It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Figure \ref{Sbreakpoint_resolution}): SR methods have higher resolution than RP methods.
-Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we strive however to increase sensitivity and specificity to actual variants in the hopes that such calls could still be useful even if their resolution is limited.
-This seems reasonable to us, especially given the possibility of pipelines in which RP calls are validated \emph{in silico} by local assembly methods or split-read methods.
+Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we believe, however, that increasing sensitivity and specificity is of greatest utility, especially given the emergence of pipelines in which RP calls are validated \emph{in silico} by local assembly.
 
@@
 
-We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive.
-This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage.
-To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same sample.
+We downloaded a set of reads from Yoruban individual NA18507, experiment ERX009609, from the Sequence Read Archive.
+This sample was sequenced by Illumina Inc.
+on the Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage.
+To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same individual.
 These included data from the Human Genome Structural Variation Project reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
 We merged any overlapping calls of the same type into the region spanned by their unions.
-It should be noted that the 1000 Genomes call set was partially produced using DELLY and BreakDancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor.
 We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.
@@
 
-Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard set for both deletions and insertions.
-All algorithms show far less specificity for the gold standard set than they did for the true variants in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample.
+Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm at detecting events larger than 40bp on the NA18507 data set.
+All algorithms show far less specificity for the gold standard set than they did in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample.
 For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
@@
 
-Given the high number of false positives produced by all tools at maximum sensitivity, we decided to characterize the predictions made at more stringent thresholds.
+Given the high number of novel predictions made by all tools at maximum sensitivity, we decided to characterize performance at more stringent thresholds.
 We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
-For insertions, we were again forced to use the maximum sensitivity thresholds for each tool due to the high observed FDRs in the simulated data.
-The precision and recall at these thresholds with respect to the gold standard set, as well as the performance of each algorithm at predicting variants of each size class, is shown in Table \ref{NA18507DeletionAndInsertionPreds}.
+For insertions, we used the maximum sensitivity thresholds for each tool due to the high observed FDRs in the simulated data.
+Precision and recall at these thresholds, as well as the performance of each algorithm at predicting variants of each size class, is shown in Table \ref{NA18507DeletionAndInsertionPreds}.
 For deletions, Cloudbreak has the greatest sensitivity of any tool, identifying the most variants in each size class.
 Pindel exhibits the highest precision with respect to the gold standard set.
-For insertions, Pindel again has the highest precision at maximum sensitivity, although according the ROC curve it is possible to choose a threshold for Cloudbreak with higher precision and recall.
-At maximum sensitivity, Cloudbreak identifies 120 more insertions from the gold standard set than Pindel, giving it by far the highest recall.
+For insertions, Pindel again has the highest precision at maximum sensitivity, while Cloudbreak has by far the highest recall.
 
@@
 We tested this on both the simulated and NA18507 data sets.
-Because we simulated a diploid genome, genotypes for the simulated data were readily available.
-For the real dataset, we considered the deletions from the 1000 Genomes Project data set, which had been genotyped using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
+For the NA18507 data set, we considered the deletions from the 1000 Genomes Project, which had been genotyped using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
 Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the deletions it detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
@@
 
-Over the next few years, due to advances in sequencing technology, genomics data are expected to increase both in volume and in coverage by several orders of magnitude, expanding into the realm referred to as Big Data.
+Over the next few years, due to advances in sequencing technology, genomics data are expected to increase in volume by several orders of magnitude, expanding into the realm referred to as ``big data''.
 In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines.
@@
 
-Among Big Data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
-In this paper, we introduce a novel conceptual framework for structural variation detection algorithms in MapReduce, based on computing local features along the genome.
-This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of a distributed deletion-finding algorithm based on fitting a GMM to the distribution of mapped insert sizes spanning each location in the genome.
+Among big data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
+In this paper, we introduced a novel conceptual framework for SV detection algorithms in MapReduce, based on computing local genomic features.
+This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of an algorithm for detecting insertions and deletions based on fitting a GMM to the distribution of mapped insert sizes spanning each genomic location.
 
-On simulated and real data sets, our approach performs well across a wide range of deletion sizes and achieves performance similar to or better than a variety of state-of-the-art structural variation detection algorithms that run on traditional workload clusters and servers.
-Detection of deletions is an important area of research; \textcite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
-In addition, an increased ability to detect variants in regions of the genome that are repetitive and difficult to align reads to opens up new possibilities in exploring human genetic variation.
+On simulated and real data sets, our approach exhibits high accuracy when compared to popular SV detection algorithms that run on traditional clusters and servers.
+Detection of insertions and deletions is an important area of research; \textcite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
+More SV annotations are also required to assess the impact of SVs on non-coding elements \autocite{Mu:2011br}.
 
-In addition to delivering state-of-the-art performance in a RP-based SV detection problem, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
-Using tools like Apache Whirr (\texttt{http://whirr.apache.org/}, it is becoming trivially easy to instantiate on-demand Hadoop compute clusters with cloud providers such as Amazon's EC2.
-Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who are unable to afford or unwilling to maintain large, specialized servers to analyze their data.
+In addition to delivering state-of-the-art performance in a RP-based SV detection tool, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
+With the advent of cloud service providers such as Amazon EC2, it is becoming easy to instantiate on-demand Hadoop compute clusters.
+Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who need to analyze increasing amounts of sequencing data.
 
@@
 We also thank reviewers for the Third Annual RECOMB Satellite Workshop on Massively Parallel Sequencing (RECOMB-seq) for their comments on a preliminary description of Cloudbreak.
+C.T.S.
+and A.L.A.
+are supported by the Italian Association for Cancer Research (AIRC).
+Sequencing of the AML sample was supported by the Oregon Medical Research Foundation.
+Cancer
 

### 2013-07-08 manuscript/nm.tex
edits to title and abstract; add user manual to supplementary

@@
 %\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}
-\title{Fast, Accurate, and Scalable Genomic Structural Variation Detection on the Cloud: MapReduce based {\em Cloudbreak}}
+\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
 
@@
 
-The detection of genomic structural variations (SV) remains a difficult challenge in analyzing high-throughput sequencing data, and the growing size and number of sequencing data sets have rendered SV detection a bona fide big data problem.
+The detection of genomic structural variations (SV) remains a difficult challenge in analyzing sequencing data, and the growing size and number of sequenced genomes have rendered SV detection a bona fide big data problem.
 MapReduce is a proven, scalable solution for distributed computing on huge data sets.
 We describe a conceptual framework for SV detection algorithms in MapReduce based on computing local genomic features, and use it to develop a deletion and insertion detection algorithm, Cloudbreak.
-On simulated and real data sets, Cloudbreak achieves accuracy improvements over popular SV detection algorithms across a range of variant sizes, especially those 40bp-100bp.
-Cloudbreak can accurately genotype variants from diploid samples.
-On large clusters, it provides dramatically shorter runtimes and the ability to scale to big data volumes.
+On simulated and real data sets, Cloudbreak achieves accuracy improvements over popular SV detection algorithms, and accurately genotypes variants from diploid samples.
+It provides dramatically shorter runtimes and the ability to scale to big data volumes on large compute clusters.
 Cloudbreak includes tools to set up and configure MapReduce (Hadoop) clusters on cloud services, enabling on-demand cluster computing.
@@
 However, detection of SVs with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.
-Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, and integrating them all is computationally difficult.
+Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, and integrating them to form an accurate detection measure is computationally difficult.
 As the volume of massively parallel sequencing data approaches ``big data'' scales, SV detection is becoming a time consuming component of genomics pipelines, and presents a significant challenge for research groups and clinical operations that may not be able to scale their computational infrastructure.
@@
 The use of Hadoop enables scalable implementations of this class of algorithm.
-We characterize our algorithm's performance on simulated and real data sets and compare it to several popular methods.
-Our implementation of Cloudbreak also provides logic to set up and configure Hadoop clusters on cloud service providers, making distributed SV detection accessible to those who wish to dynamically scale their computational infrastructure as their needs change.
+We characterize our algorithm's performance on simulated and real data sets and compare its performance to those of several popular methods.
+
+Finally, and quite importantly from a practical point of view, our implementation of Cloudbreak provides the functionality to easily set up and configure Hadoop clusters on cloud service providers, making dynamically scalable distributed SV detection accessible to all whose computational needs demand it.
 
@@
 This enables on demand provisioning of Hadoop clusters which can then be terminated when processing is complete, eliminating the need to invest in a standing cluster and allowing a model in which users can scale their computational infrastructure as their need for it varies over time.
+Instructions and examples describing how to leverage cloud computing with Cloudbreak are available in the user manual and Supplementary Materials.
 
@@
 Sequencing of the AML sample was supported by the Oregon Medical Research Foundation.
-Cancer
 

### 2013-07-08 manuscript/nm_supp.tex
edits to title and abstract; add user manual to supplementary

@@
 
-\title{Fast, Accurate, and Scalable Genomic Structural Variation Detection on the Cloud: MapReduce based {\em Cloudbreak} \\
+\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\
 Supplementary Material}
@@
 \affil[5]{Oregon Health \& Science University, Portland, OR, USA}
-\affil[6]{Department of Genetics and Microbiology, University of Bari, Bari, Italy}
+\affil[6]{Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy}
 
 \maketitle
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{../figures/insert_size_mixtures.pdf}
+\caption{Illustration of insert size mixtures at individual genomic locations.
+A) There is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings.
+B) A homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes.
+C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected.
+D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
+\label{insert_size_mixes}
+\end{figure}
+
+\clearpage
 
@@
 \label{algorithm_example}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth]{../figures/insert_size_mixtures.pdf}
-\caption{Illustration of insert size mixtures at individual genomic locations.
-A) There is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings.
-B) A homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes.
-C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected.
-D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
-\label{insert_size_mixes}
 \end{figure}
@@
 The alignment job uses sensitive mapping tools and settings to discover mapping locations for each read pair.
-Aligners can be executed to report multiple possible mapings for each read, or only the best possible mapping.
+Aligners can be executed to report multiple possible mappings for each read, or only the best possible mapping.
 Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits possible mapping locations under the $rpid$ key.
@@
 
-Our Cloudbreak software package can be seen as an implementation of the general framework defined above. In particular, Cloudbreak implements the two user-defined function described above as follows:
+Our Cloudbreak software package can be seen as an implementation of the general framework defined above.
+In particular, Cloudbreak implements the three user-defined function described above as follows:
 
@@
 
-\item[\sc{PostProcess}] We convert our features along the genome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
+\item[\sc{PostProcess}] We convert our features along each chromosome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
 To eliminate noise we apply a median filter with window size 5.
@@
 
-\printbibliography
+\printbibliography[title=Supplementary References]
+
+\newpage
+
+\section{Cloudbreak User Manual}
+\label{cloudbreak}
+
+Cloudbreak is a Hadoop-based structural variation (SV) caller for Illumina
+paired-end DNA sequencing data. Currently Cloudbreak calls genomic insertions
+and deletions; we are working on adding support for other types of SVs.
+
+Cloudbreak contains a full pipeline for aligning your data in the form of FASTQ
+files using alignment pipelines that generate many possible mappings for every
+read, in the Hadoop framework. It then contains Hadoop jobs for computing
+genomic features from the alignments, and for calling insertion and deletion
+variants from those features.
+
+You can get Cloudbreak by downloading a pre-packaged release from the ``releases''
+tab in the GitHub repository, or by building from source as described below.
+
+\subsection{Building From Source}
+\label{buildingfromsource}
+
+To build the latest version of Cloudbreak, clone the GitHub repository. You'll
+need to install Maven to build the executables. (http:/\slash maven.apache.org\slash )
+Enter the top level directory of the Cloudbreak repository and type the command:
+
+\texttt{mvn package}
+
+This should compile the code, execute tests, and create a distribution file,
+ \texttt{cloudbreak-\$VERSION-dist.tar.gz} in the \texttt{target\slash } directory. You can then copy
+ that distribution file to somewhere else on your system, unpack it with
+ \texttt{tar -xzvf cloudbreak-\$VERSION-dist.tar.gz} and access the Cloudbreak jar file
+ and related scripts and properties files.
+
+\subsection{Dependencies}
+\label{dependencies}
+
+Cloudbreak requires a cluster Hadoop 0.20.2 or Cloudera CDH3 to run (the older
+mapreduce API). If you don't have a Hadoop cluster, Cloudbreak can also use the
+Apache Whirr API to automatically provision a cluster on the Amazon Elastic
+Compute Cloud (EC2). See the section on using WHIRR below.
+
+If you wish to run alignments using Cloudbreak, you will need one of the following
+supported aligners:
+
+\begin{itemize}
+\item BWA (Recommended): http:/\slash bio-bwa.sourceforge.net\slash
+
+\item GEM: http:/\slash algorithms.cnag.cat\slash wiki\slash The\_GEM\_library
+
+\item RazerS 3: http:/\slash www.seqan.de\slash projects\slash razers\slash
+
+\item Bowtie2: http:/\slash bowtie-bio.sourceforge.net\slash bowtie2\slash index.shtml
+
+\item Novoalign: http:/\slash www.novocraft.com
+
+\end{itemize}
+
+\subsection{User Guide}
+\label{userguide}
+
+You can use Cloudbreak in several different ways, depending on whether you want
+to start with FASTQ files and use Hadoop to help parallelize your alignments, or if you already
+have an aligned BAM file and just want to use Cloudbreak to call variants. In addition,
+the workflow is slightly different depending on whether you want to run on a local
+hadoop cluster or want to run using a cloud provider like Amazon EC2. Later in this
+file, we've listed a set of scenarios to describe options for running the Cloudbreak
+pipeline. Find the scenario that best fits your use case for more details on how to
+run that workflow. For each scenario, we have created a template script that contains
+all of the steps and parameters you need, which you can modify for your particular data set.
+
+\subsection{Running on a cloud provider like Amazon EC2 with Whirr}
+\label{runningonacloudproviderlikeamazonec2withwhirr}
+
+Cloudbreak has support for automatically deploying a Hadoop cluster on
+cloud providers such as Amazon EC2, transferring your data there, running the Cloudbreak algorithm, and
+downloading the results.
+
+Of course, renting compute time on EC2 or other clouds costs money, so please be
+familiar with the appropriate usage and billing policies of your cloud provider
+before attempting this.
+
+WE ARE NOT RESPONSIBLE FOR UNEXPECTED CHARGES THAT YOU MIGHT INCUR ON EC2 OR
+OTHER CLOUD PROVIDERS.
+
+Many properties that affect the cluster created can be set in the file
+\texttt{cloudbreak-whirr.properties} in this distribution. You will need to edit this file
+to set your AWS access key and secret access key (or your credentials for other
+cloud provider services), and to tell it the location of the public and
+private SSH keys to use to access the cluster. You can also control the number
+and type of nodes to include in the cluster. The default settings in the file
+create 15 nodes of type m1.xlarge, which is sufficient to fully process a 30X
+simulation of human chromosome 2, including read alignment and data transfer time,
+in under an hour. We have only tested this capability using EC2; other cloud providers
+may not work as well. You can also direct Whirr to use Amazon EC2's spot instances, which are
+dramatically cheaper than on-demand instances, although they carry the risk of
+being terminated if your price is out-bid. Using recent spot pricing, it cost
+us about \$5 to run the aforementioned chromosome 2 simulation. We recommend
+setting your spot bid price to be the on demand price for the instance type you
+are using to minimize the chance of having your instances terminated.
+
+Please consult Amazon's EC2 documentation and the documentation for Whirr for
+more information on how to configure and deploy clusters in the cloud.
+
+\subsection{Running on a Small Example Data Set}
+\label{runningonasmallexampledataset}
+
+To facilitate testing of Cloudbreak, we have publicly hosted the reads from the simulated
+data example described in the Cloudbreak manuscript on a bucket in Amazon's S3 storage
+ service at s3:/\slash cloudbreak-example\slash . We have also provided an example script that creates
+ a cluster in Amazon EC2, copies the data to the cluster, runs the full Cloudbreak
+ workflow including alignments with BWA, and copies the variant calls back to the
+ local machine before destroying the cluster. The script, called \texttt{Cloudbreak-EC2-whirr-example.sh}
+ is in the scripts directory of the Cloudbreak distribution. Of course, you will still
+ need to edit the \texttt{cloudbreak-whirr.properties} file with your EC2 credentials, and verify
+ that the cluster size, instance types, and spot price are to your liking before
+ executing the example.
+
+\subsubsection{Scenario 1: Compute alignments in Hadoop, using a local Hadoop cluster}
+\label{scenario1:computealignmentsinhadoopusingalocalhadoopcluster}
+
+To install aligner dependencies for use by Cloudbreak, first generate the index
+for the genome reference you would like to run against. Then, copy all of the
+required index files, and the executable files for the aligner into HDFS using
+the \texttt{hadoop dfs -copyFromLocal} command. For BWA you will need all of the index files
+created by running \texttt{bwa index}. You will also need an `fai' file for the reference,
+containing chromosome names and lengths, generated by \texttt{samtools faidx}.
+
+If your reference file is \texttt{reference.fa}, and \texttt{bwa aln} has created the files
+
+\begin{verbatim}
+reference.fa.amb
+reference.fa.ann
+reference.fa.bwt
+reference.fa.pac
+reference.fa.sa
+\end{verbatim}
+
+and \texttt{reference.fa.fai} as described above, issue the following
+commands to load the necessary files into HDFS:
+
+\begin{verbatim}
+hdfs -mkdir indices/
+hdfs -mkdir executables/
+hdfs -copyFromLocal reference.fa.amb indices/
+hdfs -copyFromLocal reference.fa.ann indices/
+hdfs -copyFromLocal reference.fa.bwt indices/
+hdfs -copyFromLocal reference.fa.pac indices/
+hdfs -copyFromLocal reference.fa.sa indices/
+hdfs -copyFromLocal reference.fa.fai indices/
+hdfs -copyFromLocal /path/to/bwa/executables/bwa executables/
+hdfs -copyFromLocal /path/to/bwa/executables/xa2multi.pl executables/
+\end{verbatim}
+
+The basic workflow is:
+
+\begin{enumerate}
+\item Load the FASTQ files into HDFS
+
+\item Run one of the Cloudbreak alignment commands to align your reads
+
+\item Create a readGroup file to describe the location and insert size characteristics of your reads, and copy it into HDFS.
+
+\item Run the GMM fitting feature generation step of the Cloudbreak process.
+
+\item Extract deletion calls from the features created in step 4.
+
+\item Copy the deletion calls from HDFS to a local directory.
+
+\item Extract insertion calls from the features created in step 4.
+
+\item Copy the insertion calls from HDFS to a local directory.
+
+\item Optionally, export the alignments back into a BAM file in your local filesystem.
+
+\end{enumerate}
+
+We have created a script to run through the full process of executing the
+Cloudbreak pipeline from FASTQ files to insertion and deletion calls. The
+script is named \texttt{Cloudbreak-full.sh} and can be found in the scripts directory
+of the Cloudbreak distribution. To customize the script for your needs, copy
+it to a new location and edit the variables in the first three sections:
+``EXPERIMENT DETAILS'', ``LOCAL FILES AND DIRECTORIES'', and
+``HDFS FILES AND DIRECTORIES''.
+
+\subsubsection{Scenario 2: Call variants on existing alignments, using a local Hadoop cluster}
+\label{scenario2:callvariantsonexistingalignmentsusingalocalhadoopcluster}
+
+For this scenario you don't need to worry about having an aligner executable or
+aligner-generated reference in HDFS. You will however, need a chromosome length
+`fai' file, which you can generate by running \texttt{samtools faidx} on your reference
+FASTA files and then copying to HDFS:
+
+\begin{verbatim}
+hdfs -copyFromLocal reference.fa.fai indices/
+\end{verbatim}
+
+After that, the workflow is:
+
+\begin{enumerate}
+\item Load your BAM file into HDFS and prepare it for Cloudbreak
+
+\item Create a readGroup file to describe the location and insert size characteristics of your reads.
+
+\item Run the GMM fitting feature generation step of the Cloudbreak process.
+
+\item Extract deletion calls from the features created in step 3.
+
+\item Copy the deletion calls from HDFS to a local directory.
+
+\item Extract insertion calls from the features created in step 3.
+
+\item Copy the insertion calls from HDFS to a local directory.
+
+\end{enumerate}
+
+To prepare alignments for Cloudbreak, they must be sorted by read name. You can then use the
+ \texttt{readSAMFileIntoHDFS} Cloudbreak command.
+
+A templates for this scenario is available in the script \texttt{Cloudbreak-variants-only.sh}
+located in the scripts directory of the Cloudbreak distribution.
+
+\subsubsection{Scenario 3: Compute alignments in Hadoop, using a cloud provider like EC2}
+\label{scenario3:computealignmentsinhadoopusingacloudproviderlikeec2}
+
+First, see the section ``Running on a Cloud Provider like Amazon EC2 with Whirr'' above, and modify the file
+\texttt{cloudbreak-whirr.properties} to include your access credentials and the appropriate cluster
+specifications. After that, the workflow is similar to the workflow described for scenario \#1
+above, with the additional first steps of copying your reads and dependency files to the cloud and
+creating a cluster before processing begins, and then destroying the cluster after processing has
+completed.
+
+You can see an example workflow involving EC2 by examining the script
+\texttt{Cloudbreak-EC2-whirr.sh}. This begins by transferring your reads to Amazon S3. It then
+uses Apache Whirr to launch an EC2 Hadoop cluster, copies the necessary executable files
+to EC2, and runs the algorithm.
+
+\subsubsection{Scenario 4: Call variants on existing alignments, using a cloud provider like EC2}
+\label{scenario4:callvariantsonexistingalignmentsusingacloudproviderlikeec2}
+
+Again, please read the section ``Running on a Cloud Provider like Amazon EC2 with Whirr'' above to learn how to
+update the \texttt{cloudbreak-whirr.properties} file with your credentials and cluster specifications. After that,
+follow the template in the script \texttt{Cloudbreak-EC2-whirr-variants-only.sh} to create a workflow
+involving calling variants in the cloud.
+
+\subsection{Output Files}
+\label{outputfiles}
+
+The output from running Cloudbreak using one of the scripts above will be found in the files named
+
+\begin{verbatim}
+READ_GROUP_LIBRARY_dels_genotyped.bed
+READ_GROUP_LIBRARY_ins_genotyped.bed
+\end{verbatim}
+
+where READ\_GROUP and LIBRARY are the names of the reads in your experiment. The
+format of the files is tab-delimited with the following columns:
+
+\begin{itemize}
+\item CHROMOSOME: The chromosome of the deletion call
+
+\item START: The start coordinate of the deletion call
+
+\item END: The end coordinate of the deletion call
+
+\item NUMBER: The cloudbreak identifier of the deletion call
+
+\item LR: The likelihood ratio of the deletion (higher indicates a call more likely to be true)
+
+\item TYPE: Either ``INS'' or ``DEL''
+
+\item W: The average weight of the estimated GMM mixing parameter alpha, used in genotyping
+
+\item GENOTYPE: The predicted genotype of the call
+
+\end{itemize}
+
+\subsection{Contact information}
+\label{contactinformation}
+
+Please contact cwhelan at gmail.com with any questions on running cloudbreak.
+
+\subsection{Reference Guide}
+\label{referenceguide}
+
+All of Cloudbreak's functionality is contained in the executable jar file in the
+directory where you unpacked the Cloudbreak distribution. Use the `hadoop'
+command to run the jar file to ensure that the necessary Hadoop dependencies
+are available to Cloudbreak.
+
+To invoke any Cloudbreak command, use a command line in this format:
+
+\texttt{hadoop cloudbreak-\$\{project.version\}.jar [options] [command] [command options]}
+
+Where \texttt{command} is the name of the command, \texttt{command options} are the arguments specific
+to that command, and \texttt{options} are general options, including options for how to run
+Hadoop jobs. For example, if you'd like to specify 50 reduce tasks
+for one of your commands, pass in \texttt{-Dmapred.reduce.tasks=50} as one of the general options.
+
+Each command is detailed below and its options are listed below. You can view this information by typing
+\texttt{hadoop jar cloudbreak-\$\{project.version\}.jar} without any additional parameters.
+
+\begin{verbatim}
+    readPairedEndFilesIntoHDFS      Load paired FASTQ files into HDFS
+      Usage: readPairedEndFilesIntoHDFS [options]
+  Options:
+        *     --HDFSDataDir                  HDFS directory to load reads into
+              --clipReadIdsAtWhitespace      Whether to clip all readnames at
+                                             the first whitespace (prevents trouble
+                                             with some aligners)
+                                             Default: true
+              --compress                     Compression codec to use on the
+                                             reads stored in HDFS
+                                             Default: snappy
+        *     --fastqFile1                   File containing the first read in
+                                             each pair
+        *     --fastqFile2                   File containing the second read in
+                                             each pair
+              --filesInHDFS                  Use this flag if the BAM file has
+                                             already been copied into HDFS
+                                             Default: false
+              --filterBasedOnCasava18Flags   Use the CASAVA 1.8 QC filter to
+                                             filter out read pairs
+                                             Default: false
+              --outFileName                  Filename of the prepped reads in
+                                             HDFS
+                                             Default: reads
+              --trigramEntropyFilter         Filter out read pairs where at
+                                             least one read has a trigram entropy less
+                                             than this value. -1 = no filter
+                                             Default: -1.0
+
+    readSAMFileIntoHDFS      Load a SAM/BAM file into HDFS
+      Usage: readSAMFileIntoHDFS [options]
+  Options:
+        *     --HDFSDataDir   HDFS Directory to hold the alignment data
+              --compress      Compression codec to use for the data
+                              Default: snappy
+              --outFileName   Filename to give the file in HDFS
+                              Default: alignments
+        *     --samFile       Path to the SAM/BAM file on the local filesystem
+
+    bwaPairedEnds      Run a BWA paired-end alignment
+      Usage: bwaPairedEnds [options]
+  Options:
+        *     --HDFSAlignmentsDir    HDFS directory to hold the alignment data
+        *     --HDFSDataDir          HDFS directory that holds the read data
+        *     --HDFSPathToBWA        HDFS path to the bwa executable
+              --HDFSPathToXA2multi   HDFS path to the bwa xa2multi.pl executable
+        *     --maxProcessesOnNode   Ensure that only a max of this many BWA
+                                     processes are running on each node at once.
+                                     Default: 6
+              --numExtraReports      If > 0, set -n and -N params to bwa sampe,
+                                     and use xa2multi.pl to report multiple hits
+                                     Default: 0
+        *     --referenceBasename    HDFS path of the FASTA file from which the
+                                     BWA index files were generated.
+
+    novoalignSingleEnds      Run a Novoalign alignment in single ended mode
+      Usage: novoalignSingleEnds [options]
+  Options:
+        *     --HDFSAlignmentsDir            HDFS directory to hold the
+                                             alignment data
+        *     --HDFSDataDir                  HDFS directory that holds the read
+                                             data
+        *     --HDFSPathToNovoalign          HDFS path to the Novoalign
+                                             executable
+              --HDFSPathToNovoalignLicense   HDFS path to the Novoalign license
+                                             filez
+              --qualityFormat                Quality score format of the FASTQ
+                                             files
+                                             Default: ILMFQ
+        *     --reference                    HDFS path to the Novoalign
+                                             reference index file
+        *     --threshold                    Quality threshold to use for the -t
+                                             parameter
+
+    bowtie2SingleEnds      Run a bowtie2 alignment in single ended mode
+      Usage: bowtie2SingleEnds [options]
+  Options:
+        *     --HDFSAlignmentsDir       HDFS directory to hold the alignment
+                                        data
+        *     --HDFSDataDir             HDFS directory that holds the read data
+        *     --HDFSPathToBowtieAlign   HDFS path to the bowtie2 executable
+        *     --numReports              Max number of alignment hits to report
+                                        with the -k option
+        *     --reference               HDFS path to the bowtie 2 fasta
+                                        reference file
+
+    gemSingleEnds      Run a GEM alignment
+      Usage: gemSingleEnds [options]
+  Options:
+        *     --HDFSAlignmentsDir     HDFS directory to hold the alignment data
+        *     --HDFSDataDir           HDFS directory that holds the read data
+        *     --HDFSPathToGEM2SAM     HDFS path to the gem-2-sam executable
+        *     --HDFSPathToGEMMapper   HDFS path to the gem-mapper executable
+        *     --editDistance          Edit distance parameter (-e) to use in the
+                                      GEM mapping
+                                      Default: 0
+        *     --maxProcessesOnNode    Maximum number of GEM mapping processes to
+                                      run on one node simultaneously
+                                      Default: 6
+        *     --numReports            Max number of hits to report from GEM
+        *     --reference             HDFS path to the GEM reference file
+              --strata                Strata parameter (-s) to use in the GEM
+                                      mapping
+                                      Default: all
+
+    razerS3SingleEnds      Run a razerS3 alignment
+      Usage: razerS3SingleEnds [options]
+  Options:
+        *     --HDFSAlignmentsDir   HDFS directory to hold the alignment data
+        *     --HDFSDataDir         HDFS directory that holds the read data
+        *     --HDFSPathToRazerS3   HDFS path to the razers3 executable file
+        *     --numReports          Max number of alignments to report for each
+                                    read
+        *     --pctIdentity         RazerS 3 percent identity parameter (-i)
+                                    Default: 0
+        *     --reference           HDFS path to the reference (FASTA) file for
+                                    the RazerS 3 mapper
+        *     --sensitivity         RazerS 3 sensitivity parameter (-rr)
+                                    Default: 0
+
+    mrfastSingleEnds      Run a novoalign mate pair alignment
+      Usage: mrfastSingleEnds [options]
+  Options:
+        *     --HDFSAlignmentsDir   HDFS directory to hold the alignment data
+        *     --HDFSDataDir         HDFS directory that holds the read data
+        *     --HDFSPathToMrfast    HDFS path to the mrfast executable file
+        *     --reference           HDFS path to the mrfast reference index file
+              --threshold           MrFAST threshold parameter (-e)
+                                    Default: -1
+
+    exportAlignmentsFromHDFS      Export alignments in SAM format
+      Usage: exportAlignmentsFromHDFS [options]
+  Options:
+              --aligner        Format of the alignment records
+                               (sam|mrfast|novoalign)
+                               Default: sam
+        *     --inputHDFSDir   HDFS path to the directory holding the alignment
+                               reccords
+
+    GMMFitSingleEndInsertSizes      Compute GMM features in each bin across the genome
+      Usage: GMMFitSingleEndInsertSizes [options]
+  Options:
+              --aligner                            Format of the alignment
+                                                   records (sam|mrfast|novoalign)
+                                                   Default: sam
+              --chrFilter                          If filter params are used,
+                                                   only consider alignments in the
+                                                   region
+                                                   chrFilter:startFilter-endFilter
+              --endFilter                          See chrFilter
+              --excludePairsMappingIn              HDFS path to a BED file. Any
+                                                   reads mapped within those intervals
+                                                   will be excluded from the
+                                                   processing
+        *     --faidx                              HDFS path to the chromosome
+                                                   length file for the reference genome
+        *     --inputFileDescriptor                HDFS path to the directory
+                                                   that holds the alignment records
+              --legacyAlignments                   Use data generated with an
+                                                   older version of Cloudbreak
+                                                   Default: false
+              --mapabilityWeighting                HDFS path to a BigWig file
+                                                   containing genome uniqness scores. If
+                                                   specified, Cloudbreak will weight reads
+                                                   by the uniqueness of the regions
+                                                   they mapped to
+              --maxInsertSize                      Maximum insert size to
+                                                   consider (= max size of deletion
+                                                   detectable)
+                                                   Default: 25000
+              --maxLogMapqDiff                     Adaptive quality score cutoff
+                                                   Default: 5.0
+              --maxMismatches                      Max number of mismatches
+                                                   allowed in an alignment; all other
+                                                   will be ignored
+                                                   Default: -1
+              --minCleanCoverage                   Minimum number of spanning
+                                                   read pairs for a bin to run the
+                                                   GMM fitting procedure
+                                                   Default: 3
+              --minScore                           Minimum alignment score (SAM
+                                                   tag AS); all reads with lower AS
+                                                   will be ignored
+                                                   Default: -1
+        *     --outputHDFSDir                      HDFS path to the directory
+                                                   that will hold the output of the
+                                                   GMM procedure
+              --resolution                         Size of the bins to tile the
+                                                   genome with
+                                                   Default: 25
+              --startFilter                        See chrFilter
+              --stripChromosomeNamesAtWhitespace   Clip chromosome names from
+                                                   the reference at the first
+                                                   whitespace so they match with alignment
+                                                   fields
+                                                   Default: false
+
+    extractDeletionCalls      Extract deletion calls into a BED file
+      Usage: extractDeletionCalls [options]
+  Options:
+        *     --faidx                Chromosome length file for the reference
+        *     --inputHDFSDir         HDFS path to the GMM fit feature results
+              --medianFilterWindow   Use a median filter of this size to clean
+                                     up the results
+                                     Default: 5
+        *     --outputHDFSDir        HDFS Directory to store the variant calls
+                                     in
+              --resolution           Size of the bins to tile the genome with
+                                     Default: 25
+        *     --targetIsize          Mean insert size of the library
+                                     Default: 0
+        *     --targetIsizeSD        Standard deviation of the insert size of
+                                     the library
+                                     Default: 0
+              --threshold            Likelihood ratio threshold to call a
+                                     variant
+                                     Default: 1.68
+
+    extractInsertionCalls      Extract insertion calls into a BED file
+      Usage: extractInsertionCalls [options]
+  Options:
+        *     --faidx                Chromosome length file for the reference
+        *     --inputHDFSDir         HDFS path to the GMM fit feature results
+              --medianFilterWindow   Use a median filter of this size to clean
+                                     up the results
+                                     Default: 5
+              --noCovFilter          filter out calls next to a bin with no
+                                     coverage - recommend on for BWA alignments, off for
+                                     other aligners
+                                     Default: true
+        *     --outputHDFSDir        HDFS Directory to store the variant calls
+                                     in
+              --resolution           Size of the bins to tile the genome with
+                                     Default: 25
+        *     --targetIsize          Mean insert size of the library
+                                     Default: 0
+        *     --targetIsizeSD        Standard deviation of the insert size of
+                                     the library
+                                     Default: 0
+              --threshold            Likelihood ratio threshold to call a
+                                     variant
+                                     Default: 1.68
+
+    copyToS3      Upload a file to Amazon S3 using multi-part upload
+      Usage: copyToS3 [options]
+  Options:
+        *     --S3Bucket   S3 Bucket to upload to
+        *     --fileName   Path to the file to be uploaded on the local
+                           filesystem
+
+    launchCluster      Use whirr to create a new cluster in the cloud using 
+                       whirr/cloudbreak-whirr.properties
+      Usage: launchCluster [options]
+
+    runScriptOnCluster      Execute a script on one node of the currently running 
+                            cloud cluster
+      Usage: runScriptOnCluster [options]
+  Options:
+        *     --fileName   Path on the local filesystem of the script to run
+
+    destroyCluster      Destroy the currently running whirr cluster
+      Usage: destroyCluster [options]
+
+    summarizeAlignments      Gather statistics about a set of alignments: number of reads, 
+                             number of mappings, and total number of mismatches
+      Usage: summarizeAlignments [options]
+  Options:
+              --aligner        Format of the alignment records
+                               (sam|mrfast|novoalign)
+                               Default: sam
+        *     --inputHDFSDir   HDFS path of the directory that holds the
+                               alignments
+
+    exportGMMResults      Export wig files that contain the GMM features across 
+                          the entire genome
+      Usage: exportGMMResults [options]
+  Options:
+        *     --faidx          Local path to the chromosome length file
+        *     --inputHDFSDir   HDFS path to the directory holding the GMM
+                               features
+        *     --outputPrefix   Prefix of the names of the files to create
+              --resolution     Bin size that the GMM features were computed for
+                               Default: 25
+
+    dumpReadsWithScores      Dump all read pairs that span the given region with their 
+                             deletion scores to BED format (debugging)
+      Usage: dumpReadsWithScores [options]
+  Options:
+              --aligner                            Format of the alignment
+                                                   records (sam|mrfast|novoalign)
+                                                   Default: sam
+        *     --inputFileDescriptor                HDFS path to the directory
+                                                   that holds the alignment records
+              --maxInsertSize                      Maximum possible insert size
+                                                   to consider
+                                                   Default: 500000
+              --minScore                           Minimum alignment score (SAM
+                                                   tag AS); all reads with lower AS
+                                                   will be ignored
+                                                   Default: -1
+        *     --outputHDFSDir                      HDFS path to the directory
+                                                   that will hold the output
+        *     --region                             region to find read pairs
+                                                   for, in chr:start-end format
+              --stripChromosomeNamesAtWhitespace   Clip chromosome names from
+                                                   the reference at the first
+                                                   whitespace so they match with alignment
+                                                   fields
+                                                   Default: false
+
+    debugReadPairInfo      Compute the raw data that goes into the GMM fit procedure for 
+                           each bin (use with filter to debug a particular locus)
+      Usage: debugReadPairInfo [options]
+  Options:
+              --aligner                 Format of the alignment records
+                                        (sam|mrfast|novoalign)
+                                        Default: sam
+        *     --chrFilter               Print info for alignments in the region
+                                        chrFilter:startFilter-endFilter
+        *     --endFilter               see chrFilter
+              --excludePairsMappingIn   HDFS path to a BED file. Any reads
+                                        mapped within those intervals will be excluded
+                                        from the processing
+        *     --faidx                   HDFS path to the chromosome length file
+                                        for the reference genome
+        *     --inputFileDescriptor     HDFS path to the directory that holds
+                                        the alignment records
+              --mapabilityWeighting     HDFS path to a BigWig file containing
+                                        genome uniqness scores. If specified,
+                                        Cloudbreak will weight reads by the uniqueness of
+                                        the regions they mapped to
+              --maxInsertSize           Maximum insert size to consider (= max
+                                        size of deletion detectable)
+                                        Default: 500000
+              --minScore                Minimum alignment score (SAM tag AS);
+                                        all reads with lower AS will be ignored
+                                        Default: -1
+        *     --outputHDFSDir           HDFS directory to hold the output
+              --resolution              Size of the bins to tile the genome with
+                                        Default: 25
+        *     --startFilter             see chrFilter
+
+    findAlignment      Find an alignment record that matches the input string
+      Usage: findAlignment [options]
+  Options:
+        *     --HDFSAlignmentsDir   HDFS path to the directory that stores the
+                                    alignment data
+        *     --outputHDFSDir       HDFS path to the directory in which to put
+                                    the results
+        *     --read                Read name or portion of the read name to
+                                    search for
+
+    sortGMMResults      Sort and merge GMM Results (use with one reducer to get all GMM feature 
+                        results into a single file
+      Usage: sortGMMResults [options]
+  Options:
+        *     --inputHDFSDir    HDFS path to the directory holding the GMM
+                                features
+        *     --outputHDFSDir   Directory in which to put the results
+\end{verbatim}
 

### 2013-07-08 manuscript/arxiv.tex
version to submit to arxiv

--- 2013-07-08 manuscript/nm_supp.tex
@@
 \documentclass[11pt]{article}
+\pdfoutput=1
 
@@
 
-% \renewcommand{\thetable}{S\arabic{table}}   
-% \renewcommand{\thefigure}{S\arabic{figure}}
-% \renewcommand{\thealgorithm}{\arabic{algorithm}}
-
 \usepackage[style=nature,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
@@
 
-\renewcommand{\figurename}{Supplementary Figure}
-\renewcommand{\tablename}{Supplementary Table}
-
-\newfloat{suppalgorithm}{p}{cap}
-\floatname{suppalgorithm}{Supplementary Algorithm}
+\newcommand{\degree}{\ensuremath{^\circ}}
 
@@
 
-\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\
-Supplementary Material}
+%\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}
+\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
 
@@
 
+ \begin{abstract}
+
+The detection of genomic structural variations (SV) remains a difficult challenge in analyzing sequencing data, and the growing size and number of sequenced genomes have rendered SV detection a bona fide big data problem.
+MapReduce is a proven, scalable solution for distributed computing on huge data sets.
+We describe a conceptual framework for SV detection algorithms in MapReduce based on computing local genomic features, and use it to develop a deletion and insertion detection algorithm, Cloudbreak.
+On simulated and real data sets, Cloudbreak achieves accuracy improvements over popular SV detection algorithms, and accurately genotypes variants from diploid samples.
+It provides dramatically shorter runtimes and the ability to scale to big data volumes on large compute clusters.
+Cloudbreak includes tools to set up and configure MapReduce (Hadoop) clusters on cloud services, enabling on-demand cluster computing.
+Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.
+
+ \medskip
+ \noindent\textbf{Keywords:} genomic structural variation; distributed computing; copy number variation; high-throughput sequencing; genotyping.
+ \end{abstract}
+
+\newpage
+
+\section{Introduction}
+
+Genomic structural variations (SVs) such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}.
+However, detection of SVs with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.
+Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, and integrating them to form an accurate detection measure is computationally difficult.
+As the volume of massively parallel sequencing data approaches ``big data'' scales, SV detection is becoming a time consuming component of genomics pipelines, and presents a significant challenge for research groups and clinical operations that may not be able to scale their computational infrastructure.
+Here we present a distributed software solution that is scalable and readily available on the cloud.
+
+In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data.
+In particular, Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large data sets across clusters of commodity servers.
+Hadoop is an open source project of the Apache Foundation which provides an implementation of MapReduce.
+MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention.
+The Hadoop framework has been shown to be effective in sequencing-related tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.
+
+Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing problems like SV detection.
+MapReduce divides computation across a cluster into three phases.
+In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
+The framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
+Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
+
+Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets \textcite{Alkan:2011p547}.
+Read-pair (RP) based methods use the distance between and orientation of the mappings of paired reads to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
+Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation, and then clustering the discordant mappings to find SVs supported by multiple discordantly mapped read pairs.
+Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
+Finally, split-read (SR) methods look for breakpoints by mapping portions of individual reads to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.
+
+Many RP methods consider only unambiguously discordantly mapped read pairs.
+Some approaches also include ambiguous mappings of discordant read pairs to improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}.
+Several recent RP approaches have also considered concordant read pairs, either to integrate RD signals for improved accuracy \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds that separate concordant from discordant mappings and thus detect smaller events \autocite{Marschall:2012ek}.
+Increasing the number of read mappings considered, however, increases the computational burden of SV detection.
+
+Our goal is to leverage the strengths of the MapReduce computational framework in order to provide fast, accurate and readily scalable SV-detection pipelines.
+The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches.
+MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
+In sequencing applications, for example, MapReduce based SNV-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations on partitions of the genome.
+SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} compute scores or features along the genome and then produce SV predictions from those features in a post-processing step.
+We will show that this strategy can be translated into the MapReduce architecture.
+
+In this paper, we describe a framework for solving SV detection problems in Hadoop based on the computation of local genomic features from paired end mappings.
+In this framework we have developed a software package, Cloudbreak, that discovers genomic deletions up to 25,000bp long, and short insertions.
+Cloudbreak computes local features based on modeling the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM), an idea first implemented in MoDIL \autocite{Lee:2009da}.
+The use of Hadoop enables scalable implementations of this class of algorithm.
+We characterize our algorithm's performance on simulated and real data sets and compare its performance to those of several popular methods.
+
+Finally, and quite importantly from a practical point of view, our implementation of Cloudbreak provides the functionality to easily set up and configure Hadoop clusters on cloud service providers, making dynamically scalable distributed SV detection accessible to all whose computational needs demand it.
+
+\section{Results}
+
+\subsection{Cloudbreak: A Hadoop/MapReduce Software Package for SV Detection}
+
 \begin{figure}
 \centering
-\includegraphics[width=1\textwidth]{../figures/insert_size_mixtures.pdf}
+\includegraphics[width=.8\textwidth]{workflow_with_whirr.png}
+\caption{An overview of the Cloudbreak workflow.
+Reads are first uploaded to a Hadoop cluster from local storage.
+Cloudbreak then executes three MapReduce jobs to process the data: 1) Mapping with sensitive settings.
+2) Computation of features across the genome.
+3) Calling structural variations based on the features computed in the previous step.
+Finally, SV predictions can be downloaded from the Hadoop cluster and examined.
+Cloudbreak can also use the Apache Whirr library to automatically provision Hadoop clusters on and deploy data to cloud providers such as Amazon Elastic Compute Cloud.}
+\label{cloudbreak_workflow}
+\end{figure}
+
+Our framework for SV detection in MapReduce divides processing into three distinct MapReduce jobs (Figure \ref{cloudbreak_workflow}): a job that can align reads to the reference using a variety of mapping algorithms; a job that computes a set of features along the genome; and a job which calls structural variations based on those features.
+Our description of a general MapReduce SV detection algorithmic framework and how Cloudbreak is implemented within that framework are provided in Supplementary Algorithm \ref{cb_algo} and the Supplementary Materials; here we proceed with a high-level description of the Cloudbreak implementation.
+
+Cloudbreak's alignment job can run a variety of alignment tools that report reads in SAM format (Supplementary Materials).
+In the map phase, mappers align reads in either single-end or paired-end mode to the reference genome in parallel, outputting mapping locations as values under a key identifying the read pair.
+In the reduce phase, the framework combines the reported mapping locations for the two ends of each read pair.
+This job can also be skipped in favor of importing a pre-existing set of mappings directly into the Hadoop cluster.
+
+In the next job, Cloudbreak computes a set of features for each location in the genome.
+To begin, we tile the genome with small fixed-width, non-overlapping windows.
+For the experiments reported in this paper we use a window size of 25bp.
+Within each window, we examine the distribution of insert sizes of mappings that span that window, and compute features by fitting a GMM to that distribution (Supplementary Materials, Supplementary Figure \ref{insert_size_mixes}).
+To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials); these procedures also allow us to process multiple mappings for each read if they are reported by the aligner.
+
+Finally, the third MapReduce job is responsible for making SV calls based on these features.
+In this job, we search for contiguous blocks of genomic locations with similar features and merge them into individual insertion and deletion SV calls after applying noise reduction (Supplementary Materials).
+Reducers process each chromosome in parallel after mappers find and organize its features.
+An illustration of the Cloudbreak algorithm working on a simple example is shown in Supplementary Figure \ref{algorithm_example}.
+
+Cloudbreak can be executed on any Hadoop cluster; Hadoop abstracts away the details of cluster configuration, making distributed applications portable.
+In addition, our Cloudbreak implementation can leverage the Apache Whirr library to automatically create clusters with cloud service providers such as the Amazon Elastic Compute Cloud (EC2).
+This enables on demand provisioning of Hadoop clusters which can then be terminated when processing is complete, eliminating the need to invest in a standing cluster and allowing a model in which users can scale their computational infrastructure as their need for it varies over time.
+
+\subsection{Tests with Simulated Data}
+
+We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}.
+DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence; we refer to these sets of calls as DELLY-RP and DELLY-SR.
+We also attempted to evaluate MoDIL on the same data.
+All of these methods detect deletions.
+Insertions can be detected by BreakDancer, Pindel, and MoDIL.
+See Methods for details on how reads were aligned and each program was invoked.
+For all alignments we used BWA \autocite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Figure \ref{alignment_comparison}).
+
+There is no available test set of real Illumina sequencing data from a sample that has a complete annotation of SVs.
+Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics.
+On the other hand, any simulated data should contain realistic SVs that follow patterns observed in real data.
+Therefore, we took one of the most complete lists of SVs from an individual, the list of homozygous insertions and deletions from the genome of J.
+Craig Venter \autocite{Levy:2007fb}, and used it to simulate a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants, with 100bp reads and a mean fragment size of 300bp (See Methods).
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{chr2_sim_rocs_runtime.pdf}
+\caption{Accuracy and runtime performance on a simulated data set.
+(a) Receiver Operating Characteristic (ROC) curves showing the specificity and sensitivity of each tool to deletions larger than 40bp on a simulated set of reads giving diploid coverage of 30X on human chromosome 2.
+Deletions from the Venter genome were randomly added to one or both haplotypes.
+Each point on a curve represents a different threshold on the confidence of predictions made by that tool.
+Thresholds vary by: Cloudbreak - likelihood ratio; BreakDancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.
+(b) ROC curves for insertion predictions.
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).}
+\label{chr2CombinedRoc}
+\end{figure}
+
+Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment.
+See Methods for a description of how we identified correct predictions.
+All approaches show excellent specificity at high thresholds in this simulation.
+Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY.
+For insertions, Cloudbreak's performance is similar to or slightly better than Pindel.
+Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes.
+(Of course, Cloudbreak uses many more CPUs as distributed algorithm.
+See Supplementary Material and Supplementary Table \ref{runtimes} for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster.
+Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow.
+To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Figure \ref{scalability}).
+
+Choosing the correct threshold to set on the output of an SV calling algorithm can be difficult.
+The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying thresholds.
+First, we characterized the predictions made by each algorithm at the threshold that gives them maximum sensitivity.
+For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.
+The results are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}.
+MoDIL and Cloudbreak exhibited the greatest recall for deletions.
+Cloudbreak has also has high precision at this threshold, and discovers many small variants.
+For insertions, Cloudbreak has the highest recall, although recall is low for all four approaches.
+Cloudbreak again identifies many small variants.
+Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by RP mapping.
+We also used the ROC curves to characterize algorithm performance when a low false discovery rate is required.
+Supplementary Table \ref{chr2DeletionPredsFDR10} shows the total number of deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
+At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool.
+Insertions performance never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.
+We also examined Cloudbreak's ability to detect events in repetitive regions of the genome, and found that it was similar to the other methods tested (Supplementary Tables \ref{deletionRepmaskpreds} and \ref{insertionRepmaskpreds}).
+
+It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Figure \ref{breakpoint_resolution}): SR methods have higher resolution than RP methods.
+Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we believe, however, that increasing sensitivity and specificity is of greatest utility, especially given the emergence of pipelines in which RP calls are validated \emph{in silico} by local assembly.
+
+\begin{table}[t]
+\begin{center}
+\resizebox{\textwidth}{!}{
+\begin{tabular}{r|rrr|rrrrr}
+  \cline{2-9}
+   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\hline
+\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
+  \hline
+\cline{2-9}
+&  Cloudbreak    &  0.638 & \textbf{0.678} & \textbf{153} (9)  & 61 (0) &  62 (0) & 12 (0) & 15 (0) \\ 
+&  BreakDancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
+&  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
+&  DELLY-RP           & 0.457 & 0.613 & 114 (3)  & \textbf{68} (0) &  \textbf{66} (0) & 9 (1) & 17 (0) \\ 
+&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
+&  Pindel           & 0.462 & 0.421 & 96 (\textbf{11})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
+&  MoDIL           & 0.132  & 0.66 & 123 (6)  & 66 (\textbf{3}) &  \textbf{66} (\textbf{11}) & \textbf{17} (\textbf{7}) & \textbf{23} (\textbf{8})\\ 
+   \hline
+\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
+\cline{2-9}
+&  Cloudbreak   &0.451 & \textbf{0.305}  & \textbf{79} (\textbf{32})  & \textbf{32} (\textbf{18}) &  \textbf{11} (8) & 1 (0) & 0 (0) \\ 
+&  BreakDancer & 0.262 & 0.0968  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
+&  Pindel          & \textbf{0.572} & 0.196 & 52 (25)  & 5 (1) &  10 (\textbf{9}) & \textbf{3} (\textbf{2}) & \textbf{9} (\textbf{9})\\ 
+&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (0) &  1 (0) & 2 (\textbf{2}) & 0 (0)\\ 
+\hline
+\end{tabular}
+}
+\end{center}
+\caption{The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses).
+The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section.}
+\label{chr2DeletionAndInsertionPredsMaxSensitivity}
+\end{table}
+
+\subsection{Tests with Biological Data}
+
+We downloaded a set of reads from Yoruban individual NA18507, experiment ERX009609, from the Sequence Read Archive.
+This sample was sequenced by Illumina Inc.
+on the Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage.
+To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same individual.
+These included data from the Human Genome Structural Variation Project reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
+We merged any overlapping calls of the same type into the region spanned by their unions.
+We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{NA18507_rocs_runtime.pdf}
+\caption{Accuracy and performance on the 37X NA18507 sample.
+(a) ROC curve for deletion prediction performance, tested against the combined gold standard sets of deletions taken from \textcite{Kidd:2008p926}, \textcite{Mills:2011fi}, and \textcite{GenomesProjectConsortium:2012co}.
+(b) ROC curve for insertion prediction performance.
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).
+}
+\label{NA18507CombinedRoc}
+\end{figure}
+
+Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm at detecting events larger than 40bp on the NA18507 data set.
+All algorithms show far less specificity for the gold standard set than they did in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample.
+For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
+Cloudbreak has comparable accuracy for insertions to other tools, and can identify the most variants at higher levels of sensitivity.
+Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, BreakDancer, even when BreakDancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).
+
+Given the high number of novel predictions made by all tools at maximum sensitivity, we decided to characterize performance at more stringent thresholds.
+We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
+For insertions, we used the maximum sensitivity thresholds for each tool due to the high observed FDRs in the simulated data.
+Precision and recall at these thresholds, as well as the performance of each algorithm at predicting variants of each size class, is shown in Table \ref{NA18507DeletionAndInsertionPreds}.
+For deletions, Cloudbreak has the greatest sensitivity of any tool, identifying the most variants in each size class.
+Pindel exhibits the highest precision with respect to the gold standard set.
+For insertions, Pindel again has the highest precision at maximum sensitivity, while Cloudbreak has by far the highest recall.
+
+\begin{table}
+\begin{center}
+\resizebox{\textwidth}{!}{
+\begin{tabular}{r|rrr|rrrrr}
+  \cline{2-9}
+&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\hline
+\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,462 & 240 & 232 & 147 & 540 \\
+  \hline
+\cline{2-9}
+& Cloudbreak & 0.0943 & \textbf{0.17} & \textbf{573} (\textbf{277})  & \textbf{176} (\textbf{30}) &  \textbf{197} (\textbf{18}) & \textbf{121} (\textbf{6}) & \textbf{399} (\textbf{24}) \\ 
+& BreakDancer & 0.137 & 0.123 & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
+&  GASVPro & 0.147 & 0.0474 & 120 (21)  & 40 (2) &  85 (0) & 36 (0) & 128 (0) \\ 
+&  DELLY-RP & 0.0931 & 0.1 & 143 (6)  & 128 (3) &  167 (1) & 103 (0) & 323 (1) \\ 
+&  DELLY-SR & 0.153 & 0.0485 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
+&  Pindel & \textbf{0.179} & 0.0748 & 149 (8)  & 61 (0) &  149 (0) & 69 (1) & 217 (0) \\ 
+\hline
+\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 & 0 \\
+\cline{2-9}
+& Cloudbreak & 0.0323 & \textbf{0.455} & \textbf{265} (\textbf{104})  & \textbf{49} (\textbf{24}) &  3 (1) & 0 (0)  & 0 (0)  \\ 
+& BreakDancer & 0.0281 & 0.181 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
+&  Pindel & \textbf{0.0387} & 0.239 & 144 (45)  & 14 (7) &  \textbf{7} (\textbf{6}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
+\hline
+\end{tabular}
+}
+\end{center}
+\caption{The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class found.
+Exclusive predictions are in parentheses.
+For deletions, the same cutoffs were used as for the simulated data as in Supplementary Table \ref{chr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.
+For insertions, the maximum sensitivity cutoff was used.}
+\label{NA18507DeletionAndInsertionPreds}
+\end{table}
+
+\subsection{Performance on a Low-Coverage Cancer Data Set}
+
+We also tested Cloudbreak on a sequencing data set obtained from a patient with acute myeloid leukemia (AML).
+This data set consisted of 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
+Using a pipeline consisting of Novoalign, BreakDancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions.
+Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even when using lower coverage data sets with a greater variance of insert sizes.
+The variants identified include deletions in the gene CTDSPL/RBPS3, an AML tumor suppressor \autocite{Zheng:2012kk}, and NBEAL1, a gene up-regulated in some cancers \autocite{Chen:2004jo}.
+We are currently investigating these deletions to determine their functional impact on this patient.
+
+\subsection{Genotyping Variants}
+
+Because Cloudbreak explicitly models zygosity in its feature generation algorithm, it can predict the genotypes of identified variants.
+We tested this on both the simulated and NA18507 data sets.
+For the NA18507 data set, we considered the deletions from the 1000 Genomes Project, which had been genotyped using the population-scale SV detection algorithm Genome STRiP \autocite{Handsaker:2011ki}.
+Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the deletions it detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
+Supplementary Table \ref{deletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
+None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that met our size threshold and also had genotyping information.
+Of the 123 insertions detected by Cloudbreak on the simulated data set, 43 were heterozygous.
+Cloudbreak correctly classified 78 of the 80 homozygous insertions and 31 of the 43 heterozygous insertions, for an overall accuracy of 88.6\%.
+
+\section{Discussion}\label{Discussion}
+
+Over the next few years, due to advances in sequencing technology, genomics data are expected to increase in volume by several orders of magnitude, expanding into the realm referred to as ``big data''.
+In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines.
+In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.
+
+Among big data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
+In this paper, we introduced a novel conceptual framework for SV detection algorithms in MapReduce, based on computing local genomic features.
+This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of an algorithm for detecting insertions and deletions based on fitting a GMM to the distribution of mapped insert sizes spanning each genomic location.
+
+On simulated and real data sets, our approach exhibits high accuracy when compared to popular SV detection algorithms that run on traditional clusters and servers.
+Detection of insertions and deletions is an important area of research; \textcite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
+More SV annotations are also required to assess the impact of SVs on non-coding elements \autocite{Mu:2011br}.
+
+In addition to delivering state-of-the-art performance in a RP-based SV detection tool, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
+With the advent of cloud service providers such as Amazon EC2, it is becoming easy to instantiate on-demand Hadoop compute clusters.
+Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who need to analyze increasing amounts of sequencing data.
+
+\newpage
+
+\section{Methods}\label{discussion}
+
+\subsection{Cloudbreak Implementation}
+Cloudbreak is a native Java Hadoop application.
+We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4.
+We use snappy compression for MapReduce data.
+Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
+To execute the other tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine (\url{http://research.cs.wisc.edu/htcondor/}) with directed acyclic graphs to describe dependencies between jobs.
+
+\subsection{Cloudbreak Genotyping}
+We used the parameters of the fit GMM to infer the genotype of each predicted variant.
+Assuming that our pipeline is capturing all relevant read mappings near the locus of the variant, the genotype should be indicated by the estimated parameter $\alpha$, the mixing parameter that controls the weight of the two components in the GMM.
+We setting a simple cutoff of .35 on the average value of $\alpha$ for each prediction to call the predicted variant homozygous or heterozygous.
+We used the same cutoff for deletion and insertion predictions.
+
+\subsection{Read Simulation}
+Since there are relatively few heterozygous insertions and deletions annotated in the Venter genome, we used the set of homozygous indels contained in the HuRef data (\texttt{HuRef.homozygous\_indels.061109.gff}) and randomly assigned each variant to be either homozygous or heterozygous.
+Based on this genotype, we applied each variant to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
+We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package (\url{http://sourceforge.net/apps/mediawiki/dnaa/}).
+We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence.
+Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous insertions and deletions.
+
+\subsection{Read Alignments}
+Simulated reads were aligned to hg18 chromosome 2, and NA18507 reads were aligned to the hg19 assembly.
+Alignments for all programs, unless otherwise noted, were aligned using BWA \autocite{Li:2009p836} \texttt{aln} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
+GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \texttt{-a -r -Ex 1100 -t 250}.
+
+\subsection{SV Tool Execution}
+We ran BreakDancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
+To run Breakdancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer\_max} for each chromosome using the \texttt{-o} parameter.
+We ran DELLY version 0.0.9 with the \texttt{-p} parameter and default values for other parameters.
+For the parallel run of DELLY we first split the original BAM file with BamTools \autocite{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
+We ran GASVPro version 1.2 using the \texttt{GASVPro.sh} script and default parameters.
+Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \texttt{-c} option.
+We executed MoDIL with default parameters except for a \texttt{MAX\_DEL\_SIZE} of 25000, and processed it in parallel on our cluster with a step size of 121475.
+
+\subsection{SV Prediction Evaluation}
+We use the following criteria to define a true prediction given a gold standard set of deletion and insertion variants to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method.
+For evaluating insertions, each algorithm produces insertion predictions that define an interval in which the insertion is predicted to have occurred with start and end coordinates $s$ and $e$ as well as the predicted length of the insertion, $l$.
+The true insertions are defined in terms of their actual insertion coordinate $i$ and their actual length $l_a$.
+Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$.
+In this study we are interested in detecting events larger than 40bp, because with longer reads, smaller events can be more easily discovered by examining gaps in individual reads.
+Both Pindel and MoDIL make many calls with a predicted event size of under 40bp, so we remove those calls from the output sets of those programs.
+Finally, we exclude from consideration calls from all approaches that match a true deletion of less than 40bp where the predicted variant length is less than or equal to 75bp in length.
+
+\subsection{Preparation of AML Sample}
+
+Peripheral blood was collected from a patient with acute myelomonocytic leukemia (previously designated acute myeloid leukemia FAB M4) under a written and oral informed consent process reviewed and approved by the Institutional Review Board of Oregon Health \& Science University.
+Known cytogenetic abnormalities associated with this specimen included trisomy 8 and internal tandem duplications within the FLT3 gene.
+Mononuclear cells were separated on a Ficoll gradient, followed by red cell lysis.
+Mononuclear cells were immunostained using antibodies specific for CD3, CD14, CD34, and CD117 (all from BD Biosciences) and cell fractions were sorted using a BD FACSAria flow cytometer.
+Cell fractions isolated included T-cells (CD3-positive), malignant monocytes (CD14-positive), and malignant blasts (CD34, CD117-positive).
+We sequenced CD14+ cells on an Illumina Genome Analyzer II, producing 128,819,200 paired-end reads.
+
+\subsection{Validation of AML Deletions by PCR}
+
+Deletions identified in the AML dataset were validated by PCR.
+SVs to validate were selected based on calls made by BreakDancer, relying on the score and the number of the reads supporting each single event.
+Appropriate primers were designed with an internet-based interface, Primer3 (\url{http://frodo.wi.mit.edu/}), considering the chromosome localization and orientation of the interval involved in the candidate rearrangement.
+The primers were checked for specificity using the BLAT tool of the UCSC Human Genome Browser (\url{http://genome.ucsc.edu/cgi-bin/hgBlat}).
+All the primer pairs were preliminarily tested on the patient genomic DNA and a normal genomic DNA as control.
+The PCR conditions were as follows: 2 min at 95\degree C followed by 35 cycles of 30 sec at 95\degree C, 20 sec at 60\degree C, and 2 min at 72\degree C.
+All the obtained PCR products were sequenced and analyzed by BLAT for sequence specificity.
+
+\newpage
+
+\section{Author Contributions}
+
+C.W.W.
+and K.S.
+designed the algorithmic approaches.
+C.W.W.
+implemented the software and conducted the evaluations.
+J.T.
+obtained the AML sample and sorted the cells.
+L.C.
+sequenced the AML sample.
+C.T.S and A.L.A.
+performed PCR validations on the AML sample.
+C.W.W.
+and K.S.
+wrote the manuscript.
+
+\section*{Acknowledgments}
+
+We would like to thank Izhak Shafran at the Center for Spoken Language Understanding for advice, support, and shared computational resources, and Bob Handsaker for helpful conversations.
+We also thank reviewers for the Third Annual RECOMB Satellite Workshop on Massively Parallel Sequencing (RECOMB-seq) for their comments on a preliminary description of Cloudbreak.
+C.T.S.
+and A.L.A.
+are supported by the Italian Association for Cancer Research (AIRC).
+Sequencing of the AML sample was supported by the Oregon Medical Research Foundation.
+
+\printbibliography
+
+\setcounter{table}{1}
+\setcounter{figure}{0}
+\renewcommand{\figurename}{Supplementary Figure}
+\renewcommand{\tablename}{Supplementary Table}
+
+\newfloat{suppalgorithm}{p}{cap}
+\floatname{suppalgorithm}{Supplementary Algorithm}
+
+\newrefsection
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{insert_size_mixtures.pdf}
 \caption{Illustration of insert size mixtures at individual genomic locations.
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/cloudbreak_mapred_diagram.pdf}
+\includegraphics[width=1\textwidth]{cloudbreak_mapred_diagram.pdf}
 \caption{An example of the Cloudbreak MapReduce algorithm.
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
+\includegraphics[width=1\textwidth]{CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
 \caption{Cloudbreak performance on the chromosome 2 simulation using different alignment strategies.
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/runtimeByNodes.pdf}
+\includegraphics[width=1\textwidth]{runtimeByNodes.pdf}
 \caption{Scalability of the Cloudbreak algorithm.
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/breakpoint_resolution.pdf}
+\includegraphics[width=1\textwidth]{breakpoint_resolution.pdf}
 \caption{Breakpoint resolution for each tool for deletions on the Chromosome 2 simulated data.
@@
 
-\item[\sc{PostProcess}] We convert our features along each chromosome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
+\item[\sc{PostProcess}] We convert our features along the genome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
 To eliminate noise we apply a median filter with window size 5.
@@
 
-\printbibliography[title=Supplementary References]
+\printbibliography[title={Supplementary References}]
 
@@
 the workflow is slightly different depending on whether you want to run on a local
-hadoop cluster or want to run using a cloud provider like Amazon EC2. Later in this
+Hadoop cluster or want to run using a cloud provider like Amazon EC2. Later in this
 file, we've listed a set of scenarios to describe options for running the Cloudbreak

### 2013-07-10 manuscript/arxiv.tex
final arxiv version

@@
 We describe a conceptual framework for SV detection algorithms in MapReduce based on computing local genomic features, and use it to develop a deletion and insertion detection algorithm, Cloudbreak.
-On simulated and real data sets, Cloudbreak achieves accuracy improvements over popular SV detection algorithms, and accurately genotypes variants from diploid samples.
+On simulated and real data sets, Cloudbreak achieves accuracy improvements over popular SV detection algorithms, and genotypes variants from diploid samples.
 It provides dramatically shorter runtimes and the ability to scale to big data volumes on large compute clusters.
@@
 
-\setcounter{table}{1}
+\setcounter{table}{0}
 \setcounter{figure}{0}
@@
 \begin{center}
-\begin{tabular}{rrrrrr}
-  \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 224 &  84 & 82 &  31 & 26\\ 
-  \hline
-  Cloudbreak  & \textbf{68} (17)  & \textbf{67} (\textbf{10}) &  \textbf{56} (\textbf{5}) & \textbf{11} (\textbf{3}) & \textbf{15} (\textbf{0}) \\ 
-  Breakdancer & 52 (8)  & 49 (2) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
-  GASVPro     & 35 (2)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
-  DELLY-RP       & 22 (1)  & 56 (1) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
-  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
-  Pindel      & 60 (\textbf{32})  & 16 (0) &  41 (2) & 1 (0) & 12 (\textbf{0})\\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPredsFDR10}
-\end{table}
-
-\clearpage
-
-\begin{table}[h]
-\begin{center}
 \begin{tabular}{r|r|rrr|rrr}
@@
 \label{runtimes}
+\end{table}
+
+\clearpage
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{rrrrrr}
+  \hline
+ & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number & 224 &  84 & 82 &  31 & 26\\ 
+  \hline
+  Cloudbreak  & \textbf{68} (17)  & \textbf{67} (\textbf{10}) &  \textbf{56} (\textbf{5}) & \textbf{11} (\textbf{3}) & \textbf{15} (\textbf{0}) \\ 
+  Breakdancer & 52 (8)  & 49 (2) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
+  GASVPro     & 35 (2)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
+  DELLY-RP       & 22 (1)  & 56 (1) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
+  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
+  Pindel      & 60 (\textbf{32})  & 16 (0) &  41 (2) & 1 (0) & 12 (\textbf{0})\\ 
+   \hline
+\end{tabular}
+\end{center}
+\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
+The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
+\label{chr2DeletionPredsFDR10}
 \end{table}
@@
 
-This should compile the code, execute tests, and create a distribution file,
- \texttt{cloudbreak-\$VERSION-dist.tar.gz} in the \texttt{target\slash } directory. You can then copy
- that distribution file to somewhere else on your system, unpack it with
- \texttt{tar -xzvf cloudbreak-\$VERSION-dist.tar.gz} and access the Cloudbreak jar file
- and related scripts and properties files.
+This should compile the code, execute tests, and create a distribution file, \linebreak
+ \texttt{cloudbreak-\$VERSION-dist.tar.gz}, in the \texttt{target\slash } directory. You can then copy
+ that distribution file to somewhere else on your system, unpack it with:
+
+ \texttt{tar -xzvf cloudbreak-\$VERSION-dist.tar.gz}
+
+and access the Cloudbreak jar file and related scripts and properties files.
 
@@
 
-    sortGMMResults      Sort and merge GMM Results (use with one reducer to get all GMM feature 
-                        results into a single file
+    sortGMMResults      Sort and merge GMM Results (use with one reducer to get all 
+                        GMM feature results into a single file
       Usage: sortGMMResults [options]

### 2013-07-10 src/main/scripts/manuscript.R
messing around with figure formats and coloring

@@ -152,6 +152,9 @@ dellyChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_
 dellyBRChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_br.perf.txt'
 pindelChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/human_b36_male_chr2_venterindels_c15_i100_s30_rl100_sort_pindel_D_gte40.perf.txt'
 
+# bwa mem meappings
+breakdancerBWAMEMChr2DeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/bwa_mem_human_b36_male_chr2_venterindels_c15_i100_s30_rl100_breakdancer.dels.perf.txt'
+
 cloudbreakGEMChr2DeletionPredictionsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_del_calls_gt2p29.bed'
 cloudbreakBWAChr2DeletionPredictionsFileFDR10 <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/bwa_del_1p98.bed'
 
@@ -202,14 +205,16 @@ pindelChr2LAInsertionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_c
 
 cloudbreakGEMChr2InsertionPredictionsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_ins_calls_gt0p26.bed'
 cloudbreakBWAChr2InsertionPredictionsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/bwa_ins_0p26.bed'
-cloudbreakBWATweakChr2InsertionPredictionsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/bwa_ins_tweak_0p26.bed'
[…]

### 2013-07-10 manuscript/nm.tex
recent versions for submission

@@
 For the experiments reported in this paper we use a window size of 25bp.
-Within each window, we examine the distribution of insert sizes of mappings that span that window, and compute features by fitting a GMM to that distribution (Supplementary Materials, Supplementary Figure \ref{Sinsert_size_mixes}).
+Within each window, we examine the distribution of insert sizes of mappings that span that window, and compute features by fitting a GMM to that distribution (Supplementary Materials, Supplementary Fig.
+\ref{Sinsert_size_mixes}).
 To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials); these procedures also allow us to process multiple mappings for each read if they are reported by the aligner.
@@
 See Methods for details on how reads were aligned and each program was invoked.
-For all alignments we used BWA \autocite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Figure \ref{Salignment_comparison}).
+For all alignments we used BWA \autocite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Fig.
+\ref{Salignment_comparison}).
 
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/chr2_sim_rocs_runtime.pdf}
+\includegraphics[width=1\textwidth]{../figures/chr2_rocs_runtimes_journal.pdf}
 \caption{Accuracy and runtime performance on a simulated data set.
@@
 Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow.
-To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Figure \ref{Sscalability}).
+To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Fig.
+\ref{Sscalability}).
 
@@
 
-It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Figure \ref{Sbreakpoint_resolution}): SR methods have higher resolution than RP methods.
+It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Fig.
+\ref{Sbreakpoint_resolution}): SR methods have higher resolution than RP methods.
 Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we believe, however, that increasing sensitivity and specificity is of greatest utility, especially given the emergence of pipelines in which RP calls are validated \emph{in silico} by local assembly.
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/NA18507_rocs_runtime.pdf}
+\includegraphics[width=1\textwidth]{../figures/NA18507_rocs_runtimes_journal.pdf}
 \caption{Accuracy and performance on the 37X NA18507 sample.

### 2013-07-10 manuscript/nm_supp.tex
recent versions for submission

@@
 
-    sortGMMResults      Sort and merge GMM Results (use with one reducer to get all GMM feature 
-                        results into a single file
+    sortGMMResults      Sort and merge GMM Results (use with one reducer to get all
+                        GMM feature results into a single file
       Usage: sortGMMResults [options]

### 2013-07-22 src/main/scripts/manuscript.R
figure format generation

@@ -319,8 +319,16 @@ plotROC(perfsListDelsChr2,
         colLtyMapping=myColLtyMapping)
 dev.off()
 
+postscript("~/Documents/svpipeline/figures/CHR2SIM_DELS_ROC.eps", horizontal = FALSE, onefile = FALSE, paper = "special", height = 10, width = 12)
+par(xpd=T, mar=par()$mar+c(0,0,0,7))
+plotROC(perfsListDelsChr2, 
+        c("Cloudbreak", "Breakdancer","Pindel", "GASVPro", "DELLY-RP", "DELLY-SR"), 
+        totalDelsChr2, "Deletions in Venter diploid chr2 simulation",legendLoc=xy.coords(425,200), maxTP=350, 
+        colLtyMapping=myColLtyMapping)
+dev.off()
+
 perfsListDelsChr2 <- list(cloudbreak=cloudbreakBWAChr2DeletionPerf, breakdancer=breakdancerChr2DeletionPerf, pindel=pindelChr2DeletionPerf, gasv=gasvChr2DeletionPerf, delly=dellyChr2DeletionPerf, dellyBR=dellyBRChr2DeletionPerf)
-cairo_ps("~/Documents/svpipeline/figures/chr2_del_roc_cairo.ps", width=10, height=7)
+cairo_ps("~/Documents/svpipeline/figures/chr2_del_roc_cairo.ps", width=1, height=7)
 par(xpd=T, mar=par()$mar+c(.5,.5,.5,13))
 plotROC(perfsListDelsChr2, 
         c("Cloudbreak", "Breakdancer","Pindel", "GASVPro", "DELLY-RP", "DELLY-SR"), 
@@ -502,6 +510,14 @@ perfsListInsertionsMultimapChr2 <- list(breakdancer=breakdancerChr2InsertionPerf
                                         cloudbreakBWA=cloudbreakBWATweakChr2InsertionPerfSmoothed, cloudbreakBWAM=cloudbreakBWAXATweakChr2InsertionPerfSmoothed, cloudbreakGEM=cloudbreakGEMChr2InsertionPerfSmoothed)
 
 pdf(chr2InsertionsROCOutputFile, width=10)
+par(xpd=T, mar=par()$mar+c(0,0,0,8))
+plotROC(perfsListInsertionsChr2, 
+        c("Cloudbreak", "Breakdancer","Pindel"), 
+        totalInsertionsChr2, "Insertions in Venter diploid chr2 simulation",
[…]

### 2013-07-31 manuscript/bmc_additional1.tex
Manuscript in BMC format

--- 2013-07-10 manuscript/nm_supp.tex
@@
-\documentclass[11pt]{article}
+\documentclass[10pt]{bmc_article}
 
-\usepackage[margin=1in]{geometry}
+\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
+\usepackage{url}  % Formatting web addresses  
+\usepackage{ifthen}  % Conditional 
+\usepackage{multicol}   %Columns
+\usepackage[utf8]{inputenc} %unicode support
+%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
+%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+
 \usepackage{amsfonts}
 \usepackage{amsmath}
-\usepackage{graphicx}
 \usepackage{algorithm}
@@
 \usepackage{multirow}
-\usepackage{authblk}
 \usepackage{rotating}
@@
 
-% \renewcommand{\thetable}{S\arabic{table}}   
-% \renewcommand{\thefigure}{S\arabic{figure}}
-% \renewcommand{\thealgorithm}{\arabic{algorithm}}
+\renewcommand{\thefigure}{A\arabic{figure}}
+\renewcommand{\thetable}{A\arabic{table}}
 
-\usepackage[style=nature,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
-\renewbibmacro{in:}{%
-  \ifentrytype{article}{}{%
-  \printtext{\bibstring{in}\intitlepunct}}}
-\addbibresource{manuscript.bib}
+\urlstyle{rm}
 
-\renewcommand{\figurename}{Supplementary Figure}
-\renewcommand{\tablename}{Supplementary Table}
+\setlength{\topmargin}{0.0cm}
+\setlength{\textheight}{21.5cm}
+\setlength{\oddsidemargin}{0cm} 
+\setlength{\textwidth}{16.5cm}
+\setlength{\columnsep}{0.6cm}
 
-\newfloat{suppalgorithm}{p}{cap}
-\floatname{suppalgorithm}{Supplementary Algorithm}
+\newboolean{publ}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                              %%
+%% You may change the following style settings  %%
+%% Should you wish to format your article       %%
+%% in a publication style for printing out and  %%
+%% sharing with colleagues, but ensure that     %%
+%% before submitting to BMC that the style is   %%
+%% returned to the Review style setting.        %%
+%%                                              %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%Review style settings
+%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+
+%Publication style settings
+\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+
+\newcommand{\degree}{\ensuremath{^\circ}}
 
 \begin{document}
+\begin{bmcformat}
 
-\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\
-Supplementary Material}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the title of your article here     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\author[1,5]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
-\author[3,4]{Jeffrey Tyner}
-\author[6]{Alberto L'Abbate}
-\author[6]{Clelia Tiziana Storlazzi}
-\author[4,5]{Lucia Carbone}
-\author[1,2,5]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
-\affil[1]{Institute on Development and Disability and Center for Spoken Language Understanding}
-\affil[2]{Department of Medical Informatics \& Clinical Epidemiology}
-\affil[3]{Program in Molecular and Cellular Biosciences}
-\affil[4]{Behavioral Neuroscience Department}
-\affil[5]{Oregon Health \& Science University, Portland, OR, USA}
-\affil[6]{Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy}
+\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors here                   %%
+%%                                          %%
+%% Ensure \and is entered between all but   %%
+%% the last two authors. This will be       %%
+%% replaced by a comma in the final article %%
+%%                                          %%
+%% Ensure there are no trailing spaces at   %% 
+%% the ends of the lines                    %%     	
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\author{Christopher W Whelan\correspondingauthor$^{1}$%
+       \email{Christopher W Whelan\correspondingauthor - cwhelan@gmail.com}%
+      \and
+         Jeffrey Tyner$^2$%
+      \and
+         Alberto L'Abbate$^3$%
+      \and
+         Clelia Tiziana Storlazzi$^3$
+      \and
+         Lucia Carbone$^4$
+       and 
+         Kemal S\"onmez\correspondingauthor$^{4,5}$%
+         \email{Kemal S\"onmez\correspondingauthor - sonmezk@ohsu.edu}%
+      }
+
+\address{%
+    \iid(1)Institute on Development and Disability and Center for Spoken Language Understanding, Oregon Health \& Science University, %
+    3181 SW Sam Jackson Park Rd, GH40, Portland, OR 97239-3098\\
+    \iid(2)Program in Molecular and Cellular Biosciences, Oregon Health \& Science University, %
+    3181 SW Sam Jackson Park Rd, L102, Portland, OR 97239\\
+    \iid(3)Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy\\
+    \iid(4)Department of Behavioral Neuroscience, Oregon Health \& Science University, %
+    3181 SW Sam Jackson Park Rd, L470, Portland, Oregon, U.S.A. 97239-3098\\
+    \iid(5)Department of Medical Informatics \& Clinical Epidemiology, Oregon Health \& Science University, %
+    3181 S.W. Sam Jackson Park Rd., BICC, Portland, Oregon, U.S.A. 97239-3098
+}%
 
@@
 
-\begin{suppalgorithm}[h]
-\algrenewcommand\algorithmicprocedure{\textbf{job}}
-  \begin{algorithmic}[1]
-    \Procedure{Alignment}{}
-    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
-    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
-    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
-    \EndFor
-    \EndFunction
-    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
-    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
-    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
-    \EndFunction
-    \EndProcedure
-
-    \Procedure{Compute SV Features}{}
-    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
-    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
-    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
-    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
-    \EndFor
-    \EndFor
-    \EndFunction
-    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
-    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
-    \EndFunction
-    \EndProcedure
-
-    \Procedure{Call SVs}{}
-    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
-    \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
-    \EndFunction
-    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
-    \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
-    \EndFunction
-    \EndProcedure
-  \end{algorithmic}
-\caption{The algorithmic framework for SV calling in MapReduce.}
-\label{cb_algo}
-\end{suppalgorithm}
-
-\clearpage
-
 \begin{table}[h]
@@
 \clearpage
-
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{rrrrrr}
-  \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 224 &  84 & 82 &  31 & 26\\ 
-  \hline
-  Cloudbreak  & \textbf{68} (17)  & \textbf{67} (\textbf{10}) &  \textbf{56} (\textbf{5}) & \textbf{11} (\textbf{3}) & \textbf{15} (\textbf{0}) \\ 
-  Breakdancer & 52 (8)  & 49 (2) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
-  GASVPro     & 35 (2)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
-  DELLY-RP       & 22 (1)  & 56 (1) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
-  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
-  Pindel      & 60 (\textbf{32})  & 16 (0) &  41 (2) & 1 (0) & 12 (\textbf{0})\\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPredsFDR10}
-\end{table}
-
-\newpage
 
@@
 \end{center}
-\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
+\caption{Table A4. Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
 \label{deletionGenotypeaccuracy}
@@
 
-\section{Supplementary Discussion}
-
-\subsection{A general framework for SV Detection in MapReduce}
-
-We have developed a conceptual algorithmic framework for SV detection in MapReduce, which is outlined in Supplementary Algorithm \ref{cb_algo}.
-As described in the main text, this framework divides processing into three separate MapReduce jobs: an alignment job, a feature computation job, and an SV calling job.
-
-The alignment job uses sensitive mapping tools and settings to discover mapping locations for each read pair.
-Aligners can be executed to report multiple possible mappings for each read, or only the best possible mapping.
-Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits possible mapping locations under the $rpid$ key.
-Reducers then collect the alignments for each paired end, making them available under one key for the next job.
-Our implementation contains wrappers to execute the aligners BWA, GEM \autocite{MarcoSola:2012hm}, Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
-
-In the second job, we compute a set of features for each location in the genome.
-To begin, we tile the genome with small fixed-width, non-overlapping intervals.
-For the experiments reported in this paper we use an interval size of 25bp.
-Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
-Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
-Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
-For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
-the fragment size implied by the alignments and the likelihood the alignments are correct.
-We then leave two functions to be implemented depending on the application:
-\begin{flalign*}
- \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
- \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
-\end{flalign*}
-
-The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
-We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
-The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
-
-Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
-It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
-We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
-
-\subsection{Cloudbreak: An Implementation of the MapReduce SV Detection Framework}
-
-Our Cloudbreak software package can be seen as an implementation of the general framework defined above.
-In particular, Cloudbreak implements the three user-defined function described above as follows:
-
-\begin{description}
-\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
-For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
-\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure \ref{insert_size_mixes}).
-Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
-If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant.
-Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components.
-Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
-
-To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
-We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
-In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
-This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
-Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
-
-\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
-
-And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
-
-We fit the parameters of the GMM using the Expectation-Maximization algorithm.
-Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
-We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
-In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
-We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions.
-In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$.
-We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$.
-We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.
-
-The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
-
-\item[\sc{PostProcess}] We convert our features along each chromosome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
-To eliminate noise we apply a median filter with window size 5.
-We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
-\end{description}
-
-\subsection{Runtime Analysis}
+\section{Runtime Analysis}
 
@@
 
-In Supplementary Table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work.
+In Table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work.
 Each tool varies in the amount of parallelization supported.
@@
 This distribution of processing across the entire cluster also serves to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set during a run where one of the compute nodes was rebooted midway through the feature generation job.
-
-\printbibliography[title=Supplementary References]
 
@@
 
+\end{bmcformat}
 \end{document}

### 2013-07-31 manuscript/bmc_article.tex
Manuscript in BMC format

@@
 %\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+
+\usepackage{amsfonts}
+\usepackage{amsmath}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
+\usepackage{url}
+\usepackage{multirow}
+\usepackage{rotating}
+\usepackage{xr}
+
 \urlstyle{rm}
@@
 
+\newcommand{\degree}{\ensuremath{^\circ}}
+
 % Begin ...
@@
 
-\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}
+\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
 
@@
 
-\author{Christopher W. Whelan\correspondingauthor$^{1}$%
-       \email{Christopher W. Whelan\correspondingauthor -whelanch@ohsu.edu}%
+\author{Christopher W Whelan\correspondingauthor$^{1}$%
+       \email{Christopher W Whelan\correspondingauthor - cwhelan@gmail.com}%
       \and
-         Kemal S\"onmez\correspondingauthor$^{1,2}$%
-         \email{Kemal S\"onmez\correspondingauthor - sonmezk@ohsu.edu}
+         Jeffrey Tyner$^2$%
+      \and
+         Alberto L'Abbate$^3$%
+      \and
+         Clelia Tiziana Storlazzi$^3$
+      \and
+         Lucia Carbone$^4$
+       and 
+         Kemal S\"onmez\correspondingauthor$^{4,5}$%
+         \email{Kemal S\"onmez\correspondingauthor - sonmezk@ohsu.edu}%
       }
@@
 \address{%
-    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
-        Waterloo Road, London, UK\\
-    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK
+    \iid(1)Institute on Development and Disability and Center for Spoken Language Understanding, Oregon Health \& Science University, %
+    3181 SW Sam Jackson Park Rd, GH40, Portland, OR 97239-3098\\
+    \iid(2)Program in Molecular and Cellular Biosciences, Oregon Health \& Science University, %
+    3181 SW Sam Jackson Park Rd, L102, Portland, OR 97239\\
+    \iid(3)Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy\\
+    \iid(4)Department of Behavioral Neuroscience, Oregon Health \& Science University, %
+    3181 SW Sam Jackson Park Rd, L470, Portland, Oregon, U.S.A. 97239-3098\\
+    \iid(5)Department of Medical Informatics \& Clinical Epidemiology, Oregon Health \& Science University, %
+    3181 S.W. Sam Jackson Park Rd., BICC, Portland, Oregon, U.S.A. 97239-3098
 }%
@@
 \begin{abstract}
-        % Do not use inserted blank lines (ie \\) until main body of text.
-        \paragraph*{Background:} 
-The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data.
-Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods.
-However, the computational requirements for storing and processing data sets with multiple mappings can be formidable.
-Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers.
-MapReduce is becoming a standard framework for distributing processing across such compute clusters.
-        \paragraph*{Results:} We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome.
-In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome.
-On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes.
-In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome.
-In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples.
-        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
+The detection of genomic structural variations (SV) remains a difficult challenge in analyzing sequencing data, and the growing size and number of sequenced genomes have rendered SV detection a bona fide big data problem.
+MapReduce is a proven, scalable solution for distributed computing on huge data sets.
+We describe a conceptual framework for SV detection algorithms in MapReduce based on computing local genomic features, and use it to develop a deletion and insertion detection algorithm, Cloudbreak.
+On simulated and real data sets, Cloudbreak achieves accuracy improvements over popular SV detection algorithms, and genotypes variants from diploid samples.
+It provides dramatically shorter runtimes and the ability to scale to big data volumes on large compute clusters.
+Cloudbreak includes tools to set up and configure MapReduce (Hadoop) clusters on cloud services, enabling on-demand cluster computing.
+Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.
 \end{abstract}
@@
 \section*{Background}
- Text for this section.\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}
+
+Genomic structural variations (SVs) such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \cite{Mills:2011p1611, Conrad:2010ja}.
+However, detection of SVs with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \cite{Mills:2011p1611}.
+Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, and integrating them to form an accurate detection measure is computationally difficult.
+As the volume of massively parallel sequencing data approaches ``big data'' scales, SV detection is becoming a time consuming component of genomics pipelines, and presents a significant challenge for research groups and clinical operations that may not be able to scale their computational infrastructure.
+Here we present a distributed software solution that is scalable and readily available on the cloud.
+
+In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data.
+In particular, Google's MapReduce \cite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large data sets across clusters of commodity servers.
+Hadoop is an open source project of the Apache Foundation which provides an implementation of MapReduce.
+MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention.
+The Hadoop framework has been shown to be effective in sequencing-related tasks including short read mapping \cite{Schatz:2009p278}, SNP calling \cite{Langmead:2009p1225}, and RNA-seq differential expression analysis \cite{Langmead:2010p1268}.
+
+Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing problems like SV detection.
+MapReduce divides computation across a cluster into three phases.
+In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
+The framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
+Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
+
+Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets\cite{Alkan:2011p547}.
+Read-pair (RP) based methods use the distance between and orientation of the mappings of paired reads to identify the signatures of SVs \cite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
+Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation, and then clustering the discordant mappings to find SVs supported by multiple discordantly mapped read pairs.
+Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome \cite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
+Finally, split-read (SR) methods look for breakpoints by mapping portions of individual reads to different genomic locations \cite{Wang:2011p1607,Ye:2009p2}.
+
+Many RP methods consider only unambiguously discordantly mapped read pairs.
+Some approaches also include ambiguous mappings of discordant read pairs to improve sensitivity in repetitive regions of the genome \cite{Hormozdiari:2009p284,Quinlan:2010gf}.
+Several recent RP approaches have also considered concordant read pairs, either to integrate RD signals for improved accuracy \cite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds that separate concordant from discordant mappings and thus detect smaller events \cite{Marschall:2012ek}.
+Increasing the number of read mappings considered, however, increases the computational burden of SV detection.
+
+Our goal is to leverage the strengths of the MapReduce computational framework in order to provide fast, accurate and readily scalable SV-detection pipelines.
+The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches.
+MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
+In sequencing applications, for example, MapReduce based SNV-callers Crossbow \cite{Langmead:2009p1225} and GATK \cite{McKenna:2010p1051} perform independent calculations on partitions of the genome.
+SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \cite{Lee:2009da} and forestSV \cite{Michaelson:2012fj} compute scores or features along the genome and then produce SV predictions from those features in a post-processing step.
+We will show that this strategy can be translated into the MapReduce architecture.
+
+In this paper, we describe a framework for solving SV detection problems in Hadoop based on the computation of local genomic features from paired end mappings.
+In this framework we have developed a software package, Cloudbreak, that discovers genomic deletions up to 25,000bp long, and short insertions.
+Cloudbreak computes local features based on modeling the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM), an idea first implemented in MoDIL \cite{Lee:2009da}.
+The use of Hadoop enables scalable implementations of this class of algorithm.
+We characterize our algorithm's performance on simulated and real data sets and compare its performance to those of several popular methods.
+
+Finally, and quite importantly from a practical point of view, our implementation of Cloudbreak provides the functionality to easily set up and configure Hadoop clusters on cloud service providers, making dynamically scalable distributed SV detection accessible to all whose computational needs demand it.
 
@@
 \section*{Results and Discussion}
-  \subsection*{Results sub-heading}
-    \subsubsection*{This is a sub-sub-heading}
-      Sub-sub-sub-headings are made with the \textsl{\\subsubsection} command. \pb
-      pb at end of lines ensures correct paragraph spacing.\pb
-	  Text for this sub-sub-section \ldots
-    \subsubsection*{Another sub-sub-sub-heading}
-      Text for this sub-sub-section \ldots
 
-  \subsection*{Another results sub-heading}
-    Text for this sub-section \ldots
+\subsection*{A general framework for SV Detection in MapReduce}
 
-  \subsection*{Yet another results sub-heading}
-    Text for this sub-section.  More results \ldots
+We have developed a generalized algorithmic framework for SV detection in MapReduce.
+Our framework divides processing into three distinct MapReduce jobs (Figure 1): a job that can align reads to the reference using a variety of mapping algorithms; a job that computes a set of features along the genome; and a job which calls structural variations based on those features.
+The algorithm is listed in Table 1 and each job is described in more detail below.
+
+The alignment job uses sensitive mapping tools and settings to discover mapping locations for each read pair.
+Aligners can be executed to report multiple possible mappings for each read, or only the best possible mapping.
+Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits possible mapping locations under the $rpid$ key.
+Reducers then collect the alignments for each paired end, making them available under one key for the next job.
+Our implementation contains wrappers to execute the aligners BWA, GEM \cite{MarcoSola:2012hm}, Novoalign \cite{novoalign}, RazerS 3 \cite{Weese:2012by}, mrFAST \cite{Alkan:2009cr}, and Bowtie 2 \cite{Langmead:2012jh}.
+This job can also be skipped in favor of importing a pre-aligned BAM file directly to HDFS.
+
+In the second job, we compute a set of features for each location in the genome.
+To begin, we tile the genome with small fixed-width, non-overlapping intervals.
+For the experiments reported in this paper we use an interval size of 25bp.
+Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
+Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
+Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
+For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
+the fragment size implied by the alignments and the likelihood the alignments are correct.
+We then leave two functions to be implemented depending on the application:
+\begin{flalign*}
+ \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
+ \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
+\end{flalign*}
+
+The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
+We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
+The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
+
+Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
+It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
+We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
+
+\subsection*{Cloudbreak: A Hadoop/MapReduce Software Package for SV Detection}
+
+Cloudbreak is our implementation of the infrastructure for the framework described above.
+It demonstrates this framework by implementing a detection algorithm for small-to-medium sized deletions (40bp-25,000bp) and small insertions based on examining the insert sizes of paired end mappings.
+In particular, we implement the three user-defined functions specified above as follows:
+
+\begin{description}
+\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
+For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
+\item[$\Phi$] To compute features for each genomic location, we follow \cite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Figure A1).
+Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
+If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant.
+Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components.
+Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
+The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
+See Methods for more details on the GMM fitting procedure.
+Prior to fiting the GMM at each location, we attempt to filter our incorrect mappings for that location using an outlier-detection based clustering scheme and an adaptive mapping quality cutoff (see Methods for details).
+
+\item[\sc{PostProcess}] The third MapReduce job is responsible for making SV calls based on the feature defined above.
+We convert our features along each chromosome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
+To eliminate noise we apply a median filter with window size 5.
+We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
+\end{description}
+
+An illustration of the Cloudbreak algorithm working on a simple example is shown in Figure A2.
+
+Cloudbreak can be executed on any Hadoop cluster; Hadoop abstracts away the details of cluster configuration, making distributed applications portable.
+In addition, our Cloudbreak implementation can leverage the Apache Whirr library to automatically create clusters with cloud service providers such as the Amazon Elastic Compute Cloud (EC2).
+This enables on demand provisioning of Hadoop clusters which can then be terminated when processing is complete, eliminating the need to invest in a standing cluster and allowing a model in which users can scale their computational infrastructure as their need for it varies over time.
+Instructions and examples describing how to leverage cloud computing with Cloudbreak are available in the user manual.
+
+\subsection*{Tests with Simulated Data}
+
+We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer \cite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings \cite{Sindi:2012kk}, the SR method Pindel \cite{Ye:2009p2}, and the hybrid RP-SR method DELLY \cite{Rausch:2012he}.
+DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence; we refer to these sets of calls as DELLY-RP and DELLY-SR.
+We also attempted to evaluate MoDIL on the same data.
+All of these methods detect deletions.
+Insertions can be detected by BreakDancer, Pindel, and MoDIL.
+See Methods for details on how reads were aligned and each program was invoked.
+For all alignments we used BWA \cite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Figure A3).
+
+There is no available test set of real Illumina sequencing data from a sample that has a complete annotation of SVs.
+Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics.
+On the other hand, any simulated data should contain realistic SVs that follow patterns observed in real data.
+Therefore, we took one of the most complete lists of SVs from an individual, the list of homozygous insertions and deletions from the genome of J.
+Craig Venter \cite{Levy:2007fb}, and used it to simulate a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants, with 100bp reads and a mean fragment size of 300bp (See Methods).
+
+Figure 2 shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment.
+See Methods for a description of how we identified correct predictions.
+All approaches show excellent specificity at high thresholds in this simulation.
+Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY.
+For insertions, Cloudbreak's performance is similar to or slightly better than Pindel.
+Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes.
+(Of course, Cloudbreak uses many more CPUs as distributed algorithm.
+See Additional File 1 and Table A1 for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster.
+Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow.
+To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Figure A4).
+
+Choosing the correct threshold to set on the output of an SV calling algorithm can be difficult.
+The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying thresholds.
+First, we characterized the predictions made by each algorithm at the threshold that gives them maximum sensitivity.
+For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.
+The results are summarized in Table 2.
+MoDIL and Cloudbreak exhibited the greatest recall for deletions.
+Cloudbreak has also has high precision at this threshold, and discovers many small variants.
+For insertions, Cloudbreak has the highest recall, although recall is low for all four approaches.
+Cloudbreak again identifies many small variants.
+Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by RP mapping.
+We also used the ROC curves to characterize algorithm performance when a low false discovery rate is required.
+Table 3 shows the total number of deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
+At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool.
+Insertions performance never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.
+We also examined Cloudbreak's ability to detect events in repetitive regions of the genome, and found that it was similar to the other methods tested (Tables A2 and A3).
+
+It should be noted that the methods tested here vary in their breakpoint resolution (Figure A5): SR methods have higher resolution than RP methods.
+Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we believe, however, that increasing sensitivity and specificity is of greatest utility, especially given the emergence of pipelines in which RP calls are validated \emph{in silico} by local assembly.
+
+\subsection*{Tests with Biological Data}
+
+We downloaded a set of reads from Yoruban individual NA18507, experiment ERX009609, from the Sequence Read Archive.
+This sample was sequenced by Illumina Inc.
+on the Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage.
+To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same individual.
+These included data from the Human Genome Structural Variation Project reported by \cite{Kidd:2008p926}, a survey of small indels conducted by \cite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \cite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
+We merged any overlapping calls of the same type into the region spanned by their unions.
+We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.
+
+Figure 3 shows the performance of each algorithm at detecting events larger than 40bp on the NA18507 data set.
+All algorithms show far less specificity for the gold standard set than they did in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample.
+For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
+Cloudbreak has comparable accuracy for insertions to other tools, and can identify the most variants at higher levels of sensitivity.
+Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, BreakDancer, even when BreakDancer is run in parallel for each chromosome on different nodes in the cluster (see Additional File 1 for a discussion of parallelizing existing tools).
+
+Given the high number of novel predictions made by all tools at maximum sensitivity, we decided to characterize performance at more stringent thresholds.
+We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
+For insertions, we used the maximum sensitivity thresholds for each tool due to the high observed FDRs in the simulated data.
+Precision and recall at these thresholds, as well as the performance of each algorithm at predicting variants of each size class, is shown in Table 4.
+For deletions, Cloudbreak has the greatest sensitivity of any tool, identifying the most variants in each size class.
+Pindel exhibits the highest precision with respect to the gold standard set.
+For insertions, Pindel again has the highest precision at maximum sensitivity, while Cloudbreak has by far the highest recall.
+
+\subsection*{Performance on a Low-Coverage Cancer Data Set}
+
+We also tested Cloudbreak on a sequencing data set obtained from a patient with acute myeloid leukemia (AML).
+This data set consisted of 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
+Using a pipeline consisting of Novoalign, BreakDancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions.
+Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even when using lower coverage data sets with a greater variance of insert sizes.
+The variants identified include deletions in the gene CTDSPL/RBPS3, an AML tumor suppressor \cite{Zheng:2012kk}, and NBEAL1, a gene up-regulated in some cancers \cite{Chen:2004jo}.
+We are currently investigating these deletions to determine their functional impact on this patient.
+
+\subsection*{Genotyping Variants}
+
+Because Cloudbreak explicitly models zygosity in its feature generation algorithm, it can predict the genotypes of identified variants.
+We tested this on both the simulated and NA18507 data sets.
+For the NA18507 data set, we considered the deletions from the 1000 Genomes Project, which had been genotyped using the population-scale SV detection algorithm Genome STRiP \cite{Handsaker:2011ki}.
+Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the deletions it detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
+Table A4 shows confusion matrices for the two samples using this classifier.
+None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that met our size threshold and also had genotyping information.
+Of the 123 insertions detected by Cloudbreak on the simulated data set, 43 were heterozygous.
+Cloudbreak correctly classified 78 of the 80 homozygous insertions and 31 of the 43 heterozygous insertions, for an overall accuracy of 88.6\%.
+
+\subsection*{Discussion}
+
+Over the next few years, due to advances in sequencing technology, genomics data are expected to increase in volume by several orders of magnitude, expanding into the realm referred to as ``big data''.
+In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines.
+In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.
+
+Among big data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
+In this paper, we introduced a novel conceptual framework for SV detection algorithms in MapReduce, based on computing local genomic features.
+This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of an algorithm for detecting insertions and deletions based on fitting a GMM to the distribution of mapped insert sizes spanning each genomic location.
 
@@
 \section*{Conclusions}
-  Text for this section \ldots
+
+On simulated and real data sets, our approach exhibits high accuracy when compared to popular SV detection algorithms that run on traditional clusters and servers.
+Detection of insertions and deletions is an important area of research; \cite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
+More SV annotations are also required to assess the impact of SVs on non-coding elements \cite{Mu:2011br}.In addition to delivering state-of-the-art performance in a RP-based SV detection tool, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
+With the advent of cloud service providers such as Amazon EC2, it is becoming easy to instantiate on-demand Hadoop compute clusters.
+Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who need to analyze increasing amounts of sequencing data.
 
@@
 \section*{Methods}
-  \subsection*{Methods sub-heading for this section}
-    Text for this sub-section \ldots
 
-  \subsection*{Another methods sub-heading for this section}
-    Text for this sub-section \ldots
+\subsection*{Cloudbreak Implementation}
+Cloudbreak is a native Java Hadoop application.
+We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4.
+We use snappy compression for MapReduce data.
+Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
+To execute the other tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine (\url{http://research.cs.wisc.edu/htcondor/}) with directed acyclic graphs to describe dependencies between jobs.
 
-  \subsection*{Yet another sub-heading for this section}
-    Text for this sub-section \ldots
+\subsection*{Fitting a GMM to Observed Insert Size Distributions}
+At each genome location, we fit the parameters of the GMM using the Expectation-Maximization algorithm.
+Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
+We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
+In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
+We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions.
+In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$.
+We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$.
+We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.
+
+\subsection*{Filtering Ambiguous and Incorrect Mappings}
+To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
+These procedures also allow us to process multiple mappings for each read if they are reported by the aligner.
+We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
+In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
+This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
+Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
+
+\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
+
+And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
+
+\subsection*{Cloudbreak Genotyping}
+We used the parameters of the fit GMM to infer the genotype of each predicted variant.
+Assuming that our pipeline is capturing all relevant read mappings near the locus of the variant, the genotype should be indicated by the estimated parameter $\alpha$, the mixing parameter that controls the weight of the two components in the GMM.
+We setting a simple cutoff of .35 on the average value of $\alpha$ for each prediction to call the predicted variant homozygous or heterozygous.
+We used the same cutoff for deletion and insertion predictions.
+
+\subsection*{Read Simulation}
+Since there are relatively few heterozygous insertions and deletions annotated in the Venter genome, we used the set of homozygous indels contained in the HuRef data (\texttt{HuRef.homozygous\_indels.061109.gff}) and randomly assigned each variant to be either homozygous or heterozygous.
+Based on this genotype, we applied each variant to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
+We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package (\url{http://sourceforge.net/apps/mediawiki/dnaa/}).
+We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence.
+Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous insertions and deletions.
+
+\subsection*{Read Alignments}
+Simulated reads were aligned to hg18 chromosome 2, and NA18507 reads were aligned to the hg19 assembly.
+Alignments for all programs, unless otherwise noted, were aligned using BWA \cite{Li:2009p836} \texttt{aln} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
+GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \texttt{-a -r -Ex 1100 -t 250}.
+
+\subsection*{SV Tool Execution}
+We ran BreakDancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
+To run Breakdancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer\_max} for each chromosome using the \texttt{-o} parameter.
+We ran DELLY version 0.0.9 with the \texttt{-p} parameter and default values for other parameters.
+For the parallel run of DELLY we first split the original BAM file with BamTools \cite{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
+We ran GASVPro version 1.2 using the \texttt{GASVPro.sh} script and default parameters.
+Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \texttt{-c} option.
+We executed MoDIL with default parameters except for a \texttt{MAX\_DEL\_SIZE} of 25000, and processed it in parallel on our cluster with a step size of 121475.
+
+\subsection*{SV Prediction Evaluation}
+We use the following criteria to define a true prediction given a gold standard set of deletion and insertion variants to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method.
+For evaluating insertions, each algorithm produces insertion predictions that define an interval in which the insertion is predicted to have occurred with start and end coordinates $s$ and $e$ as well as the predicted length of the insertion, $l$.
+The true insertions are defined in terms of their actual insertion coordinate $i$ and their actual length $l_a$.
+Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$.
+In this study we are interested in detecting events larger than 40bp, because with longer reads, smaller events can be more easily discovered by examining gaps in individual reads.
+Both Pindel and MoDIL make many calls with a predicted event size of under 40bp, so we remove those calls from the output sets of those programs.
+Finally, we exclude from consideration calls from all approaches that match a true deletion of less than 40bp where the predicted variant length is less than or equal to 75bp in length.
+
+\subsection*{Preparation of AML Sample}
+
+Peripheral blood was collected from a patient with acute myelomonocytic leukemia (previously designated acute myeloid leukemia FAB M4) under a written and oral informed consent process reviewed and approved by the Institutional Review Board of Oregon Health \& Science University.
+Known cytogenetic abnormalities associated with this specimen included trisomy 8 and internal tandem duplications within the FLT3 gene.
+Mononuclear cells were separated on a Ficoll gradient, followed by red cell lysis.
+Mononuclear cells were immunostained using antibodies specific for CD3, CD14, CD34, and CD117 (all from BD Biosciences) and cell fractions were sorted using a BD FACSAria flow cytometer.
+Cell fractions isolated included T-cells (CD3-positive), malignant monocytes (CD14-positive), and malignant blasts (CD34, CD117-positive).
+We sequenced CD14+ cells on an Illumina Genome Analyzer II, producing 128,819,200 paired-end reads.
+
+\subsection*{Validation of AML Deletions by PCR}
+
+Deletions identified in the AML dataset were validated by PCR.
+SVs to validate were selected based on calls made by BreakDancer, relying on the score and the number of the reads supporting each single event.
+Appropriate primers were designed with an internet-based interface, Primer3 (\url{http://frodo.wi.mit.edu/}), considering the chromosome localization and orientation of the interval involved in the candidate rearrangement.
+The primers were checked for specificity using the BLAT tool of the UCSC Human Genome Browser (\url{http://genome.ucsc.edu/cgi-bin/hgBlat}).
+All the primer pairs were preliminarily tested on the patient genomic DNA and a normal genomic DNA as control.
+The PCR conditions were as follows: 2 min at 95\degree C followed by 35 cycles of 30 sec at 95\degree C, 20 sec at 60\degree C, and 2 min at 72\degree C.
+All the obtained PCR products were sequenced and analyzed by BLAT for sequence specificity.
 
@@
 \section*{Authors contributions}
-    Text for this section \ldots
+
+C.W.W.
+and K.S.
+designed the algorithmic approaches.
+C.W.W.
+implemented the software and conducted the evaluations.
+J.T.
+obtained the AML sample and sorted the cells.
+L.C.
+sequenced the AML sample.
+C.T.S and A.L.A.
+performed PCR validations on the AML sample.
+C.W.W.
+and K.S.
+wrote the manuscript.
 
@@
   \ifthenelse{\boolean{publ}}{\small}{}
-  Text for this section \ldots
+
+We would like to thank Izhak Shafran at the Center for Spoken Language Understanding for advice, support, and shared computational resources, and Bob Handsaker for helpful conversations.
+We also thank reviewers for the Third Annual RECOMB Satellite Workshop on Massively Parallel Sequencing (RECOMB-seq) for their comments on a preliminary description of Cloudbreak.
+C.T.S.
+and A.L.A.
+are supported by the Italian Association for Cancer Research (AIRC).
+Sequencing of the AML sample was supported by the Oregon Medical Research Foundation.
 
@@
 \section*{Figures}
-  \subsection*{Figure 1 - Sample figure title}
-      A short description of the figure content
-      should go here.
+  \subsection*{Figure 1 - An Overview of the Cloudbreak Workflow}
+An overview of the Cloudbreak workflow.
+Reads are first uploaded to a Hadoop cluster from local storage.
+Cloudbreak then executes three MapReduce jobs to process the data: 1) Mapping with sensitive settings.
+2) Computation of features across the genome.
+3) Calling structural variations based on the features computed in the previous step.
+Finally, SV predictions can be downloaded from the Hadoop cluster and examined.
+Cloudbreak can also use the Apache Whirr library to automatically provision Hadoop clusters on and deploy data to cloud providers such as Amazon Elastic Compute Cloud.
 
-  \subsection*{Figure 2 - Sample figure title}
-      Figure legend text.
+  \subsection*{Figure 2 - Accuracy and Runtime on a Simulated Data Set.}
+Accuracy and runtime performance on a simulated data set.
+(a) Receiver Operating Characteristic (ROC) curves showing the specificity and sensitivity of each tool to deletions larger than 40bp on a simulated set of reads giving diploid coverage of 30X on human chromosome 2.
+Deletions from the Venter genome were randomly added to one or both haplotypes.
+Each point on a curve represents a different threshold on the confidence of predictions made by that tool.
+Thresholds vary by: Cloudbreak - likelihood ratio; BreakDancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.
+(b) ROC curves for insertion predictions.
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Additional File 1).
+
+\subsection*{Figure 3 - Accuracy and Runtime on the NA18507 Sample.}
+Accuracy and performance on the 37X NA18507 sample.
+(a) ROC curve for deletion prediction performance, tested against the combined gold standard sets of deletions taken from \cite{Kidd:2008p926}, \cite{Mills:2011fi}, and \cite{GenomesProjectConsortium:2012co}.
+(b) ROC curve for insertion prediction performance.
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Additional File 1).
 
@@
 \section*{Tables}
-  \subsection*{Table 1 - Sample table title}
-    Here is an example of a \emph{small} table in \LaTeX\ using  
-    \verb|\tabular{...}|. This is where the description of the table 
-    should go. \par \mbox{}
+
+  \subsection*{Table 1 - A General MapReduce Algoritm for SV Detection}
+\algrenewcommand\algorithmicprocedure{\textbf{job}}
+  \begin{algorithmic}[1]
+    \Procedure{Alignment}{}
+    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
+    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
+    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
+    \EndFor
+    \EndFunction
+    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
+    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
+    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
+    \EndFunction
+    \EndProcedure
+
+    \Procedure{Compute SV Features}{}
+    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
+    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
+    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
+    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
+    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
+    \EndFor
+    \EndFor
+    \EndFunction
+    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
+    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
+    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
+    \EndFunction
+    \EndProcedure
+
+    \Procedure{Call SVs}{}
+    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
+    \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
+    \EndFunction
+    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
+    \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
+    \EndFunction
+    \EndProcedure
+  \end{algorithmic}
+
+  \subsection*{Table 2 - Variants detected in the simulated data set}
+The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses).
+The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section.
+\par \mbox{}
 \par
 \mbox{
-      \begin{tabular}{|c|c|c|}
-        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
-        A1 & B2  & C3 \\ \hline
-        A2 & ... & .. \\ \hline
-        A3 & ..  & .  \\ \hline
+\begin{tabular}{r|rrr|rrrrr}
+  \cline{2-9}
+   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\hline
+\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
+  \hline
+\cline{2-9}
+&  Cloudbreak    &  0.638 & \textbf{0.678} & \textbf{153} (9)  & 61 (0) &  62 (0) & 12 (0) & 15 (0) \\ 
+&  BreakDancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
+&  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
+&  DELLY-RP           & 0.457 & 0.613 & 114 (3)  & \textbf{68} (0) &  \textbf{66} (0) & 9 (1) & 17 (0) \\ 
+&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
+&  Pindel           & 0.462 & 0.421 & 96 (\textbf{11})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
+&  MoDIL           & 0.132  & 0.66 & 123 (6)  & 66 (\textbf{3}) &  \textbf{66} (\textbf{11}) & \textbf{17} (\textbf{7}) & \textbf{23} (\textbf{8})\\ 
+   \hline
+\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
+\cline{2-9}
+&  Cloudbreak   &0.451 & \textbf{0.305}  & \textbf{79} (\textbf{32})  & \textbf{32} (\textbf{18}) &  \textbf{11} (8) & 1 (0) & 0 (0) \\ 
+&  BreakDancer & 0.262 & 0.0968  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
+&  Pindel          & \textbf{0.572} & 0.196 & 52 (25)  & 5 (1) &  10 (\textbf{9}) & \textbf{3} (\textbf{2}) & \textbf{9} (\textbf{9})\\ 
+&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (0) &  1 (0) & 2 (\textbf{2}) & 0 (0)\\ 
+\hline
 \end{tabular}
 }
-  \subsection*{Table 2 - Sample table title}
-    Large tables are attached as separate files but should
-    still be described here.
+
+\subsection*{Table 3 - Variants Detected in the Simulated Data at 10\% FDR}
+The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
+The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.
+\par \mbox{}
+
+\par
+\mbox{
+\begin{tabular}{rrrrrr}
+  \hline
+ & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number & 224 &  84 & 82 &  31 & 26\\ 
+  \hline
+  Cloudbreak  & \textbf{68} (17)  & \textbf{67} (\textbf{10}) &  \textbf{56} (\textbf{5}) & \textbf{11} (\textbf{3}) & \textbf{15} (\textbf{0}) \\ 
+  Breakdancer & 52 (8)  & 49 (2) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
+  GASVPro     & 35 (2)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
+  DELLY-RP       & 22 (1)  & 56 (1) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
+  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
+  Pindel      & 60 (\textbf{32})  & 16 (0) &  41 (2) & 1 (0) & 12 (\textbf{0})\\ 
+   \hline
+\end{tabular}
+}
+
+  \subsection*{Table 4 - Variants Detected on the NA18507 Data Set}
+
+The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class found.
+Exclusive predictions are in parentheses.
+For deletions, the same cutoffs were used as for the simulated data as in Table 2, adjusted for the difference in coverage from 30X to 37X.
+For insertions, the maximum sensitivity cutoff was used.
+\par \mbox{}
+\par
+\mbox{
+\begin{tabular}{r|rrr|rrrrr}
+  \cline{2-9}
+&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\hline
+\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,462 & 240 & 232 & 147 & 540 \\
+  \hline
+\cline{2-9}
+& Cloudbreak & 0.0943 & \textbf{0.17} & \textbf{573} (\textbf{277})  & \textbf{176} (\textbf{30}) &  \textbf{197} (\textbf{18}) & \textbf{121} (\textbf{6}) & \textbf{399} (\textbf{24}) \\ 
+& BreakDancer & 0.137 & 0.123 & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
+&  GASVPro & 0.147 & 0.0474 & 120 (21)  & 40 (2) &  85 (0) & 36 (0) & 128 (0) \\ 
+&  DELLY-RP & 0.0931 & 0.1 & 143 (6)  & 128 (3) &  167 (1) & 103 (0) & 323 (1) \\ 
+&  DELLY-SR & 0.153 & 0.0485 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
+&  Pindel & \textbf{0.179} & 0.0748 & 149 (8)  & 61 (0) &  149 (0) & 69 (1) & 217 (0) \\ 
+\hline
+\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 & 0 \\
+\cline{2-9}
+& Cloudbreak & 0.0323 & \textbf{0.455} & \textbf{265} (\textbf{104})  & \textbf{49} (\textbf{24}) &  3 (1) & 0 (0)  & 0 (0)  \\ 
+& BreakDancer & 0.0281 & 0.181 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
+&  Pindel & \textbf{0.0387} & 0.239 & 144 (45)  & 14 (7) &  \textbf{7} (\textbf{6}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
+\hline
+\end{tabular}
+}
 
@@
 \section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
+  \subsection*{Additional file 1 --- Appendix}
+  An appendix containing additional figures, tables, and a discussion of techniques for parallelizing exisisting tools and measuring runtime performance.
 

### 2013-08-01 manuscript/bmc_article.tex
named jobs as suggested by kemal

@@
 We have developed a generalized algorithmic framework for SV detection in MapReduce.
-Our framework divides processing into three distinct MapReduce jobs (Figure 1): a job that can align reads to the reference using a variety of mapping algorithms; a job that computes a set of features along the genome; and a job which calls structural variations based on those features.
-The algorithm is listed in Table 1 and each job is described in more detail below.
+Our framework divides processing into three distinct MapReduce jobs (Figure 1): \textsc{Align Reads}, a job that can align reads to the reference using a variety of mapping algorithms; \textsc{Compute Features}, a job that computes a set of features along the genome; and \textsc{Call Variants}, a job which calls structural variations based on those features.
+Our algorithmic framework is listed in Table 1 and each job is described in more detail below.
 
-The alignment job uses sensitive mapping tools and settings to discover mapping locations for each read pair.
+The \textsc{Align Reads} job uses sensitive mapping tools and settings to discover mapping locations for each read pair.
 Aligners can be executed to report multiple possible mappings for each read, or only the best possible mapping.
@@
 Reducers then collect the alignments for each paired end, making them available under one key for the next job.
-Our implementation contains wrappers to execute the aligners BWA, GEM \cite{MarcoSola:2012hm}, Novoalign \cite{novoalign}, RazerS 3 \cite{Weese:2012by}, mrFAST \cite{Alkan:2009cr}, and Bowtie 2 \cite{Langmead:2012jh}.
+Our implementation of the framework contains wrappers to execute the aligners BWA \cite{Li:2009p836}, GEM \cite{MarcoSola:2012hm}, Novoalign \cite{novoalign}, RazerS 3 \cite{Weese:2012by}, mrFAST \cite{Alkan:2009cr}, and Bowtie 2 \cite{Langmead:2012jh}.
 This job can also be skipped in favor of importing a pre-aligned BAM file directly to HDFS.
 
-In the second job, we compute a set of features for each location in the genome.
+In the \textsc{Compute Features} job, we compute a set of features for each location in the genome.
 To begin, we tile the genome with small fixed-width, non-overlapping intervals.
@@
 
-Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
+Finally, the third MapReduce job, \textsc{Call Variants} is responsible for making SV calls based on the features computed at each genomic location.
 It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
@@
 
-Cloudbreak is our implementation of the infrastructure for the framework described above.
-It demonstrates this framework by implementing a detection algorithm for small-to-medium sized deletions (40bp-25,000bp) and small insertions based on examining the insert sizes of paired end mappings.
+Cloudbreak is our implementation of the infrastructure for the framework described above, as well as an implemention of a detection algorithm for genomic deletions (40bp-25,000bp) and small insertions based on examining the insert sizes of paired end mappings.
+An illustration of the Cloudbreak algorithm working on a simple example is shown in Figure A2.
 In particular, we implement the three user-defined functions specified above as follows:
@@
 Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components.
-Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
 The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
@@
 \end{description}
-
-An illustration of the Cloudbreak algorithm working on a simple example is shown in Figure A2.
 
@@
 See Methods for details on how reads were aligned and each program was invoked.
-For all alignments we used BWA \cite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Figure A3).
+For all alignments we used BWA, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Figure A3).
 
@@
 Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
-We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
+Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
+Therefore, we initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
 In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
@@
 Simulated reads were aligned to hg18 chromosome 2, and NA18507 reads were aligned to the hg19 assembly.
-Alignments for all programs, unless otherwise noted, were aligned using BWA \cite{Li:2009p836} \texttt{aln} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
+Alignments for all programs, unless otherwise noted, were aligned using BWA \texttt{aln} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
 GASVPro also accepts ambiguous mappings; we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \texttt{-a -r -Ex 1100 -t 250}.
@@
   \begin{algorithmic}[1]
-    \Procedure{Alignment}{}
+    \Procedure{Align Reads}{}
     \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
@@
 
-    \Procedure{Compute SV Features}{}
+    \Procedure{Compute Features}{}
     \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
@@
 
-    \Procedure{Call SVs}{}
+    \Procedure{Call Variants}{}
     \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}

### 2013-08-01 manuscript/bmc_additional1.tex
moved algorithm example figure to main text

@@
 \label{insert_size_mixes}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=1\textwidth]{../figures/cloudbreak_mapred_diagram.pdf}
-\caption{An example of the Cloudbreak MapReduce algorithm.
-A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in either paired-end or single-ended mode to generate read mappings.
-Reducers gather all alignments for both reads in each pair.
-B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) under keys indicating the genomic location spanned by that pair.
-Only one genomic location is diagrammed here for simplicity.
-Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes.
-C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.}
-\label{algorithm_example}
 \end{figure}

### 2013-08-01 manuscript/bmc_article.tex
moved algorithm example figure to main text

@@
 
-\subsection*{A general framework for SV Detection in MapReduce}
+\subsection*{A Framework for SV Detection in MapReduce}
 
@@
 Cloudbreak is our implementation of the infrastructure for the framework described above, as well as an implemention of a detection algorithm for genomic deletions (40bp-25,000bp) and small insertions based on examining the insert sizes of paired end mappings.
-An illustration of the Cloudbreak algorithm working on a simple example is shown in Figure A2.
+An illustration of the Cloudbreak algorithm working on a simple example is shown in Figure 2.
 In particular, we implement the three user-defined functions specified above as follows:
@@
 
-Figure 2 shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment.
+Figure 3 shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment.
 See Methods for a description of how we identified correct predictions.
@@
 
-Figure 3 shows the performance of each algorithm at detecting events larger than 40bp on the NA18507 data set.
+Figure 4 shows the performance of each algorithm at detecting events larger than 40bp on the NA18507 data set.
 All algorithms show far less specificity for the gold standard set than they did in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample.
@@
 
-  \subsection*{Figure 2 - Accuracy and Runtime on a Simulated Data Set.}
+  \subsection*{Figure 2 - An Example of the Cloudbreak MapReduce Algorithm}
+An Example of the Cloudbreak MapReduce Algorithm.
+A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in either paired-end or single-ended mode to generate read mappings.
+Reducers gather all alignments for both reads in each pair.
+B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) under keys indicating the genomic location spanned by that pair.
+Only one genomic location is diagrammed here for simplicity.
+Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes.
+C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.
+
+  \subsection*{Figure 3 - Accuracy and Runtime on a Simulated Data Set.}
 Accuracy and runtime performance on a simulated data set.
@@
 
-\subsection*{Figure 3 - Accuracy and Runtime on the NA18507 Sample.}
+\subsection*{Figure 4 - Accuracy and Runtime on the NA18507 Sample.}
 Accuracy and performance on the 37X NA18507 sample.

### 2013-08-06 manuscript/bmc_article.tex
my updates

@@
 %Review style settings
-%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
 
 %Publication style settings
-\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
 
@@
 \begin{abstract}
-The detection of genomic structural variations (SV) remains a difficult challenge in analyzing sequencing data, and the growing size and number of sequenced genomes have rendered SV detection a bona fide big data problem.
+The detection of genomic structural variations (SV) remains a difficult challenge in analyzing short-read sequencing data, and the growing size and number of sequenced genomes have rendered SV detection a bona fide ``big data'' problem.
 MapReduce is a proven, scalable solution for distributed computing on huge data sets.
@@
 It provides dramatically shorter runtimes and the ability to scale to big data volumes on large compute clusters.
-Cloudbreak includes tools to set up and configure MapReduce (Hadoop) clusters on cloud services, enabling on-demand cluster computing.
+Cloudbreak includes tools to automatically set up MapReduce (Hadoop) clusters on cloud services, enabling on-demand cluster computing.
 Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.
+
+\medskip
+\noindent\textbf{Keywords:} genomic structural variation; distributed computing; copy number variation; high-throughput sequencing; genotyping.
 \end{abstract}
@@
 
-Genomic structural variations (SVs) such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \cite{Mills:2011p1611, Conrad:2010ja}.
-However, detection of SVs with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \cite{Mills:2011p1611}.
+Genomic structural variations (SVs), such as deletions, insertions, and inversions of DNA, are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \cite{Mills:2011p1611, Conrad:2010ja}.
+However, detection of SVs with current high-throughput sequencing technology remains a difficult challenge, with limited concordance between available algorithms and high false discovery rates \cite{Mills:2011p1611}.
 Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, and integrating them to form an accurate detection measure is computationally difficult.
 As the volume of massively parallel sequencing data approaches ``big data'' scales, SV detection is becoming a time consuming component of genomics pipelines, and presents a significant challenge for research groups and clinical operations that may not be able to scale their computational infrastructure.
-Here we present a distributed software solution that is scalable and readily available on the cloud.
-
+In this paper we present a distributed software solution that is scalable and readily available on the cloud.
+\par
 In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data.
 In particular, Google's MapReduce \cite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large data sets across clusters of commodity servers.
-Hadoop is an open source project of the Apache Foundation which provides an implementation of MapReduce.
+A widely adopted open-source implementation of MapReduce is provided by Hadoop, a project of the Apache Software Foundation.
 MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention.
+Hadoop also provides a distributed file system (HDFS) for managing the redundant storage of large data sets across a cluster.
 The Hadoop framework has been shown to be effective in sequencing-related tasks including short read mapping \cite{Schatz:2009p278}, SNP calling \cite{Langmead:2009p1225}, and RNA-seq differential expression analysis \cite{Langmead:2010p1268}.
@@
 
-Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets\cite{Alkan:2011p547}.
+% todo: keep this line?
+To design MapReduce SV detection algorithms we must consider the types of variant signatures present in sequencing data sets.
+Popular SV detection algorithms use three main signals \cite{Alkan:2011p547}.
 Read-pair (RP) based methods use the distance between and orientation of the mappings of paired reads to identify the signatures of SVs \cite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
@@
 Our goal is to leverage the strengths of the MapReduce computational framework in order to provide fast, accurate and readily scalable SV-detection pipelines.
-The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches.
+The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce.
+In particular, the global clustering of paired end mappings at the heart of many RP approaches is hard to translate into MapReduce.
 MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
@@
 We characterize our algorithm's performance on simulated and real data sets and compare its performance to those of several popular methods.
-
 Finally, and quite importantly from a practical point of view, our implementation of Cloudbreak provides the functionality to easily set up and configure Hadoop clusters on cloud service providers, making dynamically scalable distributed SV detection accessible to all whose computational needs demand it.
@@
 
-The \textsc{Align Reads} job uses sensitive mapping tools and settings to discover mapping locations for each read pair.
+The \textsc{Align Reads} job uses existing alignment tools to discover mapping locations for each read pair.
 Aligners can be executed to report multiple possible mappings for each read, or only the best possible mapping.
@@
 Our implementation of the framework contains wrappers to execute the aligners BWA \cite{Li:2009p836}, GEM \cite{MarcoSola:2012hm}, Novoalign \cite{novoalign}, RazerS 3 \cite{Weese:2012by}, mrFAST \cite{Alkan:2009cr}, and Bowtie 2 \cite{Langmead:2012jh}.
-This job can also be skipped in favor of importing a pre-aligned BAM file directly to HDFS.
+This job can also be skipped in favor of importing a pre-aligned BAM file directly into HDFS.
 
@@
 The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
-We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
 The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
 
-Finally, the third MapReduce job, \textsc{Call Variants} is responsible for making SV calls based on the features computed at each genomic location.
+Finally, the third MapReduce job, \textsc{Call Variants}, is responsible for making SV calls based on the features computed at each genomic location.
 It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
@@
 
-Cloudbreak is our implementation of the infrastructure for the framework described above, as well as an implemention of a detection algorithm for genomic deletions (40bp-25,000bp) and small insertions based on examining the insert sizes of paired end mappings.
+Cloudbreak is our implementation, in the framework described above, of a detection algorithm for genomic deletions (40bp-25,000bp) and small insertions based on examining the insert sizes of paired end mappings.
 An illustration of the Cloudbreak algorithm working on a simple example is shown in Figure 2.
@@
 For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
-\item[$\Phi$] To compute features for each genomic location, we follow \cite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Figure A1).
+In addition, if there are multiple possible mappings for each read in the input set, we optimize this step by assuming that if there exists a concordant mapping for a read pair, defined as a mapping pair in which the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, it is likely to be correct and therefore we do not consider any discordant alignments of the pair.
+\item[$\Phi$] To compute features for each genomic location, we follow Lee et al.
+\cite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Figure A1).
 Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
@@
 See Methods for more details on the GMM fitting procedure.
-Prior to fiting the GMM at each location, we attempt to filter our incorrect mappings for that location using an outlier-detection based clustering scheme and an adaptive mapping quality cutoff (see Methods for details).
+Prior to fitting the GMM at each location, we attempt to filter our incorrect mappings for that location using an outlier-detection based clustering scheme and an adaptive mapping quality cutoff (see Methods for details).
 
@@
 On the other hand, any simulated data should contain realistic SVs that follow patterns observed in real data.
-Therefore, we took one of the most complete lists of SVs from an individual, the list of homozygous insertions and deletions from the genome of J.
+To simulate a realistic set of variants, we took one of the most complete lists of SVs from an individual, the list of homozygous insertions and deletions from the genome of J.
 Craig Venter \cite{Levy:2007fb}, and used it to simulate a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants, with 100bp reads and a mean fragment size of 300bp (See Methods).
@@
 Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes.
-(Of course, Cloudbreak uses many more CPUs as distributed algorithm.
-See Additional File 1 and Table A1 for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster.
+(Of course, Cloudbreak uses many more CPUs as a distributed algorithm.
+See Additional File 1 and Table A1 for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster so it is not included in the runtime figure.
 Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow.
-To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Figure A4).
+To test its scalability based on the number of compute nodes available, we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Figure A4).
 
@@
 Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by RP mapping.
+
 We also used the ROC curves to characterize algorithm performance when a low false discovery rate is required.
-Table 3 shows the total number of deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
+Table 3 shows the total number of deletions found by each tool when choosing a threshold that gives a false discovery rate (FDR) closest to 10\% based on the ROC curve.
 At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool.
@@
 To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same individual.
-These included data from the Human Genome Structural Variation Project reported by \cite{Kidd:2008p926}, a survey of small indels conducted by \cite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \cite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
+These included data from the Human Genome Structural Variation Project reported by Kidd et al.
+\cite{Kidd:2008p926}, a survey of small indels conducted by Mills et al.
+\cite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \cite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
 We merged any overlapping calls of the same type into the region spanned by their unions.
-We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.
+We were unable to run MoDIL on the whole-genome data set due to its estimated runtime and storage requirements.
 
 Figure 4 shows the performance of each algorithm at detecting events larger than 40bp on the NA18507 data set.
-All algorithms show far less specificity for the gold standard set than they did in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample.
+All algorithms show far less specificity for the gold standard set than they did in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and aligning reads to a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample.
 For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
@@
 
-\subsection*{Discussion}
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Conclusions}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Conclusions}
+On simulated and real data sets, our approach exhibits high accuracy when compared to popular SV detection algorithms that run on traditional clusters and servers.
+Detection of insertions and deletions is an important area of research; Mills et al.
+\cite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
+More SV annotations are also required to assess the impact of SVs on non-coding elements \cite{Mu:2011br}.
 
-On simulated and real data sets, our approach exhibits high accuracy when compared to popular SV detection algorithms that run on traditional clusters and servers.
-Detection of insertions and deletions is an important area of research; \cite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
-More SV annotations are also required to assess the impact of SVs on non-coding elements \cite{Mu:2011br}.In addition to delivering state-of-the-art performance in a RP-based SV detection tool, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
+In addition to delivering state-of-the-art performance in a RP-based SV detection tool, our approach offers a basis for developing a variety of SV algorithms that are capable of running in a MapReduce pipeline with the power to process vast amounts of data in a cloud or commodity server setting.
 With the advent of cloud service providers such as Amazon EC2, it is becoming easy to instantiate on-demand Hadoop compute clusters.
@@
 We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4.
-We use snappy compression for MapReduce data.
-Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
-To execute the other tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine (\url{http://research.cs.wisc.edu/htcondor/}) with directed acyclic graphs to describe dependencies between jobs.
+We use snappy compression for MapReduce data, and Hadoop's distributed cache mechanism to share the executable files and indices needed for mapping tasks to the nodes in the cluster.
 
@@
 Assuming that our pipeline is capturing all relevant read mappings near the locus of the variant, the genotype should be indicated by the estimated parameter $\alpha$, the mixing parameter that controls the weight of the two components in the GMM.
-We setting a simple cutoff of .35 on the average value of $\alpha$ for each prediction to call the predicted variant homozygous or heterozygous.
-We used the same cutoff for deletion and insertion predictions.
+We set a simple cutoff of .35 on the average value of $\alpha$ for each prediction to call the predicted variant homozygous or heterozygous, and used the same cutoff for deletion and insertion predictions.
 
@@
 Based on this genotype, we applied each variant to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
-We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package (\url{http://sourceforge.net/apps/mediawiki/dnaa/}).
+We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package \cite{DNAA}.
 We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence.
@@
 We executed MoDIL with default parameters except for a \texttt{MAX\_DEL\_SIZE} of 25000, and processed it in parallel on our cluster with a step size of 121475.
+To execute parallelize other SV detection tools we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine \cite{condor} with directed acyclic graphs to describe dependencies between jobs.
 
@@
 SVs to validate were selected based on calls made by BreakDancer, relying on the score and the number of the reads supporting each single event.
-Appropriate primers were designed with an internet-based interface, Primer3 (\url{http://frodo.wi.mit.edu/}), considering the chromosome localization and orientation of the interval involved in the candidate rearrangement.
-The primers were checked for specificity using the BLAT tool of the UCSC Human Genome Browser (\url{http://genome.ucsc.edu/cgi-bin/hgBlat}).
+Appropriate primers were designed with an internet-based interface, Primer3 \cite{primer3}, considering the chromosome localization and orientation of the interval involved in the candidate rearrangement.
+The primers were checked for specificity using the BLAT tool of the UCSC Human Genome Browser \cite{BLAT}.
 All the primer pairs were preliminarily tested on the patient genomic DNA and a normal genomic DNA as control.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{List of Abbreviations}
+
+SV: Structural Variation; RP: read-pair; RD: read-depth; SR: split-read; SNV: single nucleotide variations; GMM: Gaussian Mixture Model; HDFS: Hadoop File System; ROC: Receiver Operating Characteristic; FDR: False Discovery Rate; AML: acute myeloid leukemia.
+
+\section*{Competing Interests}
+
+The authors declare that they have no competing interests.
+
 \section*{Authors contributions}
 
-C.W.W.
-and K.S.
-designed the algorithmic approaches.
-C.W.W.
-implemented the software and conducted the evaluations.
-J.T.
-obtained the AML sample and sorted the cells.
-L.C.
-sequenced the AML sample.
-C.T.S and A.L.A.
-performed PCR validations on the AML sample.
-C.W.W.
-and K.S.
-wrote the manuscript.
+CWW and KS designed the algorithmic approaches.
+CWW implemented the software and conducted the evaluations.
+JT obtained the AML sample and sorted the cells.
+LC sequenced the AML sample.
+CTS and ALA performed PCR validations on the AML sample.
+CWW and KS wrote the manuscript.
 
@@
 
-We would like to thank Izhak Shafran at the Center for Spoken Language Understanding for advice, support, and shared computational resources, and Bob Handsaker for helpful conversations.
+We would like to thank Izhak Shafran at the Center for Spoken Language Understanding for advice, support, and shared computational resources.
 We also thank reviewers for the Third Annual RECOMB Satellite Workshop on Massively Parallel Sequencing (RECOMB-seq) for their comments on a preliminary description of Cloudbreak.
@@
 Reads are first uploaded to a Hadoop cluster from local storage.
-Cloudbreak then executes three MapReduce jobs to process the data: 1) Mapping with sensitive settings.
+Cloudbreak then executes three MapReduce jobs to process the data: 1) Read Alignment.
 2) Computation of features across the genome.
-3) Calling structural variations based on the features computed in the previous step.
+3) Calling SVs based on the features computed in the previous step.
 Finally, SV predictions can be downloaded from the Hadoop cluster and examined.
@@
 Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes.
-C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.
+C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a variant.
 
@@
 (b) ROC curves for insertion predictions.
-(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Additional File 1).
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Additional File 1 for a discussion of parallelization).
 
@@
 (b) ROC curve for insertion prediction performance.
-(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Additional File 1).
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Additional File 1 for a discussion of parallelization).
 
@@
 
-  \subsection*{Table 1 - A General MapReduce Algoritm for SV Detection}
+  \subsection*{Table 1 - A General MapReduce Algorithm for SV Detection}
 \algrenewcommand\algorithmicprocedure{\textbf{job}}
@@
 Exclusive predictions are in parentheses.
-For deletions, the same cutoffs were used as for the simulated data as in Table 2, adjusted for the difference in coverage from 30X to 37X.
+For deletions, the same cutoffs were used as for the simulated data as in Table 3, adjusted for the difference in coverage from 30X to 37X.
 For insertions, the maximum sensitivity cutoff was used.
@@
   \subsection*{Additional file 1 --- Appendix}
-  An appendix containing additional figures, tables, and a discussion of techniques for parallelizing exisisting tools and measuring runtime performance.
+  An appendix containing additional figures, tables, a discussion of techniques for parallelizing existing tools and measuring runtime performance, and the user manual for Cloudbreak.
 

### 2013-08-06 manuscript/bmc_article.tex
edits from kemal

@@
 In this paper we present a distributed software solution that is scalable and readily available on the cloud.
-\par
+
 In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data.
@@
 A widely adopted open-source implementation of MapReduce is provided by Hadoop, a project of the Apache Software Foundation.
-MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention.
+MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible to the data they require, minimizing network traffic and I/O contention.
 Hadoop also provides a distributed file system (HDFS) for managing the redundant storage of large data sets across a cluster.
@@
 
-Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing problems like SV detection.
+Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing data processing problems such as SV detection.
 MapReduce divides computation across a cluster into three phases.
@@
 % todo: keep this line?
-To design MapReduce SV detection algorithms we must consider the types of variant signatures present in sequencing data sets.
-Popular SV detection algorithms use three main signals \cite{Alkan:2011p547}.
+To design MapReduce SV detection algorithms, we must consider the types of variant signatures present in sequencing data sets.
+Popular SV detection algorithms use three main types of signals \cite{Alkan:2011p547}.
 Read-pair (RP) based methods use the distance between and orientation of the mappings of paired reads to identify the signatures of SVs \cite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}.
@@
 For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
-the fragment size implied by the alignments and the likelihood the alignments are correct.
+the fragment size implied by the alignments and the likelihood that the alignments are correct.
 We then leave two functions to be implemented depending on the application:
@@
 
-SV: Structural Variation; RP: read-pair; RD: read-depth; SR: split-read; SNV: single nucleotide variations; GMM: Gaussian Mixture Model; HDFS: Hadoop File System; ROC: Receiver Operating Characteristic; FDR: False Discovery Rate; AML: acute myeloid leukemia.
+SV: Structural Variation; RP: read-pair; RD: read-depth; SR: split-read; SNV: single nucleotide variation; GMM: Gaussian Mixture Model; HDFS: Hadoop File System; ROC: Receiver Operating Characteristic; FDR: False Discovery Rate; AML: acute myeloid leukemia.
 
@@
 
-  \subsection*{Table 2 - Variants detected in the simulated data set}
+  \subsection*{Table 2 - Variants Detected in the Simulated Data Set}
 The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses).
@@
 
-The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class found.
-Exclusive predictions are in parentheses.
+The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class.
+Predictions exclusive to the particular algorithm are in parentheses.
 For deletions, the same cutoffs were used as for the simulated data as in Table 3, adjusted for the difference in coverage from 30X to 37X.

### 2013-12-13 src/main/scripts/manuscript.R
updates to figure plotting, etc

@@ -189,6 +189,13 @@ cloudbreakGEMChr2Perf5000ReportPerfFile <- '~/Documents/gene_rearrange/svpipelin
 cloudbreakGEMChr2Perf50000ReportPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_e6_s2_del_d50000.perf.txt'
 cloudbreakGEMChr2Perf100000ReportPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_e6_s2_del_d100000.perf.txt'
 
+cloudbreakRes1PerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/res_test_1_del.perf.txt'
+cloudbreakRes5PerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/res_test_5_del.perf.txt'
+cloudbreakRes10PerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/res_test_10_del.perf.txt'
+cloudbreakRes25PerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/res_test_25_del.perf.txt'
+cloudbreakRes50PerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/res_test_50_del.perf.txt'
+cloudbreakRes100PerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/res_test_100_del.perf.txt'
+
 # insertions
 cloudbreakGEMChr2InsertionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/gem_e6_s2_ins.perf.txt'
 cloudbreakBWAChr2InsertionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_chr2_allindels_100bp_dip/bwa_ins.perf.txt'
@@ -317,7 +324,7 @@ pindelNA18507InsertionHitsFile <- '~/Documents/gene_rearrange/svpipeline/NA18507
 cloudbreakBWATweakNA18507InsertionHitsFile <- '~/Documents/gene_rearrange/svpipeline/NA18507/bwa_instweak_calls_sparseve_gt1p16.hits.txt'
 
 # cloud crf
[…]

### 2013-12-13 manuscript/NA18507DeletionAndInsertionPredsBySize.table.brew.tex
updates to table to add F1 measures and window size effect on accuracy

--- 2013-05-01 manuscript/NA18507DeletionAndInsertionPredsBySize.table.brew.tex
@@
 \begin{center}
-\begin{tabular}{r|rrr|rrrrr}
-  \cline{2-9}
-&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\begin{tabular}{r|rrrr|rrrrr}
+  \cline{2-10}
+&  & Prec. & Recall & F1 & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 \hline
-\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & <%= format(totalDels['40-100bp'], big.mark=",") %> & <%= format(totalDels['101-250bp'], big.mark=",") %> & <%= format(totalDels['251-500bp'], big.mark=",") %> & <%= format(totalDels['501-1000bp'], big.mark=",") %> & <%= format(totalDels['> 1000bp'], big.mark=",") %> \\
+\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & & <%= format(totalDels['40-100bp'], big.mark=",") %> & <%= format(totalDels['101-250bp'], big.mark=",") %> & <%= format(totalDels['251-500bp'], big.mark=",") %> & <%= format(totalDels['501-1000bp'], big.mark=",") %> & <%= format(totalDels['> 1000bp'], big.mark=",") %> \\
   \hline
-\cline{2-9}
-& Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', delPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '> 1000bp') %> \\
-& Breakdancer & <%= printPredTableValueFromList('Breakdancer', delPrecision) %> & <%= printPredTableValueFromList('Breakdancer', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '> 1000bp') %> \\
-&  GASVPro & <%= printPredTableValueFromList('GASVPro', delPrecision) %> & <%= printPredTableValueFromList('GASVPro',delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '> 1000bp') %> \\
-&  DELLY-RP & <%= printPredTableValueFromList('DELLY', delPrecision) %> & <%= printPredTableValueFromList('DELLY', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '> 1000bp') %> \\
-&  DELLY-SR & <%= printPredTableValueFromList('DELLYSR', delPrecision) %> & <%= printPredTableValueFromList('DELLYSR', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '> 1000bp') %> \\
-&  Pindel & <%= printPredTableValueFromList('Pindel', delPrecision) %> & <%= printPredTableValueFromList('Pindel', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '> 1000bp') %> \\
+\cline{2-10}
+& Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', delPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', delRecall) %> & <%= printPredTableValueFromList('Cloudbreak', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '> 1000bp') %> \\
+& Breakdancer & <%= printPredTableValueFromList('Breakdancer', delPrecision) %> & <%= printPredTableValueFromList('Breakdancer', delRecall) %> &  <%= printPredTableValueFromList('Breakdancer', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '> 1000bp') %> \\
+&  GASVPro & <%= printPredTableValueFromList('GASVPro', delPrecision) %> & <%= printPredTableValueFromList('GASVPro',delRecall) %> &  <%= printPredTableValueFromList('GASVPro',delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '> 1000bp') %> \\
+&  DELLY-RP & <%= printPredTableValueFromList('DELLY', delPrecision) %> & <%= printPredTableValueFromList('DELLY', delRecall) %> &  <%= printPredTableValueFromList('DELLY', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '> 1000bp') %> \\
+&  DELLY-SR & <%= printPredTableValueFromList('DELLYSR', delPrecision) %> & <%= printPredTableValueFromList('DELLYSR', delRecall) %> &  <%= printPredTableValueFromList('DELLYSR', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '> 1000bp') %> \\
+&  Pindel & <%= printPredTableValueFromList('Pindel', delPrecision) %> & <%= printPredTableValueFromList('Pindel', delRecall) %> &  <%= printPredTableValueFromList('Pindel', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '> 1000bp') %> \\
 \hline
-\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & <%= format(totalIns['40-100bp'], big.mark=",") %> & <%= format(totalIns['101-250bp'], big.mark=",") %> & <%= format(totalIns['251-500bp'], big.mark=",") %> & <%= format(totalIns['501-1000bp'], big.mark=",") %> \\
-\cline{2-9}
-& Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', insPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '501-1000bp') %> \\
-& Breakdancer & <%= printPredTableValueFromList('Breakdancer', insPrecision) %> & <%= printPredTableValueFromList('Breakdancer', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '501-1000bp') %> \\
-&  Pindel & <%= printPredTableValueFromList('Pindel', insPrecision) %> & <%= printPredTableValueFromList('Pindel', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '501-1000bp') %>  \\
+\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & & <%= format(totalIns['40-100bp'], big.mark=",") %> & <%= format(totalIns['101-250bp'], big.mark=",") %> & <%= format(totalIns['251-500bp'], big.mark=",") %> & <%= format(totalIns['501-1000bp'], big.mark=",") %> & 0 \\
+\cline{2-10}
+& Cloudbreak & <%= printPredTableValueFromList('Cloudbreak', insPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', insRecall) %> &  <%= printPredTableValueFromList('Cloudbreak', insf1) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '501-1000bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '> 1000bp') %>  \\
+& Breakdancer & <%= printPredTableValueFromList('Breakdancer', insPrecision) %> & <%= printPredTableValueFromList('Breakdancer', insRecall) %> &  <%= printPredTableValueFromList('Breakdancer', insf1) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '501-1000bp')  %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '> 1000bp') %> \\
+&  Pindel & <%= printPredTableValueFromList('Pindel', insPrecision) %> & <%= printPredTableValueFromList('Pindel', insRecall) %> &  <%= printPredTableValueFromList('Pindel', insf1) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '501-1000bp')  %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '> 1000bp')  %> \\
 \hline

### 2013-12-13 manuscript/chr2AccuracyByWindowSize.table.brew.tex
updates to table to add F1 measures and window size effect on accuracy

\begin{table}
\begin{center}
\begin{tabular}{r|rrrrr}
 \hline
 Window Size & Calls & True Positives & Precision & Recall & F \\ 
 \hline
   <%= formatNumber(windowTable[1,'Resolution']) %> & <%= formatNumber(windowTable[1,'Calls']) %> & <%= formatNumber(windowTable[1,'TP']) %> & <%= formatNumber(windowTable[1,'Precision']) %> & <%= formatNumber(windowTable[1,'Recall']) %> & <%= formatNumber(windowTable[1,'F']) %> \\ 
   <%= formatNumber(windowTable[2,'Resolution']) %> & <%= formatNumber(windowTable[2,'Calls']) %> & <%= formatNumber(windowTable[2,'TP']) %> & <%= formatNumber(windowTable[2,'Precision']) %> & <%= formatNumber(windowTable[2,'Recall']) %> & <%= formatNumber(windowTable[2,'F']) %> \\ 
   <%= formatNumber(windowTable[3,'Resolution']) %> & <%= formatNumber(windowTable[3,'Calls']) %> & <%= formatNumber(windowTable[3,'TP']) %> & <%= formatNumber(windowTable[3,'Precision']) %> & <%= formatNumber(windowTable[3,'Recall']) %> & <%= formatNumber(windowTable[3,'F']) %> \\ 
   <%= formatNumber(windowTable[4,'Resolution']) %> & <%= formatNumber(windowTable[4,'Calls']) %> & <%= formatNumber(windowTable[4,'TP']) %> & <%= formatNumber(windowTable[4,'Precision']) %> & <%= formatNumber(windowTable[4,'Recall']) %> & <%= formatNumber(windowTable[4,'F']) %> \\ 
   <%= formatNumber(windowTable[5,'Resolution']) %> & <%= formatNumber(windowTable[5,'Calls']) %> & <%= formatNumber(windowTable[5,'TP']) %> & <%= formatNumber(windowTable[5,'Precision']) %> & <%= formatNumber(windowTable[5,'Recall']) %> & <%= formatNumber(windowTable[5,'F']) %> \\ 
   <%= formatNumber(windowTable[6,'Resolution']) %> & <%= formatNumber(windowTable[6,'Calls']) %> & <%= formatNumber(windowTable[6,'TP']) %> & <%= formatNumber(windowTable[6,'Precision']) %> & <%= formatNumber(windowTable[6,'Recall']) %> & <%= formatNumber(windowTable[6,'F']) %> \\  
\end{tabular}
\end{center}
\caption{Accuracy measures for Cloudbreak on the Chromosome 2 simulated data set using different choices of window size. For each window size the same Cloudbreak score threshold was used (1.98).}
\label{chr2AccuracyByWindowSize}
\end{table}

### 2013-12-13 manuscript/chr2DeletionAndInsertionPredsMaxSensitivity.table.brew.tex
updates to table to add F1 measures and window size effect on accuracy

@@
 \begin{center}
-\begin{tabular}{r|rrr|rrrrr}
-  \cline{2-9}
-   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\begin{tabular}{r|rrrr|rrrrr}
+  \cline{2-10}
+   &                     & Prec. & Recall & F1 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 \hline
-\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & <%= totalDels['40-100bp'] %> &  <%= totalDels['101-250bp'] %> & <%= totalDels['251-500bp'] %> &  <%= totalDels['501-1000bp'] %> & <%= totalDels['> 1000bp'] %>\\
+\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & &  <%= totalDels['40-100bp'] %> &  <%= totalDels['101-250bp'] %> & <%= totalDels['251-500bp'] %> &  <%= totalDels['501-1000bp'] %> & <%= totalDels['> 1000bp'] %>\\
   \hline
-\cline{2-9}
-&  Cloudbreak    &  <%= printPredTableValueFromList('Cloudbreak', delPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '> 1000bp') %> \\
-&  Breakdancer   &  <%= printPredTableValueFromList('Breakdancer', delPrecision) %> & <%= printPredTableValueFromList('Breakdancer', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '> 1000bp') %> \\
-&  GASVPro        & <%= printPredTableValueFromList('GASVPro', delPrecision) %> & <%= printPredTableValueFromList('GASVPro', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '> 1000bp') %> \\
-&  DELLY-RP           & <%= printPredTableValueFromList('DELLY', delPrecision) %> & <%= printPredTableValueFromList('DELLY', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '> 1000bp') %> \\
-&  DELLY-SR           & <%= printPredTableValueFromList('DELLYSR', delPrecision) %> & <%= printPredTableValueFromList('DELLYSR', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '> 1000bp') %> \\
-&  Pindel           & <%= printPredTableValueFromList('Pindel', delPrecision) %> & <%= printPredTableValueFromList('Pindel', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '> 1000bp') %>\\
-&  MoDIL           & <%= printPredTableValueFromList('MoDIL', delPrecision) %>  & <%= printPredTableValueFromList('MoDIL', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '> 1000bp') %>\\
+\cline{2-10}
+&  Cloudbreak    &  <%= printPredTableValueFromList('Cloudbreak', delPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', delRecall) %> & <%= printPredTableValueFromList('Cloudbreak', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Cloudbreak', '> 1000bp') %> \\
+&  Breakdancer   &  <%= printPredTableValueFromList('Breakdancer', delPrecision) %> & <%= printPredTableValueFromList('Breakdancer', delRecall) %> & <%= printPredTableValueFromList('Breakdancer', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Breakdancer', '> 1000bp') %> \\
+&  GASVPro        & <%= printPredTableValueFromList('GASVPro', delPrecision) %> & <%= printPredTableValueFromList('GASVPro', delRecall) %> & <%= printPredTableValueFromList('GASVPro', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'GASVPro', '> 1000bp') %> \\
+&  DELLY-RP           & <%= printPredTableValueFromList('DELLY', delPrecision) %> & <%= printPredTableValueFromList('DELLY', delRecall) %> & <%= printPredTableValueFromList('DELLY', delRecall) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-RP', '> 1000bp') %> \\
+&  DELLY-SR           & <%= printPredTableValueFromList('DELLYSR', delPrecision) %> & <%= printPredTableValueFromList('DELLYSR', delRecall) %> &  <%= printPredTableValueFromList('DELLYSR', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'DELLY-SR', '> 1000bp') %> \\
+&  Pindel           & <%= printPredTableValueFromList('Pindel', delPrecision) %> & <%= printPredTableValueFromList('Pindel', delRecall) %> &  <%= printPredTableValueFromList('Pindel', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'Pindel', '> 1000bp') %>\\
+&  MoDIL           & <%= printPredTableValueFromList('MoDIL', delPrecision) %>  & <%= printPredTableValueFromList('MoDIL', delRecall) %> & <%= printPredTableValueFromList('MoDIL', delf1) %> & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(delPreds, exDelPreds, 'MoDIL', '> 1000bp') %>\\
    \hline
-\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & <%= totalIns['40-100bp'] %> &  <%= totalIns['101-250bp'] %> & <%= totalIns['251-500bp'] %> &  <%= totalIns['501-1000bp'] %> & <%= totalIns['> 1000bp'] %>\\
-\cline{2-9}
-&  Cloudbreak   &<%= printPredTableValueFromList('Cloudbreak', insPrecision) %> & <%= printPredTableValueFromList('Cloudbreak', insRecall) %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '> 1000bp') %> \\
-&  Breakdancer & <%= printPredTableValueFromList('Breakdancer', insPrecision) %> & <%= printPredTableValueFromList('Breakdancer', insRecall) %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '> 1000bp') %> \\
-&  Pindel          & <%= printPredTableValueFromList('Pindel', insPrecision) %> & <%= printPredTableValueFromList('Pindel', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '> 1000bp') %>\\
-&  MoDIL          & <%= printPredTableValueFromList('MoDIL', insPrecision) %> & <%= printPredTableValueFromList('MoDIL', insRecall) %> & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '> 1000bp') %>\\
+\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & & <%= totalIns['40-100bp'] %> &  <%= totalIns['101-250bp'] %> & <%= totalIns['251-500bp'] %> &  <%= totalIns['501-1000bp'] %> & <%= totalIns['> 1000bp'] %>\\
+\cline{2-10}
+&  Cloudbreak   &<%= printPredTableValueFromList('Cloudbreak', insPrecisionf) %> & <%= printPredTableValueFromList('Cloudbreak', insRecall) %>  & <%= printPredTableValueFromList('Cloudbreak', insf1) %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Cloudbreak', '> 1000bp') %> \\
+&  Breakdancer & <%= printPredTableValueFromList('Breakdancer', insPrecision) %> & <%= printPredTableValueFromList('Breakdancer', insRecall) %>  & <%= printPredTableValueFromList('Breakdancer', insf1) %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Breakdancer', '> 1000bp') %> \\
+&  Pindel          & <%= printPredTableValueFromList('Pindel', insPrecision) %> & <%= printPredTableValueFromList('Pindel', insRecall) %> &  <%= printPredTableValueFromList('Pindel', insf1) %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'Pindel', '> 1000bp') %>\\
+&  MoDIL          & <%= printPredTableValueFromList('MoDIL', insPrecision) %> & <%= printPredTableValueFromList('MoDIL', insRecall) %> &  <%= printPredTableValueFromList('MoDIL', insf1) %> & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(insPreds, exInsPreds, 'MoDIL', '> 1000bp') %>\\
 \hline

### 2013-12-13 manuscript/chr2DeletionPredsMaxSensitivity.table.brew.tex
updates to table to add F1 measures and window size effect on accuracy

--- 2013-03-04 manuscript/chr2DeletionPredsMaxSensitivity.table.brew.tex
@@
 \begin{center}
-\begin{tabular}{rrrrrr}
+\begin{tabular}{rrrr|rrrrr}
   \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
+                        & Precf. & Recall & F1 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+ Total Number &          &           & <%= totals['40-100bp'] %> &  <%= totals['101-250bp'] %> & <%= totals['251-500bp'] %> &  <%= totals['501-1000bp'] %> & <%= totals['> 1000bp'] %>\\ 
   \hline
-  Cloudbreak  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\
-  Breakdancer & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\
-  GASVPro     & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\
-  DELLY       & <%= printPredTableCell(preds, exPreds, 'DELLY', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY', '> 1000bp') %> \\
-  Pindel      & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\
-  MoDIL      & <%= printPredTableCell(preds, exPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '> 1000bp') %>\\
+  Cloudbreak    &  <%= printPredTableValueFromList('Cloudbreak', precision) %> & <%= printPredTableValueFromList('Cloudbreak', recall) %> & <%= printPredTableValueFromList('Cloudbreak', f1) %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Cloudbreak', '> 1000bp') %> \\
+  Breakdancer   &  <%= printPredTableValueFromList('Breakdancer', precision) %> & <%= printPredTableValueFromList('Breakdancer', recall) %> & & <%= printPredTableValueFromList('Breakdancer', f1) %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Breakdancer', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Breakdancer', '> 1000bp') %> \\
+  GASVPro        & <%= printPredTableValueFromList('GASVPro', precision) %> & <%= printPredTableValueFromList('GASVPro', recall) %> & <%= printPredTableValueFromList('GASVPro', f1) %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'GASVPro', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'GASVPro', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'GASVPro', '> 1000bp') %> \\
+  DELLY-RP           & <%= printPredTableValueFromList('DELLY', precision) %> & <%= printPredTableValueFromList('DELLY', recall) %> & & <%= printPredTableValueFromList('DELLY', f1) %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-RP', '> 1000bp') %> \\
+  DELLY-SR           & <%= printPredTableValueFromList('DELLYSR', precision) %> & <%= printPredTableValueFromList('DELLYSR', recall) %> & <%= printPredTableValueFromList('DELLYSR', f1) %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'DELLY-SR', '> 1000bp') %> \\
+  Pindel           & <%= printPredTableValueFromList('Pindel', precision) %> & <%= printPredTableValueFromList('Pindel', recall) %> & <%= printPredTableValueFromList('Pindel', f1) %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'Pindel', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'Pindel', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'Pindel', '> 1000bp') %>\\
+  MoDIL           & <%= printPredTableValueFromList('MoDIL', precision) %>  & <%= printPredTableValueFromList('MoDIL', recall) %> & <%= printPredTableValueFromList('MoDIL', f1) %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '40-100bp') %>  & <%= printPredTableCell(preds, exPreds, 'MoDIL', '101-250bp') %> &  <%= printPredTableCell(preds, exPreds, 'MoDIL', '251-500bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '501-1000bp') %> & <%= printPredTableCell(preds, exPreds, 'MoDIL', '> 1000bp') %>\\
    \hline

### 2014-01-13 manuscript/bioinfo/document.tex
draft

--- 2013-07-08 manuscript/nm_supp.tex
@@
-\documentclass[11pt]{article}
+\documentclass{bioinfo}
 
-\usepackage[margin=1in]{geometry}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
 \usepackage{url}
@@
 \usepackage{xr}
+\usepackage{amsfonts}
+\usepackage{amsmath}
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
 
-% \renewcommand{\thetable}{S\arabic{table}}   
-% \renewcommand{\thefigure}{S\arabic{figure}}
-% \renewcommand{\thealgorithm}{\arabic{algorithm}}
+\newcommand{\todo}[1]{}
 
-\usepackage[style=nature,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
-\renewbibmacro{in:}{%
-  \ifentrytype{article}{}{%
-  \printtext{\bibstring{in}\intitlepunct}}}
-\addbibresource{manuscript.bib}
+% added from http://tex.stackexchange.com/questions/78776/forced-indentation-in-algorithmicx to help with wrapping lines in algorithmic enviro
+\makeatletter
+\newlength{\continueindent}
+\setlength{\continueindent}{6em}
 
-\renewcommand{\figurename}{Supplementary Figure}
-\renewcommand{\tablename}{Supplementary Table}
+\renewenvironment{algorithmic}[1][0]%
+   {%
+   \edef\ALG@numberfreq{#1}%
+   \def\@currentlabel{\theALG@line}%
+   %
+   \setcounter{ALG@line}{0}%
+   \setcounter{ALG@rem}{0}%
+   %
+   \let\\\algbreak%
+   %
+   \expandafter\edef\csname ALG@currentblock@\theALG@nested\endcsname{0}%
+   \expandafter\let\csname ALG@currentlifetime@\theALG@nested\endcsname\relax%
+   %
+   \begin{list}%
+      {\ALG@step}%
+      {%
+      \rightmargin\z@%
+      \itemsep\z@ \itemindent\z@ \listparindent2em%
+      \partopsep\z@ \parskip\z@ \parsep\z@%
+      \labelsep 0.5em \topsep 0.2em%\skip 1.2em 
+      \ifthenelse{\equal{#1}{0}}%
+         {\labelwidth 0.5em}%
+         {\labelwidth 1.2em}%
+       \leftmargin\labelwidth \addtolength{\leftmargin}{\labelsep}
+      \ALG@tlm\z@%
+      }%
+      \parshape 2 \leftmargin \linewidth \continueindent \dimexpr\linewidth-\continueindent\relax
+   \setcounter{ALG@nested}{0}%
+   \ALG@beginalgorithmic%
+   }%
+   {% end{algorithmic}
+   % check if all blocks are closed
+   \ALG@closeloops%
+   \expandafter\ifnum\csname ALG@currentblock@\theALG@nested\endcsname=0\relax%
+   \else%
+      \PackageError{algorithmicx}{Some blocks are not closed!!!}{}%
+   \fi%
+   \ALG@endalgorithmic%
+   \end{list}%
+   }%
+\makeatother
 
-\newfloat{suppalgorithm}{p}{cap}
-\floatname{suppalgorithm}{Supplementary Algorithm}
+\externaldocument[S]{supplementary}
+
+\copyrightyear{2014}
+\pubyear{2014}
 
 \begin{document}
+\firstpage{1}
 
-\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\
-Supplementary Material}
+\title[Cloudbreak]{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
+\author[Whelan \textit{et~al}]{Christopher W.
+Whelan\,$^{1}$, Jeffrey Tyner\,$^{2}$, Alberto L'Abbate\,$^{3}$, Clelia Tiziana Storlazzi,$^{3}$, Lucia Carbone\,$^{4}$, and Kemal S\"onmez\,$^{1,5,*}$\footnote{to whom correspondence should be addressed}}
+\address{$^{1}$Institute on Development and Disability and Center for Spoken Language Understanding, Oregon Health \& Science University, Portland, OR, USA \\
+$^{2}$ Program in Molecular and Cellular Biosciences, Oregon Health \& Science University, Portland, OR, USA \\
+$^{3}$ Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy \\
+$^{4}$ Behavioral Neuroscience Department, Oregon Health \& Science University, Portland, OR, USA \\
+$^{5}$ Department of Medical Informatics \& Clinical Epidemiology, Oregon Health \& Science University, Portland, OR, USA 
+}
 
-\author[1,5]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
-\author[3,4]{Jeffrey Tyner}
-\author[6]{Alberto L'Abbate}
-\author[6]{Clelia Tiziana Storlazzi}
-\author[4,5]{Lucia Carbone}
-\author[1,2,5]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
-\affil[1]{Institute on Development and Disability and Center for Spoken Language Understanding}
-\affil[2]{Department of Medical Informatics \& Clinical Epidemiology}
-\affil[3]{Program in Molecular and Cellular Biosciences}
-\affil[4]{Behavioral Neuroscience Department}
-\affil[5]{Oregon Health \& Science University, Portland, OR, USA}
-\affil[6]{Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy}
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\editor{Associate Editor: XXXXXXX}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth]{../figures/insert_size_mixtures.pdf}
-\caption{Illustration of insert size mixtures at individual genomic locations.
-A) There is no variant present at the location indicated by the vertical line (left), so the mix of insert sizes (right) follows the expected distribution of the library centered at 200bp, with a small amount of noise coming from low-quality mappings.
-B) A homozygous deletion of 50bp at the location has shifted the distribution of observed insert sizes.
-C) A heterozygous deletion at the location causes a mixture of normal and long insert sizes to be detected.
-D) A heterozygous small insertion shifts a portion of the mixture to have lower insert sizes.}
-\label{insert_size_mixes}
-\end{figure}
+\begin{abstract}
 
-\clearpage
+\section{Motivation:}
+The detection of genomic structural variations (SV) remains a difficult challenge in analyzing sequencing data, and the growing size and number of sequenced genomes have rendered SV detection a bona fide big data problem.
+MapReduce is a proven, scalable solution for distributed computing on huge data sets.
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=1\textwidth]{../figures/cloudbreak_mapred_diagram.pdf}
-\caption{An example of the Cloudbreak MapReduce algorithm.
-A) In the first MapReduce job, mappers scan input reads in FASTQ format and execute an alignment program in either paired-end or single-ended mode to generate read mappings.
-Reducers gather all alignments for both reads in each pair.
-B) In the second MapReduce job, mappers first emit information about each read pair (in this case the insert size and quality) under keys indicating the genomic location spanned by that pair.
-Only one genomic location is diagrammed here for simplicity.
-Reducers then compute features for each location on the genome by fitting a GMM to the distribution of spanning insert sizes.
-C) Mappers group all emitted features by their chromosome, and reducers find contiguous blocks of features that indicate the presence of a deletion.}
-\label{algorithm_example}
-\end{figure}
+\section{Results:}
+We describe an algorithmic framework for SV detection algorithms in MapReduce based on computing local genomic features, and use it to develop a deletion and insertion detection algorithm, Cloudbreak.
+On simulated and real data sets, Cloudbreak achieves accuracy improvements over popular SV detection algorithms, and genotypes variants from diploid samples.
+It provides dramatically shorter runtimes and the ability to scale to big data volumes on large compute clusters.
+Cloudbreak includes tools to set up and configure MapReduce (Hadoop) clusters on cloud services, enabling on-demand cluster computing.
 
-\clearpage
+\section{Availability:}
+Source code, binary releases, example scripts, and a full user manual are available at \url{http://github.com/cwhelan/cloudbreak}
+
+\section{Contact:} \href{sonmezk@ohsu.edu}{sonmezk@ohsu.edu}
+\end{abstract}
+
+\section{Introduction}
+
+Genomic structural variations (SVs) such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes~\citep{Mills:2011p1611, Conrad:2010ja}.
+However, detection of SVs with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates~\citep{Mills:2011p1611}.
+Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, making the problem computationally intensive.
+As the volume of massively parallel sequencing data approaches ``big data'' scales, SV detection is becoming a time consuming component of genomics pipelines, and presents a significant challenge for research groups and clinical operations that may not be able to scale their computational infrastructure.
+Here we present a distributed software solution that is scalable and readily available on the cloud.
+
+In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data.
+In particular, Google's MapReduce~\citep{Dean:2008p277} was designed to manage the storage and efficient processing of big data across clusters of commodity servers.
+Hadoop is an open source project of the Apache Foundation that provides an implementation of MapReduce.
+MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention.
+The Hadoop framework has been successfully applied to sequencing-related tasks including short read mapping~\citep{Schatz:2009p278} and SNP calling~\citep{Langmead:2009p1225}\todo{, and RNA-seq differential expression analysis~\citep{Langmead:2010p1268}}.
+
+Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing problems like SV detection.
+MapReduce divides computation across a cluster into three phases.
+In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined.
+The framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key.
+Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.
+
+SV detection algorithms use three main signals present in high-throughput sequencing data sets~\citep{Alkan:2011p547}.
+Read-pair (RP) based methods use the distance between and orientation of the mappings of paired reads to identify the signatures of SVs~\citep{Campbell:2008p539,Chen:2009p3,Sindi:2009gu,Korbel:2009dy}.
+Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation, and then clustering the discordant mappings to find SVs supported by multiple discordantly mapped read pairs.
+Read-depth (RD) approaches consider the changing depth of coverage of concordantly mapped reads along the genome~\citep{Alkan:2009cr,Yoon:2009kb,Chiang:2009di}.
+Finally, split-read (SR) methods look for breakpoints by mapping portions of individual reads to different genomic locations~\citep{Wang:2011p1607,Ye:2009p2}.
+
+Many RP methods consider only unambiguously discordantly mapped read pairs.
+Some approaches also include ambiguous mappings of discordant read pairs to improve sensitivity in repetitive regions of the genome~\citep{Hormozdiari:2009p284,Quinlan:2010gf}.
+Several recent RP approaches have also considered concordant read pairs, either to integrate RD signals for improved accuracy~\citep{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds that separate concordant from discordant mappings and thus detect smaller events~\citep{Marschall:2012ek}.
+Increasing the number of mappings considered, however, increases the computational burden of SV detection.
+
+Our goal is to leverage the strengths of MapReduce to provide fast, accurate and readily scalable SV-detection pipelines.
+The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches.
+MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel.
+In sequencing, for example, MapReduce based SNV-callers Crossbow~\citep{Langmead:2009p1225} and GATK~\citep{McKenna:2010p1051} perform independent calculations on partitions of the genome.
+SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL~\citep{Lee:2009da} and forestSV~\citep{Michaelson:2012fj} compute features along the genome and then produce SV predictions from those features in a post-processing step.
+We will show that this strategy can be translated into the MapReduce architecture.
+
+In this paper, we describe a framework for solving SV detection problems in Hadoop based on the computation of local genomic features from paired end mappings.
+We have used this framework to develop a software package, Cloudbreak, that discovers genomic deletions up to 25,000bp long, and short insertions.
+Cloudbreak computes local features based on modeling the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM), an idea first implemented in MoDIL~\citep{Lee:2009da}.
+The use of Hadoop enables scalable implementations of this class of algorithm.
+We characterize our algorithm's performance on a simulated data set, as well as a high coverage real data set from a normal individual and a low coverage cancer data set, and compare its performance to those of several popular methods.
+
+Finally, and quite importantly from a practical point of view, our implementation of Cloudbreak provides the functionality to easily set up and configure Hadoop clusters on cloud service providers, making dynamically scalable distributed SV detection accessible to all whose computational needs demand it.
+
+% \begin{equation}
+% \sum x+ y =Z\label{eq:01}
+% \end{equation}
+
+\begin{methods}
+\section{Methods}
+
+\subsection{A Framework for SV Detection in MapReduce}
 
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
-\caption{Cloudbreak performance on the chromosome 2 simulation using different alignment strategies.
-ROC curves show the number of true positives and false positives for each operating point for deletions and insertions.
-The Cloudbreak alignment strategies are: 1) ``Cloudbreak'': Alignments generated with BWA in paired-end mode, reporting the best hit for each pair.
-2) ``Cloudbreak-BWA-PE-MM'': Alignments generated with BWA in paired-end mode, reporting up to 25 additional hits for each mapping in SAM format using the \texttt{-n} and \texttt{-N} parameters for \texttt{bwa sampe} and the script \texttt{xa2multi.pl}.
-3) ``Cloudbreak-GEM-SE-MM'': Alignments generated by running the GEM aligner in single-ended mode, reporting up to 1000 additional hits per alignment.
-GEM was executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta), with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0},  requesting all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read.
-Considering multiple mappings improves Cloudbreak's specificity for insertions but decreases sensitivity to deletions.}
-\label{alignment_comparison}
+\includegraphics[width=.5\textwidth]{../../figures/workflow_with_whirr.png}
+\caption{An overview of the Cloudbreak workflow.
+Reads are first uploaded to a Hadoop cluster from local storage.
+Cloudbreak then executes three MapReduce jobs to process the data: 1) Mapping with sensitive settings.
+2) Computation of features across the genome.
+3) Calling structural variations based on the features computed in the previous step.
+Finally, SV predictions can be downloaded from the Hadoop cluster and examined.
+Cloudbreak can also use the Apache Whirr library to automatically provision Hadoop clusters on and deploy data to cloud providers such as Amazon Elastic Compute Cloud.}
+\label{cloudbreak_workflow}
 \end{figure}
 
-\clearpage
+Our framework for SV detection in MapReduce divides processing into three distinct MapReduce jobs (Figure \ref{cloudbreak_workflow}): a job that can align reads to the reference using a variety of mapping algorithms; a job that computes a set of features along the genome; and a job which calls structural variations based on those features.
 
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth]{../figures/runtimeByNodes.pdf}
-\caption{Scalability of the Cloudbreak algorithm.
-Runtime of the Cloudbreak feature generation job for the simulated Chromosome 2 data is shown on Hadoop clusters consisting of varying numbers of compute nodes.
-Clusters were created in the Amazon Elastic Compute Cloud.}
-\label{scalability}
-\end{figure}
+The \textsc{Align Reads} job uses existing alignment tools to discover mapping locations for each read pair.
+Aligners can be executed to report multiple possible mappings for each read, or only the best possible mapping.
+Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits possible mapping locations under the $rpid$ key.
+Reducers then collect the alignments for each paired end, making them available under one key for the next job.
 
-\clearpage
+In the \textsc{Compute Features} job, we compute a set of features for each location in the genome.
+To begin, we tile the genome with small fixed-width, non-overlapping intervals.
+For the experiments reported below we use an interval size of 25bp (see Supplementary Materials for an exploration of the effects of different window sizes on accuracy and runtime).
+Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
+Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
+Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
+For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
+the fragment size implied by the alignments and the likelihood that the alignments are correct.
+We then leave two functions to be implemented depending on the application:
+\begin{flalign*}
+ \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
+ \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N 
+\end{flalign*}
 
-\begin{figure}
-\centering
-\includegraphics[width=1\textwidth]{../figures/breakpoint_resolution.pdf}
-\caption{Breakpoint resolution for each tool for deletions on the Chromosome 2 simulated data.
-For each correctly predicted deletion, we calculated the difference in length between the true deletion and the prediction.}
-\label{breakpoint_resolution}
-\end{figure}
+The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
+\todo{This} We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
+The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
 
-\clearpage
+Finally, the third MapReduce job, \textsc{Call Variants}, is responsible for making SV calls based on the features computed at each genomic location. It calls another application-specific function: 
+ \[ \textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\} \]
 
-\begin{suppalgorithm}[h]
+This function maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
+We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
+
+\begin{algorithm}[t]
 \algrenewcommand\algorithmicprocedure{\textbf{job}}
-  \begin{algorithmic}[1]
+ \begin{algorithmic}[1]\raggedright
+\footnotesize{
  \Procedure{Alignment}{}
-    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
-    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
+ \Function{Map}{$\textrm{ReadPairId }rpid$, $\textrm{ReadId }r$, $\textrm{ReadSequence }s$, $\textrm{ReadQuality }q$}
+ \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(s,q)$}
  \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
@@
  \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
-    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
-    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
-    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
+ \ForAll{$ \textrm{AlignmentPairs }\langle a_1,a_2 \rangle \in ap$}
+ \ForAll{$ \textrm{GenomicLoci }l \in \textsc{Loci }(a_1,a_2)$}
+ \State $ \textrm{ReadPairInfo }rpi \gets \break\hspace*{2em} \langle\textrm{InsertSize}(a_1,a_2), \textrm{AlignScore}(a_1,a_2)\rangle$
+ \State $\textsc{Emit}(\textrm{GenomicLocus }l, \textrm{ReadPairInfo }rpi)$
  \EndFor
@@
  \EndFunction
-    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
+ \Function{Reduce}{$\textrm{GenomicLocus }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
  \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
+ \State $\textsc{Emit}(\textrm{GenomicLocus }l, \textrm{SVFeatures } \phi_l)$
  \EndFunction
@@
  \Procedure{Call SVs}{}
-    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
+ \Function{Map}{$\textrm{GenomicLocus }l, \textrm{SVFeatures } \phi_l$}
  \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
  \EndFunction
-    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
+ \Function{Reduce}{$\textrm{Chromosome }c, \break \textrm{GenomicLocus } l_{1,2,\ldots}, \phi_{1,2,\ldots}$}
  \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
@@
  \EndProcedure
+}
  \end{algorithmic}
-\caption{The algorithmic framework for SV calling in MapReduce.}
+\caption{A framework for SV calling in MapReduce.}
 \label{cb_algo}
-\end{suppalgorithm}
+\end{algorithm}
 
-\clearpage
+\subsection{Cloudbreak: Detecting Deletions and Insertions}
 
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{rrrrrr}
-  \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 224 &  84 & 82 &  31 & 26\\ 
-  \hline
-  Cloudbreak  & \textbf{68} (17)  & \textbf{67} (\textbf{10}) &  \textbf{56} (\textbf{5}) & \textbf{11} (\textbf{3}) & \textbf{15} (\textbf{0}) \\ 
-  Breakdancer & 52 (8)  & 49 (2) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
-  GASVPro     & 35 (2)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
-  DELLY-RP       & 22 (1)  & 56 (1) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
-  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
-  Pindel      & 60 (\textbf{32})  & 16 (0) &  41 (2) & 1 (0) & 12 (\textbf{0})\\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPredsFDR10}
-\end{table}
+We have implemented an RP-based SV detection package, Cloudbreak, in the framework described above.
+Cloudbreak detects deletions of size 40bp to 25kb, and small insertions.
+We accomplish this by providing the following implementions of the user-defined functions described above:
 
-\clearpage
+\textbf{\sc{Loci}}: Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
+For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
 
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{r|r|rrr|rrr}
-\multicolumn{2}{c}{}  & \multicolumn{3}{c}{Simulated Data} & \multicolumn{3}{c}{NA18507} \\
-\hline
- & SV Types &  Single CPU & Parallel & Proc. &  Single CPU & Parallel & Proc.  \\ 
-  \hline
-  Cloudbreak & D,I &   NA    & 290 & 312    & NA         & 824 & 636 \\ 
-  Breakdancer & D,I,V,T &  653   & NA       & NA          & 134,170 &  5,586 & 84 \\
-  GASVPro & D,V   &  3,339  & NA       & NA         & 52,385  & NA & NA \\
-  DELLY & D         &  1,964 & NA          & NA      & 30,311  & 20,224 & 84 \\
-  Pindel & D,I,V,P         & 37,006 &  4,885     & 8          &  284,932  & 28,587 & 84 \\ 
-  MoDIL & D,I        &  NA      & 52,547 & 250 & NA         & NA  & NA\\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Runtimes (elapsed) on both data sets of each tool tested, in single-processor and parallel mode.
-For parallel runs, Proc.
-is the maximum number of simultaneously running processes or threads.
-All times are in seconds.
-The types of variants detected by each program are listed with the abbreviations: D - deletion; I - insertion; V - Inversion; P - duplication; T - translocation.
-Interchromosomal translocations are only detected by Breakdancer in single CPU mode.
-}
-\label{runtimes}
-\end{table}
-
-\newpage
-
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\hline
- &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 120 & 327 & 562 & 8059 \\ 
-  \hline
-  Cloudbreak  & \textbf{37} (\textbf{10}) & \textbf{180} (\textbf{25}) & \textbf{300} (\textbf{85}) & \textbf{1166} (\textbf{270}) \\ 
-  Breakdancer & 29 (7) & 142 (3) & 192 (11) & 868 (21) \\
-  GASVPro     & 16 (1) & 79 (1) & 79 (7) & 330 (16) \\
-  DELLY-RP       & 21 (1) & 117 (1) & 152 (6) & 712 (5) \\
-  DELLY-SR       & 0 (0) & 42 (0) & 27 (0) & 391 (0) \\
-  Pindel      & 18 (9) & 112 (\textbf{25}) & 109 (2) & 536 (7) \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.
-For both data sets we used the thresholds derived from finding the 10\% FDR level in the simulated data set.
-The proportion of deletions that overlap repetitive elements discovered by Cloudbreak is similar to that of the other methods.}
-\label{deletionRepmaskpreds}
-\end{table}
-
-\newpage
-
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\hline
- &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 133 & 270 & 341 & 355 \\ 
-  \hline
-  Cloudbreak  & \textbf{32} (11) & \textbf{91} (\textbf{47}) & \textbf{169} (\textbf{61}) & \textbf{148} (\textbf{68}) \\ 
-  Breakdancer & 17 (5) & 22 (6) & 82 (7) & 44 (9) \\
-  Pindel      & 25 (\textbf{16}) & 54 (30) & 84 (35) & 82 (24) \\ 
-  MoDIL      & 5 (0) & 16 (3) & NA & NA \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.
-The maximum sensitivity cutoffs were used for both data sets.
-The proportion of insertions in repetitive elements discovered by Cloudbreak is similar to that of the other methods.}
-\label{insertionRepmaskpreds}
-\end{table}
-
-\newpage
-
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{r|r|rr|rr|}
-\multicolumn{2}{c}{}  & \multicolumn{4}{c}{Actual Genotypes} \\
-\multicolumn{2}{c}{}  & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\cline{3-6}
-\multicolumn{2}{c|}{} &  Homozygous & Heterozygous & Homozygous & Heterozygous \\ 
-\cline{2-6}
-\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 35 & 2 &  96 & 21 \\
- & Heterozygous & 0 & 39 &  2 & 448 \\
-\cline{2-6}
-\end{tabular}
-\end{center}
-\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
-\label{deletionGenotypeaccuracy}
-\end{table}
-
-\newpage
-
-\section{Supplementary Discussion}
-
-\subsection{A general framework for SV Detection in MapReduce}
-
-We have developed a conceptual algorithmic framework for SV detection in MapReduce, which is outlined in Supplementary Algorithm \ref{cb_algo}.
-As described in the main text, this framework divides processing into three separate MapReduce jobs: an alignment job, a feature computation job, and an SV calling job.
-
-The alignment job uses sensitive mapping tools and settings to discover mapping locations for each read pair.
-Aligners can be executed to report multiple possible mappings for each read, or only the best possible mapping.
-Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits possible mapping locations under the $rpid$ key.
-Reducers then collect the alignments for each paired end, making them available under one key for the next job.
-Our implementation contains wrappers to execute the aligners BWA, GEM \autocite{MarcoSola:2012hm}, Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
-
-In the second job, we compute a set of features for each location in the genome.
-To begin, we tile the genome with small fixed-width, non-overlapping intervals.
-For the experiments reported in this paper we use an interval size of 25bp.
-Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
-Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
-Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
-For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
-the fragment size implied by the alignments and the likelihood the alignments are correct.
-We then leave two functions to be implemented depending on the application:
-\begin{flalign*}
- \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
- \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
-\end{flalign*}
-
-The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
-We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
-The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
-
-Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
-It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
-We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
-
-\subsection{Cloudbreak: An Implementation of the MapReduce SV Detection Framework}
-
-Our Cloudbreak software package can be seen as an implementation of the general framework defined above.
-In particular, Cloudbreak implements the three user-defined function described above as follows:
-
-\begin{description}
-\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
-For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
-\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure \ref{insert_size_mixes}).
+\textbf{$\Phi$}: To compute features for each genomic location, we follow \cite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure \ref{Sinsert_size_mixes}).
 Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
@@
 Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
-
-To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
-We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
-In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
-This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
-Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
-
-\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
-
-And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
-
-We fit the parameters of the GMM using the Expectation-Maximization algorithm.
-Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$.
-We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$.
-In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$.
-We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions.
-In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$.
-We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$.
-We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.
-
+We fit these parameters using an Expectation-Maximization algorithm (see Supplementary Material).
+To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique; these procedures also allow us to process multiple mappings for each read if they are reported by the aligner (see Supplementary Material).
 The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
 
-\item[\sc{PostProcess}] We convert our features along each chromosome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
-To eliminate noise we apply a median filter with window size 5.
-We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
-\end{description}
+\textbf{\sc{PostProcess}}: We convert our features along each chromosome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
+To eliminate noise we apply a median filter.
+We end regions when $\mu'$ changes by more than $2\sigma$, and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
 
-\subsection{Runtime Analysis}
+An illustration of the Cloudbreak algorithm working on a simple example is shown in Supplementary Figure \ref{Salgorithm_example}.
 
-We implemented and executed Cloudbreak on a 56-node Hadoop cluster, with 636 map slots and 477 reduce slots.
-Not including alignment time, we were able to process the Chromosome 2 simulated data in under five minutes, and the the NA18507 data set in under 15 minutes.
-For the simulated data set we used 100 reducers for the compute SV features job; for the real data set we used 300.
-The bulk of Cloudbreak's execution is spent in the feature generation step.
-Extracting deletion and insertion calls take under two minutes each for both the real and simulated data sets; the times are equal because each reducer is responsible for processing a single chromosome, and so the runtime is bounded by the length of time it takes to process the largest chromosome.
+\subsection{Implementation Details}
 
-In Supplementary Table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work.
-Each tool varies in the amount of parallelization supported.
-We report runtimes for tools run in their default single-threaded mode, as well as for levels of parallelization achievable with basic scripting, noting that one of the key advantages of Hadoop/MapReduce is the ability to scale parallel execution to the size of the available compute cluster without any custom programming.
-Pindel allows multi-threaded operation on multicore servers.
-Pindel and Breakdancer allow processing of a single chromosome in one process, so it is possible to execute all chromosomes in parallel on a cluster that has a job scheduler and shared filesystem.
-Breakdancer has an additional preprocessing step (\texttt{bam2cfg.pl}) which runs in a single thread.
-DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode.
-GASVPro allows parallelization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting, and we did not find that the MCMC module consumed most of the runtime in our experiments, so we do not attempt to parallelize this component.
-The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers; we adapted these for use in our cluster.
+Cloudbreak is a native Java Hadoop application.
+We deployed Cloudbreak on a 56-node cluster running the Cloudera CDH3 Hadoop distribution, version 0.20.2-cdh3u4.
+We use snappy compression for MapReduce data.
+Hadoop's distributed cache mechanism shares the executable files and indices needed for mapping tasks to the nodes in the cluster.
 
-In parallel execution, the total time to execute is bounded by the runtime of the longest-running process.
-In the case of chromosome-parallelizable tools including Breakdancer, Pindel, and DELLY, this is typically the process working on the largest chromosome.\footnote{We note that one Breakdancer process, handling an unplaced contig in the hg19 reference genome, never completed in our runs and had to be killed manually; we exclude that process from our results.} In the case of MoDIL's run on the simulated data, we found that the different processes varied widely in their execution times, likely caused by regions of high coverage or with many ambiguously mapped reads.
-Cloudbreak mitigates this problem during the time-consuming feature generation process by using Hadoop partitioners to randomly assign each genomic location to one of the set of reducers, ensuring that the work is evenly distributed across all processes.
-This distribution of processing across the entire cluster also serves to protect against server slowdowns and hardware failures - for example, we were still able to complete processing of the NA18507 data set during a run where one of the compute nodes was rebooted midway through the feature generation job.
+Cloudbreak's alignment job can run a variety of alignment tools that report reads in SAM format (Supplementary Materials).
+In the map phase, mappers align reads in either single-end or paired-end mode to the reference genome in parallel, outputting mapping locations as values under a key identifying the read pair.
+In the reduce phase, the framework combines the reported mapping locations for the two ends of each read pair.
+This job can also be skipped in favor of importing a pre-existing set of mappings directly into the Hadoop cluster.
 
-\printbibliography[title=Supplementary References]
+\subsection{Cloudbreak Genotyping}
+We used the parameters of the fit GMM to infer the genotype of each predicted variant.
+Assuming that our pipeline is capturing all relevant read mappings near the locus of the variant, the genotype should be indicated by the estimated parameter $\alpha$, the mixing parameter that controls the weight of the two components in the GMM.
+By tuning on the simulated data, we determined that by setting a simple cutoff of .35 on the average value of $\alpha$ for each prediction, we were able to accurately (see Section~\ref{section_results_genotyping}) call the predicted variant homozygous or heterozygous.
+We used the same cutoff for deletion and insertion predictions.
 
-\newpage
+\subsection{Running Locally or in the Cloud}
 
-\section{Cloudbreak User Manual}
-\label{cloudbreak}
+Cloudbreak can be executed on any Hadoop cluster; Hadoop abstracts away the details of cluster configuration, making distributed applications portable.
+In addition, our Cloudbreak implementation can leverage the Apache Whirr library to automatically create clusters with cloud service providers such as the Amazon Elastic Compute Cloud (EC2).
+This enables on demand provisioning of Hadoop clusters which can then be terminated when processing is complete, eliminating the need to invest in a standing cluster and allowing a model in which users can scale their computational infrastructure as their need for it varies over time.
+In addition, Whirr supports multiple cloud service providers, making it possible to leverage many public or private clouds and avoid vendor lock-in.
+Instructions and examples describing how to leverage cloud computing with Cloudbreak are available in the user manual and Supplementary Materials.
 
-Cloudbreak is a Hadoop-based structural variation (SV) caller for Illumina
-paired-end DNA sequencing data. Currently Cloudbreak calls genomic insertions
-and deletions; we are working on adding support for other types of SVs.
+% \begin{table}[!t]
+% \processtable{This is table caption\label{Tab:01}}
+% {\begin{tabular}{llll}\toprule
+% head1 & head2 & head3 & head4\\\midrule
+% row1 & row1 & row1 & row1\\
+% row2 & row2 & row2 & row2\\
+% row3 & row3 & row3 & row3\\
+% row4 & row4 & row4 & row4\\\botrule
+% \end{tabular}}{This is a footnote}
+% \end{table}
 
-Cloudbreak contains a full pipeline for aligning your data in the form of FASTQ
-files using alignment pipelines that generate many possible mappings for every
-read, in the Hadoop framework. It then contains Hadoop jobs for computing
-genomic features from the alignments, and for calling insertion and deletion
-variants from those features.
+\end{methods}
 
-You can get Cloudbreak by downloading a pre-packaged release from the ``releases''
-tab in the GitHub repository, or by building from source as described below.
+% \begin{figure}[!tpb]%figure1
+% %\centerline{\includegraphics{fig01.eps}}
+% \caption{Caption, caption.}\label{fig:01}
+% \end{figure}
 
-\subsection{Building From Source}
-\label{buildingfromsource}
+% \begin{figure}[!tpb]%figure2
+% %\centerline{\includegraphics{fig02.eps}}
+% \caption{Caption, caption.}\label{fig:02}
+% \end{figure}
 
-To build the latest version of Cloudbreak, clone the GitHub repository. You'll
-need to install Maven to build the executables. (http:/\slash maven.apache.org\slash )
-Enter the top level directory of the Cloudbreak repository and type the command:
+\section{Results}
 
-\texttt{mvn package}
+\subsection{Tests with Simulated Data}
 
-This should compile the code, execute tests, and create a distribution file,
- \texttt{cloudbreak-\$VERSION-dist.tar.gz} in the \texttt{target\slash } directory. You can then copy
- that distribution file to somewhere else on your system, unpack it with
- \texttt{tar -xzvf cloudbreak-\$VERSION-dist.tar.gz} and access the Cloudbreak jar file
- and related scripts and properties files.
+We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer~\citep{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings~\citep{Sindi:2012kk}, the SR method Pindel~\citep{Ye:2009p2}, and the hybrid RP-SR method DELLY~\citep{Rausch:2012he}.
+DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence; we refer to these sets of calls as DELLY-RP and DELLY-SR.
+We also attempted to evaluate MoDIL on the same data.
+All of these methods detect deletions.
+Insertions can be detected by BreakDancer, Pindel, and MoDIL.
+See Supplementary Materials for details on how reads were aligned and each program was invoked.
+For all alignments we used BWA~\citep{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Fig.
+\ref{Salignment_comparison}).
 
-\subsection{Dependencies}
-\label{dependencies}
+There is no available test set of real Illumina sequencing data from a sample that has a complete annotation of SVs.
+Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics.
+On the other hand, any simulated data should contain realistic SVs that follow patterns observed in real data.
+We took one of the most complete lists of SVs from an individual, the list of homozygous insertions and deletions from the genome of J.
+Craig Venter~\citep{Levy:2007fb}, and used it to simulate a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants, with 100bp reads and a mean fragment size of 300bp (See Supplementary Material).
 
-Cloudbreak requires a cluster Hadoop 0.20.2 or Cloudera CDH3 to run (the older
-mapreduce API). If you don't have a Hadoop cluster, Cloudbreak can also use the
-Apache Whirr API to automatically provision a cluster on the Amazon Elastic
-Compute Cloud (EC2). See the section on using WHIRR below.
+\begin{figure*}[!tpb]
+\centering
+\includegraphics[width=1\textwidth]{../../figures/chr2_rocs_runtimes_journal.pdf}
+\caption{Accuracy and runtime performance on a simulated data set.
+(a) Receiver Operating Characteristic (ROC) curves showing the specificity and sensitivity of each tool to deletions larger than 40bp on a simulated set of reads giving diploid coverage of 30X on human chromosome 2.
+Deletions from the Venter genome were randomly added to one or both haplotypes.
+Each point on a curve represents a different threshold on the confidence of predictions made by that tool.
+Thresholds vary by: Cloudbreak - likelihood ratio; BreakDancer, DELLY, GASVPro - number of supporting read pairs; Pindel - simple score.
+(b) ROC curves for insertion predictions.
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).}
+\label{chr2CombinedRoc}
+\end{figure*}
 
-If you wish to run alignments using Cloudbreak, you will need one of the following
-supported aligners:
+Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment.
+See Supplementary Material for a description of how we identified correct predictions.
+All approaches show high specificity at stringent thresholds in this simulation.
+Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY.
+For insertions, Cloudbreak's performance is similar to Pindel, although Cloudbreak can achieve greater sensitivity at the same false positive rate by decreasing its threshold.
+Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes.
+(Of course, Cloudbreak uses many more CPUs as distributed algorithm.
+See Supplementary Material for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster.
+Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow.
+To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Fig.
+\ref{Sscalability}).
 
-\begin{itemize}
-\item BWA (Recommended): http:/\slash bio-bwa.sourceforge.net\slash
+Choosing the correct threshold to set on the output of an SV calling algorithm can be difficult.
+The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying thresholds.
+First, we characterized the predictions made by each algorithm at the threshold that gives them maximum sensitivity.
+\todo{For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.} The results are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}.
+MoDIL and Cloudbreak exhibited the greatest recall for deletions.
+Cloudbreak also has high precision at this threshold, and discovers many small variants.
+For insertions, Cloudbreak has the highest recall, although recall is low for all four approaches.
+Cloudbreak again identifies many small variants.
+Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by RP mapping.
+We also used the ROC curves to characterize algorithm performance when a low false discovery rate is required.
+Supplementary Table \ref{Schr2DeletionPredsFDR10} shows the total number of deletions found when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
+At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool.
+Insertion performance never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.\todo{ We also examined Cloudbreak's ability to detect events in repetitive regions of the genome, and found that it was similar to the other methods tested (Supplementary Tables \ref{SdeletionRepmaskpreds} and \ref{SinsertionRepmaskpreds}).}
 
-\item GEM: http:/\slash algorithms.cnag.cat\slash wiki\slash The\_GEM\_library
+It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Fig.
+\ref{Sbreakpoint_resolution}).
+SR methods have higher resolution than RP methods.
+Cloudbreak's resolution suffers because the independent calculation of features at each window from the distribution of overlapping insert sizes does not allow the algorithm to keep track of the actual coordinates of the mapped reads from each pair, which other tools use to set bounds on the true breakpoint locations.
+We believe, however, that it is worth increasing sensitivity and specificity in SV detection tools even at the cost of breakpoint resolution.
+\todo{It would also be possible to extend our feature set to include information useful for increasing resolution, such as the location of nearby read mappings or soft-clipped alignments.
+}In addition, several recent pipelines have validated RP calls \emph{in silico} by local assembly, a process that could potentially be parallelized in a subsequent MapReduce job.
 
-\item RazerS 3: http:/\slash www.seqan.de\slash projects\slash razers\slash
+\begin{table*}[b]
+\begin{center}
+%\resizebox{\textwidth}{!}{
+\footnotesize
+\begin{tabular}{r|rrr|rrrrr}
+  \cline{2-9}
+   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\hline
+\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
+  \hline
+\cline{2-9}
+&  Cloudbreak    &  0.638 & \textbf{0.678} & \textbf{153} (9)  & 61 (0) &  62 (0) & 12 (0) & 15 (0) \\ 
+&  BreakDancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
+&  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
+&  DELLY-RP           & 0.457 & 0.613 & 114 (3)  & \textbf{68} (0) &  \textbf{66} (0) & 9 (1) & 17 (0) \\ 
+&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
+&  Pindel           & 0.462 & 0.421 & 96 (\textbf{11})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
+&  MoDIL           & 0.132  & 0.66 & 123 (6)  & 66 (\textbf{3}) &  \textbf{66} (\textbf{11}) & \textbf{17} (\textbf{7}) & \textbf{23} (\textbf{8})\\ 
+   \hline
+\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
+\cline{2-9}
+&  Cloudbreak   &0.451 & \textbf{0.305}  & \textbf{79} (\textbf{32})  & \textbf{32} (\textbf{18}) &  \textbf{11} (8) & 1 (0) & 0 (0) \\ 
+&  BreakDancer & 0.262 & 0.0968  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
+&  Pindel          & \textbf{0.572} & 0.196 & 52 (25)  & 5 (1) &  10 (\textbf{9}) & \textbf{3} (\textbf{2}) & \textbf{9} (\textbf{9})\\ 
+&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (0) &  1 (0) & 2 (\textbf{2}) & 0 (0)\\ 
+\hline
+\end{tabular}
+%}
+\end{center}
+\caption{The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses).
+The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section.}
+\label{chr2DeletionAndInsertionPredsMaxSensitivity}
+\end{table*}
 
-\item Bowtie2: http:/\slash bowtie-bio.sourceforge.net\slash bowtie2\slash index.shtml
+\subsection{Tests with Biological Data}
 
-\item Novoalign: http:/\slash www.novocraft.com
+We downloaded a set of reads from Yoruban individual NA18507, experiment ERX009609, from the Sequence Read Archive.
+This sample was sequenced by Illumina Inc.
+on the Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage.
+To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same individual.
+These included data from the Human Genome Structural Variation Project reported by~\cite{Kidd:2008p926}, a survey of small indels conducted by~\cite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project~\citep{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
+We merged any overlapping calls of the same type into the region spanned by their unions.
+We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.
 
-\end{itemize}
+\begin{figure*}
+\centering
+\includegraphics[width=1\textwidth]{../../figures/NA18507_rocs_runtimes_journal.pdf}
+\caption{Accuracy and performance on the 37X NA18507 sample.
+(a) ROC curve for deletion prediction performance, tested against the combined gold standard sets of deletions taken from~\cite{Kidd:2008p926},~\cite{Mills:2011fi}, and~\cite{GenomesProjectConsortium:2012co}.
+(b) ROC curve for insertion prediction performance.
+(c) Runtimes for each tool, not including alignment time, parallelized when possible (see Supplementary Material).
+}
+\label{NA18507CombinedRoc}
+\end{figure*}
 
-\subsection{User Guide}
-\label{userguide}
+Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm on the NA18507 data set.
+All algorithms show far less specificity for the gold standard set than they did in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample.
+For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels.
+Cloudbreak has comparable accuracy for insertions to other tools, and can identify the most variants at higher levels of sensitivity.
+Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, BreakDancer, even when BreakDancer is run in parallel for each chromosome on different nodes in the cluster.
 
-You can use Cloudbreak in several different ways, depending on whether you want
-to start with FASTQ files and use Hadoop to help parallelize your alignments, or if you already
-have an aligned BAM file and just want to use Cloudbreak to call variants. In addition,
-the workflow is slightly different depending on whether you want to run on a local
-hadoop cluster or want to run using a cloud provider like Amazon EC2. Later in this
-file, we've listed a set of scenarios to describe options for running the Cloudbreak
-pipeline. Find the scenario that best fits your use case for more details on how to
-run that workflow. For each scenario, we have created a template script that contains
-all of the steps and parameters you need, which you can modify for your particular data set.
+Given the high number of novel predictions made by all tools at maximum sensitivity, we decided to characterize performance at more stringent thresholds.
+We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X.
+For insertions, we used the maximum sensitivity thresholds for each tool due to the high observed FDRs in the simulated data.
+Precision and recall at these thresholds, as well as the performance of each algorithm at predicting variants of each size class, is shown in Table \ref{NA18507DeletionAndInsertionPreds}.
+For deletions, Cloudbreak has the greatest sensitivity of any tool, identifying the most variants in each size class.
+Pindel exhibits the highest precision with respect to the gold standard set.
+For insertions, Pindel again has the highest precision at maximum sensitivity, while Cloudbreak has by far the highest recall.
 
-\subsection{Running on a cloud provider like Amazon EC2 with Whirr}
-\label{runningonacloudproviderlikeamazonec2withwhirr}
+\begin{table*}
+\begin{center}
+%\resizebox{\textwidth}{!}{
+\footnotesize
+\begin{tabular}{r|rrr|rrrrr}
+  \cline{2-9}
+&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\hline
+\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,462 & 240 & 232 & 147 & 540 \\
+  \hline
+\cline{2-9}
+& Cloudbreak & 0.0943 & \textbf{0.17} & \textbf{573} (\textbf{277})  & \textbf{176} (\textbf{30}) &  \textbf{197} (\textbf{18}) & \textbf{121} (\textbf{6}) & \textbf{399} (\textbf{24}) \\ 
+& BreakDancer & 0.137 & 0.123 & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
+&  GASVPro & 0.147 & 0.0474 & 120 (21)  & 40 (2) &  85 (0) & 36 (0) & 128 (0) \\ 
+&  DELLY-RP & 0.0931 & 0.1 & 143 (6)  & 128 (3) &  167 (1) & 103 (0) & 323 (1) \\ 
+&  DELLY-SR & 0.153 & 0.0485 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
+&  Pindel & \textbf{0.179} & 0.0748 & 149 (8)  & 61 (0) &  149 (0) & 69 (1) & 217 (0) \\ 
+\hline
+\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 & 0 \\
+\cline{2-9}
+& Cloudbreak & 0.0323 & \textbf{0.455} & \textbf{265} (\textbf{104})  & \textbf{49} (\textbf{24}) &  3 (1) & 0 (0)  & 0 (0)  \\ 
+& BreakDancer & 0.0281 & 0.181 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
+&  Pindel & \textbf{0.0387} & 0.239 & 144 (45)  & 14 (7) &  \textbf{7} (\textbf{6}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
+\hline
+\end{tabular}
+%}
+\end{center}
+\caption{The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class found.
+Exclusive predictions are in parentheses.
+For deletions, the same cutoffs were used as for the simulated data as in Supplementary Table \ref{Schr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.
+For insertions, the maximum sensitivity cutoff was used.}
+\label{NA18507DeletionAndInsertionPreds}
+\end{table*}
 
-Cloudbreak has support for automatically deploying a Hadoop cluster on
-cloud providers such as Amazon EC2, transferring your data there, running the Cloudbreak algorithm, and
-downloading the results.
+\subsection{Performance on a Low-Coverage Cancer Data Set}
 
-Of course, renting compute time on EC2 or other clouds costs money, so please be
-familiar with the appropriate usage and billing policies of your cloud provider
-before attempting this.
+We also tested Cloudbreak on a sequencing data set obtained from a patient with acute myeloid leukemia (AML).
+This data set consisted of 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
+Using a pipeline consisting of Novoalign, BreakDancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions.
+Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even when using lower coverage data sets with a greater variance of insert sizes.
+The variants identified include deletions in the gene CTDSPL/RBPS3, an AML tumor suppressor~\citep{Zheng:2012kk}, and NBEAL1, a gene up-regulated in some cancers~\citep{Chen:2004jo}.
+We are currently investigating these deletions to determine their functional impact on this patient.
 
-WE ARE NOT RESPONSIBLE FOR UNEXPECTED CHARGES THAT YOU MIGHT INCUR ON EC2 OR
-OTHER CLOUD PROVIDERS.
+\subsection{Genotyping Deletions}\label{section_results_genotyping}
 
-Many properties that affect the cluster created can be set in the file
-\texttt{cloudbreak-whirr.properties} in this distribution. You will need to edit this file
-to set your AWS access key and secret access key (or your credentials for other
-cloud provider services), and to tell it the location of the public and
-private SSH keys to use to access the cluster. You can also control the number
-and type of nodes to include in the cluster. The default settings in the file
-create 15 nodes of type m1.xlarge, which is sufficient to fully process a 30X
-simulation of human chromosome 2, including read alignment and data transfer time,
-in under an hour. We have only tested this capability using EC2; other cloud providers
-may not work as well. You can also direct Whirr to use Amazon EC2's spot instances, which are
-dramatically cheaper than on-demand instances, although they carry the risk of
-being terminated if your price is out-bid. Using recent spot pricing, it cost
-us about \$5 to run the aforementioned chromosome 2 simulation. We recommend
-setting your spot bid price to be the on demand price for the instance type you
-are using to minimize the chance of having your instances terminated.
+Because Cloudbreak explicitly models zygosity in its feature generation algorithm, it can predict the genotypes of identified variants.
+We tested this on both the simulated and NA18507 data sets.
+For the NA18507 data set, we considered the deletions from the 1000 Genomes Project, which had been genotyped using the population-scale SV detection algorithm Genome STRiP~\citep{Handsaker:2011ki}.
+Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the deletions it detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
+Supplementary Table \ref{SdeletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
+\todo{None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that met our size threshold and also had genotyping information.
+Of the 123 insertions detected by Cloudbreak on the simulated data set, 43 were heterozygous.
+Cloudbreak correctly classified 78 of the 80 homozygous insertions and 31 of the 43 heterozygous insertions, for an overall accuracy of 88.6\%.}
 
-Please consult Amazon's EC2 documentation and the documentation for Whirr for
-more information on how to configure and deploy clusters in the cloud.
+\section{Discussion}
 
-\subsection{Running on a Small Example Data Set}
-\label{runningonasmallexampledataset}
+Over the next few years, due to advances in sequencing technology, genomics data are expected to increase in volume by several orders of magnitude, expanding into the realm referred to as ``big data''.
+In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines.
+In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.
 
-To facilitate testing of Cloudbreak, we have publicly hosted the reads from the simulated
-data example described in the Cloudbreak manuscript on a bucket in Amazon's S3 storage
- service at s3:/\slash cloudbreak-example\slash . We have also provided an example script that creates
- a cluster in Amazon EC2, copies the data to the cluster, runs the full Cloudbreak
- workflow including alignments with BWA, and copies the variant calls back to the
- local machine before destroying the cluster. The script, called \texttt{Cloudbreak-EC2-whirr-example.sh}
- is in the scripts directory of the Cloudbreak distribution. Of course, you will still
- need to edit the \texttt{cloudbreak-whirr.properties} file with your EC2 credentials, and verify
- that the cluster size, instance types, and spot price are to your liking before
- executing the example.
+Among big data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters.
+In this paper, we introduced a novel conceptual framework for SV detection algorithms in MapReduce, based on computing local genomic features.
+This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of an algorithm for detecting insertions and deletions based on fitting a GMM to the distribution of mapped insert sizes spanning each genomic location.
 
-\subsubsection{Scenario 1: Compute alignments in Hadoop, using a local Hadoop cluster}
-\label{scenario1:computealignmentsinhadoopusingalocalhadoopcluster}
+On simulated and real data sets, our approach exhibits high accuracy when compared to popular SV detection algorithms that run on traditional clusters and servers.
+Detection of insertions and deletions is an important area of research; \cite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
+\todo{More SV annotations are also required to assess the impact of SVs on non-coding elements~\citep{Mu:2011br}.}
 
-To install aligner dependencies for use by Cloudbreak, first generate the index
-for the genome reference you would like to run against. Then, copy all of the
-required index files, and the executable files for the aligner into HDFS using
-the \texttt{hadoop dfs -copyFromLocal} command. For BWA you will need all of the index files
-created by running \texttt{bwa index}. You will also need an `fai' file for the reference,
-containing chromosome names and lengths, generated by \texttt{samtools faidx}.
+In addition to delivering high accuracy and fast runtimes for a RP-based SV detection tool, our approach offers a basis for developing a variety of SV algorithms can run in MapReduce pipelines with the power to process vast amounts of data in a cloud or commodity server setting.
+With the advent of cloud service providers such as Amazon EC2, it is becoming easy to instantiate on-demand Hadoop compute clusters.
+Having computational approaches that can harness this capability will become increasingly important for researchers or clinicians who need to analyze increasing amounts of sequencing data.
 
-If your reference file is \texttt{reference.fa}, and \texttt{bwa aln} has created the files
+% Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
+\citepalp{Boffelli03} might want to know about  text text text text
+% Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
+\citepalp{Boffelli03} might want to know about  text text text text
+% Text Text Text Text Text Text  Text Text Text Text.
 
-\begin{verbatim}
-reference.fa.amb
-reference.fa.ann
-reference.fa.bwt
-reference.fa.pac
-reference.fa.sa
-\end{verbatim}
+% Table~\ref{Tab:01} shows that Text Text Text Text Text  Text Text Text Text Text Text. Figure \ref{fig:02} shows that
+% the above method Text Text. Text Text Text  Text Text Text Text Text Text. Figure \ref{fig:02} shows that
+% the above method Text Text. Text Text Text  Text Text Text Text Text Text. Figure \ref{fig:02} shows that
+% the above method Text Text.
 
-and \texttt{reference.fa.fai} as described above, issue the following
-commands to load the necessary files into HDFS:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
+%     as it may ignore the figures.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{verbatim}
-hdfs -mkdir indices/
-hdfs -mkdir executables/
-hdfs -copyFromLocal reference.fa.amb indices/
-hdfs -copyFromLocal reference.fa.ann indices/
-hdfs -copyFromLocal reference.fa.bwt indices/
-hdfs -copyFromLocal reference.fa.pac indices/
-hdfs -copyFromLocal reference.fa.sa indices/
-hdfs -copyFromLocal reference.fa.fai indices/
-hdfs -copyFromLocal /path/to/bwa/executables/bwa executables/
-hdfs -copyFromLocal /path/to/bwa/executables/xa2multi.pl executables/
-\end{verbatim}
+\section{Conclusion}
 
-The basic workflow is:
+% (Table~\ref{Tab:01}) Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
+\citepalp{Boffelli03} might want to know about  text text text text
+% Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
+\citepalp{Boffelli03} might want to know about  text text text text
+% Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
 
-\begin{enumerate}
-\item Load the FASTQ files into HDFS
+% Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
+\citepalp{Boffelli03} might want to know about  text text text text
 
-\item Run one of the Cloudbreak alignment commands to align your reads
+% \begin{enumerate}
+% \item this is item, use enumerate
+% \item this is item, use enumerate
+% \item this is item, use enumerate
+% \end{enumerate}
 
-\item Create a readGroup file to describe the location and insert size characteristics of your reads, and copy it into HDFS.
+% Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
+\citepalp{Boffelli03} might want to know about  text text text text
+% Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
+\citepalp{Boffelli03} might want to know about  text text text text
+% Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
 
-\item Run the GMM fitting feature generation step of the Cloudbreak process.
+% Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text. Figure \ref{fig:02} shows that the above method  Text Text Text Text
 
-\item Extract deletion calls from the features created in step 4.
+\section*{Acknowledgement}
+We thank Izhak Shafran at the Center for Spoken Language Understanding for advice, support, and shared computational resources, and Bob Handsaker for helpful conversations.
 
-\item Copy the deletion calls from HDFS to a local directory.
+\paragraph{Funding\textcolon} C.T.S.
+and A.L.A.
+are supported by the Italian Association for Cancer Research (AIRC).
+Sequencing of the AML sample was supported by the Oregon Medical Research Foundation.
 
-\item Extract insertion calls from the features created in step 4.
+\bibliographystyle{natbib}
+%\bibliographystyle{achemnat}
+%\bibliographystyle{plainnat}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{bioinformatics}
+%
+%\bibliographystyle{plain}
+%
+\bibliography{Document}
 
-\item Copy the insertion calls from HDFS to a local directory.
+% \begin{thebibliography}{}
+% \bibitem[Bofelli {\it et~al}., 2000]{Boffelli03} Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
 
-\item Optionally, export the alignments back into a BAM file in your local filesystem.
+% \bibitem[Bag {\it et~al}., 2001]{Bag01} Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
 
-\end{enumerate}
+% \bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
+% Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
+% in nigral dopaminergic neurnol cell: correlation with the known
+% pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
+% Res.}, \textbf{110}(Suppl. 1), 76--84.
 
-We have created a script to run through the full process of executing the
-Cloudbreak pipeline from FASTQ files to insertion and deletion calls. The
-script is named \texttt{Cloudbreak-full.sh} and can be found in the scripts directory
-of the Cloudbreak distribution. To customize the script for your needs, copy
-it to a new location and edit the variables in the first three sections:
-``EXPERIMENT DETAILS'', ``LOCAL FILES AND DIRECTORIES'', and
-``HDFS FILES AND DIRECTORIES''.
+% \bibitem[Lehmann, 1986]{Leh86}
+% Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
 
-\subsubsection{Scenario 2: Call variants on existing alignments, using a local Hadoop cluster}
-\label{scenario2:callvariantsonexistingalignmentsusingalocalhadoopcluster}
+% \bibitem[Crenshaw and Jones, 2003]{Cre03}
+% Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
+% cancer management: one tumor, one chip. \textit{Bioinformatics},
+% doi:10.1093/bioinformatics/btn000.
 
-For this scenario you don't need to worry about having an aligner executable or
-aligner-generated reference in HDFS. You will however, need a chromosome length
-`fai' file, which you can generate by running \texttt{samtools faidx} on your reference
-FASTA files and then copying to HDFS:
+% \bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
+% Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
+% (ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
+% ???--???.
 
-\begin{verbatim}
-hdfs -copyFromLocal reference.fa.fai indices/
-\end{verbatim}
+% \bibitem[Bardet, 1920]{Bar20}
+% Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
+% polydactylie et retinite pigmentaire (contribution a l'etude des
+% formes cliniques 
+% @article{Dean:2008p277,
+% author = {Dean, J and Ghemawat, S},
+% title = {{MapReduce: Simplified data processing on large clusters}},
+% journal = {Commun. ACM},
+% year = {2008},
+% volume = {51},
+% number = {1},
+% pages = {107--113}
+% }
+% de l'obesite hypophysaire). PhD Thesis, name of
+% institution, Paris, France.
 
-After that, the workflow is:
-
-\begin{enumerate}
-\item Load your BAM file into HDFS and prepare it for Cloudbreak
-
-\item Create a readGroup file to describe the location and insert size characteristics of your reads.
-
-\item Run the GMM fitting feature generation step of the Cloudbreak process.
-
-\item Extract deletion calls from the features created in step 3.
-
-\item Copy the deletion calls from HDFS to a local directory.
-
-\item Extract insertion calls from the features created in step 3.
-
-\item Copy the insertion calls from HDFS to a local directory.
-
-\end{enumerate}
-
-To prepare alignments for Cloudbreak, they must be sorted by read name. You can then use the
- \texttt{readSAMFileIntoHDFS} Cloudbreak command.
-
-A templates for this scenario is available in the script \texttt{Cloudbreak-variants-only.sh}
-located in the scripts directory of the Cloudbreak distribution.
-
-\subsubsection{Scenario 3: Compute alignments in Hadoop, using a cloud provider like EC2}
-\label{scenario3:computealignmentsinhadoopusingacloudproviderlikeec2}
-
-First, see the section ``Running on a Cloud Provider like Amazon EC2 with Whirr'' above, and modify the file
-\texttt{cloudbreak-whirr.properties} to include your access credentials and the appropriate cluster
-specifications. After that, the workflow is similar to the workflow described for scenario \#1
-above, with the additional first steps of copying your reads and dependency files to the cloud and
-creating a cluster before processing begins, and then destroying the cluster after processing has
-completed.
-
-You can see an example workflow involving EC2 by examining the script
-\texttt{Cloudbreak-EC2-whirr.sh}. This begins by transferring your reads to Amazon S3. It then
-uses Apache Whirr to launch an EC2 Hadoop cluster, copies the necessary executable files
-to EC2, and runs the algorithm.
-
-\subsubsection{Scenario 4: Call variants on existing alignments, using a cloud provider like EC2}
-\label{scenario4:callvariantsonexistingalignmentsusingacloudproviderlikeec2}
-
-Again, please read the section ``Running on a Cloud Provider like Amazon EC2 with Whirr'' above to learn how to
-update the \texttt{cloudbreak-whirr.properties} file with your credentials and cluster specifications. After that,
-follow the template in the script \texttt{Cloudbreak-EC2-whirr-variants-only.sh} to create a workflow
-involving calling variants in the cloud.
-
-\subsection{Output Files}
-\label{outputfiles}
-
-The output from running Cloudbreak using one of the scripts above will be found in the files named
-
-\begin{verbatim}
-READ_GROUP_LIBRARY_dels_genotyped.bed
-READ_GROUP_LIBRARY_ins_genotyped.bed
-\end{verbatim}
-
-where READ\_GROUP and LIBRARY are the names of the reads in your experiment. The
-format of the files is tab-delimited with the following columns:
-
-\begin{itemize}
-\item CHROMOSOME: The chromosome of the deletion call
-
-\item START: The start coordinate of the deletion call
-
-\item END: The end coordinate of the deletion call
-
-\item NUMBER: The cloudbreak identifier of the deletion call
-
-\item LR: The likelihood ratio of the deletion (higher indicates a call more likely to be true)
-
-\item TYPE: Either ``INS'' or ``DEL''
-
-\item W: The average weight of the estimated GMM mixing parameter alpha, used in genotyping
-
-\item GENOTYPE: The predicted genotype of the call
-
-\end{itemize}
-
-\subsection{Contact information}
-\label{contactinformation}
-
-Please contact cwhelan at gmail.com with any questions on running cloudbreak.
-
-\subsection{Reference Guide}
-\label{referenceguide}
-
-All of Cloudbreak's functionality is contained in the executable jar file in the
-directory where you unpacked the Cloudbreak distribution. Use the `hadoop'
-command to run the jar file to ensure that the necessary Hadoop dependencies
-are available to Cloudbreak.
-
-To invoke any Cloudbreak command, use a command line in this format:
-
-\texttt{hadoop cloudbreak-\$\{project.version\}.jar [options] [command] [command options]}
-
-Where \texttt{command} is the name of the command, \texttt{command options} are the arguments specific
-to that command, and \texttt{options} are general options, including options for how to run
-Hadoop jobs. For example, if you'd like to specify 50 reduce tasks
-for one of your commands, pass in \texttt{-Dmapred.reduce.tasks=50} as one of the general options.
-
-Each command is detailed below and its options are listed below. You can view this information by typing
-\texttt{hadoop jar cloudbreak-\$\{project.version\}.jar} without any additional parameters.
-
-\begin{verbatim}
-    readPairedEndFilesIntoHDFS      Load paired FASTQ files into HDFS
-      Usage: readPairedEndFilesIntoHDFS [options]
-  Options:
-        *     --HDFSDataDir                  HDFS directory to load reads into
-              --clipReadIdsAtWhitespace      Whether to clip all readnames at
-                                             the first whitespace (prevents trouble
-                                             with some aligners)
-                                             Default: true
-              --compress                     Compression codec to use on the
-                                             reads stored in HDFS
-                                             Default: snappy
-        *     --fastqFile1                   File containing the first read in
-                                             each pair
-        *     --fastqFile2                   File containing the second read in
-                                             each pair
-              --filesInHDFS                  Use this flag if the BAM file has
-                                             already been copied into HDFS
-                                             Default: false
-              --filterBasedOnCasava18Flags   Use the CASAVA 1.8 QC filter to
-                                             filter out read pairs
-                                             Default: false
-              --outFileName                  Filename of the prepped reads in
-                                             HDFS
-                                             Default: reads
-              --trigramEntropyFilter         Filter out read pairs where at
-                                             least one read has a trigram entropy less
-                                             than this value. -1 = no filter
-                                             Default: -1.0
-
-    readSAMFileIntoHDFS      Load a SAM/BAM file into HDFS
-      Usage: readSAMFileIntoHDFS [options]
-  Options:
-        *     --HDFSDataDir   HDFS Directory to hold the alignment data
-              --compress      Compression codec to use for the data
-                              Default: snappy
-              --outFileName   Filename to give the file in HDFS
-                              Default: alignments
-        *     --samFile       Path to the SAM/BAM file on the local filesystem
-
-    bwaPairedEnds      Run a BWA paired-end alignment
-      Usage: bwaPairedEnds [options]
-  Options:
-        *     --HDFSAlignmentsDir    HDFS directory to hold the alignment data
-        *     --HDFSDataDir          HDFS directory that holds the read data
-        *     --HDFSPathToBWA        HDFS path to the bwa executable
-              --HDFSPathToXA2multi   HDFS path to the bwa xa2multi.pl executable
-        *     --maxProcessesOnNode   Ensure that only a max of this many BWA
-                                     processes are running on each node at once.
-                                     Default: 6
-              --numExtraReports      If > 0, set -n and -N params to bwa sampe,
-                                     and use xa2multi.pl to report multiple hits
-                                     Default: 0
-        *     --referenceBasename    HDFS path of the FASTA file from which the
-                                     BWA index files were generated.
-
-    novoalignSingleEnds      Run a Novoalign alignment in single ended mode
-      Usage: novoalignSingleEnds [options]
-  Options:
-        *     --HDFSAlignmentsDir            HDFS directory to hold the
-                                             alignment data
-        *     --HDFSDataDir                  HDFS directory that holds the read
-                                             data
-        *     --HDFSPathToNovoalign          HDFS path to the Novoalign
-                                             executable
-              --HDFSPathToNovoalignLicense   HDFS path to the Novoalign license
-                                             filez
-              --qualityFormat                Quality score format of the FASTQ
-                                             files
-                                             Default: ILMFQ
-        *     --reference                    HDFS path to the Novoalign
-                                             reference index file
-        *     --threshold                    Quality threshold to use for the -t
-                                             parameter
-
-    bowtie2SingleEnds      Run a bowtie2 alignment in single ended mode
-      Usage: bowtie2SingleEnds [options]
-  Options:
-        *     --HDFSAlignmentsDir       HDFS directory to hold the alignment
-                                        data
-        *     --HDFSDataDir             HDFS directory that holds the read data
-        *     --HDFSPathToBowtieAlign   HDFS path to the bowtie2 executable
-        *     --numReports              Max number of alignment hits to report
-                                        with the -k option
-        *     --reference               HDFS path to the bowtie 2 fasta
-                                        reference file
-
-    gemSingleEnds      Run a GEM alignment
-      Usage: gemSingleEnds [options]
-  Options:
-        *     --HDFSAlignmentsDir     HDFS directory to hold the alignment data
-        *     --HDFSDataDir           HDFS directory that holds the read data
-        *     --HDFSPathToGEM2SAM     HDFS path to the gem-2-sam executable
-        *     --HDFSPathToGEMMapper   HDFS path to the gem-mapper executable
-        *     --editDistance          Edit distance parameter (-e) to use in the
-                                      GEM mapping
-                                      Default: 0
-        *     --maxProcessesOnNode    Maximum number of GEM mapping processes to
-                                      run on one node simultaneously
-                                      Default: 6
-        *     --numReports            Max number of hits to report from GEM
-        *     --reference             HDFS path to the GEM reference file
-              --strata                Strata parameter (-s) to use in the GEM
-                                      mapping
-                                      Default: all
-
-    razerS3SingleEnds      Run a razerS3 alignment
-      Usage: razerS3SingleEnds [options]
-  Options:
-        *     --HDFSAlignmentsDir   HDFS directory to hold the alignment data
-        *     --HDFSDataDir         HDFS directory that holds the read data
-        *     --HDFSPathToRazerS3   HDFS path to the razers3 executable file
-        *     --numReports          Max number of alignments to report for each
-                                    read
-        *     --pctIdentity         RazerS 3 percent identity parameter (-i)
-                                    Default: 0
-        *     --reference           HDFS path to the reference (FASTA) file for
-                                    the RazerS 3 mapper
-        *     --sensitivity         RazerS 3 sensitivity parameter (-rr)
-                                    Default: 0
-
-    mrfastSingleEnds      Run a novoalign mate pair alignment
-      Usage: mrfastSingleEnds [options]
-  Options:
-        *     --HDFSAlignmentsDir   HDFS directory to hold the alignment data
-        *     --HDFSDataDir         HDFS directory that holds the read data
-        *     --HDFSPathToMrfast    HDFS path to the mrfast executable file
-        *     --reference           HDFS path to the mrfast reference index file
-              --threshold           MrFAST threshold parameter (-e)
-                                    Default: -1
-
-    exportAlignmentsFromHDFS      Export alignments in SAM format
-      Usage: exportAlignmentsFromHDFS [options]
-  Options:
-              --aligner        Format of the alignment records
-                               (sam|mrfast|novoalign)
-                               Default: sam
-        *     --inputHDFSDir   HDFS path to the directory holding the alignment
-                               reccords
-
-    GMMFitSingleEndInsertSizes      Compute GMM features in each bin across the genome
-      Usage: GMMFitSingleEndInsertSizes [options]
-  Options:
-              --aligner                            Format of the alignment
-                                                   records (sam|mrfast|novoalign)
-                                                   Default: sam
-              --chrFilter                          If filter params are used,
-                                                   only consider alignments in the
-                                                   region
-                                                   chrFilter:startFilter-endFilter
-              --endFilter                          See chrFilter
-              --excludePairsMappingIn              HDFS path to a BED file. Any
-                                                   reads mapped within those intervals
-                                                   will be excluded from the
-                                                   processing
-        *     --faidx                              HDFS path to the chromosome
-                                                   length file for the reference genome
-        *     --inputFileDescriptor                HDFS path to the directory
-                                                   that holds the alignment records
-              --legacyAlignments                   Use data generated with an
-                                                   older version of Cloudbreak
-                                                   Default: false
-              --mapabilityWeighting                HDFS path to a BigWig file
-                                                   containing genome uniqness scores. If
-                                                   specified, Cloudbreak will weight reads
-                                                   by the uniqueness of the regions
-                                                   they mapped to
-              --maxInsertSize                      Maximum insert size to
-                                                   consider (= max size of deletion
-                                                   detectable)
-                                                   Default: 25000
-              --maxLogMapqDiff                     Adaptive quality score cutoff
-                                                   Default: 5.0
-              --maxMismatches                      Max number of mismatches
-                                                   allowed in an alignment; all other
-                                                   will be ignored
-                                                   Default: -1
-              --minCleanCoverage                   Minimum number of spanning
-                                                   read pairs for a bin to run the
-                                                   GMM fitting procedure
-                                                   Default: 3
-              --minScore                           Minimum alignment score (SAM
-                                                   tag AS); all reads with lower AS
-                                                   will be ignored
-                                                   Default: -1
-        *     --outputHDFSDir                      HDFS path to the directory
-                                                   that will hold the output of the
-                                                   GMM procedure
-              --resolution                         Size of the bins to tile the
-                                                   genome with
-                                                   Default: 25
-              --startFilter                        See chrFilter
-              --stripChromosomeNamesAtWhitespace   Clip chromosome names from
-                                                   the reference at the first
-                                                   whitespace so they match with alignment
-                                                   fields
-                                                   Default: false
-
-    extractDeletionCalls      Extract deletion calls into a BED file
-      Usage: extractDeletionCalls [options]
-  Options:
-        *     --faidx                Chromosome length file for the reference
-        *     --inputHDFSDir         HDFS path to the GMM fit feature results
-              --medianFilterWindow   Use a median filter of this size to clean
-                                     up the results
-                                     Default: 5
-        *     --outputHDFSDir        HDFS Directory to store the variant calls
-                                     in
-              --resolution           Size of the bins to tile the genome with
-                                     Default: 25
-        *     --targetIsize          Mean insert size of the library
-                                     Default: 0
-        *     --targetIsizeSD        Standard deviation of the insert size of
-                                     the library
-                                     Default: 0
-              --threshold            Likelihood ratio threshold to call a
-                                     variant
-                                     Default: 1.68
-
-    extractInsertionCalls      Extract insertion calls into a BED file
-      Usage: extractInsertionCalls [options]
-  Options:
-        *     --faidx                Chromosome length file for the reference
-        *     --inputHDFSDir         HDFS path to the GMM fit feature results
-              --medianFilterWindow   Use a median filter of this size to clean
-                                     up the results
-                                     Default: 5
-              --noCovFilter          filter out calls next to a bin with no
-                                     coverage - recommend on for BWA alignments, off for
-                                     other aligners
-                                     Default: true
-        *     --outputHDFSDir        HDFS Directory to store the variant calls
-                                     in
-              --resolution           Size of the bins to tile the genome with
-                                     Default: 25
-        *     --targetIsize          Mean insert size of the library
-                                     Default: 0
-        *     --targetIsizeSD        Standard deviation of the insert size of
-                                     the library
-                                     Default: 0
-              --threshold            Likelihood ratio threshold to call a
-                                     variant
-                                     Default: 1.68
-
-    copyToS3      Upload a file to Amazon S3 using multi-part upload
-      Usage: copyToS3 [options]
-  Options:
-        *     --S3Bucket   S3 Bucket to upload to
-        *     --fileName   Path to the file to be uploaded on the local
-                           filesystem
-
-    launchCluster      Use whirr to create a new cluster in the cloud using 
-                       whirr/cloudbreak-whirr.properties
-      Usage: launchCluster [options]
-
-    runScriptOnCluster      Execute a script on one node of the currently running 
-                            cloud cluster
-      Usage: runScriptOnCluster [options]
-  Options:
-        *     --fileName   Path on the local filesystem of the script to run
-
-    destroyCluster      Destroy the currently running whirr cluster
-      Usage: destroyCluster [options]
-
-    summarizeAlignments      Gather statistics about a set of alignments: number of reads, 
-                             number of mappings, and total number of mismatches
-      Usage: summarizeAlignments [options]
-  Options:
-              --aligner        Format of the alignment records
-                               (sam|mrfast|novoalign)
-                               Default: sam
-        *     --inputHDFSDir   HDFS path of the directory that holds the
-                               alignments
-
-    exportGMMResults      Export wig files that contain the GMM features across 
-                          the entire genome
-      Usage: exportGMMResults [options]
-  Options:
-        *     --faidx          Local path to the chromosome length file
-        *     --inputHDFSDir   HDFS path to the directory holding the GMM
-                               features
-        *     --outputPrefix   Prefix of the names of the files to create
-              --resolution     Bin size that the GMM features were computed for
-                               Default: 25
-
-    dumpReadsWithScores      Dump all read pairs that span the given region with their 
-                             deletion scores to BED format (debugging)
-      Usage: dumpReadsWithScores [options]
-  Options:
-              --aligner                            Format of the alignment
-                                                   records (sam|mrfast|novoalign)
-                                                   Default: sam
-        *     --inputFileDescriptor                HDFS path to the directory
-                                                   that holds the alignment records
-              --maxInsertSize                      Maximum possible insert size
-                                                   to consider
-                                                   Default: 500000
-              --minScore                           Minimum alignment score (SAM
-                                                   tag AS); all reads with lower AS
-                                                   will be ignored
-                                                   Default: -1
-        *     --outputHDFSDir                      HDFS path to the directory
-                                                   that will hold the output
-        *     --region                             region to find read pairs
-                                                   for, in chr:start-end format
-              --stripChromosomeNamesAtWhitespace   Clip chromosome names from
-                                                   the reference at the first
-                                                   whitespace so they match with alignment
-                                                   fields
-                                                   Default: false
-
-    debugReadPairInfo      Compute the raw data that goes into the GMM fit procedure for 
-                           each bin (use with filter to debug a particular locus)
-      Usage: debugReadPairInfo [options]
-  Options:
-              --aligner                 Format of the alignment records
-                                        (sam|mrfast|novoalign)
-                                        Default: sam
-        *     --chrFilter               Print info for alignments in the region
-                                        chrFilter:startFilter-endFilter
-        *     --endFilter               see chrFilter
-              --excludePairsMappingIn   HDFS path to a BED file. Any reads
-                                        mapped within those intervals will be excluded
-                                        from the processing
-        *     --faidx                   HDFS path to the chromosome length file
-                                        for the reference genome
-        *     --inputFileDescriptor     HDFS path to the directory that holds
-                                        the alignment records
-              --mapabilityWeighting     HDFS path to a BigWig file containing
-                                        genome uniqness scores. If specified,
-                                        Cloudbreak will weight reads by the uniqueness of
-                                        the regions they mapped to
-              --maxInsertSize           Maximum insert size to consider (= max
-                                        size of deletion detectable)
-                                        Default: 500000
-              --minScore                Minimum alignment score (SAM tag AS);
-                                        all reads with lower AS will be ignored
-                                        Default: -1
-        *     --outputHDFSDir           HDFS directory to hold the output
-              --resolution              Size of the bins to tile the genome with
-                                        Default: 25
-        *     --startFilter             see chrFilter
-
-    findAlignment      Find an alignment record that matches the input string
-      Usage: findAlignment [options]
-  Options:
-        *     --HDFSAlignmentsDir   HDFS path to the directory that stores the
-                                    alignment data
-        *     --outputHDFSDir       HDFS path to the directory in which to put
-                                    the results
-        *     --read                Read name or portion of the read name to
-                                    search for
-
-    sortGMMResults      Sort and merge GMM Results (use with one reducer to get all GMM feature 
-                        results into a single file
-      Usage: sortGMMResults [options]
-  Options:
-        *     --inputHDFSDir    HDFS path to the directory holding the GMM
-                                features
-        *     --outputHDFSDir   Directory in which to put the results
-\end{verbatim}
-
+% \end{thebibliography}
 \end{document}

### 2014-01-13 manuscript/bioinfo/supplementary.tex
draft

--- 2013-07-08 manuscript/nm_supp.tex
@@
 \usepackage{xr}
+\usepackage{natbib}
 
@@
 
-\usepackage[style=nature,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
-\renewbibmacro{in:}{%
-  \ifentrytype{article}{}{%
-  \printtext{\bibstring{in}\intitlepunct}}}
-\addbibresource{manuscript.bib}
+%\usepackage[style=nature,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
+% \renewbibmacro{in:}{%
+%   \ifentrytype{article}{}{%
+%   \printtext{\bibstring{in}\intitlepunct}}}
+% \addbibresource{manuscript.bib}
+
+\newcommand{\degree}{\ensuremath{^\circ}}
 
@@
 
-\begin{figure}
+\tableofcontents
+
+\newpage
+
+\section{Supplementary Figures}
+
+\begin{figure}[h]
 \centering
-\includegraphics[width=1\textwidth]{../figures/insert_size_mixtures.pdf}
+\includegraphics[width=1\textwidth]{../../figures/insert_size_mixtures.pdf}
 \caption{Illustration of insert size mixtures at individual genomic locations.
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/cloudbreak_mapred_diagram.pdf}
+\includegraphics[width=1\textwidth]{../../figures/cloudbreak_mapred_diagram.pdf}
 \caption{An example of the Cloudbreak MapReduce algorithm.
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
+\includegraphics[width=1\textwidth]{../../figures/CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
 \caption{Cloudbreak performance on the chromosome 2 simulation using different alignment strategies.
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/runtimeByNodes.pdf}
+\includegraphics[width=1\textwidth]{../../figures/runtimeByNodes.pdf}
 \caption{Scalability of the Cloudbreak algorithm.
@@
 \centering
-\includegraphics[width=1\textwidth]{../figures/breakpoint_resolution.pdf}
+\includegraphics[width=1\textwidth]{../../figures/breakpoint_resolution.pdf}
 \caption{Breakpoint resolution for each tool for deletions on the Chromosome 2 simulated data.
@@
 
-\begin{suppalgorithm}[h]
-\algrenewcommand\algorithmicprocedure{\textbf{job}}
-  \begin{algorithmic}[1]
-    \Procedure{Alignment}{}
-    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
-    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
-    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
-    \EndFor
-    \EndFunction
-    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
-    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
-    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
-    \EndFunction
-    \EndProcedure
-
-    \Procedure{Compute SV Features}{}
-    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
-    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
-    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
-    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
-    \EndFor
-    \EndFor
-    \EndFunction
-    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
-    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
-    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
-    \EndFunction
-    \EndProcedure
-
-    \Procedure{Call SVs}{}
-    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
-    \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
-    \EndFunction
-    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
-    \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
-    \EndFunction
-    \EndProcedure
-  \end{algorithmic}
-\caption{The algorithmic framework for SV calling in MapReduce.}
-\label{cb_algo}
-\end{suppalgorithm}
+\begin{figure}
+\centering
+\includegraphics[width=.8\textwidth]{../../figures/runtime_by_windowSize.pdf}
+\caption[Cloudbreak runtimes with varying window sizes.]{Runtimes for Cloudbreak on the Chromosome 2 simulated data set using differing choices of window size.
+Runtimes include the feature computation and variant calling Cloudbreak Hadoop jobs.}
+\label{figure_runtime_by_window_size}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\textwidth]{../../figures/breakpoint_resolution_by_windowSize.pdf}
+\caption[Cloudbreak breakpoint resolution with varying window sizes.]{Breakpoint resolution for Cloudbreak on the Chromosome 2 simulated data set using differing choices of window size.}
+\label{breakpoint_resolution_by_windowSize}
+\end{figure}
 
@@
 
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{rrrrrr}
-  \hline
- & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
- Total Number & 224 &  84 & 82 &  31 & 26\\ 
-  \hline
-  Cloudbreak  & \textbf{68} (17)  & \textbf{67} (\textbf{10}) &  \textbf{56} (\textbf{5}) & \textbf{11} (\textbf{3}) & \textbf{15} (\textbf{0}) \\ 
-  Breakdancer & 52 (8)  & 49 (2) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
-  GASVPro     & 35 (2)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
-  DELLY-RP       & 22 (1)  & 56 (1) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
-  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
-  Pindel      & 60 (\textbf{32})  & 16 (0) &  41 (2) & 1 (0) & 12 (\textbf{0})\\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses).
-The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
-\label{chr2DeletionPredsFDR10}
-\end{table}
-
-\clearpage
+\section{Supplementary Tables}
 
@@
 
+\begin{table}
+\begin{center}
+\begin{tabular}{r|rrrrr}
+ \hline
+ Window Size & Calls & True Positives & Precision & Recall & F1 \\ 
+ \hline
+   1 & 274 & 228 & 0.832 & 0.57 & 0.677 \\ 
+   5 & 268 & 228 & 0.851 & 0.57 & 0.683 \\ 
+   10 & 258 & 223 & 0.864 & 0.557 & 0.678 \\ 
+   25 & 240 & 217 & 0.904 & 0.542 & 0.678 \\ 
+   50 & 215 & 194 & 0.902 & 0.485 & 0.631 \\ 
+   100 & 162 & 141 & 0.87 & 0.352 & 0.502 \\  
+\end{tabular}
+\end{center}
+\caption[Cloudbreak accuracy with varying window sizes.]{Accuracy measures for Cloudbreak on the Chromosome 2 simulated data set using different choices of window size.
+For each window size the same Cloudbreak score threshold was used (1.98).}
+\label{chr2AccuracyByWindowSize}
+\end{table}
+
+\clearpage
+
 \section{Supplementary Discussion}
 
-\subsection{A general framework for SV Detection in MapReduce}
-
-We have developed a conceptual algorithmic framework for SV detection in MapReduce, which is outlined in Supplementary Algorithm \ref{cb_algo}.
-As described in the main text, this framework divides processing into three separate MapReduce jobs: an alignment job, a feature computation job, and an SV calling job.
-
-The alignment job uses sensitive mapping tools and settings to discover mapping locations for each read pair.
-Aligners can be executed to report multiple possible mappings for each read, or only the best possible mapping.
-Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits possible mapping locations under the $rpid$ key.
-Reducers then collect the alignments for each paired end, making them available under one key for the next job.
-Our implementation contains wrappers to execute the aligners BWA, GEM \autocite{MarcoSola:2012hm}, Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.
-
-In the second job, we compute a set of features for each location in the genome.
-To begin, we tile the genome with small fixed-width, non-overlapping intervals.
-For the experiments reported in this paper we use an interval size of 25bp.
-Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome.
-Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads.
-Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$.
-For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g.
-the fragment size implied by the alignments and the likelihood the alignments are correct.
-We then leave two functions to be implemented depending on the application:
-\begin{flalign*}
- \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
- \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
-\end{flalign*}
-
-The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
-We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
-The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
-
-Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location.
-It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.
-We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.
-
-\subsection{Cloudbreak: An Implementation of the MapReduce SV Detection Framework}
-
-Our Cloudbreak software package can be seen as an implementation of the general framework defined above.
-In particular, Cloudbreak implements the three user-defined function described above as follows:
-
-\begin{description}
-\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads.
-For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
-\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure \ref{insert_size_mixes}).
-Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
-If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant.
-Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components.
-Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.
-
-To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
-We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
-In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
-This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
-Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
-
-\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
-
-And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
+\subsection{Fitting a GMM to the Observed Overlapping Insert Sizes}
 
@@
 
-The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.
-
-\item[\sc{PostProcess}] We convert our features along each chromosome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold.
-To eliminate noise we apply a median filter with window size 5.
-We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
-\end{description}
+\subsection{Filtering Incorrect and Ambiguous Mappings}
+
+To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location.
+We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor, using differences between insert sizes as a measure of distance, is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
+In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
+This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
+Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
+
+\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]
+
+And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
+
+\subsection{Simulated and Biological Data Sets}\label{section_eval_data_sets}
+
+As has been observed elsewhere, there is no available test set of real Illumina sequencing data from a sample that has a complete annotation of SVs from the reference.
+Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics.
+On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data.
+To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous insertions and deletions from the genome of J.
+Craig Venter~\citep{Levy:2007fb}.
+Using these variants, we simulated a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants.
+Since there are relatively few heterozygous insertions and deletions annotated in the Venter genome, we used the set of homozygous indels contained in the HuRef data (\texttt{HuRef.homozygous\_indels.061109.gff}) and randomly assigned each variant to be either homozygous or heterozygous with uniform probability.
+Based on this genotype, we applied each variant to one or both of two copies of the human GRCh36 chromosome 2 reference sequence.
+We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package (\url{http://sourceforge.net/apps/mediawiki/dnaa/}).
+We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence.
+Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous insertions and deletions.
+
+We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive.
+This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage.
+
+To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same sample.
+These included data from the Human Genome Structural Variation Project reported by Kidd et al.~\citep{Kidd:2008p926}, a survey of small indels conducted by Mills et al.~\citep{Mills:2011fi}, and insertions and deletions from the merged call set of the Phase 1 release of the 1000 Genomes Project~\citep{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507.
+We merged any overlapping calls of the same type into the region spanned by their unions.
+It should be noted that the 1000 Genomes call set was partially produced using DELLY and BreakDancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor.
+
+\subsection{Parameters Used for Alignment and SV Detection}
+
+We aligned simulated reads to hg18 chromosome 2, NA18507 reads to the hg19 assembly.
+Alignments for all programs, unless otherwise noted, were found using BWA \texttt{aln} version 0.6.2-r126, with parameter \texttt{-e 5} to allow for longer gaps in alignments due to the number of small indels near the ends of larger indels in the Venter data set.
+We also tested the effect of including multiple possible mapping locations for ambiguously mapped reads in results reported in Supplementary Figure~\ref{alignment_comparison}.
+For those tests, we used two different sets of reads with multiple mapping locations reported.
+The first used alignments generated with BWA in paired-end mode, reporting up to 25 additional hits for each mapping  using the \texttt{-n} and \texttt{-N} parameters for \texttt{bwa sampe} and the script \texttt{xa2multi.pl}.
+For the second, we attempted to generate an exhaustive list of possible mapping locations by running the GEM aligner in single-ended mode on each read in each pair individually, reporting up to 1000 additional hits per alignment.
+GEM was executed in parallel using Hadoop tasks which wrap GEM version 1.362 (beta), with parameters \texttt{-e 6 -m 6 -s 2 -q ignore -d 1000 --max-big-indel-length 0}.
+These parameters request all hits for a read that are within an edit distance of 6 of the reference, within 2 strata of the best hit, with a maximum of 1000 possible alignments reported for each read.
+GASVPro also accepts ambiguous mappings but expects them to be realigned with a more sensitive alignment tool; following the details given in their manuscript we extracted read pairs that did not align concordantly with BWA and re-aligned them with Novoalign V2.08.01, with parameters \texttt{-a -r -Ex 1100 -t 250}.
+
+We ran Cloudbreak with a median filter window size of 5 genomic windows.
+
+We ran BreakDancer version 1.1\_2011\_02\_21 in single threaded mode by first executing \texttt{bam2cfg.pl} and then running \texttt{breakdancer\_max} with the default parameter values.
+To run BreakDancer in parallel mode we first ran \texttt{bam2cfg.pl} and then launched parallel instances of \texttt{breakdancer\_max} for each chromosome using the \texttt{-o} parameter.
+We ran DELLY version 0.0.9 with the \texttt{-p} parameter and default values for other parameters.
+For the parallel run of DELLY we first split the original BAM file with BamTools \citep{Barnett:2011hm}, and then ran instances of DELLY in parallel for each BAM file.
+We ran GASVPro version 1.2 using the \texttt{GASVPro.sh} script and default parameters.
+Pindel 0.2.4t was executed with default parameters in single CPU mode, and executed in parallel mode for each chromosome using the \texttt{-c} option.
+We executed MoDIL with default parameters except for a \texttt{MAX\_DEL\_SIZE} of 25000, and processed it in parallel on our cluster with a step size of 121475.
+To execute other SV detection tools in parallel we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine \citep{condor-practice} with directed acyclic graphs to describe dependencies between jobs.
+
+\subsection{SV Prediction Evaluation}
+
+We use the following criteria to define a true prediction given a gold standard set of deletion and insertion variants to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not already been discovered by another prediction from the same method.
+For evaluating insertions, each algorithm produces insertion predictions that define an interval in which the insertion is predicted to have occurred with start and end coordinates $s$ and $e$ as well as the predicted length of the insertion, $l$.
+The true insertions are defined in terms of their actual insertion coordinate $i$ and their actual length $l_a$.
+Given this information, we modify the overlap criteria in a) to include overlaps of the intervals $\langle s,\max{\left(e,s+l\right)} \rangle$ and $\langle i,i+l_a \rangle$.
+In this study we are interested in detecting events larger than 40bp, because with longer reads, smaller events can be more easily discovered by examining gaps in individual reads.
+Both Pindel and MoDIL make many calls with a predicted event size of under 40bp, so we remove those calls from the output sets of those programs.
+Finally, we exclude from consideration calls from all approaches that match a true deletion of less than 40bp where the predicted variant length is less than or equal to 75bp in length.
 
@@
 We report runtimes for tools run in their default single-threaded mode, as well as for levels of parallelization achievable with basic scripting, noting that one of the key advantages of Hadoop/MapReduce is the ability to scale parallel execution to the size of the available compute cluster without any custom programming.
+To execute the tools in parallel mode we wrote simple scripts to submit jobs to the cluster using the HTCondor scheduling engine (\url{http://research.cs.wisc.edu/htcondor/}) with directed acyclic graphs to describe dependencies between jobs.
 Pindel allows multi-threaded operation on multicore servers.
@@
 
-\printbibliography[title=Supplementary References]
+\subsection{Choice of Window Size}\label{section_window_size}
+
+A key parameter choice for Cloudbreak is the size of the fixed-width, non-overlapping windows used for local feature computation.
+The experiments reported in the main paper used a window size of 25bp.
+Using the simulated data set, we evaluated the effect of choosing differing window sizes on runtime, breakpoint resolution, and accuracy.
+Supplementary Figure~\ref{figure_runtime_by_window_size} shows the runtime of the feature computation and variant calling steps of Cloudbreak on the Chromosome 2 data set using differing window sizes.
+Runtime decreases dramatically with larger window sizes.
+For very small window sizes, the runtime is dominated by the variant extraction job, which is only parallelized by chromosome and therefore only runs on one core when running on the simulated data set.
+Table~\ref{chr2AccuracyByWindowSize} shows several accuracy measures for Cloudbreak running with varying window sizes, using the same threshold for each run.
+The greatest recall is achieved at very low window sizes, probably because evidence for smaller variants is less likely to be mixed in with non-variant supporting read pairs from the flanking regions.
+However, overall accuracy is high until window sizes reach 50bp, at which point recall decreases dramatically.
+Finally, we examined the effect of breakpoint accuracy on window size and found that it had little effect (Supplementary Figure~\ref{breakpoint_resolution_by_windowSize}).
+
+\subsection{Preparation of AML Sample}
+
+Peripheral blood was collected from a patient with acute myelomonocytic leukemia (previously designated acute myeloid leukemia FAB M4) under a written and oral informed consent process reviewed and approved by the Institutional Review Board of Oregon Health \& Science University.
+Known cytogenetic abnormalities associated with this specimen included trisomy 8 and internal tandem duplications within the FLT3 gene.
+Mononuclear cells were separated on a Ficoll gradient, followed by red cell lysis.
+Mononuclear cells were immunostained using antibodies specific for CD3, CD14, CD34, and CD117 (all from BD Biosciences) and cell fractions were sorted using a BD FACSAria flow cytometer.
+Cell fractions isolated included T-cells (CD3-positive), malignant monocytes (CD14-positive), and malignant blasts (CD34, CD117-positive).
+We sequenced CD14+ cells on an Illumina Genome Analyzer II, producing 128,819,200 paired-end reads.
+
+\subsection{Validation of AML Deletions by PCR}
+
+Deletions identified in the AML dataset were validated by PCR.
+SVs to validate were selected based on calls made by BreakDancer, relying on the score and the number of the reads supporting each single event.
+Appropriate primers were designed with an internet-based interface, Primer3 (\url{http://frodo.wi.mit.edu/}), considering the chromosome localization and orientation of the interval involved in the candidate rearrangement.
+The primers were checked for specificity using the BLAT tool of the UCSC Human Genome Browser (\url{http://genome.ucsc.edu/cgi-bin/hgBlat}).
+All the primer pairs were preliminarily tested on the patient genomic DNA and a normal genomic DNA as control.
+The PCR conditions were as follows: 2 min at 95\degree C followed by 35 cycles of 30 sec at 95\degree C, 20 sec at 60\degree C, and 2 min at 72\degree C.
+All the obtained PCR products were sequenced, and analyzed by BLAT for sequence specificity.
+
+\bibliographystyle{natbib}
+%\bibliographystyle{achemnat}
+%\bibliographystyle{plainnat}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{bioinformatics}
+%
+%\bibliographystyle{plain}
+%
+\bibliography{Document}
 

### 2014-01-13 manuscript/bioinfo/document.tex
update tables

@@
 \footnotesize
-\begin{tabular}{r|rrr|rrrrr}
-  \cline{2-9}
-   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\begin{tabular}{r|rrrr|rrrrr}
+  \cline{2-10}
+   &                     & Prec. & Recall & F1 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 \hline
-\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
+\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           &  &  224 &  84 & 82 &  31 & 26\\ 
   \hline
-\cline{2-9}
-&  Cloudbreak    &  0.638 & \textbf{0.678} & \textbf{153} (9)  & 61 (0) &  62 (0) & 12 (0) & 15 (0) \\ 
-&  BreakDancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
-&  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
-&  DELLY-RP           & 0.457 & 0.613 & 114 (3)  & \textbf{68} (0) &  \textbf{66} (0) & 9 (1) & 17 (0) \\ 
-&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
-&  Pindel           & 0.462 & 0.421 & 96 (\textbf{11})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
-&  MoDIL           & 0.132  & 0.66 & 123 (6)  & 66 (\textbf{3}) &  \textbf{66} (\textbf{11}) & \textbf{17} (\textbf{7}) & \textbf{23} (\textbf{8})\\ 
+\cline{2-10}
+&  Cloudbreak    &  0.638 & \textbf{0.678} & \textbf{0.657} & \textbf{153} (9)  & 61 (0) &  62 (0) & 12 (0) & 15 (0) \\ 
+&  BreakDancer   &  0.356 & 0.49 & 0.412 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
+&  GASVPro        & 0.146 & 0.432 & 0.218 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
+&  DELLY-RP           & 0.457 & 0.613 & 0.613 & 114 (3)  & \textbf{68} (0) &  \textbf{66} (0) & 9 (1) & 17 (0) \\ 
+&  DELLY-SR           & \textbf{0.679} & 0.166 &  0.266 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
+&  Pindel           & 0.462 & 0.421 &  0.44 & 96 (\textbf{11})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
+&  MoDIL           & 0.132  & 0.66 & 0.22 & 123 (6)  & 66 (\textbf{3}) &  \textbf{66} (\textbf{11}) & \textbf{17} (\textbf{7}) & \textbf{23} (\textbf{8})\\ 
    \hline
-\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
-\cline{2-9}
-&  Cloudbreak   &0.451 & \textbf{0.305}  & \textbf{79} (\textbf{32})  & \textbf{32} (\textbf{18}) &  \textbf{11} (8) & 1 (0) & 0 (0) \\ 
-&  BreakDancer & 0.262 & 0.0968  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
-&  Pindel          & \textbf{0.572} & 0.196 & 52 (25)  & 5 (1) &  10 (\textbf{9}) & \textbf{3} (\textbf{2}) & \textbf{9} (\textbf{9})\\ 
-&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (0) &  1 (0) & 2 (\textbf{2}) & 0 (0)\\ 
+\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & & 199 &  83 & 79 &  21 & 21\\ 
+\cline{2-10}
+&  Cloudbreak   &0.451 & \textbf{0.305}  & \textbf{0.364}  & \textbf{79} (\textbf{32})  & \textbf{32} (\textbf{18}) &  \textbf{11} (8) & 1 (0) & 0 (0) \\ 
+&  BreakDancer & 0.262 & 0.0968  & 0.141  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
+&  Pindel          & \textbf{0.572} & 0.196 &  0.292 & 52 (25)  & 5 (1) &  10 (\textbf{9}) & \textbf{3} (\textbf{2}) & \textbf{9} (\textbf{9})\\ 
+&  MoDIL          & 0.186 & 0.0521 &  0.0814 & 14 (1)  & 4 (0) &  1 (0) & 2 (\textbf{2}) & 0 (0)\\ 
 \hline
@@
 \footnotesize
-\begin{tabular}{r|rrr|rrrrr}
-  \cline{2-9}
-&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
+\begin{tabular}{r|rrrr|rrrrr}
+  \cline{2-10}
+&  & Prec. & Recall & F1 & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 \hline
-\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,462 & 240 & 232 & 147 & 540 \\
+\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & & 7,462 & 240 & 232 & 147 & 540 \\
   \hline
-\cline{2-9}
-& Cloudbreak & 0.0943 & \textbf{0.17} & \textbf{573} (\textbf{277})  & \textbf{176} (\textbf{30}) &  \textbf{197} (\textbf{18}) & \textbf{121} (\textbf{6}) & \textbf{399} (\textbf{24}) \\ 
-& BreakDancer & 0.137 & 0.123 & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
-&  GASVPro & 0.147 & 0.0474 & 120 (21)  & 40 (2) &  85 (0) & 36 (0) & 128 (0) \\ 
-&  DELLY-RP & 0.0931 & 0.1 & 143 (6)  & 128 (3) &  167 (1) & 103 (0) & 323 (1) \\ 
-&  DELLY-SR & 0.153 & 0.0485 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
-&  Pindel & \textbf{0.179} & 0.0748 & 149 (8)  & 61 (0) &  149 (0) & 69 (1) & 217 (0) \\ 
+\cline{2-10}
+& Cloudbreak & 0.0943 & \textbf{0.17} & 0.121 & \textbf{573} (\textbf{277})  & \textbf{176} (\textbf{30}) &  \textbf{197} (\textbf{18}) & \textbf{121} (\textbf{6}) & \textbf{399} (\textbf{24}) \\ 
+& BreakDancer & 0.137 & 0.123 &  \textbf{0.13} & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
+&  GASVPro & 0.147 & 0.0474 &  0.0717 & 120 (21)  & 40 (2) &  85 (0) & 36 (0) & 128 (0) \\ 
+&  DELLY-RP & 0.0931 & 0.1 &  0.0965 & 143 (6)  & 128 (3) &  167 (1) & 103 (0) & 323 (1) \\ 
+&  DELLY-SR & 0.153 & 0.0485 &  0.0736 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
+&  Pindel & \textbf{0.179} & 0.0748 &  0.106 & 149 (8)  & 61 (0) &  149 (0) & 69 (1) & 217 (0) \\ 
 \hline
-\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 & 0 \\
-\cline{2-9}
-& Cloudbreak & 0.0323 & \textbf{0.455} & \textbf{265} (\textbf{104})  & \textbf{49} (\textbf{24}) &  3 (1) & 0 (0)  & 0 (0)  \\ 
-& BreakDancer & 0.0281 & 0.181 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
-&  Pindel & \textbf{0.0387} & 0.239 & 144 (45)  & 14 (7) &  \textbf{7} (\textbf{6}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
+\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & & 536 & 114 & 45 & 1 & 0 \\
+\cline{2-10}
+& Cloudbreak & 0.0323 & \textbf{0.455} &  0.0604 & \textbf{265} (\textbf{104})  & \textbf{49} (\textbf{24}) &  3 (1) & 0 (0)  & 0 (0)  \\ 
+& BreakDancer & 0.0281 & 0.181 &  0.0487 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
+&  Pindel & \textbf{0.0387} & 0.239 &  \textbf{0.0666} & 144 (45)  & 14 (7) &  \textbf{7} (\textbf{6}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
 \hline

### 2014-01-24 manuscript/bioinfo/document.tex
comments from kemal

@@
 
-\title[Cloudbreak]{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
+\title[Cloudbreak]{Cloudbreak: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
 \author[Whelan \textit{et~al}]{Christopher W.
 Whelan\,$^{1}$, Jeffrey Tyner\,$^{2}$, Alberto L'Abbate\,$^{3}$, Clelia Tiziana Storlazzi,$^{3}$, Lucia Carbone\,$^{4}$, and Kemal S\"onmez\,$^{1,5,*}$\footnote{to whom correspondence should be addressed}}
-\address{$^{1}$Institute on Development and Disability and Center for Spoken Language Understanding, Oregon Health \& Science University, Portland, OR, USA \\
-$^{2}$ Program in Molecular and Cellular Biosciences, Oregon Health \& Science University, Portland, OR, USA \\
+\address{$^{1}$Institute on Development and Disability and Center for Spoken Language Understanding, Oregon Health \& Science University (OHSU), Portland, OR, USA \\
+$^{2}$ Program in Molecular and Cellular Biosciences, OHSU\todo{Oregon Health \& Science University}, Portland, OR, USA \\
 $^{3}$ Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy \\
-$^{4}$ Behavioral Neuroscience Department, Oregon Health \& Science University, Portland, OR, USA \\
-$^{5}$ Department of Medical Informatics \& Clinical Epidemiology, Oregon Health \& Science University, Portland, OR, USA 
+$^{4}$ Behavioral Neuroscience Department, OHSU\todo{Oregon Health \& Science University}, Portland, OR, USA \\
+$^{5}$ Department of Medical Informatics \& Clinical Epidemiology, OHSU\todo{Oregon Health \& Science University}, Portland, OR, USA 
 }
@@
 \section{Availability:}
-Source code, binary releases, example scripts, and a full user manual are available at \url{http://github.com/cwhelan/cloudbreak}
+Source code, binary releases, and user manual are available at \url{http://github.com/cwhelan/cloudbreak}
+
+\section{Supplementary information:} Supplementary material is available at \emph{Bioinformatics} online.
 
@@
 \centering
-\includegraphics[width=.5\textwidth]{../../figures/workflow_with_whirr.png}
+\includegraphics[width=.5\textwidth]{workflow_with_whirr.png}
 \caption{An overview of the Cloudbreak workflow.
@@
 Cloudbreak detects deletions of size 40bp to 25kb, and small insertions.
-We accomplish this by providing the following implementions of the user-defined functions described above:
+We accomplish this by providing the following implementations of the user-defined functions described above:
 
@@
 \centering
-\includegraphics[width=1\textwidth]{../../figures/chr2_rocs_runtimes_journal.pdf}
+\includegraphics[width=1\textwidth]{chr2_rocs_runtimes_journal.pdf}
 \caption{Accuracy and runtime performance on a simulated data set.
@@
 See Supplementary Material for a description of how we identified correct predictions.
-All approaches show high specificity at stringent thresholds in this simulation.
-Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY.
+All approaches show high specificity to deletions at stringent thresholds in this simulation.
+When specificity is relaxed slightly, Cloudbreak provides the highest levels of sensitivity, followed by DELLY.
 For insertions, Cloudbreak's performance is similar to Pindel, although Cloudbreak can achieve greater sensitivity at the same false positive rate by decreasing its threshold.
@@
 
-It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Fig.
-\ref{Sbreakpoint_resolution}).
-SR methods have higher resolution than RP methods.
-Cloudbreak's resolution suffers because the independent calculation of features at each window from the distribution of overlapping insert sizes does not allow the algorithm to keep track of the actual coordinates of the mapped reads from each pair, which other tools use to set bounds on the true breakpoint locations.
-We believe, however, that it is worth increasing sensitivity and specificity in SV detection tools even at the cost of breakpoint resolution.
-\todo{It would also be possible to extend our feature set to include information useful for increasing resolution, such as the location of nearby read mappings or soft-clipped alignments.
-}In addition, several recent pipelines have validated RP calls \emph{in silico} by local assembly, a process that could potentially be parallelized in a subsequent MapReduce job.
+We designed Cloudbreak to optimize for sensitivity and specificity for SV events, as demonstrated in the ROC curves, as opposed to attempting to achieve high levels of breakpoint resolution.
+Supplementary Fig.
+\ref{Sbreakpoint_resolution} displays the breakpoint resolution of each of the methods tested on the simulated data set.
+Cloudbreak's resolution is lower than other methods because the independent calculation of features at each window does not allow the algorithm to keep track of the actual coordinates of the mapped reads from each pair, which other tools use to set bounds on the true breakpoint locations.
+In practice, RP-based predictions are typically validated using techniques such as targeted split read search, local assembly, or wet-lab methods, processes which refine their breakpoint predictions.
+Therefore, Cloudbreak attempts to maximize the accurate detection of the presence of SV events.
 
@@
 \centering
-\includegraphics[width=1\textwidth]{../../figures/NA18507_rocs_runtimes_journal.pdf}
+\includegraphics[width=1\textwidth]{NA18507_rocs_runtimes_journal.pdf}
 \caption{Accuracy and performance on the 37X NA18507 sample.
@@
 
-We also tested Cloudbreak on a sequencing data set obtained from a patient with acute myeloid leukemia (AML).
-This data set consisted of 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
+We also tested Cloudbreak on a data set obtained from a patient with acute myeloid leukemia (AML).
+This data consisted of 76bp paired end reads with a mean insert size of 285bp and standard deviation of 50bp, yielding sequence coverage of 5X and physical coverage of 8X.
 Using a pipeline consisting of Novoalign, BreakDancer, and a set of custom scripts for filtering and annotating candidate SVs, we had previously identified a set of variants present in this sample and validated several using PCR, including 8 deletions.
-Cloudbreak was able to identify all 8 of the validated deletions, showing that it is still sensitive to variants even when using lower coverage data sets with a greater variance of insert sizes.
-The variants identified include deletions in the gene CTDSPL/RBPS3, an AML tumor suppressor~\citep{Zheng:2012kk}, and NBEAL1, a gene up-regulated in some cancers~\citep{Chen:2004jo}.
-We are currently investigating these deletions to determine their functional impact on this patient.
+Cloudbreak was able to identify all of the validated deletions, showing that it is still sensitive even when using lower coverage data with a greater variance of insert sizes.
+The variants identified include deletions in the gene \emph{CTDSPL}/\emph{RBPS3}, an AML tumor suppressor~\citep{Zheng:2012kk}, and NBEAL1, a gene up-regulated in some cancers~\citep{Chen:2004jo}.
+We are currently investigating these SVs to determine their functional impact on this patient.
 
@@
 Over the next few years, due to advances in sequencing technology, genomics data are expected to increase in volume by several orders of magnitude, expanding into the realm referred to as ``big data''.
-In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines.
+In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and data is reprocessed and integrated into new pipelines.
 In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.

### 2014-01-24 manuscript/bioinfo/supplementary.tex
comments from kemal

@@
 
-\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\
+\title{Cloudbreak: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\
 Supplementary Material}
@@
 \centering
-\includegraphics[width=1\textwidth]{../../figures/insert_size_mixtures.pdf}
+\includegraphics[width=1\textwidth]{insert_size_mixtures.pdf}
 \caption{Illustration of insert size mixtures at individual genomic locations.
@@
 \centering
-\includegraphics[width=1\textwidth]{../../figures/cloudbreak_mapred_diagram.pdf}
+\includegraphics[width=1\textwidth]{cloudbreak_mapred_diagram.pdf}
 \caption{An example of the Cloudbreak MapReduce algorithm.
@@
 \centering
-\includegraphics[width=1\textwidth]{../../figures/CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
+\includegraphics[width=1\textwidth]{CHR2SIM_ROCS_MULTIPLE_MAPPINGS.pdf}
 \caption{Cloudbreak performance on the chromosome 2 simulation using different alignment strategies.
@@
 \centering
-\includegraphics[width=1\textwidth]{../../figures/runtimeByNodes.pdf}
+\includegraphics[width=1\textwidth]{runtimeByNodes.pdf}
 \caption{Scalability of the Cloudbreak algorithm.
@@
 \centering
-\includegraphics[width=1\textwidth]{../../figures/breakpoint_resolution.pdf}
+\includegraphics[width=1\textwidth]{breakpoint_resolution.pdf}
 \caption{Breakpoint resolution for each tool for deletions on the Chromosome 2 simulated data.
@@
 \centering
-\includegraphics[width=.8\textwidth]{../../figures/runtime_by_windowSize.pdf}
+\includegraphics[width=.8\textwidth]{runtime_by_windowSize.pdf}
 \caption[Cloudbreak runtimes with varying window sizes.]{Runtimes for Cloudbreak on the Chromosome 2 simulated data set using differing choices of window size.
@@
 \centering
-\includegraphics[width=.8\textwidth]{../../figures/breakpoint_resolution_by_windowSize.pdf}
+\includegraphics[width=.8\textwidth]{breakpoint_resolution_by_windowSize.pdf}
 \caption[Cloudbreak breakpoint resolution with varying window sizes.]{Breakpoint resolution for Cloudbreak on the Chromosome 2 simulated data set using differing choices of window size.}
@@
 We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor, using differences between insert sizes as a measure of distance, is more than $n\sigma$ distant, where $k = 3$ and $n = 5$.
-In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$.
+The parameter $n=5$ represents the choice that only extreme outlier sets are to be eliminated and the parameter $k=3$ represents a trade-off in selecting size of the outlier set between eliminating true modes and allowing small outlier groups to skew the inference, which has been found by observing the distribution of outliers in data.
+
+In addition, when the input data contains multiple possible mappings for each read pair, we calculate a score based on the probability that each mapping is correct relative to the other possible mappings.
+We then rank all observations at each genomic location, which in most cases include mappings from multiple read pairs, by their scores and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the score is less than the maximum score observed at that location minus a constant $c$.
 This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.
-Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:
+We compute mapping scores as follows: defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we compute the score $p^{k}_c$ of each end alignment by:
 
@@
 
-And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.
+Where $k$ is the position of the read in the pair (i.e.
+1,2), $m$ is the identifier of the read pair, and $i$ is the index of a particular possible mapping in the set of all possible mappings for that read.
+We then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to give a score for that read pair.
+This score calculation was designed to give higher scores to the set of alignments with the smallest number of mismatches, while dividing smaller amounts of score among possible alignments with more mismatches than the best alignment in the set.
+When the data only contains one possible mapping for each read pair, as is the case in the experiments reported in the main manuscript, all read pairs receive a score of 1 under this scheme and therefore the adaptive quality cutoff is not invoked.
 
@@
 
-We ran Cloudbreak with a median filter window size of 5 genomic windows.
+We ran Cloudbreak with a median filter window size of 5 genomic windows, an adaptive quality cutoff of 6, and a minimum coverage after filtering at each window of 3 spanning read pairs (parameters \texttt{--maxLogMapqDiff 6 --minCleanCoverage 3} on the feature generation step and \texttt{--medianFilterWindow 5} on the variant extraction steps).
+We chose these parameters based on limited tuning on the simulated data set.
 

### 2014-01-24 manuscript/bioinfo/document.tex
additional rewrites

@@
 
-\externaldocument[S]{supplementary}
+%\externaldocument[S]{supplementary}
 
@@
 
-\textbf{$\Phi$}: To compute features for each genomic location, we follow \cite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure \ref{Sinsert_size_mixes}).
+\textbf{$\Phi$}: To compute features for each genomic location, we follow \cite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure 1\todo{\ref{Sinsert_size_mixes}}).
 Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$.
@@
 
-An illustration of the Cloudbreak algorithm working on a simple example is shown in Supplementary Figure \ref{Salgorithm_example}.
+An illustration of the Cloudbreak algorithm working on a simple example is shown in Supplementary Figure 2\todo{\ref{Salgorithm_example}}.
 
@@
 For all alignments we used BWA~\citep{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Fig.
-\ref{Salignment_comparison}).
+3\todo{\ref{Salignment_comparison}}).
 
@@
 All approaches show high specificity to deletions at stringent thresholds in this simulation.
-When specificity is relaxed slightly, Cloudbreak provides the highest levels of sensitivity, followed by DELLY.
+If specificity is relaxed slightly, Cloudbreak provides the highest levels of sensitivity, followed by DELLY.
 For insertions, Cloudbreak's performance is similar to Pindel, although Cloudbreak can achieve greater sensitivity at the same false positive rate by decreasing its threshold.
@@
 (Of course, Cloudbreak uses many more CPUs as distributed algorithm.
-See Supplementary Material for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster.
+See Supplementary Material and Supplementary Table 1 for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs.
 Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow.
-To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Fig.
-\ref{Sscalability}).
+To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes until a point of diminishing returns is reached (Supplementary Fig.
+4\todo{\ref{Sscalability}}).
 
@@
 We also used the ROC curves to characterize algorithm performance when a low false discovery rate is required.
-Supplementary Table \ref{Schr2DeletionPredsFDR10} shows the total number of deletions found when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
+Supplementary Table 2\todo{\ref{Schr2DeletionPredsFDR10}} shows the total number of deletions found when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve.
 At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool.
@@
 Supplementary Fig.
-\ref{Sbreakpoint_resolution} displays the breakpoint resolution of each of the methods tested on the simulated data set.
+5\todo{\ref{Sbreakpoint_resolution}} displays the breakpoint resolution of each of the methods tested on the simulated data set.
 Cloudbreak's resolution is lower than other methods because the independent calculation of features at each window does not allow the algorithm to keep track of the actual coordinates of the mapped reads from each pair, which other tools use to set bounds on the true breakpoint locations.
@@
 Exclusive predictions are in parentheses.
-For deletions, the same cutoffs were used as for the simulated data as in Supplementary Table \ref{Schr2DeletionPredsFDR10}, adjusted for the difference in coverage from 30X to 37X.
+For deletions, the same cutoffs were used as for the simulated data as in Supplementary Table 2\todo{\ref{Schr2DeletionPredsFDR10}}, adjusted for the difference in coverage from 30X to 37X.
 For insertions, the maximum sensitivity cutoff was used.}
@@
 Cloudbreak was able to achieve 92.7\% and 95.9\% accuracy in predicting the genotype of the deletions it detected at our 10\% FDR threshold in the simulated and real data sets, respectively.
-Supplementary Table \ref{SdeletionGenotypeaccuracy} shows confusion matrices for the two samples using this classifier.
+Supplementary Table 3\todo{\ref{SdeletionGenotypeaccuracy}} shows confusion matrices for the two samples using this classifier.
 \todo{None of the three input sets that made up the gold standard for NA18507 contained a sufficient number of insertions that met our size threshold and also had genotyping information.

### 2014-01-24 manuscript/bioinfo/document.tex
remove most todos

@@
 MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention.
-The Hadoop framework has been successfully applied to sequencing-related tasks including short read mapping~\citep{Schatz:2009p278} and SNP calling~\citep{Langmead:2009p1225}\todo{, and RNA-seq differential expression analysis~\citep{Langmead:2010p1268}}.
+The Hadoop framework has been successfully applied to sequencing-related tasks including short read mapping~\citep{Schatz:2009p278} and SNP calling~\citep{Langmead:2009p1225}.
 
@@
 The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments.
-\todo{This} We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
+We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair.
 The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.
@@
 First, we characterized the predictions made by each algorithm at the threshold that gives them maximum sensitivity.
-\todo{For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low.} The results are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}.
+The results are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}.
 MoDIL and Cloudbreak exhibited the greatest recall for deletions.
@@
 At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool.
-Insertion performance never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.\todo{ We also examined Cloudbreak's ability to detect events in repetitive regions of the genome, and found that it was similar to the other methods tested (Supplementary Tables \ref{SdeletionRepmaskpreds} and \ref{SinsertionRepmaskpreds}).}
+Insertion performance never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table.
 
@@
 Detection of insertions and deletions is an important area of research; \cite{Mills:2011fi} recently identified over 220 coding deletions in a survey of a large number of individuals, and they note that such variants are likely to cause phenotypic variation in humans.
-\todo{More SV annotations are also required to assess the impact of SVs on non-coding elements~\citep{Mu:2011br}.}
 

### 2014-01-24 manuscript/bioinfo/supplementary.tex
additional rewrites
remove most todos

@@
 % \addbibresource{manuscript.bib}
+
+\newcommand{\todo}[1]{}
 
@@
 \begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\hline
- &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 120 & 327 & 562 & 8059 \\ 
-  \hline
-  Cloudbreak  & \textbf{37} (\textbf{10}) & \textbf{180} (\textbf{25}) & \textbf{300} (\textbf{85}) & \textbf{1166} (\textbf{270}) \\ 
-  Breakdancer & 29 (7) & 142 (3) & 192 (11) & 868 (21) \\
-  GASVPro     & 16 (1) & 79 (1) & 79 (7) & 330 (16) \\
-  DELLY-RP       & 21 (1) & 117 (1) & 152 (6) & 712 (5) \\
-  DELLY-SR       & 0 (0) & 42 (0) & 27 (0) & 391 (0) \\
-  Pindel      & 18 (9) & 112 (\textbf{25}) & 109 (2) & 536 (7) \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Detected deletions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element.
-For both data sets we used the thresholds derived from finding the 10\% FDR level in the simulated data set.
-The proportion of deletions that overlap repetitive elements discovered by Cloudbreak is similar to that of the other methods.}
-\label{deletionRepmaskpreds}
-\end{table}
-
-\newpage
-
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{rrr|rr}
- & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
-\hline
- &  Non-repeat & Repeat  &  Non-repeat & Repeat \\ 
- Total Number & 133 & 270 & 341 & 355 \\ 
-  \hline
-  Cloudbreak  & \textbf{32} (11) & \textbf{91} (\textbf{47}) & \textbf{169} (\textbf{61}) & \textbf{148} (\textbf{68}) \\ 
-  Breakdancer & 17 (5) & 22 (6) & 82 (7) & 44 (9) \\
-  Pindel      & 25 (\textbf{16}) & 54 (30) & 84 (35) & 82 (24) \\ 
-  MoDIL      & 5 (0) & 16 (3) & NA & NA \\ 
-   \hline
-\end{tabular}
-\end{center}
-\caption{Detected insertions in the simulated and NA18507 data sets identified by each tool, broken down by whether the insertion occurs in a RepeatMasker-annotated element.
-The maximum sensitivity cutoffs were used for both data sets.
-The proportion of insertions in repetitive elements discovered by Cloudbreak is similar to that of the other methods.}
-\label{insertionRepmaskpreds}
-\end{table}
-
-\newpage
-
-\begin{table}[h]
-\begin{center}
 \begin{tabular}{r|r|rr|rr|}
@@
 
-\begin{table}
+\begin{table}[h]
 \begin{center}

### 2014-04-30 manuscript/combinedRepmaskMultipleMappings.table.brew.tex
tables

\begin{table}
\begin{center}
\begin{tabular}{r|r|rr|rr}
  \cline{2-6}
 & & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\hline
 & & Non-repeat & Repeat  & Non-repeat & Repeat \\ 
\multirow{9}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & <%= chr2DelTotals['0'] %> & <%= chr2DelTotals['1'] %> & <%= NA18507DelTotals['0'] %> & <%= NA18507DelTotals['1'] %> \\ 
  \cline{2-6}
&  Cloudbreak  & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'Cloudbreak', '1') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'Cloudbreak', '1') %> \\ 
&  Cloudbreak-GEM  & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'Cloudbreak-GEM', '0') %> & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'Cloudbreak-GEM', '1') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'Cloudbreak-GEM', '0') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'Cloudbreak-GEM', '1') %> \\ 
&  BreakDancer & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'Breakdancer', '1') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'Breakdancer', '1') %> \\
&  GASVPro     & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'GASVPro', '0') %> & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'GASVPro', '1') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'GASVPro', '0') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'GASVPro', '1') %> \\
&  DELLY-RP       & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'DELLY-RP', '0') %> & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'DELLY-RP', '1') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'DELLY-RP', '0') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'DELLY-RP', '1') %> \\
&  DELLY-SR       & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'DELLY-SR', '0') %> & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'DELLY-SR', '1') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'DELLY-SR', '0') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'DELLY-SR', '1') %> \\
&  Pindel      & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'Pindel', '0') %> & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'Pindel', '1') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'Pindel', '0') %> & <%= printPredTableCell(NA18507DelPreds, NA18507ExDelPreds, 'Pindel', '1') %> \\ 
&  MoDIL      & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'MoDIL', '0') %> & <%= printPredTableCell(chr2DelPreds, chr2ExDelPreds, 'MoDIL', '1') %> & NA & NA \\ 
   \hline
\multirow{6}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & <%= chr2InsTotals['0'] %> & <%= chr2InsTotals['1'] %> & <%= NA18507InsTotals['0'] %> & <%= NA18507InsTotals['1'] %> \\ 
\cline{2-6}
&  Cloudbreak  & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'Cloudbreak', '1') %> & <%= printPredTableCell(NA18507InsPreds, NA18507ExInsPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(NA18507InsPreds, NA18507ExInsPreds, 'Cloudbreak', '1') %> \\ 
&  Cloudbreak-GEM  & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'Cloudbreak-GEM', '0') %> & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'Cloudbreak-GEM', '1') %> & <%= printPredTableCell(NA18507InsPreds, NA18507ExInsPreds, 'Cloudbreak-GEM', '0') %> & <%= printPredTableCell(NA18507InsPreds, NA18507ExInsPreds, 'Cloudbreak-GEM', '1') %> \\ 
&  Breakdancer & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'Breakdancer', '1') %> & <%= printPredTableCell(NA18507InsPreds, NA18507ExInsPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(NA18507InsPreds, NA18507ExInsPreds, 'Breakdancer', '1') %> \\
&  Pindel      & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'Pindel', '0') %> & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'Pindel', '1') %> & <%= printPredTableCell(NA18507InsPreds, NA18507ExInsPreds, 'Pindel', '0') %> & <%= printPredTableCell(NA18507InsPreds, NA18507ExInsPreds, 'Pindel', '1') %> \\ 
&  MoDIL      & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'MoDIL', '0') %> & <%= printPredTableCell(chr2InsPreds, chr2ExInsPreds, 'MoDIL', '1') %> & NA & NA \\ 
\hline
\end{tabular}
\end{center}
\caption{Detected deletions and insertions on the simulated and NA18507 data sets identified by each tool, broken down by whether the deletion overlaps with a RepeatMasker-annotated element. For all calls the maximum sensitivity cutoff was used.}
\label{predsByRepmask}
\end{table}

### 2014-04-30 manuscript/deletionRepmask.table.brew.tex
tables

--- 2013-02-27 manuscript/repmask.table.brew.tex
@@
   Cloudbreak  & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Cloudbreak', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Cloudbreak', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Cloudbreak', '1') %> \\
-  Breakdancer & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Breakdancer', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Breakdancer', '1') %> \\
+  BreakDancer & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Breakdancer', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Breakdancer', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Breakdancer', '1') %> \\
   GASVPro     & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'GASVPro', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'GASVPro', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'GASVPro', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'GASVPro', '1') %> \\
-  DELLY       & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY', '1') %> \\
+  DELLY-RP       & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY-RP', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY-RP', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY-RP', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY-RP', '1') %> \\
+  DELLY-SR       & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY-SR', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'DELLY-SR', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY-SR', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'DELLY-SR', '1') %> \\
   Pindel      & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(chr2Preds, chr2ExPreds, 'Pindel', '1') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '0') %> & <%= printPredTableCell(NA18507Preds, NA18507ExPreds, 'Pindel', '1') %> \\

### 2014-04-30 manuscript/bmc_additional1.tex
drafts

@@
 
-\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}
+\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\ (Appendix)}
 
@@
 
-\begin{figure}
+\tableofcontents
+
+\newpage
+
+\section{Additional Figures}
+
+\bigskip
+
+\begin{figure}[h]
 \centering
@@
 \clearpage
+
+\section{Additional Tables}
 
@@
 \end{center}
-\caption{Table A4. Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
+\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
 \label{deletionGenotypeaccuracy}

### 2014-04-30 manuscript/whelan_et_al_cloudbreak_ed.tex
drafts

--- 2013-08-06 manuscript/bmc_article.tex
@@
 
-Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing data processing problems such as SV detection.
+Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequence processing problems such as SV detection.
 MapReduce divides computation across a cluster into three phases.
@@
 
-  \subsection*{Table 2 - Variants Detected in the Simulated Data Set}
+  \subsection*{Table 2 - Variants detected in the simulated data set}
 The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses).

### 2014-04-30 src/main/scripts/manuscript.R
manu script

@@ -6,6 +6,10 @@ library(VariantAnnotation)
 library(brew)
 library(ggplot2)
 
+fMeasure <- function(p, r) {
+  (2 * p * r) / (p + r) 
+}
+
 drawPerfLine <- function(perf, col, lineType, maxFP) {
   calls <- c(perf$Calls, min(perf$Calls) - min(perf$TP), 0)
   tp <- c(perf$TP, 0, 0)
@@ -69,7 +73,8 @@ sizeclasses <- function(sizelens) {
   sc[sizelens > 250 & sizelens <= 500] <- "251-500bp"
   sc[sizelens > 500 & sizelens <= 1000] <- "501-1000bp"
   sc[sizelens > 1000] <- "> 1000bp"  
-  factor(sc, levels=c("< 40bp", "40-100bp","101-250bp","251-500bp","501-1000bp","> 1000bp"),ordered=TRUE)
+  #factor(sc, levels=c("< 40bp", "40-100bp","101-250bp","251-500bp","501-1000bp","> 1000bp"),ordered=TRUE)
+  factor(sc, levels=c("< 40bp", "40-100bp","101-250bp","251-500bp","501-1000bp","> 1000bp"))
 }
 
 processDeletionPredictions <- function(name, predfile, trueDels) {
@@ -229,12 +234,12 @@ modilChr2InsertionHitsFileMaxSensitivity <- '~/Documents/gene_rearrange/svpipeli
 # # JCVI SIM WGS
 # 
 # # deletions
-# cloudbreakBWAWGSDeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/_allindels_100bp_dip/bwa_del.perf.txt'
-# breakdancerWGSDeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_WGS_allindels_100bp_dip/human_b36_male_WGS_venterindels_c15_i100_s30_rl100_sort_35_2_3_del.perf.txt'
-# gasvProWGSDeletionPerfFile <-  '~/Documents/gene_rearrange/svpipeline/venter_WGS_allindels_100bp_dip/gasvpro_gte40.perf.txt'
-# dellyWGSDeletionPerfFile <- '~/Documents/gene_rearrange/svpipeline/venter_WGS_allindels_100bp_dip/human_b36_male_WGS_venterindels_c15_i100_s30_rl100_sort.delly_q0_c5_gte40.perf.txt'
[…]
